### 2022-08-21 proposal/MSECE_thesis_proposal.tex
Add initial files for thesis proposal.

\documentclass{article}

\usepackage{cite}
\usepackage{url}
\usepackage[numbib]{tocbibind}
\usepackage[lmargin=1in,rmargin=1in,tmargin=1in,bmargin=1in]{geometry}

\title{
    Master of Science in Electrical \& Computer Engineering\\
    \underline{Master's Thesis Proposal}\\
    Transformer-based Language Modeling of Air Traffic Communications
}

\author{Aaron Van De Brook---2456908}

\date{\today}

\begin{document}
    \maketitle

    % TODO: need to write abstract (probably last step)
    \begin{center}
        \section*{\textit{Abstract}}
        \textit{Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.}
    \end{center}

    \section{Introduction}
        The Transformer neural network architecture was originally designed and created for Neural Machine Translation (NMT) tasks \cite{vaswani_attention_2017} and has since seen massive success in various other Natural Language Processing (NLP) tasks such as language modeling (both casual and masked language modeling tasks) \cite{devlin_bert_2019}, prompt completion \cite{radford_improving_2018}, and sequence-to-sequence (S2S) classification \cite{lewis_bart_2019}.
        Recently, Automatic Speech Recognition (ASR) models have started to incorporate language models \cite{badrinath_automatic_2022} (sometimes also referred to as linguistic models) in addition to acoustic models \cite{li2019jasper} as the state of the art models begin to shift towards generative end-to-end approaches \cite{hannun2014deep}.
        At the time of writing this, the current state of the art ASR models are a mix between Transformer and Convolutional neural network (CNN) architectures\footnote{According to \url{https://paperswithcode.com/sota/automatic-speech-recognition-on-lrs2}} \cite{baevski2020wav2vec}.

        As ASR models begin to breakout into the aerospace domain, one of the most limiting factors in their development and deployment has been their relatively high word error rate(s) (WER) \cite{vsmidl2019air,juan2020automatic,badrinath_automatic_2022} as compared to typical state of the art models.
        This is due, at least in part, to the severe lack of transcribed data as compared to the more general ASR domain.
        For instance the best estimates of combined labeled datasets for Air Traffic Control (ATC) communications reach approximately 180 hours of speech data \cite{juan2020automatic} whereas state of the art ASR datasets typically reach approximately 1000 hours of speech data \cite{librispeech_2015}.
        This has lead to the recommendation of semi-supervised learning models wherein unlabeled air traffic communications are transcribed by a pretrained ASR model with a relatively low WER.
        The transcriptions are subsequently scored by a language model to obtain some insight into the transcription quality \cite{badrinath_automatic_2022,zuluaga2021contextual}.
        Depending on the distribution of the transcription scores, a certain percentage (best $n$ candidates) of the transcribed data is used to further fine-tune the ASR model.

        Despite the effectiveness of transformer-based architectures for language modeling NLP tasks \cite{devlin_bert_2019,lewis_bart_2019,liu2019roberta}, there is little to no implementation of them for modeling ATC communications for semi-supervised learning or otherwise in the existing literature.
        Additionally, my preliminary work into the transfer learning of BERT models in the ATC domain yield surprisingly high perplexity scores before overfitting to the training data.
        This suggests that a more sophisticated approach is needed to make use of transformer-based architectures such as BERT or RoBERTa for modeling air traffic communications.

    \section{Literature Review}

    \section{Methodology}

    \section{Deliverables \& Timeline}
        \begin{table}[h!]
            \centering
            \begin{tabular}{|c|c|c|}
                \hline
                \# & Deliverable & Date\\
                \hline
                0 & Submit thesis proposal & \today\\
                1 & & \\
                \hline
            \end{tabular}
            \label{table:timeline}
            \caption{Deliverables and approximate delivery dates.}
        \end{table}

    \newpage
    \bibliographystyle{IEEEtran}
    \bibliography{proposal_refs}
\end{document}

### 2022-10-24 proposal/MSECE_thesis_proposal.tex
Final revisions to proposal.

@@
-\documentclass{article}
+\documentclass[10pt]{article}
 
 \usepackage{cite}
-\usepackage{url}
+\usepackage{hyperref}
+\usepackage{listings}
+\usepackage{lipsum}
+\usepackage{multicol}
+\usepackage{ragged2e}
+\usepackage{newtxtext}
+\usepackage{setspace}
+\usepackage{sectsty}
 \usepackage[numbib]{tocbibind}
-\usepackage[lmargin=1in,rmargin=1in,tmargin=1in,bmargin=1in]{geometry}
+\usepackage[lmargin=1.5in,rmargin=1.5in,tmargin=1in]{geometry}
 
-\title{
-    Master of Science in Electrical \& Computer Engineering\\
-    \underline{Master's Thesis Proposal}\\
-    Transformer-based Language Modeling of Air Traffic Communications
-}
-
-\author{Aaron Van De Brook---2456908}
-
-\date{\today}
+% Conforming somewhat to the NeurIPS format
+\settowidth{\parindent}{}
+\setlength{\parskip}{5.5pt}
+\sectionfont{\large}
+\subsectionfont{\normalsize}
 
 \begin{document}
-    \maketitle
+    % Manually specify/format title, because making small formatting changes to the title is too damn frustrating
+    \begin{center}
+        \LARGE
+        \textbf{Language Modeling of Air-Traffic Communications}\\
+        \Large
+        Master of Science in Electrical \& Computer Engineering\\
+        Master's Thesis Proposal\\
 
-    % TODO: need to write abstract (probably last step)
+        \vspace*{0.25in}
+
+        \normalsize
+        \textbf{Aaron Van De Brook---2456908}\\
+        \textit{Department of Electrical Engineering \& Computer Science}\\
+        \textit{Embry-Riddle Aeronautical University}\\
+        vandebra@my.erau.edu
+    \end{center}
+
+    % An abstract. In 100 to 350 words, briefly summarize the problem to solve, the objectives of the project,
+    % and the methodology to use.
     \begin{center}
         \section*{\textit{Abstract}}
-        \textit{Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
-Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
-Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
-Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.}
+        \justifying
+        \textit{
+            Transformer neural networks have proven to be effective in several problems domains and have seen extensive use
+            in Natural Language Processing (NLP), particularly language modeling, through models such as BERT, BART, and GPT. Due
+            to the success of transformers in the general NLP domain, I propose the idea of developing, training, and evaluating
+            a transformer for aviation english which, despite having regulated phraseology through organizations such as ICAO,
+            can take many forms including that of conversational english. The variance in the structure of the language used
+            in aviation communications coupled with lack of use of transformer-based language models in the current literature
+            leaves a significant gap that can be addressed by the proposed work and expanded upon in future work.
+        }
     \end{center}
 
+    % An introduction (to the problems to be addressed). In a few paragraphs, clearly define what the problem is being
+    % addressed/solved in the thesis. Background information is needed. A literature review is needed to provide an
+    % account of the state-of-the-art solutions of this problem by other researchers or industry experts.
+    % Objectives of the thesis should be provided here as well.
     \section{Introduction}
-        The Transformer neural network architecture was originally designed and created for Neural Machine Translation (NMT) tasks \cite{vaswani_attention_2017} and has since seen massive success in various other Natural Language Processing (NLP) tasks such as language modeling (both casual and masked language modeling tasks) \cite{devlin_bert_2019}, prompt completion \cite{radford_improving_2018}, and sequence-to-sequence (S2S) classification \cite{lewis_bart_2019}.
-        Recently, Automatic Speech Recognition (ASR) models have started to incorporate language models \cite{badrinath_automatic_2022} (sometimes also referred to as linguistic models) in addition to acoustic models \cite{li2019jasper} as the state of the art models begin to shift towards generative end-to-end approaches \cite{hannun2014deep}.
-        At the time of writing this, the current state of the art ASR models are a mix between Transformer and Convolutional neural network (CNN) architectures\footnote{According to \url{https://paperswithcode.com/sota/automatic-speech-recognition-on-lrs2}} \cite{baevski2020wav2vec}.
+        The transformer neural network architecture was originally designed and created for Neural Machine
+        Translation (NMT) tasks \cite{vaswani_attention_2017} and has since seen massive success in various
+        other Natural Language Processing (NLP) tasks such as language modeling (both casual and masked language
+        modeling tasks) \cite{devlin_bert_2019}, prompt completion \cite{radford_improving_2018}, and
+        sequence-to-sequence (S2S) classification \cite{lewis_bart_2019}.
+        Recently, Automatic Speech Recognition (ASR) models have started to incorporate language models
+        \cite{badrinath_automatic_2022} (sometimes also referred to as linguistic models) in addition to
+        acoustic models \cite{li_jasper_2019} as the state-of-the-art models begin to shift towards generative
+        end-to-end approaches \cite{hannun_deep_2014}.
+        At the time of this writing, the current state-of-the-art ASR models are a mix between transformer
+        and Convolutional neural network (CNN) architectures
+        \footnote{According to \url{https://paperswithcode.com/sota/speech-recognition-on-librispeech-test-clean}}
+        \cite{baevski_wav2vec_2020}.
 
-        As ASR models begin to breakout into the aerospace domain, one of the most limiting factors in their development and deployment has been their relatively high word error rate(s) (WER) \cite{vsmidl2019air,juan2020automatic,badrinath_automatic_2022} as compared to typical state of the art models.
+        As ASR models begin to breakout into the aviation domain, one of the most limiting factors in their
+        development and deployment has been their relatively high word error rate(s) (WER)
+        \cite{smidl_air_2019,zuluaga-gomez_automatic_2020,badrinath_automatic_2022} as compared to typical state-of-the-art models.
         This is due, at least in part, to the severe lack of transcribed data as compared to the more general ASR domain.
-        For instance the best estimates of combined labeled datasets for Air Traffic Control (ATC) communications reach approximately 180 hours of speech data \cite{juan2020automatic} whereas state of the art ASR datasets typically reach approximately 1000 hours of speech data \cite{librispeech_2015}.
-        This has lead to the recommendation of semi-supervised learning models wherein unlabeled air traffic communications are transcribed by a pretrained ASR model with a relatively low WER.
-        The transcriptions are subsequently scored by a language model to obtain some insight into the transcription quality \cite{badrinath_automatic_2022,zuluaga2021contextual}.
-        Depending on the distribution of the transcription scores, a certain percentage (best $n$ candidates) of the transcribed data is used to further fine-tune the ASR model.
+        For instance the best estimates of combined labeled datasets for Air Traffic Control (ATC)
+        communications reach approximately 180 hours of speech data \cite{zuluaga-gomez_automatic_2020} whereas
+        state-of-the-art ASR datasets typically reach approximately 1000 hours of speech data \cite{panayotov_librispeech_2015}.
+        This has lead to the use of semi-supervised learning methods wherein unlabeled air
+        traffic communications are transcribed by a pretrained ASR model with a relatively low WER.
+        The transcriptions are subsequently scored by a language model to obtain some insight into the
+        transcription quality \cite{badrinath_automatic_2022,zuluaga-gomez_contextual_2021}.
+        Depending on the distribution of the transcription scores, a certain percentage of the transcribed data or candidate
+        samples under a specific level of uncertainty are added to the larger data set and used to retrain the model. This
+        process can repeat until there is no more unlabeled data to use or until a desired WER has been reached.
 
-        Despite the effectiveness of transformer-based architectures for language modeling NLP tasks \cite{devlin_bert_2019,lewis_bart_2019,liu2019roberta}, there is little to no implementation of them for modeling ATC communications for semi-supervised learning or otherwise in the existing literature.
-        Additionally, my preliminary work into the transfer learning of BERT models in the ATC domain yield surprisingly high perplexity scores before overfitting to the training data.
-        This suggests that a more sophisticated approach is needed to make use of transformer-based architectures such as BERT or RoBERTa for modeling air traffic communications.
+        Despite the effectiveness of transformer-based architectures for language modeling NLP tasks
+        \cite{devlin_bert_2019,lewis_bart_2019,liu_roberta_2019}, there is little to no implementation
+        of them for modeling ATC communications for semi-supervised learning or otherwise in the
+        existing literature.
+        Additionally, my preliminary work into the transfer learning of BERT models in the ATC
+        domain yield surprisingly high perplexity scores before overfitting to the training data.
+        This suggests that a more sophisticated approach is needed to make use of transformer-based
+        architectures such as BERT or RoBERTa for modeling air traffic communications.
 
-    \section{Literature Review}
+    % TODO: formal literature review
+    \section{Background \& Literature}
+        % transformer history
+        The transformer neural network architecture was proposed in 2017 for Neural Machine Translation tasks and immediately achieved
+        state-of-the-art (28.4 BLEU on WMT 2014 English-to-German; 41.8 BLEU on WMT English-to-French datasets)
+        \cite{vaswani_attention_2017}. Transformer architectures have been found to be extremely effective at learning representations
+        and understandings of languages to predict token probabilities as opposed to transforming one language into another
+        \cite{devlin_bert_2019,liu_roberta_2019}. Transformers have also been found to be very effective at
+        other NLP-related tasks such as prompt completion and sentiment analysis among others (i.e.~auto-regressive and sequence
+        classification tasks, respectively) \cite{lewis_bart_2019,radford_improving_2018}.
 
+        % ASR context (suggest some need for semi-supervised approaches and therefore LMs)
+        End-to-end generative models for automatic speech recognition models have made significant progress in recent years with
+        current state-of-the-art models achieving WERs as low as 2\% on LibriSpeech test sets
+        \cite{han_contextnet_2020,kriman_quartznet_2020,baevski_wav2vec_2020,li_jasper_2019}. This has led to the development of ASR
+        models for the aviation domain, specifically, in air traffic control communications
+        \cite{badrinath_automatic_2022,smidl_air_2019,zuluaga-gomez_automatic_2020,srinivasamurthy_semi-supervised_2017}.
+        However, due to the lack of transcribed data in the aviation domain
+        \cite{zuluaga-gomez_automatic_2020,srinivasamurthy_semi-supervised_2017,badrinath_automatic_2022,smidl_air_2019}
+        ASR models maintain relatively high WERs compared to their counterparts in the more generalized ASR domain
+        \cite{zuluaga-gomez_automatic_2020,badrinath_automatic_2022}. Transfer learning has even yielded limited results
+        in this domain (depending on model architecture and dataset quality)
+        \cite{badrinath_automatic_2022,zuluaga-gomez_automatic_2020}.
+
+        % Unsupervised/semi-supervised in general and LMs commonly used with them
+        Unsupervised and semi-supervised methodologies have become popular recently in attempts to address limited data availability
+        and develop new approaches towards modeling human speech (notably, wav2vec has achieved state-of-the-art performance with
+        very little training data)
+        \cite{baevski_wav2vec_2020,badrinath_automatic_2022,srinivasamurthy_semi-supervised_2017,zuluaga-gomez_contextual_2021}.
+        Language models are an integral part semi-supervised learning. They are used to obtain a certainty score (or uncertainty
+        score, as the case may be) for the predicted text, these are usually either word lattices or N-gram models
+        \cite{badrinath_automatic_2022,srinivasamurthy_semi-supervised_2017,zuluaga-gomez_contextual_2021}.
+        While these have proven to be adequate for most self-supervised learning tasks, they are hardly state-of-the-art.
+        Wav2vec, the current state-of-the-art unsupervised model (and possibly top performing overall), uses a contrastive process
+        between convolutional feature extraction and transformer context representations \cite{baevski_wav2vec_2020}.
+
+        % LMs and NLP in aero. domain
+        Various natural language processing methods have been applied to the aviation domain to help deal with miscommunications
+        and try to mitigate safety incidents \cite{ragnarsdottir_language_2003,tanguy_natural_2016,madeira_machine_2021}.
+        Some machine learning approaches have been implemented to analyze the text in aviation incident and safety reports to predict
+        contributing factors and topic models \cite{tanguy_natural_2016,madeira_machine_2021}. An ASR system with NLP post-processing
+        has also been proposed to analyze Air-Traffic communications and condense significant features (e.g.~weather, runway, and
+        heading info) into an XML language structure \cite{ragnarsdottir_language_2003}. Transformer language models such as BERT,
+        RoBERTa, and DeBERTa have been applied to notice to airmen (NOTAM) messages to perform named entity recognition (NER) tasks,
+        translation tasks (between notations; e.g.~NOTAM notation to Airlang to make parsing tasks easier) and reduce the workload
+        for pilots during long-haul flights \cite{arnold_knowledge_2022}. Transformer models have also been used for speaker role
+        identification tasks in the aviation domain (e.g.~identifying pilot versus controller in communications), specifically,
+        a pretrained BERT model was used and fine-tuned on problem specific data and compared to other models that performed
+        well at speaker and role identification tasks in general \cite{guo_comparative_2022}.
+
+    % Methodology. Describe the approaches you plan to use for solving the problem at hand. Specify the
+    % tools/instruments/procedures/designs you need to use to finish your work. Also specify other resources,
+    % including the budgets, you need to use to finish your work.
+    %
+    % TODO: CITATIONS/REFERENCES!!
     \section{Methodology}
+        This section lays out the methodology and sources for items such as data, model design, model training and testing methodology
+        such as performance metrics and loss functions, preliminary data analysis and statistics, potential application areas, and
+        project specific development and management techniques.
 
+        \subsection{Data Sources \& Means of Collection}
+        There are a few text datasets that exist that can be used either freely or accessed easily. There are also a few ASR datasets
+        in which the primary spoken language is english (or is sorted in such a way that it is easy to isolate the english transcripts
+        from the others) and the transcripts can be mined for english text for training language models. The table below lists the
+        relevant datasets (hereafter referred to as \textit{the corpus}):
+
+        % TODO: identify resources for things like NOTAMs, safety reports from relevant regulatory agencies, etc.
+        %       probably not all applicable, but worth looking into regardless.
+        \begin{table}[h!]
+            \centering
+            \begin{tabular}{c|c|c|c}
+                Dataset Name & Access & Data Type & Citation \\
+                \hline
+                Air-Traffic Control Complete & Paid (have access) & Real, transcripts & \cite{godfrey_air_1994} \\
+                ATCSpeech & Free/Paid (apply for access) & Real, transcripts & \cite{yang_atcspeech_2020} \\
+                HIWIRE & Paid (have access) & Simulated, scripts & \cite{segura_hiwire_2007} \\
+                ATCC & Free (have access) & Real, transcripts & \cite{smidl_air_2019} \\
+                ATCOSIM & Free (have access) & Simulated, transcripts & \cite{hofbauer_atcosim_2008} \\
+            \end{tabular}
+            \label{table:datasets}
+            \caption{Dataset name, access requirements (including prior availability through past projects), and citation information.
+                    \textbf{Real/Simulated} corresponds to real world communications that were transcribed in some way or communications
+                    that were recorded in a simulated environment. \textbf{Transcripts/Scripts} corresponds to real communications that were
+                    transcribed or scripts that were generated.}
+        \end{table}
+
+        Since the text in the corpus comes from a variety of sources, it will have to be processed and cleaned of special characters
+        or sequences specific to the original dataset. Since that process is so specific to the data source and original dataset(s),
+        the specific methodology used to preprocess and clean the data will be detailed later in the thesis.
+
+        % - primarily transformer based, since almost all state-of-the-art approaches are transformers. maybe Bayesian formulation as
+        %   well, depending on how "noisy" the data is.
+        % - potential optimizers to use
+        % - potential loss function(s)
+        % - baseline models for comparison: N-gram, word lattice, (maybe) RNN, CNN, LSTM
+        \subsection{Model Design(s)}
+        \textbf{Development Model(s)}. At the moment the prevailing idea for the main model architecture is the transformer
+        \cite{vaswani_attention_2017}, considering its widespread success in language modeling and other NLP tasks. Language structure
+        and grammar in aviation communications typically follows a very formal structure, however, deviations from this formality are
+        common, for this reason it may be worth considering Bayesian Deep Learning (BDL) approach to the transformer language model
+        since it has demonstrated good performance in accurately classifying data with high variance. The main tradeoffs in developing
+        a BDL-transformer hybrid model would most likely be development time, and training time---from past work, end-to-end BDL models
+        have taken an average of three times longer than ``normal'' networks to train.
+
+        \textbf{Baseline Models}. These are the models against which the performance and effectiveness of language modeling, and NLP
+                                           %establish?
+        tasks moreover, will be compared to determine the gain in effectiveness and performance between the developed models and
+        previously used models. The models that have been used in previous works in similar or identical problem domains are below:
+
+        % TODO: the vertical alignment is aggravating; fix if possible
+        \begin{multicols}{2}
+            \begin{itemize}
+                \item N-Gram
+                \item Word Lattice
+                \item RNN-based Language Models
+                \item CNN-based Language Models
+                \item LSTM-based Language Models
+                \item Pretrained models, e.g.~BERT, RoBERTa,
+            \end{itemize}
+        \end{multicols}
+
+        \subsection{Model Performance Metrics \& Evaluation}
+        \textbf{Performance Metrics}.
+        The common accuracy/error rate quantification metric used in language modeling is perplexity, which is typically formulated as
+        the normalized inverse of the sequence probability, along with the typical classification problem metrics (i.e.~accuracy,
+        precision, recall, F-1 score).
+
+        \textbf{Loss Function(s)}.
+        The most common loss function used in language modeling is Cross Entropy loss. Connectionist Temporal Classification (CTC) loss
+        can also be used if the input sequence can be formulated as a time series, but this is most commonly used in ASR, since
+        transcript labels are accompanied by audio time series.
+
+        During the training and validation phases, the performance metrics will be logged each step (or epoch, whichever is more
+        applicable) and plotted, most likely in TensorBoard, to keep track of performance during training. During the testing phase,
+        the accuracy and average model perplexity will be reported and, if applicable, a confusion matrix will be generated and plotted
+        to help illustrate the model performance.
+
+        % uncertainty scoring, different decoding heads for things like masked language modeling, sequence classification,
+        % token classification (named entity recognition), prompt completion (thinking this could be used to complete corrupted
+        % text/transmissions)
+        \subsection{Application Areas}
+        As evidenced by the \textit{transformers} API from HuggingFace\footnote{\url{https://huggingface.co/docs/transformers}} (and the models hosted
+        on their website), transformer models can be trained to a generic checkpoint through objectives such as masked language modeling to learn
+        token probabilities, and fine-tuned on downstream tasks with task specific decoding layers (also commonly referred to as heads
+        e.g.~sequence classification head, language modeling head, etc.). Due to this fact, the range of application areas is fairly
+        extensive, ranging from casual and masked language modeling to named entity recognition (NER; also referred to as token
+        classification), prompt completion, and sequence classification tasks such as sentiment analysis and infinitely more tasks and
+        problems within those categories.
+
+        There are several application areas that can be addressed by the work this thesis produces, but not all will be extensively
+        tested. One of the main areas that can be addressed easily and relatively quickly is \textbf{uncertainty scoring} i.e.~using a
+        LM to score transcripts generated by ASR models and determine a level of uncertainty for the validity of the transcript.
+        A sequence classification task can be formed to \textbf{identify communication types} or perform \textbf{role identification}
+        \footnote{There is a fair amount of research here as well, so a comparison of results wouldn't be difficult to perform.}
+        (e.g.~pilot, air-controller, ground-controller, etc.; this could also be broken down into a NER task, depending on the data).
+
+        % TODO: expand or remove (probably keep, just elaborate a bit on specifics, identify alternatives)
+        \subsection{Model Development}
+        Fortunately, there is a plethora of libraries available for developing machine learning, deep learning, and NLP models. The
+        main packages that have been chosen (listed in Appendix A) have front-end language bindings for Python and most have
+        back-end APIs written in C/C++ and Rust for streamlined computations.
+
+        % TODO: review min and max hardware required to train models i.e. bare minimum to train model,
+        % preferences for efficient training (workstation with GPU/TPU, AWS/Azure compute cluster, etc.)
+        \subsection{Resource Requirements}
+
+    \clearpage
+    % Deliverables and timeline. Provide a detailed schedule and deliverables, shown in a table like the one below.
     \section{Deliverables \& Timeline}
@@
             \centering
-            \begin{tabular}{|c|c|c|}
+            \begin{tabular}{l|l}
+                \textbf{Deliverable} & \textbf{Date} \\
                 \hline
-                \# & Deliverable & Date\\
-                \hline
-                0 & Submit thesis proposal & \today\\
-                1 & & \\
-                \hline
+                Submit thesis proposal & October, 2022 \\
+                Accumulate data sets\footnotemark[4] & December, 2022 \\
+                Finish performing data analysis\footnotemark[5] & January, 2023 \\
+                Finalize model designs and traning/testing methodologies for existing architectures & February, 2023 \\
+                Finish developing models/scripts for training, testing models & April, 2023 \\
+                Full scale training and testing of models & May, 2023 \\
+                Analysis of results\footnotemark[6] & August, 2023 \\
+                Full scale testing and analysis in designated application areas \& analysis of results & September, 2023 \\
             \end{tabular}
@@
         \end{table}
+        % apparently footnotes don't work inside of float environments, so...
+        \footnotetext[4]{Most likely hosted on OneDrive or SharePoint (associated with ERAU student account).}
+        \footnotetext[5]{This includes relevant statistics about the data, as well as mining and converting all text into a common form.}
+        \footnotetext[6]{This will include tasks such as performance analysis/comparison, generating relevant plots, and preliminary testing for
+                         designated application areas.}
 
     \newpage
+    % References. List the references in IEEE format.
     \bibliographystyle{IEEEtran}

### 2023-01-06 source/atccommunication.py
Templates for datasets.

@@ -1,16 +1,0 @@
-from data import Data
-
-class ATCCommunication(Data):
-    """
-    This class defines the format of the data for the Air Traffic Control Communication
-    and implements functions to parse the data into a common format for data analysis.
-
-    This dataset can be obtained from:
-        https://lindat.mff.cuni.cz/repository/xmlui/handle/11858/00-097C-0000-0001-CCA1-0
-    """
-
-    def __init__(self, data_root: str):
-        super(ATCCommunication, self).__init__(data_root)
-
-    def name(self) -> str:
-        return "ATCCommunication"

### 2023-01-06 source/atccomplete.py

@@ -7,6 +7,7 @@ from data import atccutils
 from pathlib import Path
 from pprint import pprint
 
+
 class ATCCompleteData(Data):
     """
     This class defines the data format of the Air Traffic Control Complete dataset and
@@ -110,9 +111,7 @@ class ATCCompleteData(Data):
             # extraneous info that could cause KeyErrors)
             for datum in transcript_data:
                 if "TEXT" in datum.keys():
-                    audio_duration = (
-                        datum["TIMES"]["end"] - datum["TIMES"]["start"]
-                    )
+                    audio_duration = datum["TIMES"]["end"] - datum["TIMES"]["start"]
                     manifest_data.append(
                         {
                             "audio_filepath": str(audio),
@@ -128,4 +127,4 @@ class ATCCompleteData(Data):
 
     @property
     def name(self):
-        return "ATCC"
+        return "Air Traffic Control Complete"

### 2023-01-06 source/atco2sim.py

@@ -1,5 +1,6 @@
 from data import Data
 
+
 class ATCO2SimData(Data):
     """
     This class defines the format for the Air Traffic Control 2 dataset and implements

### 2023-01-06 source/czechdataset.py

@@ -1,0 +1,19 @@
+import os
+import glob
+import subprocess
+from typing import *
+from data import Data
+
+
+class ZCUATCDataset(Data):
+    """ """
+
+    def __init__(self, data_root: str, **kwargs):
+        super(ZCUATCDataset, self).__init__(data_root, **kwargs)
+
+    def parse_transcripts(self) -> Dict[str, Union[str, float]]:
+        pass
+
+    @property
+    def name(self) -> str:
+        return "ZCU ATC"

### 2023-01-06 source/data/__init__.py

@@ -1,4 +1,4 @@
-from .data import Data, UtteranceStats
+from .data import Data, TokenStats
 from .utils import hiwireutils, atccutils
 
 __all__ = ["Data", "UtteranceStats" "hiwireutils", "atccutils"]

### 2023-01-06 source/data/data.py

@@ -1,7 +1,4 @@
 import os
-import json
-import librosa
-import librosa.display
 import matplotlib
 import matplotlib.collections
 import matplotlib.figure
@@ -14,15 +11,11 @@ from typing import *
 from pathlib import Path
 
 
-UtteranceStats = namedtuple(
-    "UtteranceStats", ["utterances", "sample_duration", "samples"]
-)
+TokenStats = namedtuple("TokenStats", ["tokens", "samples"])
 """
 Named tuple for ease of access and return for utterance statistics for the dataset.
 
-`utterances` corresponds to the total number of utterances found in the dataset
-
-`sample_duration` corresponds to the cumulative duration of the samples in the dataset
+`tokens` corresponds to the total number of tokens found in the dataset
 
 `samples` corresponds to the number of samples in the dataset
 """
@@ -30,6 +23,12 @@ Named tuple for ease of access and return for utterance statistics for the datas
 
 class Data:
     """
+    Top-level Data class that provides several methods for processing and analyzing text datasets for NLP processes.
+
+    This class should be extended and the following methods/properties implemented for each dataset:
+    * `parse_transcripts`
+    * `name`
+
     Attributes:
     -----------
     `_manifest_data`: list of dictionary objects. Each object corresponds to one data sample
@@ -41,11 +40,14 @@ class Data:
     * `offset` - if more than one sample is present in a single audio file, this field
     specifies its offset i.e. start time in the audio file. Type: `float`
 
-    `_random`:
+    `_random`: numpy seeded RNG instance
+
+    `_normalized`: bool indicating whether samples in the dataset have been normalized/preprocessed
     """
 
     _manifest_data: List[Dict[str, Union[float, str]]]
     _random: np.random.Generator
+    _normalized: bool
 
     def __init__(self, data_root: str, random_seed: int = None):
         """
@@ -76,7 +78,7 @@ class Data:
             "is an issue with the class that extended the Data class."
         )
 
-    def create_utterance_hist(
+    def create_token_hist(
         self,
         utterance_counts: List[int] = [],
         plot_type: Literal["matplotlib", "seaborn"] = "seaborn",
@@ -116,64 +118,7 @@ class Data:
         p.label()
         return p
 
-    def create_spectrograms(
-        self, n_plots=2, random_sample=True, plot_db=True
-    ) -> List[matplotlib.collections.QuadMesh]:
-        """
[…]

### 2023-01-06 source/hiwire.py

@@ -1,47 +1,0 @@
-import os
-import glob
-from typing import *
-from pathlib import Path
-from data import Data, hiwireutils
-
-
-class HIWIREData(Data):
-    """
-    The following attributes are defined in this class due to the specific formatting of
-    the HIWIRE dataset.
-
-    This dataset can be obtained here:
-        https://catalogue.elra.info/en-us/repository/browse/ELRA-S0293/
-
-    Attributes:
-    -----------
-    `_transcript_glob`: list of aths of transcript files that contain labels for the audio data in the dataset.
-
-    """
-
-    _transcript_glob: List[str]
-
-    def __init__(self, data_root: str, **kwargs):
-        super(HIWIREData, self).__init__(data_root, **kwargs)
-
-        # glob search string to get list of transcript paths (absolute paths)
-        search_string = os.path.join(data_root, "**/list*.txt")
-        HIWIREData._transcript_glob = glob.glob(search_string, recursive=True)
-
-    def parse_transcripts(self) -> Dict[str, Union[str, float]]:
-        # ensure there are actually transcript paths present
-        assert len(self._transcript_glob) != 0, "No transcript paths found."
-
-        manifest_data = []
-
-        for path in self._transcript_glob:
-            # parse returns a list of dictionaries
-            manifest_data.extend(hiwireutils.parse(path))
-
-        # save a copy of the manifest data to the class attribute before returning
-        HIWIREData._manifest_data = manifest_data
-        return manifest_data
-
-    @property
-    def name(self):
-        return "HIWIRE"

### 2023-01-06 source/main.py

@@ -1,7 +1,9 @@
 from typing import *
-from hiwire import HIWIREData
-from data import Data, UtteranceStats
+from data import Data, TokenStats
 from atccomplete import ATCCompleteData
+from atcosim import ATCOSimData
+from atco2sim import ATCO2SimData
+from czechdataset import ZCUATCDataset
 import matplotlib.pyplot as plt
 
 if __name__ == "__main__":
@@ -10,34 +12,24 @@ if __name__ == "__main__":
     RANDOM_SEED: int = 1
 
     # root dataset paths corresponding to data analysis classes
-    datasets: Dict[str, Data]  = {
-        "/data/s0293/S0293/speechdata": HIWIREData,
-        "/data/atc0_comp_ldc945s14a/atc0_comp_raw": ATCCompleteData,
+    datasets: Dict[str, Data] = {
+        "~/data/atc0_comp_LDC94S14A/atc0_comp": ATCCompleteData,
+        "~/data/ATCO": ATCOSimData,
+        "~/data/ATCO2": ATCO2SimData,
+        "~/data/ZCU_CZ_ATC": ZCUATCDataset,
     }
 
     for root_path, data_class in datasets.items():
-        data_analysis: Data = data_class(
-            data_root=root_path, random_seed=RANDOM_SEED
-        )
+        data_analysis: Data = data_class(data_root=root_path, random_seed=RANDOM_SEED)
 
         print(data_analysis.name)
         # parse transcripts in dataset
         data_analysis.parse_transcripts()
-        data_analysis.dump_manifest(f"{data_analysis.name}_all.json")
+        data_analysis.dump_corpus(f"{data_analysis.name}.txt")
         # utterance stats
-        # stats = data_analysis.calc_utterance_stats()
+        stats = data_analysis.calc_token_stats()
 
         # utterance distribution
-        # utterance_hist = data_analysis.create_utterance_hist()
-        # utterance_hist.plot(pyplot=True).show()
-
-        # spectrogram plots, randomly sampled (depending on the size of the dataset,
-        # this can take up a lot of time and memory)
-        spec = data_analysis.create_spectrograms(random_sample=False)
-        # reasoning for using plt.show() instead of spec.show() from matplotlib docs:
-        # "If you're running a pure python shell or executing a non-GUI python script,
-        #  you should use matplotlib.pyplot.show instead, which takes care of managing
-        #  the event loop for you."
+        utterance_hist = data_analysis.create_token_hist()
[…]

### 2023-01-06 thesis/MSECE_thesis.tex
Update gitignore, readme; setup thesis template

\documentclass[12pt]{article}

\usepackage{cite}
\usepackage{listings}
\usepackage{multicol}
\usepackage{newtxtext}
\usepackage{setspace}
\usepackage{sectsty}
\usepackage[numbib]{tocbibind}
\usepackage[lmargin=1.25in,rmargin=1.25in,tmargin=1.0in,bmargin=1.0in]{geometry}

\settowidth{\parindent}{}

\begin{document}
    % Riddle thesis title page
    \begin{center}
        \LARGE
        \textbf{
            \textit{
                Language Modeling of Aviation Communications\\
            }
        }

        \vspace*{0.2in}

        \normalsize
        by\\
        \textit{Aaron P.~Van De Brook}\\

        \vspace*{0.25in}

    \end{center}
    This thesis was prepared under the direction of the candidate's thesis
    committee chairman, \underline{Dr.~Jianhua Liu}, Department of \underline{Electrical Engineering \&
    Computer Science}, and has been approved by the members of the thesis committee.
    It was submitted to the Department of Electrical Engineering \& Computer Science
    and was accepted in partial fulfillment of the requirements for the degree of
    Masters of Electrical \& Computer Engineering.

    % Signature lines/committee members
    \begin{center}
        \begin{minipage}{3in}
            \vspace*{0.25in}
            THESIS COMMITTEE:\\
            \vspace*{0.25in}
            \hrule
            \vspace*{2pt}
            Jianhua Liu, Ph.D.\\
            Committee Chairman\\
            \vspace*{0.75in}
            \hrule
            \vspace*{2pt}
            Prashant Shekhar, Ph.D.\\
            Committee Member\\
            \vspace*{0.75in}
            \hrule
            \vspace*{2pt}
            Andrew Schneider, Ph.D.\\
            Committee Member\\
            \vspace*{0.75in}
            \hrule
            \vspace*{2pt}
            Farahzad Behi\\
            Department Chair, Electrical Engineering \& Computer Science\\
            \vspace*{0.75in}
            \hrule
            \vspace*{2pt}
            Associate Vice President for Academics
        \end{minipage}
    \end{center}
    \newpage
    \tableofcontents
    \newpage
    \doublespacing

    \section{Abstract}
    \section{Introduction}
    \section{Literature Review}
    \section{Methodology}
        \subsection{Data Source(s)}
        \subsection{Data Analysis}
        \subsection{Preprocessing}
    \section{Results}
    \section{Discussion}
    \section{Conclusion}

    \newpage
    \bibliography{refs}
    \bibliographystyle{plain}

\end{document}

### 2023-01-07 source/atccomplete.py
Small refactor in data class; implemented atco class.

@@ -96,7 +96,7 @@ class ATCCompleteData(Data):
         --------
         A list of dictionary objects.
         """
-        manifest_data = []
+        data = []
 
         for text, audio in zip(self._transcript_glob, self._audio_glob):
             # need absolute file path for compliance with NeMo manifest format
@@ -111,19 +111,11 @@ class ATCCompleteData(Data):
             # extraneous info that could cause KeyErrors)
             for datum in transcript_data:
                 if "TEXT" in datum.keys():
-                    audio_duration = datum["TIMES"]["end"] - datum["TIMES"]["start"]
-                    manifest_data.append(
-                        {
-                            "audio_filepath": str(audio),
-                            "text": datum["TEXT"],
-                            "duration": float(audio_duration),
-                            "offset": float(datum["TIMES"]["start"]),
-                        }
-                    )
+                    data.append(datum["TEXT"])
 
         # save manifest data to class attribute before returning
-        ATCCompleteData._manifest_data = manifest_data
-        return manifest_data
+        ATCCompleteData.data = data
+        return data
 
     @property
     def name(self):

### 2023-01-07 source/atcosim.py

@@ -1,4 +1,9 @@
+import os
+import glob
+import re
 from data import Data
+from typing import *
+from tokenizers.normalizers import NFD
 
 
 class ATCOSimData(Data):
@@ -8,10 +13,49 @@ class ATCOSimData(Data):
 
     This dataset is described in more depth and can be obtained here:
         https://www.spsc.tugraz.at/databases-and-tools/atcosim-air-traffic-control-simulation-speech-corpus.html
+
+    The data is shipped in iso (disk) format. To extract, mount the disk onto the file system and copy the data
+    into another directory.
+    ```
+    mount -t iso9660 -o loop atcosim.iso atcosimmount
+    cp -r atcosimmount .
+    ```
     """
 
     def __init__(self, data_root: str, **kwargs):
         super(ATCOSimData, self).__init__(data_root, **kwargs)
+        self.text_glob = glob.glob(os.path.join(data_root, "txtdata/**/*.txt"), recursive=True)
+        self.text_glob.remove(os.path.join(data_root, "txtdata/wordlist.txt"))
+
+    def parse_transcripts(self) -> List[str]:
+        data = []
+
+        # regular expressions for removing transcript annotations
+        xml_tag = re.compile(r"(<[A-Z]+>|</[A-Z]+>)")
+        annotation_tag = re.compile(r"(\[[A-Z]+\])")
+        special_chars = re.compile(r"[=~@]")
+        hesitation_tokens = re.compile(r"(ah|hm|ahm|yeah|aha|nah|ohh)")
+
+        for file in self.text_glob:
+            # read data from file
+            with open(file, "r") as f:
+                text = "".join([t.strip() for t in f.readlines()])
+
+            # remove transcript annotations
+            text = xml_tag.sub("", text)
+            text = annotation_tag.sub("", text)
+            text = special_chars.sub("", text)
+            text = hesitation_tokens.sub("", text)
+            # trim leading a trailing whitespace
+            text = text.strip()
+
+            # some transcripts are empty after removing transcriber
+            # annotations
+            if len(text) > 0:
+                data.append(text)
+
+        ATCOSimData.data = data
+        return data
 
     @property
     def name(self):

### 2023-01-07 source/data/data.py

@@ -45,7 +45,7 @@ class Data:
     `_normalized`: bool indicating whether samples in the dataset have been normalized/preprocessed
     """
 
-    _manifest_data: List[Dict[str, Union[float, str]]]
+    data: List[Dict[str, Union[float, str]]]
     _random: np.random.Generator
     _normalized: bool
 
@@ -62,7 +62,7 @@ class Data:
         # create random number generator sequence with specified seed, if applicable
         Data._random = np.random.default_rng(random_seed)
 
-    def parse_transcripts(self) -> Dict[str, Union[str, float]]:
+    def parse_transcripts(self) -> List[str]:
         """
         This method must be overridden and implemented for each implementation of this class
         for datasets.
@@ -104,13 +104,13 @@ class Data:
         """
         # check if manifest data has been generated, parse transcripts and generate
         # manifest data if not
-        if len(self._manifest_data) == 0:
+        if len(self.data) == 0:
             self.parse_transcripts()
 
         # check if utterance counts (optional arg) has been provided, calculate utterance
         # counts from transcriptions
         if len(utterance_counts) == 0:
-            for data in self._manifest_data:
+            for data in self.data:
                 words = data["text"].split(" ")
                 utterance_counts.append(len(words))
 
@@ -131,18 +131,18 @@ class Data:
         and total samples in the data
         """
         # check if manifest data has been generated
-        if len(self._manifest_data) == 0:
+        if len(self.data) == 0:
             self.parse_transcripts()
 
         total_token_count = 0
 
-        for data in self._manifest_data:
+        for data in self.data:
             utterances = data["text"].split(" ")
             total_token_count += len(utterances)
 
         return TokenStats(
             utterances=total_token_count,
-            samples=len(self._manifest_data),
+            samples=len(self.data),
         )
 
     def token_freq_analysis(self, normalize=False) -> Dict[str, Union[int, float]]:
@@ -157,12 +157,12 @@ class Data:
         --------
         `token_freqs` `dict` with tokens and number of occurrences of those tokens throughout the dataset.
         """
-        if len(self._manifest_data) == 0:
+        if len(self.data) == 0:
             self.parse_transcripts()
 
         token_freqs = {}
 
-        for sample in self._manifest_data:
+        for sample in self.data:
             sample = sample["text"]
             for token in sample.split():
                 if token in token_freqs.keys():
@@ -199,14 +199,14 @@ class Data:
         os.makedirs(str(outfile.parent), exist_ok=make_dirs)
 
[…]

### 2023-01-07 source/main.py

@@ -13,10 +13,10 @@ if __name__ == "__main__":
 
     # root dataset paths corresponding to data analysis classes
     datasets: Dict[str, Data] = {
-        "~/data/atc0_comp_LDC94S14A/atc0_comp": ATCCompleteData,
-        "~/data/ATCO": ATCOSimData,
-        "~/data/ATCO2": ATCO2SimData,
-        "~/data/ZCU_CZ_ATC": ZCUATCDataset,
+        "/home/avandebrook/data/atc0_comp_LDC94S14A/atc0_comp": ATCCompleteData,
+        "/home/avandebrook/data/ATCO/": ATCOSimData,
+        "/home/avandebrook/data/ATCO2": ATCO2SimData,
+        # "~/data/ZCU_CZ_ATC": ZCUATCDataset,
     }
 
     for root_path, data_class in datasets.items():
@@ -27,9 +27,9 @@ if __name__ == "__main__":
         data_analysis.parse_transcripts()
         data_analysis.dump_corpus(f"{data_analysis.name}.txt")
         # utterance stats
-        stats = data_analysis.calc_token_stats()
+        # stats = data_analysis.calc_token_stats()
 
         # utterance distribution
-        utterance_hist = data_analysis.create_token_hist()
-        utterance_hist.plot(pyplot=True)
-        plt.show()
+        # utterance_hist = data_analysis.create_token_hist()
+        # utterance_hist.plot(pyplot=True)
+        # plt.show()

### 2023-04-25 thesis/MSECE_thesis.tex
Update bibtex references; update lit review, data sources.

@@
 
-\settowidth{\parindent}{}
+% \settowidth{\parindent}{}
 
@@
 \section{Literature Review}
+% from proposal; revise and add to this as more references are used
+% transformer history
+The transformer neural network architecture was proposed in 2017 for Neural Machine Translation tasks and immediately achieved
+state-of-the-art (28.4 BLEU on WMT 2014 English-to-German; 41.8 BLEU on WMT English-to-French datasets)
+\cite{vaswani_attention_2017}. Transformer architectures have been found to be extremely effective at learning representations
+and understandings of languages to predict token probabilities as opposed to transforming one language into another
+\cite{devlin_bert_2019,liu_roberta_2019}. Transformers have also been found to be very effective at
+other NLP-related tasks such as prompt completion and sentiment analysis among others (i.e.~auto-regressive and sequence
+classification tasks, respectively) \cite{lewis_bart_2019,radford_improving_2018}.
+
+% ASR context (suggest some need for semi-supervised approaches and therefore LMs)
+End-to-end generative models for automatic speech recognition models have made significant progress in recent years with
+current state-of-the-art models achieving WERs as low as 2\% on LibriSpeech test sets
+\cite{han_contextnet_2020,kriman_quartznet_2020,baevski_wav2vec_2020,li_jasper_2019}. This has led to the development of ASR
+models for the aviation domain, specifically, in air traffic control communications
+\cite{badrinath_automatic_2022,smidl_air_2019,zuluaga-gomez_automatic_2020,srinivasamurthy_semi-supervised_2017}.
+However, due to the lack of transcribed data in the aviation domain
+\cite{zuluaga-gomez_automatic_2020,srinivasamurthy_semi-supervised_2017,badrinath_automatic_2022,smidl_air_2019}
+ASR models maintain relatively high WERs compared to their counterparts in the more generalized ASR domain
+\cite{zuluaga-gomez_automatic_2020,badrinath_automatic_2022}. Transfer learning has even yielded limited results
+in this domain (depending on model architecture and dataset quality)
+\cite{badrinath_automatic_2022,zuluaga-gomez_automatic_2020}.
+
+% Unsupervised/semi-supervised in general and LMs commonly used with them
+Unsupervised and semi-supervised methodologies have become popular recently in attempts to address limited data availability
+and develop new approaches towards modeling human speech (notably, wav2vec has achieved state-of-the-art performance with
+very little training data)
+\cite{baevski_wav2vec_2020,badrinath_automatic_2022,srinivasamurthy_semi-supervised_2017,zuluaga-gomez_contextual_2021}.
+Language models are an integral part semi-supervised learning. They are used to obtain a certainty score (or uncertainty
+score, as the case may be) for the predicted text, these are usually either word lattices or N-gram models
+\cite{badrinath_automatic_2022,srinivasamurthy_semi-supervised_2017,zuluaga-gomez_contextual_2021}.
+While these have proven to be adequate for most self-supervised learning tasks, they are hardly state-of-the-art.
+Wav2vec, the current state-of-the-art unsupervised model (and possibly top performing overall), uses a contrastive process
+between convolutional feature extraction and transformer context representations \cite{baevski_wav2vec_2020}.
+
+% LMs and NLP in aero. domain
+Various natural language processing methods have been applied to the aviation domain to help deal with miscommunications
+and try to mitigate safety incidents \cite{ragnarsdottir_language_2003,tanguy_natural_2016,madeira_machine_2021}.
+Some machine learning approaches have been implemented to analyze the text in aviation incident and safety reports to predict
+contributing factors and topic models \cite{tanguy_natural_2016,madeira_machine_2021}. An ASR system with NLP post-processing
+has also been proposed to analyze Air-Traffic communications and condense significant features (e.g.~weather, runway, and
+heading info) into an XML language structure \cite{ragnarsdottir_language_2003}. Transformer language models such as BERT,
+RoBERTa, and DeBERTa have been applied to notice to airmen (NOTAM) messages to perform named entity recognition (NER) tasks,
+translation tasks (between notations; e.g.~NOTAM notation to Airlang to make parsing tasks easier) and reduce the workload
+for pilots during long-haul flights \cite{arnold_knowledge_2022}. Transformer models have also been used for speaker role
+identification tasks in the aviation domain (e.g.~identifying pilot versus controller in communications), specifically,
+a pretrained BERT model was used and fine-tuned on problem specific data and compared to other models that performed
+well at speaker and role identification tasks in general \cite{guo_comparative_2022}.
 \section{Methodology}
 \subsection{Data Source(s)}
+There are four datasets used in this work. All are described in detail below. All data samples are text-based and are mined
+from transcripts intended for Automatic Speech Recognition (ASR) tasks.
+
+\textbf{Air Traffic Control Complete}. Hereafter this dataset will be referred to as ATCC. This data was collected from three major
+airports in the United States, specifically, Boston Logan, Washington National, and Dallas-Fort Worth airports. The data is intended
+to be used primarily for Automatic Speech Recognition tasks, therefore the data in this work was mined from the transcripts of those
+data samples \cite{godfrey_air_1994}.
+
+\textbf{Air Traffic Control 2}. This dataset is referred to as ATCO2 throughout the rest of this work. This data was collected as part
+of an open source initiative to collect and label audio data from airports around the world. The data was originally used for
+acoustic-based NLP tasks, such as ASR, in the aviation/aerospace domain(s). Similarly to ATCC, the data used in this work from ATCO2
+is mined from the transcripts of the samples in the dataset \cite{szoke_detecting_2021}.
+
+\textbf{Air Traffic Control Simulation}. This dataset is referred to as ATCOSIM throughout the rest of this thesis. The data in this
+corpus was collected at an air-traffic control training and simulation center used for teaching and evaluating air-traffic control
+and air-traffic management concepts. As with the other datasets in this work, this dataset was created for acoustic and audio-based
+analysis and machine learning tasks such as ASR among others. Likewise, the data used in this work is mined from the transcripts
+of the samples in the dataset \cite{hofbauer_atcosim_2008}.
+
+\textbf{Czech ATC Corpus}. This dataset is referred to throughout the rest of this work as the ZCUATC corpus/dataset. The data in this
+corpus was recorded at an airport in the Czech Republic and subsequently transcribed. This dataset differs the most from the other
+datasets used in this work due to the location in which it was collected and transcribed, however a majority of the samples, if not all
+of them, should conform the the ICAO standards of aviation communications, which is why it remains in this work. As with the other
+audio-based corpora, the data used in this thesis was derived from the transcripts that correspond to the audio data in the ZCUATC
+corpus \cite{smidl_air_2019}.
+
 \subsection{Data Analysis}
+% sequence lengths, unique tokens, total number of tokens, total number of samples, breakdown by each dataset
 \subsection{Preprocessing}
+% explain and justify what is removed from the text, how, and why
+% present and explain any relevant algorithms where necessary
+\subsection{Tokenization}
+% present and explain the tokenization algorithms used (possibly which models they correspond to)
+\section{Discussion}
 \section{Results}
-    \section{Discussion}
 \section{Conclusion}

### 2023-08-29 thesis/MSECE_thesis.tex
Refactor name property in data modules. Finish ZCU datamodule.

--- 2023-01-06 thesis/MSECE_thesis.tex
@@
 \usepackage{sectsty}
+\usepackage{booktabs}
 \usepackage[numbib]{tocbibind}
@@
 \newpage
-    \doublespacing
+\doublespacing{}
 
-    \section{Abstract}
-    \section{Introduction}
+% \section{Abstract}
+
 \section{Literature Review}
-    \section{Methodology}
-        \subsection{Data Source(s)}
+% from proposal; revise and add to this as more references are used
+% transformer history
+The transformer neural network architecture was proposed in 2017 for Neural Machine Translation tasks and immediately achieved
+state-of-the-art (28.4 BLEU on WMT 2014 English-to-German; 41.8 BLEU on WMT English-to-French datasets)\cite{vaswani_attention_2017}. Transformer
+architectures have been found to be extremely effective at learning representations and understandings of languages to predict token probabilities as
+opposed to transforming one language into another\cite{devlin_bert_2019,liu_roberta_2019}. Transformers have also been found to be very effective at
+other NLP-related tasks such as prompt completion and sentiment analysis among others (i.e.~auto-regressive and sequence classification tasks,
+respectively)\cite{lewis_bart_2019,radford_improving_2018}.
+
+% ASR context (suggest some need for semi-supervised approaches and therefore LMs)
+End-to-end generative models for automatic speech recognition models have made significant progress in recent years with current state-of-the-art
+models achieving WERs as low as 2\% on LibriSpeech test sets\cite{han_contextnet_2020,kriman_quartznet_2020,baevski_wav2vec_2020,li_jasper_2019}.
+This has led to the development of ASR models for the aviation domain, specifically, in air traffic control communications\cite{badrinath_automatic_2022,smidl_air_2019,zuluaga-gomez_automatic_2020,srinivasamurthy_semi-supervised_2017}.
+However, due to the lack of
+transcribed data in the aviation domain\cite{zuluaga-gomez_automatic_2020,srinivasamurthy_semi-supervised_2017,badrinath_automatic_2022,smidl_air_2019}
+ASR models maintain relatively high WERs compared to their counterparts in the more generalized ASR domain\cite{zuluaga-gomez_automatic_2020,badrinath_automatic_2022}.
+Transfer learning has even yielded limited results in this domain (depending on model
+architecture and dataset quality)\cite{badrinath_automatic_2022,zuluaga-gomez_automatic_2020}.
+
+% Unsupervised/semi-supervised in general and LMs commonly used with them
+Unsupervised and semi-supervised methodologies have become popular recently in attempts to address limited data availability
+and develop new approaches towards modeling human speech (notably, wav2vec has achieved state-of-the-art performance with
+very little training data)\cite{baevski_wav2vec_2020,badrinath_automatic_2022,srinivasamurthy_semi-supervised_2017,zuluaga-gomez_contextual_2021}.
+Language models are an integral part semi-supervised learning. They are used to obtain a certainty score (or uncertainty
+score, as the case may be) for the predicted text, these are usually either word lattices or N-gram models\cite{badrinath_automatic_2022,srinivasamurthy_semi-supervised_2017,zuluaga-gomez_contextual_2021}.
+While these have proven to be adequate for most self-supervised learning tasks, they are hardly state-of-the-art.
+Wav2vec, the current state-of-the-art unsupervised model (and possibly top performing overall), uses a contrastive process
+between convolutional feature extraction and transformer context representations\cite{baevski_wav2vec_2020}.
+
+% LMs and NLP in aero. domain
+Various natural language processing methods have been applied to the aviation domain to help deal with miscommunications
+and try to mitigate safety incidents\cite{ragnarsdottir_language_2003,tanguy_natural_2016,madeira_machine_2021}.
+Some machine learning approaches have been implemented to analyze the text in aviation incident and safety reports to predict
+contributing factors and topic models\cite{tanguy_natural_2016,madeira_machine_2021}. An ASR system with NLP post-processing
+has also been proposed to analyze Air-Traffic communications and condense significant features (e.g.~weather, runway, and
+heading info) into an XML language structure\cite{ragnarsdottir_language_2003}. Transformer language models such as BERT,
+RoBERTa, and DeBERTa have been applied to notice to airmen (NOTAM) messages to perform named entity recognition (NER) tasks,
+translation tasks (between notations; e.g.~NOTAM notation to Airlang to make parsing tasks easier) and reduce the workload
+for pilots during long-haul flights\cite{arnold_knowledge_2022}. Transformer models have also been used for speaker role
+identification tasks in the aviation domain (e.g.~identifying pilot versus controller in communications), specifically,
+a pretrained BERT model was used and fine-tuned on problem specific data and compared to other models that performed
+well at speaker and role identification tasks in general\cite{guo_comparative_2022}.
+
+\section{Data Source(s)}
+There are four datasets used in this work. The datasets are primarily speech corpora intended either for automatic speech recognition
+research in the domain of aviation communications, linguistics research in aviation english, or both. As a result, the text-based data
+samples that are used to train the language models in this work are transcripts of aural speech. The datasets used are described in more
+detail below. All recordings and transcripts are in english, although there are multiple instances of foreign words occuring
+
+\subsection{Air Traffic Control Complete}
+The Air Traffic Control Complete (ATCC) corpus is a speech corpus consisting of audio recordings with corresponding transcriptions.
+ATCC is a collection of three smaller corpora recorded at Dallas-Fort Worth, Logan International, and Washington National airports and
+transcribed by subject matter experts (usually controllers and/or pilots) familiar with the terminology specific to their respective
+areas\cite{godfrey_air_1994}.
+
+\subsection{ATCO2}
+The ATCO2 dataset is a speech corpus that consists of audio communications at Prague and Brno airports, in Czechia, along with corresponding
+transcriptions. The speech recordings and transcripts are crowd sourced from volunteers\cite{szoke_detecting_2021}.
+
+\subsection{Air Traffic Control Simulation}
+The Air Traffic Control Simulation (ATCOSIM) corpus is a speech corpus consisting of audio recordings and corresponding transcriptions.
+The data was recorded at a air traffic control simulation center in Bretigny-sur-Orge, France, specifically the EUROCONTROL Experimental Centre.
+In this data, only the controllers' voices are included and therefore the transcripts only include the controller's side of each interaction.
+The audio data was transcribed by one person, trained according to the guidelines and formatting requirements of the corpus. After all data was
+transcribed, it was reviewed and corrected and any remaining problems were reviewed by an operational air traffic controller and
+resolved\cite{hofbauer_atcosim_2008}.
+
+\subsection{ZCU CZ ATC Corpus}
+The ZCU CZ ATC corpus consists of audio recordings and corresponding transcripts in the Czech airspace. Both controller and pilot sides of the
+communications are included in this data. Annotations were created by experienced transcribers/annotators and labeled samples were randomly selected
+for review during the annoation process. After the dataset was completely labeled, all samples were checked, corrected, and
+unified\cite{smidl_air_2019}.
+
 \subsection{Data Analysis}
+% sequence lengths, unique tokens, total number of tokens, total number of samples, breakdown by each dataset
+\begin{table}
+    \centering
+    \begin{tabular}{l c c c c}
+        Corpus                         & Samples & Mean Sequence Length & Tokens  & Unique Tokens \\
+        \toprule
+        Air Traffic Control Complete   & 29,862  & 11.90                & 355,476 & 2,210         \\
+        ATCO2                          & 874     & 12.46                & 10,886  & 855           \\
+        Air Traffic Control Simulation & 9,544   & 11.29                & 107,759 & 824           \\
+        ZCU CZ ATC Corpus              & 14,435  & 10.32                & 149,031 & 3106          \\
+        \bottomrule
+        Total                          & 54,715  & 11.39                & 623,152 & 5343          \\
+    \end{tabular}
+    \caption[short]{title}
+\end{table}
+
 \subsection{Preprocessing}
-    \section{Results}
-    \section{Discussion}
-    \section{Conclusion}
+% explain and justify what is removed from the text, how, and why
+% present and explain any relevant algorithms where necessary
+
+\subsection{Tokenization}
+% present and explain the tokenization algorithms used (possibly which models they correspond to)
 

### 2023-08-29 source/data/__init__.py

@@ -3,7 +3,7 @@ from .utils import atccutils
 from .atccomplete import ATCCompleteData
 from .atco2sim import ATCO2SimData
 from .atcosim import ATCOSimData
-from .czechdataset import ZCUATCDataset
+from .czechdataset import ZCUCZATCDataset
 
 __all__ = [
     "Data",
@@ -13,6 +13,6 @@ __all__ = [
     "ATCCompleteData",
     "ATCO2SimData",
     "ATCOSimData",
-    "ZCUATCDataset",
+    "ZCUCZATCDataset",
     "get_train_test_split",
 ]

### 2023-08-29 source/data/atccomplete.py

@@ -53,7 +53,9 @@ class ATCCompleteData(Data):
     # _transcript_glob: List[str]
 
     def __init__(self, data_root: str, **kwargs):
-        super(ATCCompleteData, self).__init__(**kwargs)
+        super(ATCCompleteData, self).__init__(
+            dataset_name="Air Traffic Control Complete", **kwargs
+        )
         transcript_glob_string = os.path.join(data_root, "**/data/transcripts/*.txt")
         self._transcript_glob = glob.glob(transcript_glob_string, recursive=True)
 
@@ -105,7 +107,3 @@ class ATCCompleteData(Data):
         # save manifest data to class attribute before returning
         self.data = data
         return data
-
-    @property
-    def name(self):
-        return "Air Traffic Control Complete"

### 2023-08-29 source/data/atco2sim.py

@@ -27,7 +27,7 @@ class ATCO2SimData(Data):
     ]
 
     def __init__(self, data_root: str, **kwargs):
-        super(ATCO2SimData, self).__init__(**kwargs)
+        super(ATCO2SimData, self).__init__(dataset_name="ATCO2", **kwargs)
 
         self.transcripts = glob.glob(os.path.join(data_root, "DATA/*.xml"))
         assert len(self.transcripts) != 0
@@ -121,7 +121,3 @@ class ATCO2SimData(Data):
 
         self.data = data
         return data
-
-    @property
-    def name(self) -> str:
-        return "ATCO2"

### 2023-08-29 source/data/atcosim.py

@@ -24,7 +24,7 @@ class ATCOSimData(Data):
     transcription_corrections = [("kil0", "kilo"), ("ai", "air"), ("airr", "air")]
 
     def __init__(self, data_root: str, **kwargs):
-        super(ATCOSimData, self).__init__(**kwargs)
+        super(ATCOSimData, self).__init__(dataset_name="ATCOSIM", **kwargs)
         # TODO: update regex
         self.text_glob = glob.glob(
             os.path.join(data_root, "txtdata/**/*.txt"), recursive=True
@@ -77,7 +77,3 @@ class ATCOSimData(Data):
 
         self.data = data
         return data
-
-    @property
-    def name(self):
-        return "ATCO"

### 2023-08-29 source/data/czechdataset.py

@@ -6,7 +6,7 @@ from data import Data
 from xml.etree import ElementTree
 
 
-class ZCUATCDataset(Data):
+class ZCUCZATCDataset(Data):
     """
     Data is organized into `trs` files (follows an XML standard). Details in `~parse_transcripts` function.
 
@@ -26,9 +26,9 @@ class ZCUATCDataset(Data):
     """
 
     def __init__(self, data_root: str, **kwargs):
-        super(ZCUATCDataset, self).__init__(**kwargs)
-
+        super(ZCUCZATCDataset, self).__init__(dataset_name="ZCU CZ ATC", **kwargs)
         self.transcript_paths = glob.glob(os.path.join(data_root, "*.trs"))
+
         assert (
             len(self.transcript_paths) != 0
         ), f"Cannot find transcripts in data_root: {data_root}"
@@ -53,36 +53,44 @@ class ZCUATCDataset(Data):
         data = []
 
         annotation_tag = re.compile(r"(\[[A-Za-z_\|\?]+\])")
+        nonstandard_pronunciation = re.compile(
+            r"\((?P<effective_transcript>[\w\d\s\+]+)\((?P<phonetic_transcript>[\w\d\s]+)\)\)"
+        )
 
-        for path in self.transcript_paths:
+        for transcript_path in self.transcript_paths:
+            doc_data = []
             try:
                 # root node: <Trans>
-                document = ElementTree.parse(path).getroot()
+                document = ElementTree.parse(transcript_path).getroot()
             except ElementTree.ParseError as e:
                 # because not all transcripts conform to the given format
-                with open(path, "a") as dumb_path:
+                with open(transcript_path, "a") as dumb_path:
                     # there is a single file that is missing the closing tags on all nodes
                     # hard-coding the fix here, so I can forget about it
                     dumb_path.write("</Turn>\n</Section>\n</Episode>\n</Trans>\n")
                 # parse the document again
-                document = ElementTree.parse(path).getroot()
+                document = ElementTree.parse(transcript_path).getroot()
 
             # find <Sync> tags, extract text, reformat/clean
             for sync_node in document.iterfind(
                 ".//Sync"
             ):  # searches all subelements for Sync nodes
-                assert sync_node.tag == "Sync"
-                assert len(sync_node.tail) != 0
-
                 text = annotation_tag.sub("", sync_node.tail).strip()
+
+                for nonstandard_match in nonstandard_pronunciation.finditer(text):
+                    if nonstandard_match is not None:
+                        text = text.replace(
[…]

### 2023-08-29 source/data/data.py

@@ -347,6 +347,33 @@ class Data(IterableDataset):
 
         return len(unique_tokens)
 
+    @property
+    def average_sequence_length(self) -> int:
+        """
+        Average number of tokens per sequence in the dataset. Calculated as
+        the sum of the sequence lengths divided by the number of samples.
+        """
+        sum_of_seq_lens = 0
+
+        for item in self.data:
+            sum_of_seq_lens += len(item.split(" "))
+
+        return sum_of_seq_lens / self.num_samples
+
+    def summary(self, printout=True):
+        """ """
+        dataset_summary = f"""
+Name: {self.name}
+Samples: {self.num_samples}
+Mean Sequence Length: {self.average_sequence_length}
+Number of Tokens: {self.total_tokens}
+Number of Unique Tokens: {self.unique_tokens}
+"""
+        if printout:
+            print(dataset_summary)
+        else:
+            return dataset_summary
+
 
 def get_train_test_split(
     data: Data,

### 2023-08-29 source/main.py

@@ -4,7 +4,14 @@ import os
 from typing import *
 
 import matplotlib.pyplot as plt
-from data import ATCCompleteData, ATCO2SimData, ATCOSimData, Data, get_train_test_split
+from data import (
+    ATCCompleteData,
+    ATCO2SimData,
+    ATCOSimData,
+    ZCUCZATCDataset,
+    Data,
+    get_train_test_split,
+)
 from models import Model
 
 # root dataset paths corresponding to data analysis classes
@@ -15,11 +22,15 @@ datasets: Dict[str, Data] = {
     "/home/students/vandebra/programming/thesis_data/ATCO2-ASRdataset-v1_beta": ATCO2SimData,
     # TODO: this dataset has strange formats for transcriptions, need to do a lot of
     # work to clean and reformat them. Disabling this dataset for now
-    # "/home/students/vandebra/programming/thesis_data/ZCU_CZ_ATC": ZCUATCDataset,
+    "/home/students/vandebra/programming/thesis_data/ZCU_CZ_ATC": ZCUCZATCDataset,
 }
 
 
 def parse_datasets():
+    # collection of dataset stats. using a dictionary so it's easier to dump to JSON or YAML later
+    dataset_info = {"dataset_info": []}
+    data_objects: List[Data] = []
+
     # initializes each implementing class with its data which is specified by `root_path`
     # see `datasets` and implementing classes for more details
     for root_path, data_class in datasets.items():
@@ -32,11 +43,7 @@ def parse_datasets():
         data_analysis.parse_transcripts()
         print("Done")
 
-        # printing some stats to stdout just to confirm things worked correctly
-        print(
-            f"Found {data_analysis.num_samples} samples with {data_analysis.unique_tokens} "
-            f"unique tokens and {data_analysis.total_tokens} tokens in total for '{data_analysis.name}'"
-        )
+        data_analysis.summary()
         # dump corpora to a standard corpus.txt style file, we'll have one for each dataset and
         # one for everything combined together
         corpus_filename = f"{data_analysis.name.replace(' ', '_')}.txt"
@@ -55,6 +62,9 @@ def parse_datasets():
     # concatenate everything to first object
     for o in data_objects[1:]:
         data_objects[0].concat(o)
+
+    data_objects[0].dataset_name = "All"
+    data_objects[0].summary()
 
     # split data into train and test
     print("Generating train/test split", end="...")
[…]

### 2023-09-11 thesis/MSECE_thesis.tex
Finish data processing for ZCU ATC.

@@
 
-\section{Data Source(s)}
+\section{Data Source(s)}\label{sec:data_source}
 There are four datasets used in this work. The datasets are primarily speech corpora intended either for automatic speech recognition
@@
 
-\subsection{Data Analysis}
+\section{Data Analysis}
 % sequence lengths, unique tokens, total number of tokens, total number of samples, breakdown by each dataset
@@
     \centering
-    \begin{tabular}{l c c c c}
-        Corpus                         & Samples & Mean Sequence Length & Tokens  & Unique Tokens \\
+    \begin{tabular}{l r r r r}
+        Corpus                         & Samples & Mean Sequence Length & Unique Tokens & Tokens  \\
         \toprule
-        Air Traffic Control Complete   & 29,862  & 11.90                & 355,476 & 2,210         \\
-        ATCO2                          & 874     & 12.46                & 10,886  & 855           \\
-        Air Traffic Control Simulation & 9,544   & 11.29                & 107,759 & 824           \\
-        ZCU CZ ATC Corpus              & 14,435  & 10.32                & 149,031 & 3106          \\
+        Air Traffic Control Complete   & 29,862  & 11.90                & 2,210         & 355,476 \\
+        ATCO2                          & 874     & 12.46                & 855           & 10,886  \\
+        Air Traffic Control Simulation & 9,544   & 11.29                & 824           & 107,759 \\
+        ZCU CZ ATC Corpus              & 14,435  & 10.32                & 3,106         & 149,031 \\
         \bottomrule
-        Total                          & 54,715  & 11.39                & 623,152 & 5343          \\
+        Total                          & 54,715  & 11.39                & 5,343         & 623,152 \\
     \end{tabular}
-    \caption[short]{title}
+    \caption{Data statistics by corpus, including statistics across all corpora.}
 \end{table}
 
-\subsection{Preprocessing}
+\section{Data Processing}
+All of the corpora listed above (see~\ref{sec:data_source}) are created at different times, for different purposes, and by different authors.
+As a result, the data in these corpora all have different formats and need to be processed into a common format to be used together, with the goal
+being to essentially create a super-corpus that includes all the corpora listed above.
+
+All data was processed in Python using primarily built-in functions/modules\footnote{The \lstinline|re| and \lstinline|xml| modules were also used although
+    these are built into most Python distributions by default.}. The text data is extracted from the corpus transcripts i.e.~Lisp in ATCC, XML in
+ATCO2 and ZCU CZ ATC, and plain text in ATCOSIM, the transcriber annotations are removed where necessary\footnote{A copy of the original data is
+    made, then the annotations are removed from the copy.}, and the resulting text is written in the standard corpus format (an ASCII text file with
+one line in the file corresponding to one sample from the corpus) to a file.
+
+\subsection{Data Preprocessing}
 % explain and justify what is removed from the text, how, and why

### 2023-09-11 source/data/atccomplete.py

@@ -2,8 +2,7 @@ import os
 import glob
 import re
 from typing import *
-from data import Data
-from data import atccutils
+from data import Data, atccutils
 
 
 class ATCCompleteData(Data):

### 2023-09-11 source/data/atco2sim.py

@@ -4,7 +4,6 @@ import re
 from data import Data
 from typing import *
 from xml.etree import ElementTree
-from pprint import pprint
 
 
 class ATCO2SimData(Data):

### 2023-09-11 source/data/czechdataset.py

@@ -4,6 +4,7 @@ import re
 from typing import *
 from data import Data
 from xml.etree import ElementTree
+from tokenizers.normalizers import NFD
 
 
 class ZCUCZATCDataset(Data):
@@ -56,6 +57,7 @@ class ZCUCZATCDataset(Data):
         nonstandard_pronunciation = re.compile(
             r"\((?P<effective_transcript>[\w\d\s\+]+)\((?P<phonetic_transcript>[\w\d\s]+)\)\)"
         )
+        normalizer = NFD()
 
         for transcript_path in self.transcript_paths:
             doc_data = []
@@ -75,7 +77,7 @@ class ZCUCZATCDataset(Data):
             for sync_node in document.iterfind(
                 ".//Sync"
             ):  # searches all subelements for Sync nodes
-                text = annotation_tag.sub("", sync_node.tail).strip()
+                text: str = annotation_tag.sub("", sync_node.tail).strip()
 
                 for nonstandard_match in nonstandard_pronunciation.finditer(text):
                     if nonstandard_match is not None:
@@ -83,9 +85,36 @@ class ZCUCZATCDataset(Data):
                             nonstandard_match.group(0), nonstandard_match.group(1)
                         )
 
+                words = text.split()
+
+                # for word in words:
+                # if "+" in word:
+                #     words.remove(word)
+
+                text = " ".join(words).lower()
+
                 # ".." corresponds to silent segments, "" can occur when the transcript is made up
                 # of only transcriber annotations
                 if text != ".." and text != "":
+                    text = text.translate(
+                        str.maketrans(
+                            {
+                                "0": "zero",
+                                "1": "one",
+                                "2": "two",
+                                "3": "tree",
+                                "4": "four",
+                                "5": "fife",
+                                "6": "six",
+                                "7": "seven",
+                                "8": "eight",
+                                "9": "niner",
+                                ".": "dot",
+                            }
+                        )
+                    )
+
+                    text = normalizer.normalize_str(text)
                     doc_data.append(text.strip())
 
             # check to make sure useful samples were found in the document

### 2023-09-11 source/data/data.py

@@ -24,22 +24,10 @@ class Data(IterableDataset):
 
     This class should be extended and the following methods/properties implemented for each dataset:
     * `parse_transcripts`
-    * `name`
 
     Attributes:
     -----------
-    `_manifest_data`: list of dictionary objects. Each object corresponds to one data sample
-    and typically contains the following metadata:
-    * `audio_filepath` (required) - path to the audio data (input data). Type: `str`,
-    absolute file path, conforms to `os.PathLike`
-    * `duration` (required) - duration, in seconds, of the audio data. Type: `float`
-    * `text` (required) - transcript of the audio data (label/ground truth). Type: `str`
-    * `offset` - if more than one sample is present in a single audio file, this field
-    specifies its offset i.e. start time in the audio file. Type: `float`
-
     `_random`: numpy seeded RNG instance
-
-    `_normalized`: bool indicating whether samples in the dataset have been normalized/preprocessed
     """
 
     _random: np.random.Generator
@@ -55,7 +43,6 @@ class Data(IterableDataset):
         self.data: List[str] = []
         # create random number generator sequence with specified seed, if applicable
         Data._random: np.random.Generator = np.random.default_rng(random_seed)
-        self.normalized: bool = False
 
     def __iter__(self):
         assert len(self.data) != 0
@@ -195,7 +182,6 @@ class Data(IterableDataset):
         token_freqs = {}
 
         for sample in self.data:
-            sample = sample["text"]
             for token in sample.split():
                 if token in token_freqs.keys():
                     # increment occurences if there is already an entry
@@ -273,6 +259,41 @@ class Data(IterableDataset):
             "total_tokens": self.total_tokens,
             "unique_tokens": self.unique_tokens,
         }
+
+    def token_ratio(self, as_tuple=False) -> Union[float, Tuple[int, int]]:
+        """
+        The ratio of unique tokens to the total number of tokens.
+
+        :param as_tuple: Defaults to False. If True, returns a tuple
+        of the ratio of unique tokens to total tokens in the order [unique, total].
+        """
+        unique_tokens = self.unique_tokens
+        total_tokens = self.total_tokens
+
+        if as_tuple:
+            return tuple([unique_tokens, total_tokens])
+        else:
+            return unique_tokens / total_tokens
+
+    def summary(self, printout=True):
+        """
[…]

### 2023-09-12 source/data/czechdataset.py
Add helper functon for sequence lengths, properties for mean and std of sequence lengths.

@@ -86,11 +86,6 @@ class ZCUCZATCDataset(Data):
                         )
 
                 words = text.split()
-
-                # for word in words:
-                # if "+" in word:
-                #     words.remove(word)
-
                 text = " ".join(words).lower()
 
                 # ".." corresponds to silent segments, "" can occur when the transcript is made up
@@ -114,6 +109,7 @@ class ZCUCZATCDataset(Data):
                         )
                     )
 
+                    # unicode string normalization, gets rid of some but not all foreign and uncommon characters
                     text = normalizer.normalize_str(text)
                     doc_data.append(text.strip())
 

### 2023-09-12 source/data/data.py

@@ -41,6 +41,7 @@ class Data(IterableDataset):
         """
         self.dataset_name: str = dataset_name
         self.data: List[str] = []
+        self.seq_lens = []
         # create random number generator sequence with specified seed, if applicable
         Data._random: np.random.Generator = np.random.default_rng(random_seed)
 
@@ -156,48 +157,6 @@ class Data(IterableDataset):
 
         return histogram
 
-    def token_freq_analysis(self, normalize=False) -> Dict[str, Union[int, List]]:
-        """
-        Perform a token frequency analysis on the dataset (number of occurrences of each token throughout the dataset).
-
-        Arguments:
-        ----------
-        `normalize`: (optional)`bool`, whether to normalize values such that all frequencies add to 1.
-
-        Returns:
-        --------
-        `token_freqs`: `dict` with tokens and number of occurrences of those tokens throughout the dataset.
-        If `normalize` is set to `True` a dictionary with tokens corresponding to a list is returned e.g.
-        ```
-        {
-            "token": [24, 0.0486] # token with corresponding occurences and frequencies
-        }
-        ```
-
-        """
-        if len(self.data) == 0:
-            self.parse_transcripts()
-
-        # tokens corresponding to occurences/frequencies
-        token_freqs = {}
-
-        for sample in self.data:
-            for token in sample.split():
-                if token in token_freqs.keys():
-                    # increment occurences if there is already an entry
-                    token_freqs[token] += 1
-                else:
-                    # add entry if one does not exist already
-                    token_freqs[token] = 1
-
-        if normalize:
-            # convert from token occureces to frequencies: (# of token occurences / # of tokens)
-            num_tokens = len(token_freqs)
-            for token, freq in token_freqs.items():
-                token_freqs[token] = [freq, float(freq) / num_tokens]
-
-        return token_freqs
-
     def dump_corpus(
         self, outfile: str, make_dirs: bool = True, return_list: bool = False
     ) -> Union[None, List[str]]:
@@ -286,6 +245,7 @@ class Data(IterableDataset):
 Name: {self.name}
 Samples: {self.num_samples}
 Mean Sequence Length: {self.average_sequence_length}
+Std of Sequence Length: {self.std_sequence_length}
 Number of Tokens: {self.total_tokens}
 Number of Unique Tokens: {self.unique_tokens}
[…]

### 2023-09-12 thesis/MSECE_thesis.tex
Add outlier removal function.

@@
 
-% \section{Abstract}
+\section{Abstract}\label{sec:abstract}
 
-\section{Literature Review}
+\section{Introduction}\label{sec:introduction}
+
+\section{Literature Review}\label{sec:lit_review}
 % from proposal; revise and add to this as more references are used
@@
 
-\section{Data Source(s)}\label{sec:data_source}
+\section{Data Sources}\label{sec:data_source}
 There are four datasets used in this work. The datasets are primarily speech corpora intended either for automatic speech recognition
@@
 
-\subsection{Air Traffic Control Complete}
+\subsection{Air Traffic Control Complete}\label{sec:atcc}
 The Air Traffic Control Complete (ATCC) corpus is a speech corpus consisting of audio recordings with corresponding transcriptions.
@@
 
-\subsection{ATCO2}
+\subsection{ATCO2}\label{sec:atco2}
 The ATCO2 dataset is a speech corpus that consists of audio communications at Prague and Brno airports, in Czechia, along with corresponding
@@
 
-\subsection{Air Traffic Control Simulation}
+\subsection{Air Traffic Control Simulation}\label{sec:atcosim}
 The Air Traffic Control Simulation (ATCOSIM) corpus is a speech corpus consisting of audio recordings and corresponding transcriptions.
@@
 
-\subsection{ZCU CZ ATC Corpus}
+\subsection{ZCU CZ ATC Corpus}\label{sec:zcu_atc}
 The ZCU CZ ATC corpus consists of audio recordings and corresponding transcripts in the Czech airspace. Both controller and pilot sides of the
@@
 
-\section{Data Analysis}
-% sequence lengths, unique tokens, total number of tokens, total number of samples, breakdown by each dataset
-\begin{table}
-    \centering
-    \begin{tabular}{l r r r r}
-        Corpus                         & Samples & Mean Sequence Length & Unique Tokens & Tokens  \\
-        \toprule
-        Air Traffic Control Complete   & 29,862  & 11.90                & 2,210         & 355,476 \\
-        ATCO2                          & 874     & 12.46                & 855           & 10,886  \\
-        Air Traffic Control Simulation & 9,544   & 11.29                & 824           & 107,759 \\
-        ZCU CZ ATC Corpus              & 14,435  & 10.32                & 3,106         & 149,031 \\
-        \bottomrule
-        Total                          & 54,715  & 11.39                & 5,343         & 623,152 \\
-    \end{tabular}
-    \caption{Data statistics by corpus, including statistics across all corpora.}
-\end{table}
-
-\section{Data Processing}
+\section{Data Processing}\label{sec:data_processing}
 All of the corpora listed above (see~\ref{sec:data_source}) are created at different times, for different purposes, and by different authors.
@@
 one line in the file corresponding to one sample from the corpus) to a file.
+
+\section{Data Analysis}\label{sec:data_analysis}
+% sequence lengths, unique tokens, total number of tokens, total number of samples, breakdown by each dataset
+Relevant corpora statistics were observed or calculated are are presented in Table~\ref{table:corpora_stats}. This includes the number of samples in
+the corpus, mean sequence length per sample in the corpus, the number of unique tokens in the corpus, and the total number of tokens in the corpus.
+The same statistics were also calculated across corpora, since all four corpora will be combined for training and testing the language and speech
+recognition models. Note that this is based on the raw data after loading and unifying the corpora, prior to preprocessing for model training/testing
+and outlier removal.
+
+\begin{table}[h!]
+    \centering
+    \begin{tabular}{l r r r r}
+        \toprule
+        \textbf{Corpus} & \textbf{Samples} & \textbf{Mean Sequence Length (\(\pm \sigma\))} & \textbf{Unique Tokens} & \textbf{Tokens} \\
+        \midrule
+        ATCC            & 29,862           & \(11.90 \pm 7.22\)                             & 2,210                  & 355,476         \\
+        ATCO2           & 874              & \(12.46 \pm 5.93\)                             & 855                    & 10,886          \\
+        ATCOSIM         & 9,544            & \(11.29 \pm 4.18\)                             & 824                    & 107,759         \\
+        ZCU CZ ATC      & 14,435           & \(10.32 \pm 5.50\)                             & 3,106                  & 149,031         \\
+        \midrule
+        Total           & 54,715           & \(11.39 \pm 7.22\)                             & 5,343                  & 623,152         \\
+        \bottomrule
+    \end{tabular}
+    \label{table:corpora_stats}
+    \caption{The number of samples, mean sequence length, number of unique tokens, and total number of tokens by corpus, including the cumulative
+        total across all corpora.}
+\end{table}
+
+The ATCC and ZCU CZ ATC corpora make up a significant portion of the data, ATCC alone makes up more than half (approximately 54\%) of the combined
+data, and ATCC and ZCU CZ ATC together make up aprroximately 81\% of the data. The sequence lengths of the samples are fairly consistent across
+corpora with at most about an 18\% percent difference between sequence lengths. The number of unique tokens does not scale additively to the
+combined corpora, which is as expected. Considering that aviation English generally conforms to and follows ICAO phraseology standards we would expect
+that many terms would be shared across corpora with the exception of geographic/location specific terms, such as city, landmark, airport, and airline
+names, etc.~that typically vary across regions.
+
+\subsection{Outlier Removal}\label{sec:outlier_removal}
+This section analyzes the same corpora as above after outliers in the data have been removed, if any are present. Outlier removal was performed using
+the interquartile range (IQR) outlier removal method. IQR is defined as the difference between the medians of the third and first quartiles of the
+distribution: \(\mathbf{IQR} = \mathbf{Q3} - \mathbf{Q1}\), where \(\mathbf{Q3}\) and \(\mathbf{Q1}\) are the medians of the third and first quartiles,
+respectively. Outliers are then calculated as samples whose sequence lengths lie below \(\mathbf{Q1} - 1.5\mathbf{IQR}\) and above
+\(\mathbf{Q3} + 1.5\mathbf{IQR}\).
 

### 2023-09-12 source/data/data.py

@@ -8,14 +8,14 @@ import matplotlib.collections
 import matplotlib.figure
 import matplotlib.pyplot as plt
 import numpy as np
+import pandas as pd
 from numpy.random import default_rng
-from torch.utils.data import IterableDataset
+from torch.utils.data import IterableDataset, DataLoader
 from transformers import (
     PreTrainedTokenizer,
     DataCollatorForLanguageModeling,
 )
 from pytorch_lightning import LightningDataModule
-from torch.utils.data import DataLoader
 
 
 class Data(IterableDataset):
@@ -199,6 +199,10 @@ class Data(IterableDataset):
         include data from child object. Also updates any relevant common metadata
         fields.
         """
+        if len(child_dataset.seq_lens) == 0:
+            child_dataset._get_sequence_lengths()
+
+        self.seq_lens.extend(child_dataset.seq_lens)
         self.data.extend(child_dataset.data)
         self.dataset_name = f"{self.name} + {child_dataset.name}"
 
@@ -254,6 +258,39 @@ Ratio of unique tokens to the total number of tokens: {self.token_ratio()}, {sel
             print(dataset_summary)
         else:
             return dataset_summary
+
+    def remove_outliers(self) -> None:
+        """
+        Perform IQR-based outlier removal based on sequence lengths of the data.
+        Samples with sequence lengths less than Q1 - 1.5 * IQR and greater than Q3 + 1.5 * IQR
+        will be removed from the data.
+
+        The data and seq_len instance variables are updated as a result of this.
+        """
+        # using a pandas data frame because it keeps everything aligned and provides easy-to-use and verified helper functions
+        data_frame = pd.DataFrame({"seq_len": self.seq_lens, "text": self.data})
+        # divide data into quartiles
+        categories = pd.qcut(data_frame["seq_len"], 4, labels=["q1", "q2", "q3", "q4"])
+
+        q1 = float(data_frame[categories == "q1"].median())
+        q3 = float(data_frame[categories == "q3"].median())
+        iqr = q3 - q1
+
+        lower_bound = q1 - 1.5 * iqr
+        upper_bound = q3 + 1.5 * iqr
+
+        # select regions below and above the upper and lower bounds, respectively
+        # take inverse to select region in-between
+        trimmed_data = data_frame[
+            ~(
+                (data_frame["seq_len"] < lower_bound)
+                | (data_frame["seq_len"] > upper_bound)
+            )
+        ]
+
[…]

### 2023-09-12 source/main.py

@@ -65,6 +65,7 @@ def parse_datasets():
 
     data_objects[0].dataset_name = "All"
     data_objects[0].summary()
+    data_objects[0].remove_outliers()
 
     # split data into train and test
     print("Generating train/test split", end="...")

### 2023-09-13 thesis/MSECE_thesis.tex
Add outlier trimmed data table. Fix bug in remove outlier function.

@@
 
+The corpora statistics after outlier removal are presented in Table \ref{tab:corpora_stats_outlier_removed}, below.
+
+\begin{table}[h!]
+    \centering
+    \begin{tabular}{l r r r r}
+        \toprule
+        \textbf{Corpus} & \textbf{Samples} & \textbf{Mean Sequence Length (\(\pm \sigma\))} & \textbf{Unique Tokens} & \textbf{Tokens} \\
+        \midrule
+        ATCC            & 28,335           & \(10.81 \pm 5.43\)                             & 2,006                  & 306,382         \\
+        ATCO2           & 848              & \(11.91 \pm 5.07\)                             & 812                    & 10,103          \\
+        ATCOSIM         & 9,438            & \(11.13 \pm 3.86\)                             & 787                    & 105,017         \\
+        ZCU CZ ATC      & 14,275           & \(9.87 \pm 5.24\)                              & 2,967                  & 140,855         \\
+        \midrule
+        Total           & 52,896           & \(10.63 \pm 5.15\)                             & 4,884                  & 562,357         \\
+        \bottomrule
+    \end{tabular}
+    \label{tab:corpora_stats_outlier_removed}
+    \caption{Individual and cumulative corpora statistics after outlier removal.}
+\end{table}
+
 \subsection{Data Preprocessing}

### 2023-09-13 source/data/data.py

@@ -219,8 +219,11 @@ class Data(IterableDataset):
         return {
             "dataset_name": self.name,
             "samples": self.num_samples,
+            "mean_sequence_length": self.average_sequence_length,
+            "std_sequence_length": self.std_sequence_length,
             "total_tokens": self.total_tokens,
             "unique_tokens": self.unique_tokens,
+            "token_ratio": self.token_ratio()
         }
 
     def token_ratio(self, as_tuple=False) -> Union[float, Tuple[int, int]]:
@@ -259,7 +262,7 @@ Ratio of unique tokens to the total number of tokens: {self.token_ratio()}, {sel
         else:
             return dataset_summary
 
-    def remove_outliers(self) -> None:
+    def remove_outliers(self) -> pd.DataFrame:
         """
         Perform IQR-based outlier removal based on sequence lengths of the data.
         Samples with sequence lengths less than Q1 - 1.5 * IQR and greater than Q3 + 1.5 * IQR
@@ -270,16 +273,21 @@ Ratio of unique tokens to the total number of tokens: {self.token_ratio()}, {sel
         # using a pandas data frame because it keeps everything aligned and provides easy-to-use and verified helper functions
         data_frame = pd.DataFrame({"seq_len": self.seq_lens, "text": self.data})
         # divide data into quartiles
-        categories = pd.qcut(data_frame["seq_len"], 4, labels=["q1", "q2", "q3", "q4"])
+        quartiles = pd.qcut(data_frame["seq_len"], 4, labels=["q1", "q2", "q3", "q4"])
 
-        q1 = float(data_frame[categories == "q1"].median())
-        q3 = float(data_frame[categories == "q3"].median())
+        q1 = data_frame[quartiles == "q1"]
+        q3 = data_frame[quartiles == "q3"]
+
+        q1 = q1["seq_len"].median()
+        q3 = q3["seq_len"].median()
         iqr = q3 - q1
 
         lower_bound = q1 - 1.5 * iqr
         upper_bound = q3 + 1.5 * iqr
 
-        # select regions below and above the upper and lower bounds, respectively
+        print(lower_bound, upper_bound)
+
+        # select regions below and above the lower and upper bounds, respectively
         # take inverse to select region in-between
         trimmed_data = data_frame[
             ~(
@@ -291,6 +299,9 @@ Ratio of unique tokens to the total number of tokens: {self.token_ratio()}, {sel
         # update data and sequence length variables
[…]

### 2023-09-13 source/main.py

@@ -20,15 +20,13 @@ datasets: Dict[str, Data] = {
     "/home/students/vandebra/programming/thesis_data/atc0_comp": ATCCompleteData,
     "/home/students/vandebra/programming/thesis_data/atcosim/": ATCOSimData,
     "/home/students/vandebra/programming/thesis_data/ATCO2-ASRdataset-v1_beta": ATCO2SimData,
-    # TODO: this dataset has strange formats for transcriptions, need to do a lot of
-    # work to clean and reformat them. Disabling this dataset for now
     "/home/students/vandebra/programming/thesis_data/ZCU_CZ_ATC": ZCUCZATCDataset,
 }
 
 
 def parse_datasets():
     # collection of dataset stats. using a dictionary so it's easier to dump to JSON or YAML later
-    dataset_info = {"dataset_info": []}
+    dataset_info = {"dataset_info": [], "trimmed_dataset_info": []}
     data_objects: List[Data] = []
 
     # initializes each implementing class with its data which is specified by `root_path`
@@ -52,11 +50,17 @@ def parse_datasets():
         print("Done")
 
         dataset_info["dataset_info"].append(data_analysis.dump_info())
+
+        data_analysis.remove_outliers()
+        data_analysis.summary()
+
+        dataset_info["trimmed_dataset_info"].append(data_analysis.dump_info())
+
         data_objects.append(data_analysis)
 
     # write stats to a json file
     os.makedirs("corpora", exist_ok=True)
-    with open("manifests/dataset_stats.json", "w", encoding="utf-8") as f:
+    with open("corpora/dataset_stats.json", "w", encoding="utf-8") as f:
         f.write(json.dumps(dataset_info, indent=1))
 
     # concatenate everything to first object
@@ -65,7 +69,6 @@ def parse_datasets():
 
     data_objects[0].dataset_name = "All"
     data_objects[0].summary()
-    data_objects[0].remove_outliers()
 
     # split data into train and test
     print("Generating train/test split", end="...")

### 2023-09-13 source/data/data.py
Add/change method to plot histogram of sequence lengths.

@@ -3,9 +3,6 @@ from pathlib import Path
 from typing import *
 from copy import deepcopy
 
-import matplotlib
-import matplotlib.collections
-import matplotlib.figure
 import matplotlib.pyplot as plt
 import numpy as np
 import pandas as pd
@@ -112,51 +109,6 @@ class Data(IterableDataset):
 
         return self.tokenized_data
 
-    def create_token_hist(
-        self,
-        token_counts: List[int] = [],
-    ) -> matplotlib.figure.Figure:
-        """
-        Calculates the number of utterances in each sample and generates a histogram.
-
-        Utterance counts are determined by splitting each transcript on whitespace and
-        calculating the length of the resulting list.
-
-        TODO: add flexibility for plot types.
-        TODO: issue with figures (figures/axes need to be cleared between runs)
-
-        Arguments:
-        ----------
-        `utterance_counts`: (Optional) `list` of `ints` for precalculated utterance counts.
-
-        `plot_type`: (Optional) `str` type of plot tool to use to create the histogram.
-        Can be either `"seaborn"` or `"matplotlib"`. Defaults to `"seaborn"`.
-
-        Returns:
-        --------
-        Either a `matplotlib.pyplot.Figure` or `seaborn.object.Plot` instance, depending on the value of `plot_type`.
-        """
-        # Clear figure and axes
-        plt.clf(), plt.cla()
-        # check if manifest data has been generated, parse transcripts and generate
-        # manifest data if not
-        if len(self.data) == 0:
-            self.parse_transcripts()
-
-        # check if utterance counts (optional arg) has been provided, calculate utterance
-        # counts from transcriptions
-        if len(token_counts) == 0:
-            for data in self.data:
-                words = data.split(" ")
-                token_counts.append(len(words))
-
-        histogram = plt.hist(token_counts)
-        plt.xlabel("Bins")
-        plt.ylabel("Token Counts")
-        plt.title(f"Tokens per Sample in {self.name}")
-
-        return histogram
-
     def dump_corpus(
         self, outfile: str, make_dirs: bool = True, return_list: bool = False
     ) -> Union[None, List[str]]:
@@ -223,7 +175,7 @@ class Data(IterableDataset):
             "std_sequence_length": self.std_sequence_length,
             "total_tokens": self.total_tokens,
[…]

### 2023-09-13 source/main.py

@@ -50,6 +50,7 @@ def parse_datasets():
         print("Done")
 
         dataset_info["dataset_info"].append(data_analysis.dump_info())
+        data_analysis.plot_histogram()
 
         data_analysis.remove_outliers()
         data_analysis.summary()

### 2023-09-13 source/data/atccomplete.py
remove extra whitespace from samples, rewrite outlier function to use LOF.

@@ -101,6 +101,7 @@ class ATCCompleteData(Data):
                     text = annotation.sub("", text)
 
                     if text.strip() != "":
+                        text = re.sub(r"\s{2,}", " ", text)
                         data.append(text.strip())
 
         # save manifest data to class attribute before returning

### 2023-09-13 source/data/atco2sim.py

@@ -116,6 +116,7 @@ class ATCO2SimData(Data):
                         text = text.replace(typo, correction)
 
                     if text != "":
+                        text = re.sub(r"\s{2,}", " ", text)
                         data.append(text.strip())
 
         self.data = data

### 2023-09-13 source/data/atcosim.py

@@ -73,6 +73,7 @@ class ATCOSimData(Data):
             # some transcripts are empty after removing transcriber
             # annotations
             if len(text) > 0:
+                text = re.sub(r"\s{2,}", " ", text)
                 data.append(text)
 
         self.data = data

### 2023-09-13 source/data/czechdataset.py

@@ -111,6 +111,7 @@ class ZCUCZATCDataset(Data):
 
                     # unicode string normalization, gets rid of some but not all foreign and uncommon characters
                     text = normalizer.normalize_str(text)
+                    text = re.sub(r"\s{2,}", " ", text)
                     doc_data.append(text.strip())
 
             # check to make sure useful samples were found in the document

### 2023-09-13 source/data/data.py

@@ -13,6 +13,7 @@ from transformers import (
     DataCollatorForLanguageModeling,
 )
 from pytorch_lightning import LightningDataModule
+from sklearn.neighbors import LocalOutlierFactor
 
 
 class Data(IterableDataset):
@@ -223,35 +224,24 @@ Ratio of unique tokens to the total number of tokens: {self.token_ratio()}, {sel
         The data and seq_len instance variables are updated as a result of this.
         """
         # using a pandas data frame because it keeps everything aligned and provides easy-to-use and verified helper functions
+        lof = LocalOutlierFactor(n_neighbors=35)
         data_frame = pd.DataFrame({"seq_len": self.seq_lens, "text": self.data})
 
-        q1 = data_frame["seq_len"].quantile(0.25)
-        q3 = data_frame["seq_len"].quantile(0.75)
-        iqr = q3 - q1
+        # fit to data and label outliers/inliers outlier = -1, inlier = 1
+        data_frame["outlier_label"] = lof.fit_predict(
+            # nxm matrix for API compatibility
+            data_frame["seq_len"]
+            .to_numpy()
+            .reshape(-1, 1)
+        )
 
-        lower_bound = q1 - 1.5 * iqr
-        upper_bound = q3 + 1.5 * iqr
-
-        print(lower_bound, upper_bound)
-
-        # select regions below and above the lower and upper bounds, respectively
-        # take inverse to select region in-between
-        trimmed_data = data_frame[
-            ~(
-                (data_frame["seq_len"] < lower_bound)
-                | (data_frame["seq_len"] > upper_bound)
-            )
-        ]
-
-        # update data and sequence length variables
+        # data with outliers trimmed
+        trimmed_data = data_frame[data_frame["outlier_label"] == 1].dropna()
+        # update instance variables to reflect removed outliers
         self.data = trimmed_data["text"].tolist()
         self.seq_lens = trimmed_data["seq_len"].tolist()
 
-        # TODO: return the outlying data that was trimmed, but I'm not convinced this works or is the best approach yet
-        return data_frame[
-            (data_frame["seq_len"] < lower_bound)
-            | (data_frame["seq_len"] > upper_bound)
-        ]
+        return data_frame[data_frame["outlier_label"] == -1].dropna()
 
     def plot_histogram(self, savefig=False) -> None:
         # Clear figure and axes

### 2023-09-13 source/main.py

@@ -50,10 +50,7 @@ def parse_datasets():
         print("Done")
 
         dataset_info["dataset_info"].append(data_analysis.dump_info())
-        data_analysis.plot_histogram()
-
-        data_analysis.remove_outliers()
-        data_analysis.summary()
+        # data_analysis.plot_histogram()
 
         dataset_info["trimmed_dataset_info"].append(data_analysis.dump_info())
 
@@ -70,6 +67,15 @@ def parse_datasets():
 
     data_objects[0].dataset_name = "All"
     data_objects[0].summary()
+
+    outliers = data_objects[0].remove_outliers()
+    data_objects[0].summary()
+    with open("outliers.txt", "a") as f:
+        for item, length in zip(
+            outliers["text"].tolist(), outliers["seq_len"].tolist()
+        ):
+            f.write(f"{int(length)}\n")
+            f.write(item + "\n")
 
     # split data into train and test
     print("Generating train/test split", end="...")
@@ -101,6 +107,9 @@ def parse_datasets():
 if __name__ == "__main__":
     plt.style.use("ggplot")
     RANDOM_SEED: int = 1
+
+    with open("outliers.txt", "w") as f:
+        f.write("")
 
     parse_datasets()
 

### 2023-09-15 thesis/MSECE_thesis.tex
Outlier detection via LOF

--- 2023-09-12 thesis/MSECE_thesis.tex
@@
         \midrule
-        ATCC            & 29,862           & \(11.90 \pm 7.22\)                             & 2,210                  & 355,476         \\
-        ATCO2           & 874              & \(12.46 \pm 5.93\)                             & 855                    & 10,886          \\
-        ATCOSIM         & 9,544            & \(11.29 \pm 4.18\)                             & 824                    & 107,759         \\
-        ZCU CZ ATC      & 14,435           & \(10.32 \pm 5.50\)                             & 3,106                  & 149,031         \\
+        ATCC            & 29,862           & \(11.90 \pm 7.22\)                             & 2,209                  & 355,473         \\
+        ATCO2           & 874              & \(12.28 \pm 4.12\)                             & 786                    & 10,733          \\
+        ATCOSIM         & 9,544            & \(11.23 \pm 4.12\)                             & 823                    & 107,153         \\
+        ZCU CZ ATC      & 14,435           & \(10.05 \pm 5.50\)                             & 2,983                  & 145,107         \\
         \midrule
-        Total           & 54,715           & \(11.39 \pm 7.22\)                             & 5,343                  & 623,152         \\
+        Total           & 54,715           & \(11.30 \pm 6.36\)                             & 5,040                  & 618,466         \\
         \bottomrule
@@
 
-\subsection{Outlier Removal}\label{sec:outlier_removal}
-This section analyzes the same corpora as above after outliers in the data have been removed, if any are present. Outlier removal was performed using
-the interquartile range (IQR) outlier removal method. IQR is defined as the difference between the medians of the third and first quartiles of the
-distribution: \(\mathbf{IQR} = \mathbf{Q3} - \mathbf{Q1}\), where \(\mathbf{Q3}\) and \(\mathbf{Q1}\) are the medians of the third and first quartiles,
-respectively. Outliers are then calculated as samples whose sequence lengths lie below \(\mathbf{Q1} - 1.5\mathbf{IQR}\) and above
-\(\mathbf{Q3} + 1.5\mathbf{IQR}\).
+\subsection{Outlier Detection and Analysis}\label{sec:outliers}
+The Local Outlier Factor (LOF) algorithm~\cite{breunig_lof_2000} was used to detect and remove outliers in the corpora for further analysis. The
+sequence lengths of the data samples were used to determine outliers. Since all four corpora are aggregated, LOF analysis was performed on the
+aggregated data instead of on the individual corpora.
+
+Sci-Kit Learn's implementation of the LOF
+algorithm\footnote{https://scikit-learn.org/stable/modules/generated/sklearn.neighbors.LocalOutlierFactor.html\#sklearn-neighbors-localoutlierfactor}
+was used with a value of \(N = 35\) for \(N\) neighbors in the K-Nearest Neighbors algorithm and Minkowski distance for distance metrics/calculations
+(in this case, since \(p=2\), Euclidean distance is effectively used for distance calculations). A total of 47 samples were detected as outliers and
+removed for analysis and are summarized in table \ref{}
+
+\begin{table}[h!]
+    \centering
+    \begin{tabular}{l | r}
+        \toprule
+        \multicolumn{2}{c}{Outlier Properties}   \\
+        \midrule
+        \textbf{Mean Sequence Length}    & 47.30 \\
+        \textbf{Std of Sequence Length}  & 11.82 \\
+        \textbf{Minimum Sequence Length} & 38    \\
+        \textbf{Maximum Sequence Length} & 73    \\
+        \textbf{\# Samples}              & 47    \\
+        \bottomrule
+    \end{tabular}
+    \label{table:outlier_stats}
+    \caption{}
+\end{table}
 

### 2023-09-15 source/data/atccomplete.py

@@ -46,6 +46,9 @@ class ATCCompleteData(Data):
         ("ai", ""),
         ("cir-", ""),
         ("cli-", ""),
+        ("rport", "airport"),
+        ("rcraft", "aircraft"),
+        ("mntn", "maintain"),
     ]
 
     # _audio_glob: List[str]

### 2023-09-15 source/data/data.py

@@ -217,9 +217,7 @@ Ratio of unique tokens to the total number of tokens: {self.token_ratio()}, {sel
 
     def remove_outliers(self) -> pd.DataFrame:
         """
-        Perform IQR-based outlier removal based on sequence lengths of the data.
-        Samples with sequence lengths less than Q1 - 1.5 * IQR and greater than Q3 + 1.5 * IQR
-        will be removed from the data.
+        Detect and remove outliers via the Local Outlier Factor algorithm.
 
         The data and seq_len instance variables are updated as a result of this.
         """
@@ -230,10 +228,36 @@ Ratio of unique tokens to the total number of tokens: {self.token_ratio()}, {sel
         # fit to data and label outliers/inliers outlier = -1, inlier = 1
         data_frame["outlier_label"] = lof.fit_predict(
             # nxm matrix for API compatibility
-            data_frame["seq_len"]
-            .to_numpy()
+            data_frame["seq_len"].to_numpy()
+            # reshape to a vector
             .reshape(-1, 1)
         )
+
+        data_frame["negative_outlier_factor_"] = lof.negative_outlier_factor_
+
+        # inliers = data_frame[data_frame["outlier_label"] == 1].dropna()
+        # plt.scatter(
+        #     # inlier sequence lengths
+        #     inliers["seq_len"],
+        #     # inlier negative outlier factors
+        #     inliers["negative_outlier_factor_"],
+        #     color="k",
+        #     label="Inliers",
+        # )
+        # outliers = data_frame[data_frame["outlier_label"] == -1].dropna()
+        # plt.scatter(
+        #     # outlier sequence lengths
+        #     outliers["seq_len"],
+        #     # outlier negative outlier factors
+        #     outliers["negative_outlier_factor_"],
+        #     color="b",
+        #     label="Outliers",
+        # )
+        # plt.legend()
+        # plt.ylabel("Negative Outlier Factor")
+        # plt.xlabel("Sample Sequence Length")
+        # plt.ylim(-20, 5)
+        # plt.show()
 
         # data with outliers trimmed
         trimmed_data = data_frame[data_frame["outlier_label"] == 1].dropna()

### 2023-09-15 source/main.py

@@ -69,13 +69,21 @@ def parse_datasets():
     data_objects[0].summary()
 
     outliers = data_objects[0].remove_outliers()
+    print(f"Mean outlier sequence length: {outliers['seq_len'].mean()}")
+    print(f"Min outlier sequence length: {outliers['seq_len'].min()}")
+    print(f"Max outlier sequence length: {outliers['seq_len'].max()}")
+    print(f"Std outlier sequence length: {outliers['seq_len'].std()}")
+
     data_objects[0].summary()
+
     with open("outliers.txt", "a") as f:
         for item, length in zip(
             outliers["text"].tolist(), outliers["seq_len"].tolist()
         ):
             f.write(f"{int(length)}\n")
             f.write(item + "\n")
+
+    data_objects[0].plot_histogram()
 
     # split data into train and test
     print("Generating train/test split", end="...")

### 2023-09-21 thesis/MSECE_thesis.tex
More spelling corrections, thesis section on spelling errors and inconsistencies in transcripts.

@@
 \usepackage{booktabs}
+\usepackage[T1]{fontenc}
 \usepackage[numbib]{tocbibind}
@@
 As a result, the data in these corpora all have different formats and need to be processed into a common format to be used together, with the goal
-being to essentially create a super-corpus that includes all the corpora listed above.
+being to essentially create a ``super-corpus'' that includes all the corpora listed above (i.e., a superset of the other corpora).
 
-All data was processed in Python using primarily built-in functions/modules\footnote{The \lstinline|re| and \lstinline|xml| modules were also used although
-    these are built into most Python distributions by default.}. The text data is extracted from the corpus transcripts i.e.~Lisp in ATCC, XML in
-ATCO2 and ZCU CZ ATC, and plain text in ATCOSIM, the transcriber annotations are removed where necessary\footnote{A copy of the original data is
-    made, then the annotations are removed from the copy.}, and the resulting text is written in the standard corpus format (an ASCII text file with
-one line in the file corresponding to one sample from the corpus) to a file.
+All data was processed in Python using primarily built-in functions/modules\footnote{The \lstinline|re| and \lstinline|xml| modules were also used
+    although these are built into most Python distributions by default.}. The text data is extracted from the corpus transcripts parsing from the
+corpora-specific formats where necessary i.e.~Lisp in ATCC, XML in ATCO2 and ZCU CZ ATC, and plain text in ATCOSIM, the transcriber annotations
+are removed where necessary\footnote{A copy of the original data is made, then the annotations are removed from the copy.}, and the resulting text
+is written to a file in the standard corpus format (an ASCII text file with one line in the file corresponding to one sample from the corpus) to a
+file.
+
+\subsection{Handling Spelling Errors \& Inconsistencies}\label{sec:spelling_errors}
+By the nature of the data labeling process, there are bound to be spelling errors introduced by the annotators during corpora creation. This gets
+exacerbated by the aggregation of several corpora with varying controls for correcting spelling errors and inconsistencies. Nearly all of the errors
+are recoverable/correctable given context. For example, ``possibility of \textbf{tornado's}'' should clearly have been
+``possibility of \textbf{tornadoes}'', changing the posessive form of ``tornado'' to the plural form. There are also instances of spelling conventions
+that are consistent within corpora, but become inconsistent when the corpora are combined.
+
+% 22 spelling errors
+Manually reviewing every sample in the corpus is unrealistic considering that there are over 50,000 data samples and would be subject the same
+human error that introduced those errors in the first place. Instead, the frequency of occurence of tokens in the corpora combined with manual review
+thereafter, was used to detect and correct the most common errors. Tokens with the lowest frequency across corpora were manually reviewed and
+corrected where necessary. An \(A \rightarrow B\) mapping system was used to correct the errors, where \(A\) is the erroneous token and \(B\) is the
+corrected or replaced token. A total of 22 unique errors were detected and subsequently corrected or removed using this method.
+
+% hesitations, cut off words, different kinds of filled pauses
+The mapping system described above is also used to modify within corpora conventions to be consistent across corpora. Three areas are addressed here:
+hesitations, cut-off or partial words, and filled pauses within speech. These areas are defined differently between the four corpora, which makes
+unifying the transcripts difficult. Hesitations, for example, are defined by one corpus as any incomplete or non-english speech sounds and left
+completely undefined by another (despite annotators making notes about hesitations). The way in which hesitations, pauses, and partial words are
+transcribed differs between corpora as well. Two corpora prescribe special tokens to these phenomena such as ``[hes]'', ``[hesitation]'', ``<pause>'',
+or similar, while the others transcribe the data as it sounds (i.e., ``uh'', ``eh'', or ``er'' for filled pauses and ``cir-'' or ``cir+'' for
+instances that were cut-off mid pronunciation). In order to facilitate the unification of the transcripts for all four copora, all three of these
+phenomena have been redefined with accompanying translation/replacement rules below:
 
@@
 (in this case, since \(p=2\), Euclidean distance is effectively used for distance calculations). A total of 47 samples were detected as outliers and
-removed for analysis and are summarized in table \ref{}
+removed for analysis and are summarized in table~\ref{table:outlier_stats}.
 
@@
     \label{table:outlier_stats}
-    \caption{}
+    \caption{Mean, standard deviation, minimum, and maximum sequence length of samples detected as outliers.}
 \end{table}
+
+Although all of the outliers have higher sequence lengths than the typical sequence lengths of the other samples in the data, going the through
+the outlying samples manually and reviewing each reveals that, for the most part, the samples are typical albeit somewhat complex communications.
+With the exception of four outliers, all of the samples labeled as outlying by the LOF algorithm were kept in the data.
 

### 2023-09-21 source/data/atccomplete.py

@@ -40,19 +40,17 @@ class ATCCompleteData(Data):
         ("1347.85706", "one three four seven dot eight five seven zero six"),
         ("four]", "four"),
         # flight number metadata somehow made it into some of the transcripts
-        ("swift61", ""),
-        ("aal891", ""),
+        ("swift61", "swift six one"),
+        ("aal891", "american eight ninety one"),
         # repeated words/hesitations
         ("ai", ""),
-        ("cir-", ""),
-        ("cli-", ""),
+        ("cir-", "cir+"),
+        ("cli-", "cli+"),
         ("rport", "airport"),
         ("rcraft", "aircraft"),
         ("mntn", "maintain"),
+        ("tornado's", "tornadoes"),
     ]
-
-    # _audio_glob: List[str]
-    # _transcript_glob: List[str]
 
     def __init__(self, data_root: str, **kwargs):
         super(ATCCompleteData, self).__init__(

### 2023-09-21 source/data/atcosim.py

@@ -21,7 +21,15 @@ class ATCOSimData(Data):
     ```
     """
 
-    transcription_corrections = [("kil0", "kilo"), ("ai", "air"), ("airr", "air")]
+    transcription_corrections = [
+        ("kil0", "kilo"),
+        ("ai", "air"),
+        ("airr", "air"),
+        ("ahm", "hm"),
+        ("aha", "uh"),
+        ("ah", "uh"),
+        ("ohh", "uh"),
+    ]
 
     def __init__(self, data_root: str, **kwargs):
         super(ATCOSimData, self).__init__(dataset_name="ATCOSIM", **kwargs)
@@ -43,7 +51,7 @@ class ATCOSimData(Data):
         xml_tag = re.compile(r"(<[A-Z]+>|</[A-Z]+>)")
         annotation_tag = re.compile(r"(\[[A-Z]+\])")
         special_chars = re.compile(r"[=~@]")
-        hesitation_tokens = re.compile(r"(ah|hm|ahm|yeah|aha|nah|ohh)")
+        # hesitation_tokens = re.compile(r"(ah|hm|ahm|yeah|aha|nah|ohh)")
         non_english_tags = re.compile(r"(<FL>.</FL>)")
 
         for file in self.text_glob:
@@ -59,7 +67,7 @@ class ATCOSimData(Data):
             text = xml_tag.sub("", text)
             text = annotation_tag.sub("", text)
             text = special_chars.sub("", text)
-            text = hesitation_tokens.sub("", text)
+            # text = hesitation_tokens.sub("", text)
 
             # lower case, remove whitespace
             text = text.lower().strip()

### 2023-09-21 source/data/data.py

@@ -220,6 +220,8 @@ Ratio of unique tokens to the total number of tokens: {self.token_ratio()}, {sel
         Detect and remove outliers via the Local Outlier Factor algorithm.
 
         The data and seq_len instance variables are updated as a result of this.
+
+        :returns: A pandas DataFrame with entries for the `text`, `seq_len`, `negative_outlier_factor_`, and `outlier_label`
         """
         # using a pandas data frame because it keeps everything aligned and provides easy-to-use and verified helper functions
         lof = LocalOutlierFactor(n_neighbors=35)
@@ -234,30 +236,6 @@ Ratio of unique tokens to the total number of tokens: {self.token_ratio()}, {sel
         )
 
         data_frame["negative_outlier_factor_"] = lof.negative_outlier_factor_
-
-        # inliers = data_frame[data_frame["outlier_label"] == 1].dropna()
-        # plt.scatter(
-        #     # inlier sequence lengths
-        #     inliers["seq_len"],
-        #     # inlier negative outlier factors
-        #     inliers["negative_outlier_factor_"],
-        #     color="k",
-        #     label="Inliers",
-        # )
-        # outliers = data_frame[data_frame["outlier_label"] == -1].dropna()
-        # plt.scatter(
-        #     # outlier sequence lengths
-        #     outliers["seq_len"],
-        #     # outlier negative outlier factors
-        #     outliers["negative_outlier_factor_"],
-        #     color="b",
-        #     label="Outliers",
-        # )
-        # plt.legend()
-        # plt.ylabel("Negative Outlier Factor")
-        # plt.xlabel("Sample Sequence Length")
-        # plt.ylim(-20, 5)
-        # plt.show()
 
         # data with outliers trimmed
         trimmed_data = data_frame[data_frame["outlier_label"] == 1].dropna()

### 2023-09-21 source/main.py

@@ -73,6 +73,19 @@ def parse_datasets():
     print(f"Min outlier sequence length: {outliers['seq_len'].min()}")
     print(f"Max outlier sequence length: {outliers['seq_len'].max()}")
     print(f"Std outlier sequence length: {outliers['seq_len'].std()}")
+    print(
+        f"Most significant outlier:\n{outliers.where(outliers['seq_len'] == outliers['seq_len'].max()).dropna()}"
+    )
+
+    print(f"Max LOF score: {outliers['negative_outlier_factor_'].min()}")
+    print(
+        outliers["text"]
+        .where(
+            outliers["negative_outlier_factor_"]
+            == outliers["negative_outlier_factor_"].min()
+        )
+        .dropna()
+    )
 
     data_objects[0].summary()
 

### 2023-10-05 thesis/MSECE_thesis.tex
Thesis additions.

@@
 samples that are used to train the language models in this work are transcripts of aural speech. The datasets used are described in more
-detail below. All recordings and transcripts are in english, although there are multiple instances of foreign words occuring
+detail below. All recordings and transcripts are in english, although there are multiple instances of foreign words occurring
 
@@
 communications are included in this data. Annotations were created by experienced transcribers/annotators and labeled samples were randomly selected
-for review during the annoation process. After the dataset was completely labeled, all samples were checked, corrected, and
+for review during the annotation process. After the dataset was completely labeled, all samples were checked, corrected, and
 unified\cite{smidl_air_2019}.
@@
 are recoverable/correctable given context. For example, ``possibility of \textbf{tornado's}'' should clearly have been
-``possibility of \textbf{tornadoes}'', changing the posessive form of ``tornado'' to the plural form. There are also instances of spelling conventions
+``possibility of \textbf{tornadoes}'', changing the possessive form of ``tornado'' to the plural form. There are also instances of spelling conventions
 that are consistent within corpora, but become inconsistent when the corpora are combined.
@@
 Manually reviewing every sample in the corpus is unrealistic considering that there are over 50,000 data samples and would be subject the same
-human error that introduced those errors in the first place. Instead, the frequency of occurence of tokens in the corpora combined with manual review
+human error that introduced those errors in the first place. Instead, the frequency of occurrence of tokens in the corpora combined with manual review
 thereafter, was used to detect and correct the most common errors. Tokens with the lowest frequency across corpora were manually reviewed and
 corrected where necessary. An \(A \rightarrow B\) mapping system was used to correct the errors, where \(A\) is the erroneous token and \(B\) is the
-corrected or replaced token. A total of 22 unique errors were detected and subsequently corrected or removed using this method.
+corrected or replaced token. A total of 22 unique errors were detected and subsequently corrected or removed using this method, although multiple
+instances of those errors typically occurred.
 
@@
 or similar, while the others transcribe the data as it sounds (i.e., ``uh'', ``eh'', or ``er'' for filled pauses and ``cir-'' or ``cir+'' for
-instances that were cut-off mid pronunciation). In order to facilitate the unification of the transcripts for all four copora, all three of these
-phenomena have been redefined with accompanying translation/replacement rules below:
+instances that were cut-off mid pronunciation). In order to facilitate the unification of the transcripts for all four corpora, the conventions for
+transcribing these phenomena are redefined below:
+
+\begin{table}[h!]
+    \centering
+    \begin{tabular}{l p{0.65\linewidth}}
+        \toprule
+        Phenomenon               & Definition                                                                                                              \\
+        \midrule
+        Hesitations              & Instances of hesitation for which the specific realization is not defined (usually some special token is used instead). \\
+        \midrule
+        Partial \& cut-off words & Partially pronounced words are indicated by a hyphen character where the pronunciation of the word is missing.          \\
+        \midrule
+        Filled pauses            & Hesitations with a verbal component that has been described or otherwise defined by the corpus/annotator.               \\
+        \bottomrule
+    \end{tabular}
+    \label{tab:phenomena_definitions}
+    \caption{Speech/transcription phenomena with corresponding convention for transcribing or translating, depending on the corpus being processed.}
+\end{table}
+
+\begin{table}[h!]
+    \centering
+    \begin{tabular}{l l}
+        \toprule
+        Phenomenon                   & Convention for representation \\
+        \midrule
+        Hesitations                  & [HES]                         \\
+        \midrule
+        Partial \& cut-off words     & -                             \\
+        \midrule
+        Filled pauses (open-mouth)   & uh, um, ah                    \\
+        \midrule
+        Filled pauses (closed-mouth) & hm, mhm                       \\
+        \bottomrule
+    \end{tabular}
+    \label{tab:phenomena_conventions}
+    \caption{Conventions for transcript representation/translation of phenomena to keep transcripts consistent across corpora.}
+\end{table}
+
+A special token is used to represent hesitations when no other context regarding the realization of hesitation is present. If other corpora use a
+special token to represent hesitations, it is translated to match the convention in table~\ref{tab:phenomena_conventions}. Partial/cut-off words
+are represented with a hyphen (-) at the point where the pronounced word is cut off. For example, if the last syllable of ``approaching'' is cut
+mid transmission, it would be represented in the transcript as ``approa-'' or, likewise, if the first syllable is cut off, it would be written as
+``-roaching''. Lastly, if there is enough information in the original transcript to indicate the realization of a hesitation, it is represented by
+the most appropriate token from ``Filled pauses'' in table~\ref{tab:phenomena_conventions}.
 
@@
 % sequence lengths, unique tokens, total number of tokens, total number of samples, breakdown by each dataset
-Relevant corpora statistics were observed or calculated are are presented in Table~\ref{table:corpora_stats}. This includes the number of samples in
+Relevant corpora statistics were observed or calculated and are presented in Table~\ref{table:corpora_stats}. This includes the number of samples in
 the corpus, mean sequence length per sample in the corpus, the number of unique tokens in the corpus, and the total number of tokens in the corpus.
@@
 The ATCC and ZCU CZ ATC corpora make up a significant portion of the data, ATCC alone makes up more than half (approximately 54\%) of the combined
-data, and ATCC and ZCU CZ ATC together make up aprroximately 81\% of the data. The sequence lengths of the samples are fairly consistent across
+data, and ATCC and ZCU CZ ATC together make up approximately 81\% of the data. The sequence lengths of the samples are fairly consistent across
 corpora with at most about an 18\% percent difference between sequence lengths. The number of unique tokens does not scale additively to the
@@
 
-\subsection{Data Preprocessing}
-% explain and justify what is removed from the text, how, and why
-% present and explain any relevant algorithms where necessary
+\section{Language Modeling}
+% bert, roberta
+% ngram
+This section introduces the chosen language model architectures as well as their pretrained checkpoints, if one is used. Two transformer-based
+architectures are used, BERT and RoBERTa, and statistical/machine learning-based approaches such as N-gram and word lattices.
 
-\subsection{Tokenization}
+The transformers are pre-trained using the same masked language modeling (MLM) objective presented in Devlin et al.~\cite{devlin_bert_2019}
+to train BERT for bidirectional encoding representation. Both pretrained checkpoints and base models are pretrained and used with the best performing
+ASR model to analyze their impacts on performance. The pre-training data is taken from the unified transcripts as described above, tokens are masked
+at random with a probability of 0.2.
+
+\section{Speech Recognition}
+
+\section{Tokenization Algorithms}
 % present and explain the tokenization algorithms used (possibly which models they correspond to)
+Tokenization is the process of separating or segmenting samples into tokens. The rules by which the tokens are created or segmented depends entirely
+on the algorithm used. The simplest and most intuitive way to do this by creating tokens based on whitespace i.e., each ``word'' in the corpus
+is treated as a token\footnote{The Huggingface Tokenizer library implements this as the \lstinline|WordLevel| tokenizer.}. Although this is a simple
+and easy to understand method of tokenization, it does little to nothing to address the out-of-vocabulary problem that necessarily comes with any
+NLP tasks~\cite{wu_googles_2016}.
+
+\subsection{WordPiece}
+
+\subsection{(Byte-Level) Byte-Pair Encoding}
 

### 2023-10-05 source/data/atccomplete.py

@@ -37,15 +37,13 @@ class ATCCompleteData(Data):
         ("bye]", "bye"),
         (" -", ""),
         (" 3 ", "three"),
-        ("1347.85706", "one three four seven dot eight five seven zero six"),
+        ("1347.85706", "one three four seven decimal eight five seven zero six"),
         ("four]", "four"),
         # flight number metadata somehow made it into some of the transcripts
         ("swift61", "swift six one"),
         ("aal891", "american eight ninety one"),
         # repeated words/hesitations
         ("ai", ""),
-        ("cir-", "cir+"),
-        ("cli-", "cli+"),
         ("rport", "airport"),
         ("rcraft", "aircraft"),
         ("mntn", "maintain"),

### 2023-10-05 source/data/czechdataset.py

@@ -12,19 +12,11 @@ class ZCUCZATCDataset(Data):
     Data is organized into `trs` files (follows an XML standard). Details in `~parse_transcripts` function.
 
     Dataset can be obtained here: https://lindat.mff.cuni.cz/repository/xmlui/handle/11858/00-097C-0000-0001-CCA1-0
-
-    Notes:
-        - As useful as this dataset may be, the transcriptions and their format are so convoluted and different from
-          the other datasets that it may not be worth using
-        - There are a lot of non-english tokens in this data that I am not sure how to filter out
-        - All number readouts are space separated
-        - **I'll keep working on this script to normalize the data, but I'm exluding it from the training data for now**
-        - Examples of problematic lines:
-            * Austrian 1 2 8 X contact (Vienna(vín)) 1 3 1 . 3 5 0 servus
-            * Air Malta (9(najn)) 6 9 1 contact Praha 1 2 7 . 1 2 5 go+
-            * (Praha(Prague)) 1 3 5 1 3 5  7 9
-            * +el 3 8 0 plea+
     """
+
+    _transcript_corrections = [
+        ("+", "-")
+    ]
 
     def __init__(self, data_root: str, **kwargs):
         super(ZCUCZATCDataset, self).__init__(dataset_name="ZCU CZ ATC", **kwargs)
@@ -85,8 +77,10 @@ class ZCUCZATCDataset(Data):
                             nonstandard_match.group(0), nonstandard_match.group(1)
                         )
 
-                words = text.split()
-                text = " ".join(words).lower()
+                # typos in transcripts
+                for typo, correction in self._transcript_corrections:
+                    if typo in text:
+                        text = text.replace(typo, correction)
 
                 # ".." corresponds to silent segments, "" can occur when the transcript is made up
                 # of only transcriber annotations

### 2023-10-05 source/data/data.py

@@ -215,7 +215,7 @@ Ratio of unique tokens to the total number of tokens: {self.token_ratio()}, {sel
         else:
             return dataset_summary
 
-    def remove_outliers(self) -> pd.DataFrame:
+    def detect_outliers(self) -> pd.DataFrame:
         """
         Detect and remove outliers via the Local Outlier Factor algorithm.
 

### 2023-10-05 source/main.py

@@ -68,7 +68,7 @@ def parse_datasets():
     data_objects[0].dataset_name = "All"
     data_objects[0].summary()
 
-    outliers = data_objects[0].remove_outliers()
+    outliers = data_objects[0].detect_outliers()
     print(f"Mean outlier sequence length: {outliers['seq_len'].mean()}")
     print(f"Min outlier sequence length: {outliers['seq_len'].min()}")
     print(f"Max outlier sequence length: {outliers['seq_len'].max()}")

### 2023-10-24 thesis/MSECE_thesis.tex
Update thesis.

@@
 \section{Introduction}\label{sec:introduction}
+There have been noticeable and significant advancements in the general domain of artificial intelligence recently. Automatic speech recognition and
+language modeling technologies, in particular, have made tremendous advances over the past two decades achieving performance levels good enough to
+be brought to consumer facing products such as personal assistants, search engines, phones, televisions, etc. Due to the success in the general
+domain, applications have begun to spin-off into problem-specific domains, such as (and most notably for this work) aviation. For example, several
+speech recognition systems with rule-based natural language inference and semantic extraction algorithms have been implemented at EUROCONTROL
+simulation centers across Europe.
+
+At the time of this writing, the current focus of general machine learning algorithms in the automatic speech recognition (ASR) and language modeling
+domains has been a mixture of deep learning methods, namely; recurrent, convolutional, and transformer neural networks due to the availability of
+large corpora such as LibriSpeech~\cite{panayotov_librispeech_2015} and Mozilla Common Voice~\cite{ardila_common_2020} for ASR and
+WikiText~\cite{merity_pointer_2016} and IMDB~\cite{maas_learning_2011} corpora for language modeling. In contrast to the general domain, the aviation
+domain still prefers, and sees much success with traditional machine learning models for both ASR and language modeling applications. Typically, the
+most prevalent of the traditional models are Hidden Markov models and Gaussian Mixture models, in conjunction with word lattices, N-grams, and/or
+rule-based approaches depending on the application and operating environment of the models. The most likely reason for this divergence between the
+domains is the difference and relative lack of labeled data in the aviation domain. By comparison, the largest corpus in the general domain is
+LibriSpeech with approximately 1000 hours of labeled data whereas, for aviation, the largest corpus is Air Traffic Control Complete with approximately
+26 hours of labeled data\footnote{Air Traffic Control Complete has approximately 70 hours of audio data, however, about 26 hours of the 70 are
+    labeled.}.
 
 \section{Literature Review}\label{sec:lit_review}
-% from proposal; revise and add to this as more references are used
-% transformer history
+% from thesis proposal; revise and add to this as more references are used
+% transformer history (background)
 The transformer neural network architecture was proposed in 2017 for Neural Machine Translation tasks and immediately achieved
@@
 Sci-Kit Learn's implementation of the LOF
-algorithm\footnote{https://scikit-learn.org/stable/modules/generated/sklearn.neighbors.LocalOutlierFactor.html\#sklearn-neighbors-localoutlierfactor}
-was used with a value of \(N = 35\) for \(N\) neighbors in the K-Nearest Neighbors algorithm and Minkowski distance for distance metrics/calculations
-(in this case, since \(p=2\), Euclidean distance is effectively used for distance calculations). A total of 47 samples were detected as outliers and
-removed for analysis and are summarized in table~\ref{table:outlier_stats}.
+algorithm was used with a value of \(N = 35\) for \(N\) neighbors in the K-Nearest Neighbors algorithm and Minkowski distance for distance
+metrics/calculations (in this case, since \(p=2\), Euclidean distance is effectively used for distance calculations). A total of 47 samples were
+detected as outliers and removed for analysis and are summarized in table~\ref{table:outlier_stats}.
 
@@
 ASR model to analyze their impacts on performance. The pre-training data is taken from the unified transcripts as described above, tokens are masked
-at random with a probability of 0.2.
+at random with a probability of 0.2. N-grams are trained using the Natural Language Toolkit (NLTK) library in Python with values for $N$ of 1, 2,
+and 3.
 

### 2023-10-24 source/data/atccomplete.py
Fix small bugs, update thesis.

@@ -1,8 +1,69 @@
 import os
 import glob
 import re
+import json
+from pprint import pprint
 from typing import *
+from typing import List
 from data import Data, atccutils
+
+
+class TransmissionTurn(object):
+    def __init__(
+        self,
+        start_time: float,
+        end_time: float,
+        text: str,
+        to_craft: str,
+        from_craft: str,
+    ) -> "TransmissionTurn":
+        self.start_time = start_time
+        self.end_time = end_time
+        self.text = text
+        self.to_craft = to_craft
+        self.from_craft = from_craft
+
+    def to_json(self) -> str:
+        return json.dumps(self.to_dict())
+
+    def to_dict(self) -> Dict:
+        return {
+            "start_time": self.start_time,
+            "end_time": self.end_time,
+            "text": self.text,
+            "to_craft": self.to_craft,
+            "from_craft": self.from_craft,
+        }
+
+    def __repr__(self) -> str:
+        return self.to_json()
+
+
+class TransmissionGroup(object):
+    def __init__(
+        self,
+        first_transmission: TransmissionTurn,
+        transmissions: List[TransmissionTurn],
+        cls_token="[CLS]",
+        sep_token="[SEP]",
+    ) -> "TransmissionGroup":
+        self.cls_token = cls_token
+        self.sep_token = sep_token
+        self.first_transmission = first_transmission
+        self.transmissions = transmissions
+
+    def to_json(self) -> str:
+        return json.dumps(
+            {
+                "cls_token": self.cls_token,
+                "sep_token": self.sep_token,
+                "first_transmission": self.first_transmission.to_dict(),
+                "transmissions": [t.to_dict() for t in self.transmissions],
+            }
+        )
+
+    def __repr__(self) -> str:
+        return self.to_json()
 
 
 class ATCCompleteData(Data):
@@ -46,6 +107,7 @@ class ATCCompleteData(Data):
         ("ai", ""),
         ("rport", "airport"),
         ("rcraft", "aircraft"),
+        ("rborne", "airborne"),
         ("mntn", "maintain"),
         ("tornado's", "tornadoes"),
     ]
@@ -66,6 +128,7 @@ class ATCCompleteData(Data):
         --------
         A list of dictionary objects.
         """
+        assert len(self._transcript_glob) != 0
         data = []
 
         sounds_like_tag = re.compile(r"(\[sounds like(?P<guess>[A-Za-z\s\d]+)\]*)")
@@ -75,7 +138,7 @@ class ATCCompleteData(Data):
[…]

### 2023-10-24 source/data/data.py

@@ -68,6 +68,16 @@ class Data(IterableDataset):
             "is an issue with the implementing class."
         )
 
+    def parse_transcripts_for_nsp(self) -> List[str]:
+        """
+        TODO
+        """
+        raise NotImplementedError(
+            "This is an abstract method that should be implemented and overridden for "
+            "all classes that implement this one. If this method has been called, there "
+            "is an issue with the implementing class."
+        )
+
     def preprocess(
         self, tokenizer: PreTrainedTokenizer, collator: DataCollatorForLanguageModeling
     ) -> List[List[int]]:

### 2023-10-24 source/data/utils/atccutils.py

@@ -174,7 +174,7 @@ def reformat(data: List[str]) -> Dict:
     return formatted_data
 
 
-def parse(lines: List[str]) -> List[Dict]:
+def parse(lines: List[str], file_name: str = "") -> List[Dict]:
     document = ""
     for line in lines:
         # remove trailing whitespace
@@ -193,5 +193,7 @@ def parse(lines: List[str]) -> List[Dict]:
 
     for i, data in enumerate(data_blocks):
         data_blocks[i] = reformat(data)
+        if file_name != "":
+            data_blocks[i]["FILE"] = file_name
 
     return data_blocks

### 2023-10-24 source/main.py

@@ -20,10 +20,14 @@ from models import Model
 # root dataset paths corresponding to data analysis classes
 datasets: Dict[str, Data] = {
     # TODO: find a way to sync file paths across computers (shell/env var, config file?)
-    "/home/students/vandebra/programming/thesis_data/atc0_comp": ATCCompleteData,
-    "/home/students/vandebra/programming/thesis_data/atcosim/": ATCOSimData,
-    "/home/students/vandebra/programming/thesis_data/ATCO2-ASRdataset-v1_beta": ATCO2SimData,
-    "/home/students/vandebra/programming/thesis_data/ZCU_CZ_ATC": ZCUCZATCDataset,
+    "/home/vandebra/programming/thesis_data/atc0_comp": ATCCompleteData,
+    "/home/vandebra/programming/thesis_data/atcosim/": ATCOSimData,
+    "/home/vandebra/programming/thesis_data/ATCO2-ASRdataset-v1_beta": ATCO2SimData,
+    "/home/vandebra/programming/thesis_data/ZCU_CZ_ATC": ZCUCZATCDataset,
+}
+
+nsp_datasets: Dict[str, Data] = {
+    "/home/vandebra/programming/thesis_data/atc0_comp": ATCCompleteData,
 }
 
 
@@ -32,100 +36,109 @@ def parse_datasets():
     dataset_info = {"dataset_info": [], "trimmed_dataset_info": []}
     data_objects: List[Data] = []
 
-    # initializes each implementing class with its data which is specified by `root_path`
-    # see `datasets` and implementing classes for more details
-    for root_path, data_class in datasets.items():
-        # initialize class with root path and random seed
-        data_analysis: Data = data_class(data_root=root_path, random_seed=RANDOM_SEED)
+    # # initializes each implementing class with its data which is specified by `root_path`
+    # # see `datasets` and implementing classes for more details
+    # for root_path, data_class in datasets.items():
+    #     # initialize class with root path and random seed
+    #     data_analysis: Data = data_class(data_root=root_path, random_seed=RANDOM_SEED)
 
-        # parse dataset info, along with some printout so the user has some idea of what's
-        # happening
-        print(f"Parsing transcripts for '{data_analysis.name}'", end="...")
-        data_analysis.parse_transcripts()
+    #     # parse dataset info, along with some printout so the user has some idea of what's
+    #     # happening
[…]

### 2023-10-24 thesis/MSECE_thesis.tex
Typo/conflict fix.

@@
 \usepackage{booktabs}
+\usepackage{algpseudocodex}
 \usepackage[T1]{fontenc}
@@
         \vspace*{2pt}
-        Andrew Schneider, Ph.D.\\
+        Andrew Schneider\\
         Committee Member\\
@@
 
-The transformers are pre-trained using the same masked language modeling (MLM) objective presented in Devlin et al.~\cite{devlin_bert_2019}
-to train BERT for bidirectional encoding representation. Both pretrained checkpoints and base models are pretrained and used with the best performing
-ASR model to analyze their impacts on performance. The pre-training data is taken from the unified transcripts as described above, tokens are masked
-at random with a probability of 0.2. N-grams are trained using the Natural Language Toolkit (NLTK) library in Python with values for $N$ of 1, 2,
-and 3.
+The transformers are pre-trained using the same masked language modeling (MLM) objective presented in Devlin et al.~\cite{devlin_bert_2019} to train
+BERT for bidirectional encoding representation and likewise in Liu et al.~for training RoBERTa~\cite{liu_roberta_2019}. Given the results of the
+RoBERTa pretraining and downstream task training thereafter~\cite{liu_roberta_2019}, it was initially decided to eliminate the next sentence prediction
+(NSP) task as it was used in BERT, however, their results suggest that the success of the MLM-only approach was due to the increase in batch and
+overall data size~\cite{liu_roberta_2019}. Taking this into account, along with the very limited amount of data available in the domain of aviation
+an NSP pretraining task was implemented where possible\footnote{Only about half of the data (ATCC) is in a format that enables us to extract the turn
+    taking phenomenon that is present in most air-traffic communications.} to improve its pretraining results.
 

### 2023-10-25 source/data/atccomplete.py
Added transcript file for NSP; fixed bugs.

@@ -14,14 +14,14 @@ class TransmissionTurn(object):
         start_time: float,
         end_time: float,
         text: str,
-        to_craft: str,
-        from_craft: str,
+        to_key: str,
+        from_key: str,
     ) -> "TransmissionTurn":
         self.start_time = start_time
         self.end_time = end_time
         self.text = text
-        self.to_craft = to_craft
-        self.from_craft = from_craft
+        self.to_key = to_key
+        self.from_key = from_key
 
     def to_json(self) -> str:
         return json.dumps(self.to_dict())
@@ -31,8 +31,8 @@ class TransmissionTurn(object):
             "start_time": self.start_time,
             "end_time": self.end_time,
             "text": self.text,
-            "to_craft": self.to_craft,
-            "from_craft": self.from_craft,
+            "to_key": self.to_key,
+            "from_key": self.from_key,
         }
 
     def __repr__(self) -> str:
@@ -44,6 +44,7 @@ class TransmissionGroup(object):
         self,
         first_transmission: TransmissionTurn,
         transmissions: List[TransmissionTurn],
+        file: str,
         cls_token="[CLS]",
         sep_token="[SEP]",
     ) -> "TransmissionGroup":
@@ -51,12 +52,14 @@ class TransmissionGroup(object):
         self.sep_token = sep_token
         self.first_transmission = first_transmission
         self.transmissions = transmissions
+        self.file = file
 
     def to_json(self) -> str:
         return json.dumps(
             {
                 "cls_token": self.cls_token,
                 "sep_token": self.sep_token,
+                "file": self.file,
                 "first_transmission": self.first_transmission.to_dict(),
                 "transmissions": [t.to_dict() for t in self.transmissions],
             }
@@ -250,8 +253,8 @@ class ATCCompleteData(Data):
                             start_time=pair["TIMES"]["start"],
                             end_time=pair["TIMES"]["end"],
                             text=pair["TEXT"],
-                            to_craft=pair["TO"],
-                            from_craft=pair["FROM"],
+                            to_key=pair["TO"],
+                            from_key=pair["FROM"],
                         )
                     )
 
@@ -260,12 +263,16 @@ class ATCCompleteData(Data):
                     start_time=first_transmission["TIMES"]["start"],
                     end_time=first_transmission["TIMES"]["end"],
                     text=first_transmission["TEXT"],
-                    to_craft=first_transmission["TO"],
-                    from_craft=first_transmission["FROM"],
+                    to_key=first_transmission["TO"],
[…]

### 2023-11-29 thesis/MSECE_thesis.tex
Nearly final version of thesis.

--- 2023-10-24 thesis/MSECE_thesis.tex
@@
 \usepackage{booktabs}
+\usepackage{graphicx}
+\usepackage{amsmath}
+\usepackage{amssymb}
+\usepackage{algorithm}
+\usepackage{algpseudocode}
+\usepackage{xcolor}
 \usepackage[T1]{fontenc}
@@
 \usepackage[lmargin=1.25in,rmargin=1.25in,tmargin=1.0in,bmargin=1.0in]{geometry}
+
+\definecolor{codegreen}{rgb}{0,0.6,0}
+\definecolor{codegray}{rgb}{0.5,0.5,0.5}
+\definecolor{codepurple}{rgb}{0.58,0,0.82}
+\definecolor{backcolour}{rgb}{0.95,0.95,0.92}
+
+\lstdefinestyle{thesiscodelisting}{
+    keywordstyle=\color{magenta},
+    numberstyle=\tiny\color{codegray},
+    stringstyle=\color{green},
+    basicstyle=\ttfamily\footnotesize,
+    breakatwhitespace=false,
+    breaklines=true,
+    captionpos=b,
+    keepspaces=true,
+    showspaces=false,
+    showstringspaces=false,
+    showtabs=false,
+}
+
+\lstset{style=thesiscodelisting}
 
@@
     by\\
-    \textit{Aaron P.~Van De Brook}\\
+    \textit{Aaron Van De Brook}\\
 
@@
         \vspace*{2pt}
-        Farahzad Behi\\
+        Massood Towhidnejad, Ph.D.\\
         Department Chair, Electrical Engineering \& Computer Science\\
@@
 
-\section{Abstract}\label{sec:abstract}
+\section{Notation and Terminology}
+% TODO: proper writing, bulleted list for now
+This section is included for the sake of clarity throughout this thesis. In pursuit of that, there are a few terms and notations that are defined
+below.
+
+Set notation is used in parts of this thesis, particularly in equations and algorithms, to convey mathematical concepts. With that being said, the
+shorthand below is defined
+\begin{equation*}
+    A; x := A \cup \{x\}
+\end{equation*}
+
+\noindent
+as adding the element $x$, of its own set, to the set $A$.
+
+The following words (in bold) are used throughout the thesis as they relate to language modeling and are defined concretely as follows:
+\begin{itemize}
+    \item A \textbf{sequence} shall be defined as a set of tokens (distinct from a sentence)
+    \item A \textbf{token} shall be defined as the smallest, fundamental component of a sequence (distinct from characters and words)
+    \item The rules for the formation of tokens from a sentence (and into sequences) shall be defined by tokenization algorithms (section
+          \ref{sec:tokenizers})
+    \item The definition of \textbf{words} and \textbf{sentences} follow from that of the English language
+    \item The \textbf{vocabulary}, \textbf{inventory}, \textbf{index}, and \textbf{word list} of a model shall be semantically identical
+\end{itemize}
+\newpage
+
+\section*{Abstract}\label{sec:abstract}
 
 \section{Introduction}\label{sec:introduction}
-There have been noticeable and significant advancements in the general domain of artificial intelligence recently. Automatic speech recognition and
+There have been noticeable and significant advancements in the general domain of artificial intelligence (AI) recently. Automatic speech recognition and
 language modeling technologies, in particular, have made tremendous advances over the past two decades achieving performance levels good enough to
@@
 speech recognition systems with rule-based natural language inference and semantic extraction algorithms have been implemented at EUROCONTROL
-simulation centers across Europe.
+simulation centers in Europe to train air traffic controllers and study the potential effects of AI augmented air traffic management (ATM) systems
+on controller workloads. Preliminary studies have found that AI augmented ATM systems not only reduce controller workloads but also reduce fuel
+consumption of aircraft by optimizing aircraft routes in the airspace and reducing departure/arrival times thereby reducing fuel
+consumption~\cite{helmke_quantifying_2017}.
 
 At the time of this writing, the current focus of general machine learning algorithms in the automatic speech recognition (ASR) and language modeling
-domains has been a mixture of deep learning methods, namely; recurrent, convolutional, and transformer neural networks due to the availability of
-large corpora such as LibriSpeech~\cite{panayotov_librispeech_2015} and Mozilla Common Voice~\cite{ardila_common_2020} for ASR and
+domains has been on a variety of deep learning methods/architectures, namely; recurrent, convolutional, and transformer neural networks due to the
+availability of large corpora such as LibriSpeech~\cite{panayotov_librispeech_2015} and Mozilla Common Voice~\cite{ardila_common_2020} for ASR and
 WikiText~\cite{merity_pointer_2016} and IMDB~\cite{maas_learning_2011} corpora for language modeling. In contrast to the general domain, the aviation
-domain still prefers, and sees much success with traditional machine learning models for both ASR and language modeling applications. Typically, the
+domain still prefers, and sees much success with, traditional machine learning models for both ASR and language modeling applications. Typically, the
 most prevalent of the traditional models are Hidden Markov models and Gaussian Mixture models, in conjunction with word lattices, N-grams, and/or
 rule-based approaches depending on the application and operating environment of the models. The most likely reason for this divergence between the
-domains is the difference and relative lack of labeled data in the aviation domain. By comparison, the largest corpus in the general domain is
+domains is the difference and relative lack of labeled data in the aviation domain. By comparison, one of the largest corpora in the general domain is
 LibriSpeech with approximately 1000 hours of labeled data whereas, for aviation, the largest corpus is Air Traffic Control Complete with approximately
 26 hours of labeled data\footnote{Air Traffic Control Complete has approximately 70 hours of audio data, however, about 26 hours of the 70 are
-    labeled.}.
+    labeled.}. This disparity in data availability is exacerbated by the fact that aviation English uses a specialized vocabulary in addition to
+specialized pronunciation for some words to enhance understandability in a radiotelephonic (R/T) medium~\cite{paltridge_handbook_2013}, making it
+significantly more difficult to transcribe/label data as it requires transcribers to have domain knowledge and experience in order to create effective
+transcriptions. The medium of communication along with the environment in which the communication occurs i.e.~a noisy and possibly low fidelity
+medium\footnote{Depending on the equipment used to transmit and receive R/T communications.} combined with a noisy and active communication
+environment---an active flight deck (influenced by wind, engine noise, etc.~) or an active Air Traffic Control (ATC) tower (other people/controllers
+conversing or communicating in the background)---and a rapid rate of speech~\cite{paltridge_handbook_2013}, further contributing to the difficulty in
+understanding and therefore transcribing aviation English.
+
+% connect complexity of aviation english to utility/necessity of NLP models (ASR and LM) in the aviation domain
+The potential for speech recognition and language modeling algorithms to reach better than human performance (for example, the SQuAD benchmark has
+recorded results consistently better than human since 2019\footnote{https://rajpurkar.github.io/SQuAD-explorer/})~\cite{zhang_ai_2022} suggests that
+these NLP algorithms could have high utility in the aviation domain. If they can achieve better-than-human performance in the aviation domain, and be
+used in conjunction with pilot and air traffic controller systems, they could easily boost efficiency, reduce workloads, and reduce and/or mitigate
+errors during normal operations.
+
+The methods and results laid out in this work were done with the intention of creating a streamlined process for training and testing deep learning
+based NLP methods, specifically automatic speech recognition and text-based language models. The results from the language models suggest that the
+volume of text used to train the language models is not yet sufficient to produce results akin to those in the conversational and/or literary English
+domains. However, the automatic speech recognition models were able to achieve near state-of-the-art results and, according to the literature review,
+better than other deep learning-based approaches in the aviation domain.
 
@@
 \section{Data Sources}\label{sec:data_source}
-There are four datasets used in this work. The datasets are primarily speech corpora intended either for automatic speech recognition
-research in the domain of aviation communications, linguistics research in aviation english, or both. As a result, the text-based data
-samples that are used to train the language models in this work are transcripts of aural speech. The datasets used are described in more
-detail below. All recordings and transcripts are in english, although there are multiple instances of foreign words occurring
+There are four corpora used in this work. The datasets are primarily speech corpora intended either for automatic speech recognition research in the
+domain of aviation communications, linguistics research in aviation English, or both. All data in all four corpora include both recordings and
+transcriptions, audio and text, respectively. The transcriptions are used to pretrain and train the language models and tokenizers (described in
+sections \ref{sec:language_modeling} and \ref{sec:tokenizers}).
+
+The four corpora that were selected and used for this work are briefly described below. Further analysis is performed in section \ref{sec:data_analysis}.
 
@@
 ATCC is a collection of three smaller corpora recorded at Dallas-Fort Worth, Logan International, and Washington National airports and
-transcribed by subject matter experts (usually controllers and/or pilots) familiar with the terminology specific to their respective
-areas\cite{godfrey_air_1994}.
+transcribed by current or former air traffic controllers familiar with the respective areas. Audio data was recorded through the placement of VHF
+antennae configured to monitor several frequencies at each airport, such as arrival, departure, approach, ground, etc. The types of frequencies
+monitored vary by airport, but the presence of departure, approach, and ground frequencies is relatively consistent across subcorpora~\cite{godfrey_air_1994}.
+% TODO: data length/size
 
@@
 The ATCO2 dataset is a speech corpus that consists of audio communications at Prague and Brno airports, in Czechia, along with corresponding
-transcriptions. The speech recordings and transcripts are crowd sourced from volunteers\cite{szoke_detecting_2021}.
+transcriptions. The speech recordings and transcripts are crowd sourced from volunteers. The dataset was created and labeled, in part, as an English
+language detection corpus, however, the labeling also includes transcripts of speech segments which makes it practical for ASR and language model
+development in addition to language detection (although in this work it is only used for ASR and LM development)~\cite{szoke_detecting_2021}.
+% TODO: data length/size
 
@@
 resolved\cite{hofbauer_atcosim_2008}.
+% TODO: data length/size
 
@@
 unified\cite{smidl_air_2019}.
+% TODO: data length/size
 
@@
     \end{tabular}
+    \caption{Speech/transcription phenomena with corresponding convention for transcribing or translating, depending on the corpus being processed.}
     \label{tab:phenomena_definitions}
-    \caption{Speech/transcription phenomena with corresponding convention for transcribing or translating, depending on the corpus being processed.}
 \end{table}
@@
     \end{tabular}
+    \caption{Conventions for transcript representation/translation of phenomena to keep transcripts consistent across corpora.}
     \label{tab:phenomena_conventions}
-    \caption{Conventions for transcript representation/translation of phenomena to keep transcripts consistent across corpora.}
 \end{table}
@@
 % sequence lengths, unique tokens, total number of tokens, total number of samples, breakdown by each dataset
-Relevant corpora statistics were observed or calculated and are presented in Table~\ref{table:corpora_stats}. This includes the number of samples in
+Relevant corpora statistics were observed or calculated and are presented in Table~\ref{tab:corpora_stats}. This includes the number of samples in
 the corpus, mean sequence length per sample in the corpus, the number of unique tokens in the corpus, and the total number of tokens in the corpus.
@@
     \end{tabular}
-    \label{table:corpora_stats}
     \caption{The number of samples, mean sequence length, number of unique tokens, and total number of tokens by corpus, including the cumulative
         total across all corpora.}
+    \label{tab:corpora_stats}
 \end{table}
@@
 metrics/calculations (in this case, since \(p=2\), Euclidean distance is effectively used for distance calculations). A total of 47 samples were
-detected as outliers and removed for analysis and are summarized in table~\ref{table:outlier_stats}.
+detected as outliers and removed for analysis and are summarized in table~\ref{tab:outlier_stats}.
 
@@
     \end{tabular}
-    \label{table:outlier_stats}
     \caption{Mean, standard deviation, minimum, and maximum sequence length of samples detected as outliers.}
+    \label{tab:outlier_stats}
 \end{table}
@@
 
-\section{Language Modeling}
+\section{Language Modeling}\label{sec:language_modeling}
 % bert, roberta
-% ngram
+% n-gram
 This section introduces the chosen language model architectures as well as their pretrained checkpoints, if one is used. Two transformer-based
@@
 
-The transformers are pre-trained using the same masked language modeling (MLM) objective presented in Devlin et al.~\cite{devlin_bert_2019}
-to train BERT for bidirectional encoding representation. Both pretrained checkpoints and base models are pretrained and used with the best performing
-ASR model to analyze their impacts on performance. The pre-training data is taken from the unified transcripts as described above, tokens are masked
-at random with a probability of 0.2. N-grams are trained using the Natural Language Toolkit (NLTK) library in Python with values for $N$ of 1, 2,
-and 3.
+The transformers are pre-trained using the same masked language modeling (MLM) objective presented in Devlin et al.~\cite{devlin_bert_2019} to train
+BERT for bidirectional encoding representation and likewise in Liu et al.~\cite{liu_roberta_2019} for training RoBERTa. Given the results of the
+RoBERTa pretraining and downstream task training thereafter, it was initially decided to eliminate the next sentence prediction (NSP) task as it was
+used in BERT, however, their results suggest that the success of the MLM-only approach was due to the increase in batch and overall data
+size~\cite{liu_roberta_2019}. Taking this into account, along with the very limited amount of data available in the domain of aviation an NSP
+pretraining task was implemented where possible\footnote{Only about half of the data (only the data from ATCC, since it is continuous recordings of
+    the airspace) is in a format that enables us to extract the turn taking phenomenon that is present in most air-traffic communications.} to improve
+its pretraining results. MLM and NSP pretraining are explained more in-depth in section \ref{sec:bert}, below.
 
-\section{Speech Recognition}
+\subsection{N-Gram}\label{sec:ngram}
+An N-gram language model estimates the probability of a token occurring in a sequence based on the \(N\) tokens preceding or succeeding that token.
+In other words, for a sequence, $X$, of $T$ tokens
 
-\section{Tokenization Algorithms}
+\begin{equation}\label{eq:ngram_sequence}
+    X = \mbox{$w_1$, $w_2$, $w_3$, ..., $w_T$}
+\end{equation}
+
+\noindent
+the probability of a token, $w_i$, occurring in the sequence can be represented as
+
+\begin{equation}\label{eq:ngram_preceding}
+    \begin{gathered}
+        P(w_i) = P(w_i|w_{i-1},w_{i-2},w_{i-3},...,w_{i-N}) \\
+        P(w_i) = \prod_{j=i-1}^{i-N} P(w_j)
+    \end{gathered}
+\end{equation}
+
+\noindent
+or for the case of succeeding tokens (looking ahead $N$ tokens)
+
+\begin{equation}\label{eq:ngram_succeeding}
+    \begin{gathered}
+        P(w_i) = P(w_i|w_{i+1},w_{i+2},w_{i+3},...,w_{i+N}) \\
+        P(w_i) = \prod_{j=i+1}^{i+N} P(w_j)
+    \end{gathered}
+\end{equation}
+
+\noindent
+where $N$ is the number of preceding (or succeeding) tokens. Since the $N$ preceding tokens\footnote{The most common use case for an N-gram model is
+    the preceding tokens.} are needed prior to the estimation of the probability of $w_i$, N-grams are sometimes referred to as left-associative
+models (in a left-to-right language).
+
+The probability of a token occurring anywhere in a sequence (referred to as $w_j$ in equations \ref{eq:ngram_preceding} and \ref{eq:ngram_succeeding})
+is determined statistically as the number occurrences of a token in a corpus divided by the total number of tokens in the corpus.
+
+\begin{equation}\label{eq:ngram_prob_wj}
+    P(w) = \frac{\textit{occurrences}(w)}{D_{tokens}}
+\end{equation}
+
+\noindent
+where $P(w)$ is the probability of a token, $w$, occurring anywhere in the corpus, $\textit{occurrences}(w)$ is the number of times of a specific
+token appears in the corpus, and $D_{tokens}$ is total number of tokens in the corpus, $D$. Computationally, the process of determining the
+probabilities of the individual tokens in the corpus is typically referred to as training or fitting the N-gram model.
+
+$N$ is a predetermined value based on model or application requirements and while it can technically take any value from one to infinity, its
+usability becomes very limited at the minimum sequence length in the corpus and unusable at any value at or above the maximum sequence length.
+Depending on implementation, the time and resources required to train a model with a large value of $N$ can be very expensive. The most common values
+for $N$ in the literature are one, two, and three, sometimes called unigrams, bigrams, and trigrams, respectively.
+
+\subsection{BERT}\label{sec:bert}
+The purpose of BERT (Bidirectional Encoder Representations from Transformers) is to represent tokens in text sequences in relation to the tokens
+preceding and following the token in question. Whereas an N-gram model represents token probabilities based on the $N$ tokens preceding a specific
+token or, conversely, the $N$ tokens succeeding said token, BERT models are able to predict the probability of a token based on both preceding and
+succeeding tokens, thus the bidirectionality novelty.
+
+\textbf{Masked Language Modeling}. The ability of BERT to represent text in a bidirectional context is achieved through the Masked Language Modeling
+(MLM) pretraining task~\cite{devlin_bert_2019}. MLM pretraining choses a certain percentage of token positions at random for prediction, then tokens
+are replaced with either a ``mask'' token, a random token, or the original token according to preconfigured probabilities (BERT used 80\% for the mask
+token, 10\% for random tokens, and 10\% for the original/unchanged token). The original token is then predicted with cross-entropy loss.
+
+\textbf{Next Sentence Prediction}. NSP is meant to support downstream tasks, while MLM conditions the model to represent tokens within sentences, NSP
+is meant to condition the model to understand relationships between sentences/sequences. Since BERTs training set consists primarily of literary works
+and excerpts from Wikipedia, the model simply predicts whether two sentences appear together (one after the other) in the original text. Instead of a
+single input sequence, the input consists of two sequences separated by a ``separation'' token and beginning with a ``classification'' token, these
+are represented by \lstinline|[SEP]| and \lstinline|[CLS]|, respectively in Devlin et al.~\cite{devlin_bert_2019}.
+
+For symmetry between pretraining tasks, the special tokens specifying the beginning and end of input sequences are shared between tasks. For both MLM
+and NSP objectives, input sequences always start with a classification token and end with a separation token (see figure \ref{fig:bert_mlm_input_example}).
+
+\begin{figure}[h!]
+    \includegraphics[width=\linewidth]{figures/BERT_MLM_input.png}
+    \caption{An example of an input sequence for a masked language modeling pretraining task. Modified from Devlin et al.~\cite{devlin_bert_2019}.}
+    \label{fig:bert_mlm_input_example}
+\end{figure}
+
+\noindent
+Likewise, NSP inputs always begin with a classification token and each input sequence is concatenated with a separation token terminating each
+sequence (see figure \ref{fig:bert_nsp_input_example}).
+
+\begin{figure}[h!]
+    \includegraphics[width=\linewidth]{figures/BERT_NSP_input.png}
+    \caption{An example of an input sequence for a next sentence prediction pretraining task.}
+    \label{fig:bert_nsp_input_example}
+\end{figure}
+
+\subsection{RoBERTa}\label{sec:roberta}
+As the name implies, RoBERTa, is an iteration on BERT (described in section \ref{sec:bert}, above). The results from
+Liu et al.~\cite{liu_roberta_2019} indicate that BERT was initially significantly undertrained, so by training longer, with larger batch sizes, on
+longer sequences, and dynamically changing the masking pattern applied to the MLM pretraining task, the performance of BERT was improved. The new
+checkpoint of this version of BERT was therefore called RoBERTa (Robustly Optimized BERT Pretraining Approach).
+
+The pretraining methodology introduced by RoBERTa eliminated the NSP pretraining task that was originally used in BERT and opted instead to increase
+the quantity of training data along with the batch sizes in addition to training the model longer. Lastly, there is a small difference in the
+construction of input sequences as compared to BERT; as with BERT, input sequences always begin with a classification token (\lstinline|[CLS]|) and
+individual segments are separated by a separation token (\lstinline|[SEP]|). Rather than having sequences end with a separation token as well, they
+are terminated with an end of sequence token (\lstinline|[EOS]|). See equation \ref{eq:roberta_input_example} modified from Liu et
+al.~\cite{liu_roberta_2019}, below.
+
+\begin{equation}\label{eq:roberta_input_example}
+    \mbox{[CLS], $x_1$, $x_2$, ..., $x_N$, [SEP], $y_1$, $y_2$, ..., $y_M$, [EOS]}
+\end{equation}
+
+% \section{Speech Recognition}\label{sec:speech_recognition}
+% Automatic speech recognition (ASR) or simply speech recognition is the process of translating spoken language to text i.e.~transcribing spoken
+% language. There have been a myriad of approaches to ASR problems, including, but not limited to rule-based, machine learning, and deep learning
+% methods. For the purposes of this work, we will focus primarily on deep learning approaches to ASR tasks with a specific focus on transcribing
+% spoken language in aviation (following from the data accumulated, processed, and analyzed in sections \ref{sec:data_source},
+% \ref{sec:data_processing}, and \ref{sec:data_analysis}).
+
+% Three ASR architectures were chosen due to their success in the conversational English domain and ease of reproducing results. Nvidia maintains a
+% Python package called Neural Modules (shortened to NeMo) with PyTorch implementations of many well-performing ASR models\footnote{Performance of these
+%     models can be seen in their respective papers (referenced hereafter).}, including the three models used in this work and described in more detail
+% below \cite{kuchaiev_nemo_2019}.
+
+% Although the internal processing of data varies between architectures, the format of the models input and output is generally the same. ASR models
+% have seen much success and performance gains through the use of Mel-frequency Cepstral Coefficients (MFCCs), so the audio files provided by the
+% corpora used in this work were processed and resampled into the same format, then the MFCCs are extracted according to the process
+% in~\cite{sahidullah_design_2012}. The MFCCs are then used as input to the ASR models. Each model outputs a probability distribution of characters or
+% tokens (depending on the vocabulary the ASR model is initialized with) over a segment of audio. All the ASR models used are trained with a
+% Connectionist Temporal Classification (CTC) loss function~\cite{graves_connectionist_2006}.
+
+% \subsection{Jasper}\label{sec:jasper}
+% Jasper is a family of end-to-end convolutional neural networks (CNNs) developed by NVIDIA to replace traditional ASR models that use separately
+% learned components for each stage of the pipeline (acoustic, pronunciation, language modeling, etc.) \cite{li_jasper_2019}.
+
+% Jasper models are designed with $B$ blocks and $R$ sub-blocks and are named accordingly as \textit{Jasper $B\mbox{x}R$} models. The Jasper
+% architecture also contains four extra convolutional blocks, one for preprocessing and three for postprocessing (see figure \ref{fig:jasper}). Each
+% convolutional sub-block contains one dimensional convolutions. Li et al.~trained and tested a Jasper 10x5 model that produced state-of-the-art
+% results on the LibriSpeech, Wall Street Journal, and Fisher+Switchboard speech corpora \cite{li_jasper_2019}.
+
+% \begin{figure}[h!]
+%     \includegraphics[width=\linewidth]{figures/jasper_vertical.png}
+%     \caption{Block diagram of the Jasper architecture. Copied from Li et al.~\cite{li_jasper_2019}.}
+%     \label{fig:jasper}
+% \end{figure}
+
+% \subsection{QuartzNet}\label{sec:quartznet}
+% The QuartzNet architecture is based on Jasper with some key modifications, namely, that the one dimensional convolutions are replaced with ``one
+% dimensional time-channel separable convolutions''. As with Jasper, QuartzNet models are denoted as \textit{QuartzNet $B\mbox{x}R$}, however, blocks
+% are repeated $S$ times and have a certain number of input and output channels, $c_{in}$ and $c_{out}$, respectively.
+
+% In Kriman et al.~experiments with QuartzNet were conducted on the LibriSpeech and Wall Street Journal speech corpora and achieved near state-of-the-art
+% performance on both. A transfer learning experiment was also performed (transferring between corpora). The model was initially trained on LibriSpeech
+% and Mozilla Common Voice, then fine-tuned on the Wall Street Journal corpus and also achieved near state-of-the-art results \cite{kriman_quartznet_2020}.
+% These results suggest that the QuartzNet architecture not only performs well on in-domain tasks, but can generalize and fine-tune well on related
+% data.
+
+% \begin{figure}[h!]
+%     \includegraphics[width=\linewidth]{figures/quartz_vertical.png}
+%     \caption{Block diagram of the QuartzNet architecture. Copied from Kriman et al.~\cite{kriman_quartznet_2020}.}
+%     \label{fig:quartznet}
+% \end{figure}
+
+% \subsection{Citrinet}\label{sec:citrinet}
+% Citrinet is an end-to-end CNN based on QuartzNet (and thus Jasper) and augmented with one dimensional Squeeze and Excitation modules
+% \cite{hu_squeeze-and-excitation_2018}. Citrinet models are denoted as \textit{Citrinet $B$x$R$x$C$} with $B$ representing the number of blocks, $R$
+% representing the number of repeated sub-blocks, and $C$ representing the number of channels. Similarly to Jasper and QuartzNet, Citrinet has a set of
+% prolog and epilog blocks, in this case one of each, the first and last-most blocks, respectively. The core of the model architecture consists of three
+% ``mega blocks'', which are themselves made up of seven QuartzNet blocks that are repeated $R$ times along with a Squeeze and Excitation module at the
+% end of the block. The value of $C$ follows from that of QuartzNet (section \ref{sec:quartznet}, above; $c_{in}\mbox{x}c_{out}$) \cite{majumdar_citrinet_2021}.
+
+% \begin{figure}[h!]
+%     \includegraphics[width=\linewidth]{figures/citrinet_vertical.png}
+%     \caption{Block diagram of the Citrinet architecture. Copied from Majumdar et al.~\cite{majumdar_citrinet_2021}.}
+% \end{figure}
+
+% Citrinet also uses a WordPiece tokenizer (described in more detail in section \ref{sec:wordpiece}), typically with a vocabulary size of 1024 tokens
+% (or word pieces) in the experiments performed by Majumdar et al. The results from the experiments showed model performance on ASR transcription tasks
+% near or better than the other state-of-the-art models that the \textit{Citrinet-1024} was compared against \cite{majumdar_citrinet_2021}.
+
+\section{Tokenization Algorithms}\label{sec:tokenizers}
 % present and explain the tokenization algorithms used (possibly which models they correspond to)
-Tokenization is the process of separating or segmenting samples into tokens. The rules by which the tokens are created or segmented depends entirely
-on the algorithm used. The simplest and most intuitive way to do this by creating tokens based on whitespace i.e., each ``word'' in the corpus
-is treated as a token\footnote{The Huggingface Tokenizer library implements this as the \lstinline|WordLevel| tokenizer.}. Although this is a simple
-and easy to understand method of tokenization, it does little to nothing to address the out-of-vocabulary problem that necessarily comes with any
-NLP tasks~\cite{wu_googles_2016}.
+Tokenization is the process of separating or segmenting words in sentences into their component ``tokens''. The tokenized version of a sentence is
+referred to as a sequence. The rules by which the tokens are created or segmented depend entirely on the algorithm used. The simplest and most
+intuitive way to do this by creating tokens based on whitespace i.e., each ``word'' in the corpus is  treated as a token\footnote{The Huggingface
+    Tokenizer library implements this as the \lstinline|WordLevel| tokenizer.}. While this is simple and works in theory, it does not take into
+account the fact that not every word in a language will occur in a corpus. The inevitable possibility for a tokenizer to see a word that was not in
+the original training corpus is known as the out-of-vocabulary (OOV) problem~\cite{wu_googles_2016}. The OOV problem has been addressed by several
+different algorithms. For brevity's sake, only the tokenization algorithms used by the models in this work are defined and explained.
 
-\subsection{WordPiece}
+\subsection{WordLevel}\label{sec:wordlevel}
+The WordLevel tokenization algorithm, mentioned briefly as an introductory example above, is the most technically simple algorithm among those
+introduced in this section. Tokens are extrapolated from words in a sentence based on the whitespace separating each word. This means that
+contractions, hyphenations, etc.~are treated as their own words i.e.~ ``you'', ``you're'', and ``you've'' are all seen as unique, distinct words that
+are tokenized and mapped to their own integral representations.
 
-\subsection{(Byte-Level) Byte-Pair Encoding}
+In order to map individual tokens to numerical values or IDs, an index of the words that appear in a corpus must be created. The numerical values for
+each token are derived from this index. For the WordLevel algorithm, this process is referred to as the training the tokenizer model.
+
+The first step of the training procedure is to index any specified special tokens, such as mask, classification, or separation tokens (this is why
+special tokens often have low-valued token IDs). The second step is to split the sentences in the training data into individual words. This is done
+by splitting each string based on whitespace, in other words the tokenizer treats sentences as lists of words delimited by space characters. Lastly,
+the tokenizer iterates through each word in each sentence, indexing new words as it finds them, until all sentences in the training corpus have been
+analyzed. Token IDs are then derived based on the position of the word in the tokenizer index or vocabulary.
+
+\begin{algorithm}
+    \caption{Training procedure for the WordLevel tokenizer.}
+    \label{alg:wordlevel_training}
+    \begin{algorithmic}
+        \State $D = \{s_1, s_2, ..., s_N\}$
+        \Comment{$D$, $s_i$, and $N$ are corpus, sentence, and corpus length}
+        \State $V \gets \{\varnothing\}$
+        \Comment{$V$ is the tokenizer vocabulary and is initialized to the empty set}\\
+
+        \For{$s \in D$}
+        \State Split $s$ into words using whitespace as a delimiter such that
+        \State $W=\{w_1, w_2, ... w_M\}$
+        \Comment{$w$ and $M$ are words and sentence length}\\
+
+        \For{$w \in W$}
+        \If{$w \notin V$}
+        \State $V \gets V;w$
+        \Comment{Add $w$ to the vocabulary}
+        \EndIf
+        \EndFor
+        \EndFor
+    \end{algorithmic}
+\end{algorithm}
+
+Table \ref{tab:wordlevel_tokenization_example} shows the different representations of the input sentence over the course of the tokenization process.
+
+\begin{table}[h!]
+    \centering
+    \begin{tabular}{l l}
+        \toprule
+        Tokenization stage & Representation                                                                            \\
+        \midrule
+        Input string:      & ``the quick brown fox jumps over the lazy dog''                                           \\
+        Tokenized string:  & [``the'', ``quick'', ``brown'', ``fox'', ``jumps'', ``over'', ``the'', ``lazy'', ``dog''] \\
+        Integral mapping:  & [5, 16, 9, 12, 13, 15, 5, 14, 11]                                                         \\
+        \bottomrule
+    \end{tabular}
+    \caption{A simple example of the tokenization process of the WordLevel tokenization algorithm.}
+    \label{tab:wordlevel_tokenization_example}
+\end{table}
+
+\noindent
+The simplicity of the algorithm speaks for itself and serves as a good introduction to tokenization as well as an illustrative example of the OOV
+problem. Note that the values of the integers representing the tokens are determined based on the order in which the tokens are seen during the
+training process (special tokens are defined and therefore seen first), so these results are completely reproducible as long as the data appears in
+the same order.
+
+The tokenization of words with shared roots but different suffixes is shown in table \ref{tab:wordlevel_tokenization_shared_roots_example}.
+
+\begin{table}[h!]
+    \centering
+    \begin{tabular}{l l}
+        \toprule
+        Tokenization stage & Representation                    \\
+        \midrule
+        Input string:      & ``you you're you've''             \\
+        Tokenized string:  & [``you'', ``you're'', ``you've''] \\
+        Integral mapping:  & [18, 19, 20]                      \\
+        \bottomrule
+    \end{tabular}
+    \caption{The results of tokenizing three words with the same roots and differing suffixes.}
+    \label{tab:wordlevel_tokenization_shared_roots_example}
+\end{table}
+
+\noindent
+Clearly all three words are seen as independent and unique words by the tokenizer, since they are mapped to their own integer representations.
+
+The tokenizer used to generate the examples in tables \ref{tab:wordlevel_tokenization_example} and
+\ref{tab:wordlevel_tokenization_shared_roots_example} was trained on two sentences, notably lacking the word ``you'll''. Running the tokenizer on the
+word ``you'll'' alone results in the following output:
+
+\begin{table}[h!]
+    \centering
+    \begin{tabular}{l l}
+        \toprule
+        Tokenization stage & Representation \\
+        \midrule
+        Input string:      & ``you'll''     \\
+        Tokenized string:  & [``[UNK]'']    \\
+        Integral mapping:  & [0]            \\
+        \bottomrule
+    \end{tabular}
+    \caption{Results of running the WordLevel tokenizer on an unknown word similar to some of those in the training data.}
+    \label{tab:wordlevel_unk_word}
+\end{table}
+
+\noindent
+Although the word shares a root with some of the words in the training data (``you'', ``you're'', and ``you've'') and is semantically very similar,
+the tokenizer does not recognize the word at all (represented by the ``[UNK]'' or unknown token). This word would be considered out-of-vocabulary
+since it did not appear in the training data and thus reveals one of the major downsides of this algorithm, as mentioned in section
+\ref{sec:tokenizers}. This problem can be alleviated to some extent by modifying the preprocessing strategy of the tokenizer to partition contracted
+words into their component parts e.g.~the word ``you're'' would be partitioned to \lstinline|[``you'', ``''', ``re'']|, however, it eventually
+reappears, for example, when using the past, present, and future tenses of a verb (e.g.~``go'' and ``goes'' in the training data, then encountering
+the future tense: ``going'').
+
+\subsection{WordPiece}\label{sec:wordpiece}
+% Wu et al. 2016; Schuster and Nakajima 2012
+The WordPiece algorithm was developed by Schuster and Nakajima in 2012~\cite{schuster_japanese_2012} as a word segmentation algorithm for language
+modeling in Japanese and Korean voice search systems. It is further used by Wu et al.~\cite{wu_googles_2016} for neural machine translation tasks. The
+algorithm demonstrated proficiency and increased performance for the models used in both tasks and effectively addressed the OOV problem.
+
+One of the primary stipulations for this algorithm is effective and efficient handling of OOV words such that none are produced during normal
+operation. To achieve this, during training, the vocabulary of the tokenizer is initialized to a basic set of characters; since WordPiece is designed
+to be language agnostic, the initial vocabulary is specified as all basic unicode characters (for the chosen language) in addition to all ASCII
+characters.
+
+The training procedure works by iterating over the vocabulary, combining two word units to maximize the likelihood over the training data, and
+repeating until one of two stop conditions are reached:
+
+\begin{enumerate}
+    \item The predefined word limit is reached
+    \item The increase in likelihood falls below the predefined threshold
+\end{enumerate}
+
+\noindent
+The training procedure is simple, but computationally expensive. Schuster and Nakajima calculated the time complexity of the algorithm at $O(K^2)$
+where $K$ is the current size of the vocabulary~\cite{schuster_japanese_2012}. Due to the high time complexity of the algorithm, the authors suggest
+the following considerations to reduce the computational complexity:
+
+\begin{itemize}
+    \item Test only pairs that actually exist in the training data
+    \item Test only pairs with a significant chance of being best (high prior likelihood)
+    \item Combine several clustering steps into a single iteration (for groups of pairs which don't affect each other)
+    \item Only modify language model counts for the affected entries
+\end{itemize}
+
+Table \ref{tab:wordpiece_tokenization_example}, below, shows an example of a WordPiece tokenizer model output on a sentence that was seen during the
+training procedure (trained on the same data as the WordLevel tokenizer in section \ref{sec:wordlevel}).
+
+\begin{table}[h!]
+    \centering
+    \begin{tabular}{l l}
+        \toprule
+        Tokenization stage & Representation                                                                            \\
+        \midrule
+        Input string:      & ``the quick brown fox jumps over the lazy dog''                                           \\
+        Tokenized string:  & [``the'', ``quick'', ``brown'', ``fox'', ``jumps'', ``over'', ``the'', ``lazy'', ``dog''] \\
+        Integral mapping:  & [59, 98, 95, 88, 97, 91, 59, 90, 87]                                                      \\
+        \bottomrule
+    \end{tabular}
+    \caption{A simple example of the tokenization process of the WordPiece algorithm.}
+    \label{tab:wordpiece_tokenization_example}
+\end{table}
+
+\noindent
+Since this sentence was in the training data, the segmentation is identical to that of the WordLevel tokenizer in table
+\ref{tab:wordlevel_tokenization_example}, however, tokenization of similar words e.g.~words with shared roots or similar contractions are handled
+differently as in table \ref{tab:wordpiece_tokenization_shared_roots_example}.
+
+\begin{table}[h!]
+    \centering
+    \begin{tabular}{l l}
+        \toprule
+        Tokenization stage & Representation                                            \\
+        \midrule
+        Input string:      & ``you you're you've''                                     \\
+        Tokenized string:  & [``you'', ``you'', ``''', ``re'', ``you'', ``''', ``ve''] \\
+        Integral mapping:  & [56, 56, 5, 71, 56, 5, 72]                                \\
+        \bottomrule
+    \end{tabular}
+    \caption{Example of WordPiece tokenization of words with shared roots and differing suffixes.}
+    \label{tab:wordpiece_tokenization_shared_roots_example}
+\end{table}
+
+\noindent
+Since each word has a shared root word in the contractions (as well as an apostrophe separating the contraction), the tokenizer segments each word and
+represents them as groups of small, more common (thus with higher probabilities of occurring) tokens. Additionally, the WordPiece algorithm
+effectively handles OOV words. Take the OOV in table \ref{tab:wordpiece_unk_word_similar} with a shared root and similar contraction as those in table
+\ref{tab:wordpiece_tokenization_shared_roots_example}, which appear in the training data.
+
+\begin{table}[h!]
+    \centering
+    \begin{tabular}{l l}
+        \toprule
+        Tokenization stage & Representation                     \\
+        \midrule
+        Input string:      & ``you'll''                         \\
+        Tokenized string:  & [``you'', ``''', ``l'', ``\#\#l''] \\
+        Integral mapping:  & [56, 5, 17, 54]                    \\
+        \bottomrule
+    \end{tabular}
+    \caption{WordPiece tokenization of an unknown word similar to some of those in the training data.}
+    \label{tab:wordpiece_unk_word_similar}
+\end{table}
+
+\noindent
+Since the base of the contraction, ``you'', occurs in the training data as well as the apostrophe in the contraction, so the tokenizer segments those
+parts of the word as tokens. The last part of the word, ``ll'', does not appear at all in the training data, so it is broken down into its component
+characters which are represented as ``l'' (beginning of a word) and ``\#\#l'' (segmented part of a word that should be concatenated to the previous
+token when they are recombined). Comparing this to the results of the WordLevel tokenizer output on the same word in table
+\ref{tab:wordlevel_unk_word}, the WordPiece algorithm sidesteps the OOV problem for words similar to those in the trining corpus and, according to
+the output of the WordPiece tokenizer in table \ref{tab:wordpiece_unk_word}, for words unlike those in the training corpus.
+
+\begin{table}[h!]
+    \centering
+    \begin{tabular}{l l}
+        \toprule
+        Tokenization stage & Representation                                                             \\
+        \midrule
+        Input string:      & ``anything''                                                               \\
+        Tokenized string:  & [``an'', ``\#\#y'', ``\#\#t'', ``\#\#h'', ``\#\#i'', ``\#\#n'', ``\#\#g''] \\
+        Integral mapping:  & [61, 49, 37, 53, 32, 36, 39]                                               \\
+        \bottomrule
+    \end{tabular}
+    \caption{WordPiece tokenization of a word that does not appear in the training data and is not similar to any of the words in the tokenizer
+        vocabulary.}
+    \label{tab:wordpiece_unk_word}
+\end{table}
+
+\subsection{Byte-Pair Encoding}\label{sec:bpe}
+% Sennrich et al. 2016; Gage 1994
+The Byte-Pair Encoding (BPE) tokenizer algorithm was adapted from a data compression algorithm by Gage 1994~\cite{gage_feb94_1994}. The data compression
+algorithm works by replacing common pairs of bytes in data with an unused byte to reduce the overall size of the data. The tokenization algorithm
+works under the same principle, merging frequent pairs of characters or tokens into one token~\cite{sennrich_neural_2016}.
+
+The vocabulary of the tokenizer is initialized to the base set of characters (letters, numbers, punctuation, etc.~). The training procedure begins by
+counting all pairs (``A'', ``B'') of characters that appear in the training data and combines the most frequently occurring pair into one token
+(``A'', ``B'' $\rightarrow$ ``AB''). This process repeats until a specified vocabulary size has been reached (or a specified number of merge
+operations have occurred; the final vocabulary size is equal to the base character set plus the number of merge operations). Algorithm
+\ref{alg:bep_training} shows a minimal Python implementation for the BPE training procedure.
+
+\begin{algorithm}[h!]
+    \caption{BPE training algorithm implementation in Python. Modified from Sennrich et al.~\cite{sennrich_neural_2016}.}
+    \label{alg:bep_training}
+    \begin{lstlisting}[language=Python]
+import re, collections
+
+def get_stats(vocab):
+    pairs = collections.defaultdict(int)
+    for word, freq in vocab.items():
+        symbols = word.split()
+        for i in range(len(symbols)-1):
+            pairs[symbols[i], symbols[i+1]] += freq
+        return pairs
+def merge_vocab(pair, v_in):
+    v_out = {}
+    bigram = re.escape(' '.join(pair))
+    p = re.compile(r'(?<!\S)' + bigram + r'(?!\S)')
+    for word in v_in:
+        w_out = p.sub(''.join(pair), word)
+        v_out[w_out] = v_in[word]
+    return v_out
+
+vocab = {...}
+num_merges = 10
+for i in range(num_merges):
+    pairs = get_stats(vocab)
+    best = max(pairs, key=pairs.get)
+    vocab = merge_vocab(best, vocab)
+    print(best)
+    \end{lstlisting}
+\end{algorithm}
+
+Finally, comparing the tokenization process to the WordLevel and WordPiece algorithms, we can see that the outputs of the algorithm are nearly
+identical when the sentences being tokenized have words that appear in the training data (i.e.~the BPE outputs in tables
+\ref{tab:bpe_tokenization_example} and \ref{tab:bpe_shared_roots_example} are almost identical to the WordPiece output in tables
+\ref{tab:wordpiece_tokenization_example} and \ref{tab:wordpiece_tokenization_shared_roots_example} with the exception of the numerical representation
+of the tokens) and very similar to that of the WordLevel tokenizer.
+
+\begin{table}[h!]
+    \centering
+    \begin{tabular}{l l}
+        \toprule
+        Tokenization stage & Representation                                                                            \\
+        \midrule
+        Input string:      & ``the quick brown fox jumps over the lazy dog''                                           \\
+        Tokenized string:  & [``the'', ``quick'', ``brown'', ``fox'', ``jumps'', ``over'', ``the'', ``lazy'', ``dog''] \\
+        Integral mapping:  & [36, 70, 72, 66, 74, 69, 36, 68, 64]                                                      \\
+        \bottomrule
+    \end{tabular}
+    \caption{The Byte-Pair Encoding tokenizer output for a string that appears in the training data.}
+    \label{tab:bpe_tokenization_example}
+\end{table}
+
+\begin{table}[h!]
+    \centering
+    \begin{tabular}{l l}
+        \toprule
+        Tokenization stage & Representation                                            \\
+        \midrule
+        Input string:      & ``you you're you've''                                     \\
+        Tokenized string:  & [``you'', ``you'', ``''', ``re'', ``you'', ``''', ``ve''] \\
+        Integral mapping:  & [34, 34, 5, 33, 34, 5, 37]                                \\
+        \bottomrule
+    \end{tabular}
+    \caption{Byte-Pair Encoding tokenizer output for semantically similar words with shared roots.}
+    \label{tab:bpe_shared_roots_example}
+\end{table}
+
+\noindent
+Now, testing the BPE algorithm on words that did not appear in the training data, it is clear that the BPE algorithm effectively deals with and
+processes OOV words (tables \ref{tab:bpe_unk_similar} and \ref{tab:bpe_unk}), even when no subunits of the word appear in the training data (table
+\ref{tab:bpe_unk}).
+
+\begin{table}[h!]
+    \centering
+    \begin{tabular}{l l}
+        \toprule
+        Tokenization stage & Representation                 \\
+        \midrule
+        Input string:      & ``you'll''                     \\
+        Tokenized string:  & [``you'', ``''', ``l'', ``l''] \\
+        Integral mapping:  & [34, 5, 17, 17]                \\
+        \bottomrule
+    \end{tabular}
+    \caption{Byte-Pair Encoding tokenizer output for a word that did not appear in the training data, but is similar to some of the words in the
+        training data.}
+    \label{tab:bpe_unk_similar}
+\end{table}
+
+\begin{table}[h!]
+    \centering
+    \begin{tabular}{l l}
+        \toprule
+        Tokenization stage & Representation                                     \\
+        \midrule
+        Input string:      & ``anything''                                       \\
+        Tokenized string:  & [``an'', ``y'', ``t'', ``h'', ``i'', ``n'', ``g''] \\
+        Integral mapping:  & [39, 30, 25, 13, 14, 19, 12]                       \\
+        \bottomrule
+    \end{tabular}
+    \caption{Byte-Pair Encoding tokenizer output for a word that did not appear in the training data and is not similar to any words in the training
+        data.}
+    \label{tab:bpe_unk}
+\end{table}
+
+\textbf{Byte-Level Byte-Pair Encoding}. This is a subset of the BPE algorithm that is also commonly used for NLP tasks such as neural machine
+translation, language modeling, and generative tasks (GPT-2 uses a byte-level BPE tokenizer~\cite{radford_language_2019}). As Huggingface's
+\lstinline|tokenizers| library implements it, this is a preprocessing step for the tokenizer model that maps all bytes in a string/sentence to their
+own unique and visible character. Functionally, byte-level BPE is nearly identical to standard BPE models with the main exception being that the
+beginning of word symbol is actually rendered and represented by a visible character in the tokenized sequence (see table
+\ref{tab:byte_level_bpe_example}).
+
+\begin{table}[h!]
+    \centering
+    \begin{tabular}{l l}
+        \toprule
+        Tokenization stage & Representation                                            \\
+        \midrule
+        Input string:      & ``anything''                                              \\
+        Tokenized string:  & [``Ġa'', ``n'', ``y'', ``t'', ``h'', ``i'', ``n'', ``g''] \\
+        Integral mapping:  & [36, 19, 30, 25, 13, 14, 19, 12]                          \\
+        \bottomrule
+    \end{tabular}
+    \caption{Example of the output of a byte-level Byte-Pair Encoding tokenizer with the begging of a word represented with the ``Ġ'' character.}
+    \label{tab:byte_level_bpe_example}
+\end{table}
+
+\section{Discussion}\label{sec:discussion}
+Considering the success of deep learning models with transformer-based architectures for language modeling and natural language processing (NLP)
+tasks, more generally, in the conversational and literary English domains, it follows that transformer architectures should see widespread use in the
+domain of aviation English for domain-specific NLP tasks, however, according to the literature, transformer-based language models have seen sparse
+use. The most notable example of the use of a transformer deep neural network is the application of BERT for knowledge extraction from notice to
+airmen (NOTAM) messages \cite{arnold_knowledge_2022}.
+
+\subsection{Language Model Pretraining}\label{sec:lm_pretraining}
+A standard masked language modeling training procedure is used to pretrain the language models, as described in section \ref{sec:bert}. Tokens are
+masked at random with at a uniform probability of 0.2. Input sequences are padded to the maximum length allowed by the model and truncated to the
+maximum length, if greater. The maximum sequence length for BERT and RoBERTa models is 512 tokens \cite{devlin_bert_2019,liu_roberta_2019}. Based on
+the data analysis in section \ref{sec:data_analysis}, there are few, if any, samples that meet or exceed the maximum sequence length for BERT and
+RoBERTa models.
+
+The models are trained with the \textit{AdamW} optimizer \cite{loshchilov_decoupled_2019} using the hyperparameters in table \ref{tab:optim_params}
+based on the most effective pretraining configuration presented in \cite{devlin_bert_2019} and \cite{liu_roberta_2019}.
+
+\begin{table}[h!]
+    \centering
+    \begin{tabular}{l r}
+        \toprule
+        Hyperparameter & Value              \\
+        \midrule
+        $\beta_1$      & $0.9$              \\
+        $\beta_2$      & $0.98$             \\
+        $\epsilon$     & $1 \times 10^{-6}$ \\
+        Learning Rate  & $4 \times 10^{-5}$ \\
+        Weight Decay   & $0.01$             \\
+        \bottomrule
+    \end{tabular}
+    \caption{\textit{AdamW} hyperparameters for training BERT and RoBERTa models.}
+    \label{tab:optim_params}
+\end{table}
+
+A train/test split of 70\% and 30\% and a validation split of 10\% of the training data was used for each model. The blank models (the models that are
+trained from randomly initialized weights; also referred to as being trained from scratch) were trained for 200 epochs and the pretrained models were
+trained for 100 epochs. These values were chosen through trial and error, monitoring the perplexity and loss curves of the
+validation set to ensure that the pretraining does not result in net negative effects on the loss and perplexity of the model (i.e.~to ensure that the
+models are not overfitting to the training data). Batch sizes of 16 were used for training all four models; this was the maximum batch size that
+worked reliably with the hardware available to train the models\footnote{Values larger than 16 would sometimes cause CUDA Out of Memory errors, so 16
+    was used to ensure all models trained continuously without error and/or interruption.}.
+
+\begin{table}
+    \centering
+    \begin{tabular}{l r r}
+        \toprule
+        Corpus split & Split size (\%) & Samples \\
+        \midrule
+        Train        & 68.99           & 34,470  \\
+        Validation   & 6.999           & 3,830   \\
+        Test         & 30.01           & 16,415  \\
+        \bottomrule
+    \end{tabular}
+    \caption{Number of samples and percentage of overall corpus by split.}
+    \label{tab:corpus_splits}
+\end{table}
+
+As mentioned previously, perplexity and cross-entropy loss (often shortened to loss) are used to track the performance of the model on the training
+and validation sets, per epoch, while training. Cross-entropy loss is defined by the following equation:
+
+\begin{equation}\label{eq:cross_entropy_loss}
+    \begin{gathered}
+        \mathcal{L}(x, y) = \{\mbox{$l_1$, $l_2$, ..., $l_N$}\}^T\\
+        l_n = -\log \left[\frac{\exp(x_{n,y_n})}{\sum_{c=1}^{C}\exp(x_{n,c})}\right]
+    \end{gathered}
+\end{equation}
+
+\noindent
+Where $x$ and $y$ are $N \times 1$ vectors representing the inputs and labels, respectively, for each sample in the batch. $N$ is the number of
+samples in the batch, $l_n$ is the loss for an individual sample, $n$, in the batch, and $C$ is the number of class labels. For the purpose of
+language modeling, the set of class labels is equivalent to the vocabulary of the tokenizer (and thus the language model). Numerical labels will
+correspond to token IDs.
+
+Perplexity is related to the probability of a token appearing in a sequence. It is used as a measure of how well a language model can predict tokens
+in a sequence and is defined by the equation below:
+
+\begin{equation}\label{eq:perplexity}
+    \begin{gathered}
+        X = \{x_1, x_2, ..., x_N\}\\
+        \mbox{PPL} = \exp\left[\frac{\sum_{i=1}^N - \log_2 x_i}{N}\right]
+    \end{gathered}
+\end{equation}
+
+\noindent
+Where $X$ is an input sequence, $N$ is the length of the sequence, and $x_i$ is an individual token in the sequence, $X$.
+
+With the metrics above defined, an analysis of the results of the model pretraining can be performed. The objective of the pretraining procedure is to
+build the models' understanding of the structure of the text, in this case transcripts of air-traffic communications. By tracking the average
+cross-entropy loss (loss) and perplexity (PPL) during training, we can estimate how well (and how correctly) the models are predicting tokens and if
+the models are effectively ``understanding'' token placements in sequences.
+
+Graphs of the training loss (average loss over the samples/batches in the training partition of the corpus) are presented in figure
+\ref{fig:training_loss}.
+
+\begin{figure}[h!]
+    \includegraphics[width=\linewidth]{figures/training_loss.png}
+    \caption{Loss on the training partition of the corpus over time (steps).}
+    \label{fig:training_loss}
+\end{figure}
+
+\noindent
+Both classes of models (scratch and pretrained) quickly reach a steady state in terms of loss although the initial loss values are significantly
+different. It is notable that the pretrained models start with a significantly lower loss, about 1.3, compared to models trained from scratch, which
+started with a loss value of just over 5.0. This initially indicates that the models are training well and beginning to converge on either local or
+global minima.
+
+The perplexity of the models on the training data yields interesting results (figure \ref{fig:training_ppl}). Although the pretrained models have a
+more stable curve over time, the models trained from scratch result in a significantly lower perplexity. Note the difference in the scales on the
+y-axis in figure \ref{fig:training_ppl}; the final perplexity value of each language model trained from scratch is at least two orders of magnitude
+lower than that of the pretrained models.
+
+\begin{figure}[h!]
+    \includegraphics[width=\linewidth]{figures/training_ppl.png}
+    \caption{Language model perplexity on the training partition of the corpus over time (steps). Note the scales on the y axes.}
+    \label{fig:training_ppl}
+\end{figure}
+
+\noindent
+At the end of training, the BERT and RoBERTa models trained from scratch had perplexity values of $0.52\times 10^7$ and $0.16\times 10^7$,
+respectively, whereas the pretrained models reached perplexity values of $0.29\times 10^9$ and $0.17\times 10^9$, for BERT and RoBERTa, respectively.
+This indicates that best language model performance may be achieved by pretraining models on in-domain data, then continuing to fine-tune on specific
+downstream tasks. This is a direct contradiction of the current recommended practice of using one of the pretrained checkpoints from Devlin et
+al.~\cite{devlin_bert_2019} and Liu et al.~\cite{liu_roberta_2019} then fine-tuning on specific downstream tasks. This is further upheld by the
+results of the model performance on the test partition of the corpus in table \ref{tab:pretraining_results}, below.
+
+\begin{table}[h!]
+    \centering
+    \begin{tabular}{l r r}
+        \toprule
+        Model                & Loss  & Perplexity        \\
+        \midrule
+        BERT (scratch)       & 10.54 & $0.14\times 10^7$ \\ % 1435064.875
+        RoBERTa (scratch)    & 11.41 & $1.04\times 10^7$ \\ % 10376598.000
+        BERT (fine-tuned)    & 11.00 & $0.01\times 10^9$ \\ % 13805354.000
+        RoBERTa (fine-tuned) & 11.80 & $0.01\times 10^9$ \\ % 6816011.500
+        \bottomrule
+    \end{tabular}
+    \caption{Results on the test set of data after pretraining the BERT and RoBERTa models from scratch and pretrained checkpoints. Loss and
+        perplexity values are averaged over the test partition of the corpus. For both perplexity and loss; lower values are better.}
+    \label{tab:pretraining_results}
+\end{table}
+
+\noindent
+As with the training data, the perplexity of the language models trained from scratch over the test partition of the corpus is significantly lower
+than that of the pretrained language models. The loss of each model over the test set is somewhat high, but they are around the same values for each
+model, which is consistent with the loss statistics during training. A higher loss value on unseen (test) data is normal and expected. The BERT model
+trained from scratch had the best performance among the models tested despite both RoBERTa models training better within their model classes. Within
+the class of pretrained models, however, the RoBERTa model still technically performed better than the BERT model (values shown in table
+\ref{tab:pretraining_results} were rounded to three significant figures) with a perplexity value of $0.007\times 10^9$ compared to BERT's
+$0.014\times 10^9$.
+
+With the results and performance of the models established relative to each other, the high perplexity of each model must be addressed. While the loss
+values are converging nicely during training, the perplexity is somewhat erratic and consistently high, rarely converging to a consistent curve,
+although it is exhibiting a clear trend as training progresses. Adjusting the hyperparameters of the optimizer did not solve this issue and changing
+the hyperparameters of the model (e.g.~sequence length, number of hidden layers, attention heads, etc.~) was not desirable for doing a direct
+comparison between the pretrained checkpoints and the base models to study the effectiveness of the models on orthographic transcriptions of aviation
+English. It is difficult to determine what exactly is causing this phenomenon, so several theories were developed in an attempt to explain it (some
+of which will be elaborated upon in section \ref{sec:future_work}):
+
+\noindent
+\textbf{Small volume of training data}. Especially when compared to \textit{Attention is All You Need}, BERT, and RoBERTa,
+\cite{vaswani_attention_2017,devlin_bert_2019,liu_roberta_2019} the number of samples used in the training data dwarfs that of the state-of-the-art.
+It has been exhaustively demonstrated in the past that neural networks perform better with large volumes of data. The inception of RoBERTa itself is
+based on the idea that BERT was not trained on enough data \cite{liu_roberta_2019}. Increasing the volume of in-domain data used for pretraining
+(and training) would be the most effective way to immediately improve the performance of the models.
+
+\noindent
+\textbf{Different forms of English}. The BERT and RoBERTa pretraining checkpoints were created by training the models on literary works and typed
+articles such as those from \textit{Wikipedia}. The data used in this work are orthographic transcriptions of aviation English and while the two data
+sources share the common root of English, transfer learning from literary English to aviation English yields limited results, as evidenced above.
+Especially with relatively small amounts of data.
+
+\noindent
+\textbf{Mixed regional data sources}. As described in section \ref{sec:data_source}, the data aggregated for use here is comes from several different
+regions, namely international airports in the US \cite{godfrey_air_1994} and international and domestic airports in Europe
+\cite{smidl_air_2019,hofbauer_atcosim_2008,szoke_detecting_2021}. During data processing it was ensured that only English transcriptions were included
+in the corpus, however, proper nouns from each region will be present in the data regardless. Additionally, dialectical differences and varying levels
+of English proficiency will affect the structure of token sequences across regions. All of these facets combined may be contributing to the high
+perplexity of the language models relative to other works. It should be noted that the varied regional sources of data are not necessarily a bad
+thing. Depending on the application and desired robustness of the model, the ability to interpret international sources and contexts can be a
+desirable trait for model selection, however, it will require significantly more data and proportionally balanced regional contexts within the corpus
+in order for the pretraining process to be effective for downstream tasks.
+
+Up to this point we have followed a procedure for studying the effectiveness of the models that is somewhat symmetrical to that of BERT and RoBERTa,
+so at this point the models should be trained for downstream tasks in order to further measure the effectiveness of the pretraining procedures.
+In the aviation domain and for this particular data, that would mean training for tasks such as callsign identification, speaker identification,
+flight phase classification, clearance recognition, etc., however, labels for these tasks are not present in the corpora used here. Corpora with
+relevant labels for downstream tasks could not be obtained within an appropriate amount of time to run and include the experiments in this work.
+Some potential algorithms were developed in an attempt to mine labels from the corpora aggregated for this work which will be expanded upon in section
+\ref{sec:future_work}. For these reasons, testing model performance and effectiveness on downstream tasks must be left for future work.
+
+\section{Conclusion}\label{sec:conclusion}
+Four aviation English corpora are aggregated and unified for use in neural language modeling. A masked language modeling objective is used pretrain
+two classes of BERT and RoBERTa models (models trained from scratch and pretrained models transfer learned to aviation English) on the aggregated
+data.
+
+Contrary to the recommended best practice of using pretrained checkpoints and training for downstream tasks, we find that, in the domain of aviation
+English, models achieve better performance when pretrained from scratch than model checkpoints transfer learned from the available BERT and RoBERTa
+pretrained checkpoints. The BERT and RoBERTa models trained from scratch, with parameters identical to the pretrained models, reached perplexity
+values nearly two orders of magnitude lower than their corresponding pretrained checkpoints on the training partition of the corpus. Additionally,
+the BERT and RoBERTa models trained from scratch consistently outperformed the corresponding pretrained models on the test partition of the corpus,
+with BERT performing the best overall.
+
+The results from the experiments in this work indicate that, for natural language processing tasks in the aviation domain, models pretrained from
+scratch are more effective than transfer learning pretrained models. Thus, models trained from scratch should fine-tuned and used for downstream tasks
+over pretrained models. The most significant limiting factor to the performance of the models is the small volume of aviation English data available
+for pretraining and downstream task use.
+
+\section{Future Work}\label{sec:future_work}
+This section is intended to serve as a brief proposal or set of suggestions for future work based on the results of this work. The main focus of these
+suggestions are corpus specific algorithms and data augmentation techniques to enable testing and experiments to determine the effectiveness of the
+pretraining procedures covered earlier in this thesis.
+
+\textbf{Next Sentence Prediction (NSP)}. This training procedure was introduced by Devlin et al.~\cite{devlin_bert_2019} as a pretraining procedure
+for BERT and is described in more detail in section \ref{sec:bert}. This can be used on top of masked language modeling for language model
+pretraining, however, it requires a knowledge of which sequences in the corpus appear next to each other. For pretraining transformer-based neural
+language models for the aviation domain (with orthographic transcriptions), this same procedure can be used to classify whether two transmissions
+occur in the correct order. In other words; classifying a pilots response to an air-traffic controller directive or an air-traffic controllers
+response to a pilot query. Of the four corpora used in this work, ATCC is the only corpus that can be mined for an NSP-like pretraining procedure
+since the raw transcripts have indicators for time (start and end indicators), transmitting party, and intended receiver. With this information we can
+estimate the correct order of transmissions between two parties to a reasonable degree of certainty (either manually or programmatically).
+
+\textbf{Callsign Identification}. This is the process of identifying an aircraft callsign from the transcription of the spoken callsign. This is a
+popular subtask for automatic speech recognition applications in aviation, but can also be accomplished by postprocessing speech recognition outputs
+through a language model. ATCC is a good candidate corpus for this task. There are two potential methods for implementing this as a downstream task:
+(1) In transmissions from air-traffic controllers to specific aircraft, use the receiver (``TO'' label) as the label for each sequence or (2) Identify
+the sequence of tokens, in each transmission, that correspond to the callsign of the aircraft and use those tokens as the labels for each sequence.
+Each of these methods bring their own problems. The first method effectively creates an ever expanding classification problem with infinitely many
+labels, so some kind of tokenization or decoding scheme would need to be created to address this. The second method requires manually labeling every
+sequence in the corpus, which may not be feasible in all cases. Lastly, both methods may lack sufficient context to correctly predict aircraft
+callsigns. For example, general aviation (GA) callsigns in North America typically start with the letter \textit{N} followed by a series of numbers
+and/or letters, however, the spoken callsign is based on aircraft manufacturer (e.g.~Cessna, Cirrus, Gulfstream, etc.~), so the spoken callsign will
+begin with the manufacturer name instead of ``november''.
+
+\textbf{Speaker Identification}. Generally, speaker identification is a task to classify specific speakers in a series of communications. This is
+usually a subtask that falls under automatic speech recognition, but it is also feasible for language models under some circumstances as well. In
+aviation this could be a task as simple as identifying the side of the communication i.e.~an air-traffic controller or a pilot or it can be complex
+enough to be grouped under callsign identification tasks. In the simplest form, this can be accomplished easily using ATCC since controller positions
+are labeled along with transmitting/receiving aircraft.
+
+\textbf{Speech Recognition}. Finally, speech recognition system performance can be improved using a sufficiently trained language model. This approach
+has been used by several state-of-the-art automatic speech recognition models to improve sequence decoding and predictions. NVIDIA's Citrinet and
+QuartzNet models, for example, both use this approach to reduce their relative word error rates compared to the performance of other models on the
+same corpora \cite{majumdar_citrinet_2021,kriman_quartznet_2020}. Considering the relatively high word error rates of most automatic speech
+recognition models in the aviation domain, the pretraining procedures introduced here could be used to further reduce the error rates of these models.
 

### 2023-11-29 thesis/MSECE_thesis.tex
Final version of thesis/refs.

@@
 
-\section*{Abstract}\label{sec:abstract}
+\section{Abstract}\label{sec:abstract}
 
 \section{Introduction}\label{sec:introduction}
-There have been noticeable and significant advancements in the general domain of artificial intelligence (AI) recently. Automatic speech recognition and
-language modeling technologies, in particular, have made tremendous advances over the past two decades achieving performance levels good enough to
+There have been noticeable and significant advancements in the general domain of artificial intelligence (AI) recently. Natural language processing
+and language modeling technologies, in particular, have made tremendous advances over the past two decades achieving performance levels good enough to
 be brought to consumer facing products such as personal assistants, search engines, phones, televisions, etc. Due to the success in the general
 domain, applications have begun to spin-off into problem-specific domains, such as (and most notably for this work) aviation. For example, several
-speech recognition systems with rule-based natural language inference and semantic extraction algorithms have been implemented at EUROCONTROL
-simulation centers in Europe to train air traffic controllers and study the potential effects of AI augmented air traffic management (ATM) systems
-on controller workloads. Preliminary studies have found that AI augmented ATM systems not only reduce controller workloads but also reduce fuel
-consumption of aircraft by optimizing aircraft routes in the airspace and reducing departure/arrival times thereby reducing fuel
+speech recognition systems with rule- and machine learning-based natural language processing, inference, and semantic extraction algorithms have been
+implemented at EUROCONTROL simulation centers in Europe to train air traffic controllers and study the potential effects of AI augmented air traffic
+management (ATM) systems on controller workloads. Preliminary studies have found that AI augmented ATM systems not only reduce controller workloads
+but also reduce fuel consumption of aircraft by optimizing aircraft routes in the airspace and reducing departure/arrival times thereby reducing fuel
 consumption~\cite{helmke_quantifying_2017}.
 
-At the time of this writing, the current focus of general machine learning algorithms in the automatic speech recognition (ASR) and language modeling
-domains has been on a variety of deep learning methods/architectures, namely; recurrent, convolutional, and transformer neural networks due to the
-availability of large corpora such as LibriSpeech~\cite{panayotov_librispeech_2015} and Mozilla Common Voice~\cite{ardila_common_2020} for ASR and
-WikiText~\cite{merity_pointer_2016} and IMDB~\cite{maas_learning_2011} corpora for language modeling. In contrast to the general domain, the aviation
-domain still prefers, and sees much success with, traditional machine learning models for both ASR and language modeling applications. Typically, the
-most prevalent of the traditional models are Hidden Markov models and Gaussian Mixture models, in conjunction with word lattices, N-grams, and/or
+At the time of this writing, the current focus of general machine learning and deep learning algorithms in the language modeling domains has been on a
+variety of deep learning methods/architectures, namely; recurrent, convolutional, and transformer neural networks due to the availability of large
+natural language corpora corpora such as WikiText~\cite{merity_pointer_2016} and IMDB~\cite{maas_learning_2011}. In contrast to the general domain,
+the aviation domain still prefers, and sees much success with, traditional machine learning models for language modeling applications. Typically, the
+most prevalent of the traditional models are Hidden Markov models, Gaussian Mixture models, and/or in conjunction with word lattices, N-grams, and/or
 rule-based approaches depending on the application and operating environment of the models. The most likely reason for this divergence between the
-domains is the difference and relative lack of labeled data in the aviation domain. By comparison, one of the largest corpora in the general domain is
-LibriSpeech with approximately 1000 hours of labeled data whereas, for aviation, the largest corpus is Air Traffic Control Complete with approximately
-26 hours of labeled data\footnote{Air Traffic Control Complete has approximately 70 hours of audio data, however, about 26 hours of the 70 are
-    labeled.}. This disparity in data availability is exacerbated by the fact that aviation English uses a specialized vocabulary in addition to
-specialized pronunciation for some words to enhance understandability in a radiotelephonic (R/T) medium~\cite{paltridge_handbook_2013}, making it
-significantly more difficult to transcribe/label data as it requires transcribers to have domain knowledge and experience in order to create effective
-transcriptions. The medium of communication along with the environment in which the communication occurs i.e.~a noisy and possibly low fidelity
-medium\footnote{Depending on the equipment used to transmit and receive R/T communications.} combined with a noisy and active communication
-environment---an active flight deck (influenced by wind, engine noise, etc.~) or an active Air Traffic Control (ATC) tower (other people/controllers
-conversing or communicating in the background)---and a rapid rate of speech~\cite{paltridge_handbook_2013}, further contributing to the difficulty in
-understanding and therefore transcribing aviation English.
+domains is the difference and relative lack of labeled data in the aviation domain. By comparison, the largest corpora in the general domain is either
+WikiText or the BookCorpus which contain trillions of tokens whereas, for aviation, the largest corpus is Air Traffic Control Complete with
+approximately 26 hours of labeled data amounting to just over 300,000 tokens\footnote{Air Traffic Control Complete has approximately 70 hours of
+    audio data, however, about 26 hours of the 70 are labeled.}. This disparity in data availability is exacerbated by the fact that aviation English
+uses a specialized vocabulary in addition to specialized pronunciation for some words to enhance understandability in a radiotelephonic (R/T)
+medium~\cite{paltridge_handbook_2013}, making it significantly more difficult to transcribe/label data as it requires transcribers to have domain
+knowledge and experience in order to create effective transcriptions. The medium of communication along with the environment in which the
+communication occurs i.e.~a noisy and possibly low fidelity medium\footnote{Depending on the equipment used to transmit and receive R/T
+    communications.} combined with a noisy and active communication environment---an active flight deck (influenced by wind, engine noise, etc.~) or
+an active Air Traffic Control (ATC) tower (other people/controllers conversing or communicating in the background)---and a rapid rate of speech
+\cite{paltridge_handbook_2013}, further contributing to the difficulty in understanding and therefore transcribing aviation English.
 
@@
 other NLP-related tasks such as prompt completion and sentiment analysis among others (i.e.~auto-regressive and sequence classification tasks,
-respectively)\cite{lewis_bart_2019,radford_improving_2018}.
-
-% ASR context (suggest some need for semi-supervised approaches and therefore LMs)
-End-to-end generative models for automatic speech recognition models have made significant progress in recent years with current state-of-the-art
-models achieving WERs as low as 2\% on LibriSpeech test sets\cite{han_contextnet_2020,kriman_quartznet_2020,baevski_wav2vec_2020,li_jasper_2019}.
-This has led to the development of ASR models for the aviation domain, specifically, in air traffic control communications\cite{badrinath_automatic_2022,smidl_air_2019,zuluaga-gomez_automatic_2020,srinivasamurthy_semi-supervised_2017}.
-However, due to the lack of
-transcribed data in the aviation domain\cite{zuluaga-gomez_automatic_2020,srinivasamurthy_semi-supervised_2017,badrinath_automatic_2022,smidl_air_2019}
-ASR models maintain relatively high WERs compared to their counterparts in the more generalized ASR domain\cite{zuluaga-gomez_automatic_2020,badrinath_automatic_2022}.
-Transfer learning has even yielded limited results in this domain (depending on model
-architecture and dataset quality)\cite{badrinath_automatic_2022,zuluaga-gomez_automatic_2020}.
-
-% Unsupervised/semi-supervised in general and LMs commonly used with them
-Unsupervised and semi-supervised methodologies have become popular recently in attempts to address limited data availability
-and develop new approaches towards modeling human speech (notably, wav2vec has achieved state-of-the-art performance with
-very little training data)\cite{baevski_wav2vec_2020,badrinath_automatic_2022,srinivasamurthy_semi-supervised_2017,zuluaga-gomez_contextual_2021}.
-Language models are an integral part semi-supervised learning. They are used to obtain a certainty score (or uncertainty
-score, as the case may be) for the predicted text, these are usually either word lattices or N-gram models\cite{badrinath_automatic_2022,srinivasamurthy_semi-supervised_2017,zuluaga-gomez_contextual_2021}.
-While these have proven to be adequate for most self-supervised learning tasks, they are hardly state-of-the-art.
-Wav2vec, the current state-of-the-art unsupervised model (and possibly top performing overall), uses a contrastive process
-between convolutional feature extraction and transformer context representations\cite{baevski_wav2vec_2020}.
+respectively)\cite{lewis_bart_2019,radford_improving_2018}. Devlin et al.~proposed pretraining methods such as masked language
+modeling and next sentence prediction to improve natural language understanding for downstream tasks \cite{devlin_bert_2019}, which was further
+expanded upon by Liu et al.~by training longer, with larger batch sizes, and removing the next sentence prediction task \cite{liu_roberta_2019}.
+
+% tokenizer algorithms
+In language modeling and natural language processing tasks more generally, the out-of-vocabulary (OOV) problem is prevalent and has been addressed by
+the introduction of tokenizer algorithms. WordPiece and Byte Pair Encoding have been used to solve OOV problems while simultaneously minimizing
+vocabulary sizes and maximizing the likelihood of tokens in sequences \cite{wu_googles_2016,schuster_japanese_2012,sennrich_neural_2016}.
 
@@
 has also been proposed to analyze Air-Traffic communications and condense significant features (e.g.~weather, runway, and
-heading info) into an XML language structure\cite{ragnarsdottir_language_2003}. Transformer language models such as BERT,
-RoBERTa, and DeBERTa have been applied to notice to airmen (NOTAM) messages to perform named entity recognition (NER) tasks,
-translation tasks (between notations; e.g.~NOTAM notation to Airlang to make parsing tasks easier) and reduce the workload
-for pilots during long-haul flights\cite{arnold_knowledge_2022}. Transformer models have also been used for speaker role
-identification tasks in the aviation domain (e.g.~identifying pilot versus controller in communications), specifically,
-a pretrained BERT model was used and fine-tuned on problem specific data and compared to other models that performed
-well at speaker and role identification tasks in general\cite{guo_comparative_2022}.
+heading info) into an XML language structure\cite{ragnarsdottir_language_2003}. Transformer language models such as BERT and RoBERTa have been applied
+to notice to airmen (NOTAM) messages to perform named entity recognition (NER) tasks, translation tasks (between notations; e.g.~NOTAM notation to
+Airlang to make parsing tasks easier) and reduce the workload for pilots during long-haul flights\cite{arnold_knowledge_2022}. Transformer models
+have also been used for speaker role identification tasks in the aviation domain (e.g.~identifying pilot versus controller in communications),
+specifically, a pretrained BERT model was used and fine-tuned on problem specific data and compared to other models that performed well at speaker
+and role identification tasks in general\cite{guo_comparative_2022}.
+
+% ASR context (suggest some need for semi-supervised approaches and therefore LMs)
+Automatic speech recognition is seeing increased use in aviation for tasks such as transcription, callsign detection, speaker identification,
+etc. A majority of these approaches use machine learning approaches (as opposed to deep learning) where acoustic, pronunciation, and language
+modeling is used to transcribe speech
+\cite{guo_comparative_2022,smidl_air_2019,zuluaga-gomez_automatic_2020,badrinath_automatic_2022,hofbauer_atcosim_2008,helmke_quantifying_2017}.
+Language models have been used in the general domain (both neural and machine learning-based approaches) to boost the performance of speech
+recognition models \cite{han_contextnet_2020,kriman_quartznet_2020,majumdar_citrinet_2021}. Machine learning-based language models, usually N-grams
+and word lattices have also been used in the aviation domain alongside semi-supervised approaches to automatic speech recognition to leverage
+unlabeled data, however, none of the developed methodologies use neural language models for their applications
+\cite{zuluaga-gomez_contextual_2021,srinivasamurthy_semi-supervised_2017,badrinath_automatic_2022}.
+
+% differences in language (transcripts vs literary english), narrow model selection
+The use of the English language in aviation has been shown to be sufficiently different enough to be classified under its own category (aviation
+English) \cite{paltridge_handbook_2013}. The issue with using pretrained checkpoints of transformer-based language models is characterizing the
+similarity (or, conversely, the difference) between literary English and aviation English. If the two domains are sufficiently different, then
+training the pretrained models on aviation English data would amount to transfer learning between specific domains. Transfer learning between text
+domains has been shown to be possible even with data volumes significantly lower than the original data used to train the models
+\cite{raffel_exploring_2020}. Although, Yadlowsky et al.~recently showed that if problem domains are sufficiently different, model pretraining on data
+considered out-of-domain will have negative effects on model performance \cite{yadlowsky_pretraining_2023}.
 
@@
 All data was processed in Python using primarily built-in functions/modules\footnote{The \lstinline|re| and \lstinline|xml| modules were also used
-    although these are built into most Python distributions by default.}. The text data is extracted from the corpus transcripts parsing from the
+    although these are built into most Python distributions by default.}. The text data is extracted from the corpus transcripts by parsing from the
 corpora-specific formats where necessary i.e.~Lisp in ATCC, XML in ATCO2 and ZCU CZ ATC, and plain text in ATCOSIM, the transcriber annotations
@@
 exacerbated by the aggregation of several corpora with varying controls for correcting spelling errors and inconsistencies. Nearly all of the errors
-are recoverable/correctable given context. For example, ``possibility of \textbf{tornado's}'' should clearly have been
-``possibility of \textbf{tornadoes}'', changing the possessive form of ``tornado'' to the plural form. There are also instances of spelling conventions
-that are consistent within corpora, but become inconsistent when the corpora are combined.
+are recoverable/correctable given context. For example, ``possibility of \textbf{tornado's}'' should clearly have been ``possibility of
+\textbf{tornadoes}'', changing the possessive form of ``tornado'' to the plural form. There are also instances of spelling conventions that are
+consistent within corpora, but become inconsistent when the corpora are combined.
 
@@
 
+One of the intended downstream use cases of the language models developed from this work is decoding speech recognition predictions into transcripts,
+so including as much data as possible regarding the realization of spoken words is useful given the context. The \lstinline|[HES]| token is used as a
+catch-all, fallback token in the case that no knowledge can be extracted about the utterance from the transcript.
+
 \section{Data Analysis}\label{sec:data_analysis}
@@
 \subsection{RoBERTa}\label{sec:roberta}
-As the name implies, RoBERTa, is an iteration on BERT (described in section \ref{sec:bert}, above). The results from
-Liu et al.~\cite{liu_roberta_2019} indicate that BERT was initially significantly undertrained, so by training longer, with larger batch sizes, on
+As the name implies, RoBERTa, is an iteration on BERT (described in section \ref{sec:bert}, above). The results from Liu et
+al.~\cite{liu_roberta_2019} indicate that BERT was initially significantly undertrained, so by training longer, with larger batch sizes, on
 longer sequences, and dynamically changing the masking pattern applied to the MLM pretraining task, the performance of BERT was improved. The new
@@
 the quantity of training data along with the batch sizes in addition to training the model longer. Lastly, there is a small difference in the
-construction of input sequences as compared to BERT; as with BERT, input sequences always begin with a classification token (\lstinline|[CLS]|) and
-individual segments are separated by a separation token (\lstinline|[SEP]|). Rather than having sequences end with a separation token as well, they
-are terminated with an end of sequence token (\lstinline|[EOS]|). See equation \ref{eq:roberta_input_example} modified from Liu et
+construction of input sequences as compared to BERT; input sequences always begin with a classification token (\lstinline|[CLS]|) and individual
+segments are separated by a separation token (\lstinline|[SEP]|). Rather than having sequences end with a separation token as well, they are
+terminated with an end of sequence token (\lstinline|[EOS]|). See equation \ref{eq:roberta_input_example} modified from Liu et
 al.~\cite{liu_roberta_2019}, below.
@@
 
+\subsection{Performance Metrics}\label{sec:performance_metrics}
 As mentioned previously, perplexity and cross-entropy loss (often shortened to loss) are used to track the performance of the model on the training
@@
 
+\subsection{Pretraining Results}\label{sec:pretraining_results}
 With the metrics above defined, an analysis of the results of the model pretraining can be performed. The objective of the pretraining procedure is to

### 2023-12-01 thesis/MSECE_thesis.tex
Added acknowledgements and appendix A.

@@
 \usepackage[T1]{fontenc}
-\usepackage[numbib]{tocbibind}
+% \usepackage[numbib]{tocbibind}
 \usepackage[lmargin=1.25in,rmargin=1.25in,tmargin=1.0in,bmargin=1.0in]{geometry}
@@
 
-\section{Notation and Terminology}
-% TODO: proper writing, bulleted list for now
-This section is included for the sake of clarity throughout this thesis. In pursuit of that, there are a few terms and notations that are defined
-below.
-
-Set notation is used in parts of this thesis, particularly in equations and algorithms, to convey mathematical concepts. With that being said, the
-shorthand below is defined
-\begin{equation*}
-    A; x := A \cup \{x\}
-\end{equation*}
-
-\noindent
-as adding the element $x$, of its own set, to the set $A$.
-
-The following words (in bold) are used throughout the thesis as they relate to language modeling and are defined concretely as follows:
-\begin{itemize}
-    \item A \textbf{sequence} shall be defined as a set of tokens (distinct from a sentence)
-    \item A \textbf{token} shall be defined as the smallest, fundamental component of a sequence (distinct from characters and words)
-    \item The rules for the formation of tokens from a sentence (and into sequences) shall be defined by tokenization algorithms (section
-          \ref{sec:tokenizers})
-    \item The definition of \textbf{words} and \textbf{sentences} follow from that of the English language
-    \item The \textbf{vocabulary}, \textbf{inventory}, \textbf{index}, and \textbf{word list} of a model shall be semantically identical
-\end{itemize}
-\newpage
-
 \section{Abstract}\label{sec:abstract}
+With recent advances in machine learning and deep learning technologies along with the creation of larger aviation-specific corpora, the application
+of natural language processing technologies to aviation communications, especially those based on transformer neural networks, is becoming
+increasingly feasible. Previous work has been done focusing on machine learning applications to natural language processing such as N-grams and word
+lattices. This thesis will layout a process for pretraining transformer-based language models on aviation English corpora and compare the
+effectiveness and performance of language models transfer learned from pretrained checkpoints and those trained from their base weight initializations
+(trained from scratch). The results show that transformer language models trained from scratch outperform models fine-tuned from pretrained
+checkpoints, achieving perplexity values on the test partition of the corpus nearly two orders of magnitude lower than that of the pretrained models.
+The work concludes by making recommendations for future work to improve pretraining performance as well as suggestions for downstream, in-domain tasks
+such as semantic extraction, named entity recognition (callsign identification), speaker role identification and speech recognition.
 
@@
 
+\section{Notation and Terminology}
+% TODO: proper writing, bulleted list for now
+This section is included for the sake of clarity throughout this thesis. In pursuit of that, there are a few terms and notations that are defined
+below.
+
+Set notation is used in parts of this thesis, particularly in equations and algorithms, to convey mathematical concepts. With that being said, the
+shorthand below is defined
+\begin{equation*}
+    A; x := A \cup \{x\}
+\end{equation*}
+
+\noindent
+as adding the element $x$, of its own set, to the set $A$.
+
+The following words (in bold) are used throughout the thesis as they relate to language modeling and are defined concretely as follows:
+\begin{itemize}
+    \item A \textbf{sequence} shall be defined as a set of tokens (distinct from a sentence)
+    \item A \textbf{token} shall be defined as the smallest, fundamental component of a sequence (distinct from characters and words)
+    \item The rules for the formation of tokens from a sentence (and into sequences) shall be defined by tokenization algorithms (section
+          \ref{sec:tokenizers})
+    \item The definition of \textbf{words} and \textbf{sentences} follow from that of the English language
+    \item The \textbf{vocabulary}, \textbf{inventory}, \textbf{index}, and \textbf{word list} of a model shall be semantically identical
+\end{itemize}
+
 \section{Literature Review}\label{sec:lit_review}
@@
 recognition models in the aviation domain, the pretraining procedures introduced here could be used to further reduce the error rates of these models.
+This will be elaborated further in appendix A in section \ref{sec:appendix_a}.
 
 \newpage
-\bibliography{refs}
-\bibliographystyle{plain}
+\section{Acknowledgements}
+The following libraries were used to develop, train, test, and validate the models used in this work as well as perform data processing and analysis:
+\begin{itemize}
+    \item PyTorch \cite{paszke_pytorch_2019}
+    \item HuggingFace transformers (and tokenizers) \cite{wolf_transformers_2020}
+    \item PyTorch Lightning \cite{falcon_pytorchlightning_2019}
+    \item Sci-Kit Learn \cite{pedregosa_scikit-learn_2011}
+\end{itemize}
+
+\newpage
+\section{Appendix A: Neural Language Modeling Application to Speech Recognition}\label{sec:appendix_a}
+NVIDIA's NeMo (Neural Modules) framework \cite{kuchaiev_nemo_2019} is a state-of-the-art deep learning framework with implementations for dozens of
+natural language processing models and utilities. Notably among those are the automatic speech recognition (ASR) model implementations and tools for
+decoding/augmenting predictions from model outputs.
+
+For ASR models, there are generally two methods of decoding predictions: greedy decoding and decoding with beam search. Greedy decoding methods simply
+use the neuron, in the last layer of the network (the output layer), with the highest value to determine the prediction for that time step. Beam
+search-based decoding methods use a language model along with a beam search algorithm to determine predictions at each time step. The most commonly
+used language model for these types of decoding methods is an N-gram and, in fact, are what PyTorch's \lstinline|CTCDecoder|\footnote{More
+    specifically, the \lstinline|torchaudio| package of PyTorch} \cite{paszke_pytorch_2019} and NeMo's \lstinline|BeamSearchDecoderWithLM|
+\cite{kuchaiev_nemo_2019} use to decode predictions.
+
+The purpose of language models in beam search decoding algorithms is to estimate the likelihood of the decoded sequence occurring based on the
+language models knowledge of sequence structures. The beam search algorithm is then used to select tokens such that the overall likelihood of the
+sequence occurring is maximized. As mentioned above in section \ref{sec:ngram}, N-grams are left-associative language models (assuming the model is
+being applied to the English language in this case), which is an inherently limiting factor. A neural language model capable of representing
+sequences bidirectionally, such as those covered in this work, is a significantly more robust method of language modeling in speech recognition and
+has been used in the past to improve word error rates \cite{kriman_quartznet_2020,majumdar_citrinet_2021}. There is at least one instance of a deep
+learning-based language model being used for beam search decoding in aviation applications, however, it did not perform well, likely due to small
+size of the dataset that it was trained on \cite{pellegrini_airbus_2019}. The models trained in this work used considerably more data and, therefore,
+will likely have better performance in beam search decoding algorithms.
+
+\newpage
+\bibliographystyle{IEEEtran}
+\bibliography{IEEEabrv,refs}
 

### 2023-12-03 thesis/MSECE_thesis.tex
Thesis edits.

@@
 variety of deep learning methods/architectures, namely; recurrent, convolutional, and transformer neural networks due to the availability of large
-natural language corpora corpora such as WikiText~\cite{merity_pointer_2016} and IMDB~\cite{maas_learning_2011}. In contrast to the general domain,
-the aviation domain still prefers, and sees much success with, traditional machine learning models for language modeling applications. Typically, the
-most prevalent of the traditional models are Hidden Markov models, Gaussian Mixture models, and/or in conjunction with word lattices, N-grams, and/or
-rule-based approaches depending on the application and operating environment of the models. The most likely reason for this divergence between the
-domains is the difference and relative lack of labeled data in the aviation domain. By comparison, the largest corpora in the general domain is either
-WikiText or the BookCorpus which contain trillions of tokens whereas, for aviation, the largest corpus is Air Traffic Control Complete with
-approximately 26 hours of labeled data amounting to just over 300,000 tokens\footnote{Air Traffic Control Complete has approximately 70 hours of
-    audio data, however, about 26 hours of the 70 are labeled.}. This disparity in data availability is exacerbated by the fact that aviation English
-uses a specialized vocabulary in addition to specialized pronunciation for some words to enhance understandability in a radiotelephonic (R/T)
-medium~\cite{paltridge_handbook_2013}, making it significantly more difficult to transcribe/label data as it requires transcribers to have domain
-knowledge and experience in order to create effective transcriptions. The medium of communication along with the environment in which the
-communication occurs i.e.~a noisy and possibly low fidelity medium\footnote{Depending on the equipment used to transmit and receive R/T
-    communications.} combined with a noisy and active communication environment---an active flight deck (influenced by wind, engine noise, etc.~) or
-an active Air Traffic Control (ATC) tower (other people/controllers conversing or communicating in the background)---and a rapid rate of speech
-\cite{paltridge_handbook_2013}, further contributing to the difficulty in understanding and therefore transcribing aviation English.
+natural language corpora such as WikiText~\cite{merity_pointer_2016} and IMDB~\cite{maas_learning_2011}. In contrast to the general domain, the
+aviation domain still prefers, and sees much success with, traditional machine learning models for language modeling applications. Typically, the most
+prevalent of the traditional models are N-grams, word lattices, and/or rule-based approaches depending on the application and operating environment of
+the models. The most likely reason for this divergence between the domains is the difference and relative lack of labeled data in the
+aviation domain. By comparison, the largest corpora in the general domain is either WikiText or the BookCorpus which contain trillions of tokens
+whereas, for aviation, the largest corpus is Air Traffic Control Complete with approximately 26 hours of labeled data amounting to just over 300,000
+tokens\footnote{Air Traffic Control Complete has approximately 70 hours of audio data, however, about 26 hours of the 70 are labeled.}. This disparity
+in data availability is exacerbated by the fact that aviation English uses a specialized vocabulary in addition to specialized pronunciation for some
+words to enhance understandability in a radiotelephonic (R/T) medium~\cite{paltridge_handbook_2013}, making it significantly more difficult to
+transcribe/label data as it requires transcribers to have domain knowledge and experience in order to create effective transcriptions. This, in
+addition to the following facets, further contribute to the difficulty of transcription: (1) the medium of communication, (2) the environment in which
+the communication occurs\footnote{For example, a noisy and possibly low fidelity medium, depending on the equipment used to transmit and receive
+    communications}, (3) a noisy and active communication environment i.e.~an active flight deck under the influence of wind, engine noise, etc., (4)
+an active Air Traffic Control (ATC) tower and/or channels, and (5) a rapid rate of speech \cite{paltridge_handbook_2013}.
 
 % connect complexity of aviation english to utility/necessity of NLP models (ASR and LM) in the aviation domain
-The potential for speech recognition and language modeling algorithms to reach better than human performance (for example, the SQuAD benchmark has
-recorded results consistently better than human since 2019\footnote{https://rajpurkar.github.io/SQuAD-explorer/})~\cite{zhang_ai_2022} suggests that
-these NLP algorithms could have high utility in the aviation domain. If they can achieve better-than-human performance in the aviation domain, and be
-used in conjunction with pilot and air traffic controller systems, they could easily boost efficiency, reduce workloads, and reduce and/or mitigate
-errors during normal operations.
-
-The methods and results laid out in this work were done with the intention of creating a streamlined process for training and testing deep learning
-based NLP methods, specifically automatic speech recognition and text-based language models. The results from the language models suggest that the
-volume of text used to train the language models is not yet sufficient to produce results akin to those in the conversational and/or literary English
-domains. However, the automatic speech recognition models were able to achieve near state-of-the-art results and, according to the literature review,
-better than other deep learning-based approaches in the aviation domain.
+The potential for speech recognition and language modeling algorithms to reach better than human performance (for example, the Stanford Question
+Answering Dataset (SQuAD) benchmark has recorded results consistently better than human since
+2019\footnote{https://rajpurkar.github.io/SQuAD-explorer/})~\cite{zhang_ai_2022} suggests that these NLP algorithms could have high utility in the
+aviation domain. If they can achieve better-than-human performance in the aviation domain, and be used in conjunction with pilot and air traffic
+controller systems, they could easily boost efficiency, reduce workloads, and reduce and/or mitigate errors during normal operations.
+
+The methods and results laid out in this thesis were done with the intention of creating a streamlined process for training and testing deep learning
+based NLP methods, specifically language modeling. The results from the language models suggest that the volume of text used to train the language
+models is not yet sufficient to produce results akin to those in the conversational and/or literary English domains. Additionally, the results show
+that language models trained ``from scratch'' on in-domain data perform better than model trained on out-of-domain data and transfer learned on
+in-domain data i.e.~pretrained BERT and RoBERTa models transfer learned to aviation English.
 
 \section{Notation and Terminology}
-% TODO: proper writing, bulleted list for now
 This section is included for the sake of clarity throughout this thesis. In pursuit of that, there are a few terms and notations that are defined
@@
 
-Set notation is used in parts of this thesis, particularly in equations and algorithms, to convey mathematical concepts. With that being said, the
-shorthand below is defined
+Set notation (or set-builder notation)\footnote{That is, the notation used to denote sets and associated operations therein as it relates to the field
+    of set theory, in mathematics.} as defined in \cite{jech_chapter_1978} is used in parts of this thesis, particularly in equations and algorithms,
+to convey mathematical concepts. With that being said, the shorthand below is defined
 \begin{equation*}
@@
 \noindent
-as adding the element $x$, of its own set, to the set $A$.
+as the union of the set $A$ and a set containing the element $x$. The equation below can be read as adding the element $x$ to the set $A$:
+\begin{equation*}
+    A \leftarrow A; x
+\end{equation*}
 
@@
     \item A \textbf{token} shall be defined as the smallest, fundamental component of a sequence (distinct from characters and words)
-    \item The rules for the formation of tokens from a sentence (and into sequences) shall be defined by tokenization algorithms (section
+    \item The rules for the formation of tokens from a sentence (and into sequences) shall be defined by tokenization algorithms (Section
           \ref{sec:tokenizers})
@@
 The transformer neural network architecture was proposed in 2017 for Neural Machine Translation tasks and immediately achieved
-state-of-the-art (28.4 BLEU on WMT 2014 English-to-German; 41.8 BLEU on WMT English-to-French datasets)\cite{vaswani_attention_2017}. Transformer
-architectures have been found to be extremely effective at learning representations and understandings of languages to predict token probabilities as
-opposed to transforming one language into another\cite{devlin_bert_2019,liu_roberta_2019}. Transformers have also been found to be very effective at
-other NLP-related tasks such as prompt completion and sentiment analysis among others (i.e.~auto-regressive and sequence classification tasks,
-respectively)\cite{lewis_bart_2019,radford_improving_2018}. Devlin et al.~proposed pretraining methods such as masked language
-modeling and next sentence prediction to improve natural language understanding for downstream tasks \cite{devlin_bert_2019}, which was further
-expanded upon by Liu et al.~by training longer, with larger batch sizes, and removing the next sentence prediction task \cite{liu_roberta_2019}.
+state-of-the-art \cite{vaswani_attention_2017}. Transformer architectures have been found to be extremely effective at learning representations and
+understandings of languages to predict token probabilities as opposed to transforming one language into another
+\cite{devlin_bert_2019,liu_roberta_2019}. Transformers have also been found to be very effective at other NLP-related tasks such as prompt completion
+and sentiment analysis among others (i.e.~auto-regressive and sequence classification tasks, respectively)
+\cite{lewis_bart_2019,radford_improving_2018}. Devlin et al.~\cite{devlin_bert_2019}proposed pretraining methods such as masked language modeling and
+next sentence prediction to improve natural language understanding for downstream tasks, which was further expanded upon by Liu et
+al.~\cite{liu_roberta_2019} by training longer, with larger batch sizes, and removing the next sentence prediction task.
 
 % tokenizer algorithms
-In language modeling and natural language processing tasks more generally, the out-of-vocabulary (OOV) problem is prevalent and has been addressed by
-the introduction of tokenizer algorithms. WordPiece and Byte Pair Encoding have been used to solve OOV problems while simultaneously minimizing
+In language modeling and natural language processing tasks, the out-of-vocabulary (OOV) problem is prevalent and has been addressed by the
+introduction of tokenizer algorithms. WordPiece and Byte Pair Encoding (BPE) have been used to solve OOV problems while simultaneously minimizing
 vocabulary sizes and maximizing the likelihood of tokens in sequences \cite{wu_googles_2016,schuster_japanese_2012,sennrich_neural_2016}.
@@
 and word lattices have also been used in the aviation domain alongside semi-supervised approaches to automatic speech recognition to leverage
-unlabeled data, however, none of the developed methodologies use neural language models for their applications
+unlabeled data. However, none of the developed methodologies use neural language models for their applications
 \cite{zuluaga-gomez_contextual_2021,srinivasamurthy_semi-supervised_2017,badrinath_automatic_2022}.
@@
 domains has been shown to be possible even with data volumes significantly lower than the original data used to train the models
-\cite{raffel_exploring_2020}. Although, Yadlowsky et al.~recently showed that if problem domains are sufficiently different, model pretraining on data
-considered out-of-domain will have negative effects on model performance \cite{yadlowsky_pretraining_2023}.
+\cite{raffel_exploring_2020}. Yadlowsky et al.~\cite{yadlowsky_pretraining_2023} recently showed that if problem domains are sufficiently different,
+model pretraining on data considered out-of-domain will have negative effects on model performance.
 
 \section{Data Sources}\label{sec:data_source}
-There are four corpora used in this work. The datasets are primarily speech corpora intended either for automatic speech recognition research in the
+There are four corpora used in this thesis. The datasets are primarily speech corpora intended either for automatic speech recognition research in the
 domain of aviation communications, linguistics research in aviation English, or both. All data in all four corpora include both recordings and
@@
 
-The four corpora that were selected and used for this work are briefly described below. Further analysis is performed in section \ref{sec:data_analysis}.
+The four corpora that were selected and used for this work are briefly described below. Further analysis is performed in Section
+\ref{sec:data_analysis}.
 
@@
 ATCC is a collection of three smaller corpora recorded at Dallas-Fort Worth, Logan International, and Washington National airports and
-transcribed by current or former air traffic controllers familiar with the respective areas. Audio data was recorded through the placement of VHF
-antennae configured to monitor several frequencies at each airport, such as arrival, departure, approach, ground, etc. The types of frequencies
-monitored vary by airport, but the presence of departure, approach, and ground frequencies is relatively consistent across subcorpora~\cite{godfrey_air_1994}.
-% TODO: data length/size
+transcribed by current or former air traffic controllers familiar with the respective areas. Audio data was recorded through the placement of Very
+High Frequency (VHF) antennae configured to monitor several frequencies at each airport, such as arrival, departure, approach, ground, etc. The types
+of frequencies monitored vary by airport, but the presence of departure, approach, and ground frequencies is relatively consistent across subcorpora
+\cite{godfrey_air_1994}.
+% TODO: data length/size; samples, hours, tokens
 
@@
 development in addition to language detection (although in this work it is only used for ASR and LM development)~\cite{szoke_detecting_2021}.
-% TODO: data length/size
+% TODO: data length/size; samples, hours, tokens
 
@@
 The Air Traffic Control Simulation (ATCOSIM) corpus is a speech corpus consisting of audio recordings and corresponding transcriptions.
-The data was recorded at a air traffic control simulation center in Bretigny-sur-Orge, France, specifically the EUROCONTROL Experimental Centre.
-In this data, only the controllers' voices are included and therefore the transcripts only include the controller's side of each interaction.
-The audio data was transcribed by one person, trained according to the guidelines and formatting requirements of the corpus. After all data was
-transcribed, it was reviewed and corrected and any remaining problems were reviewed by an operational air traffic controller and
-resolved\cite{hofbauer_atcosim_2008}.
-% TODO: data length/size
+The data was recorded at the EUROCONTROL Experimental Centre for air traffic control simulation in Bretigny-sur-Orge, France. In this corpus, only the
+controllers' voices are included and therefore the transcripts only include the controller's side of each interaction. The audio data was transcribed
+by one person, trained according to the guidelines and formatting requirements of the corpus. After all data was transcribed, it was reviewed and
+corrected and any remaining problems were reviewed by an operational air traffic controller and resolved \cite{hofbauer_atcosim_2008}.
+% TODO: data length/size; samples, hours, tokens
 
@@
 unified\cite{smidl_air_2019}.
-% TODO: data length/size
+% TODO: data length/size; samples, hours, tokens
 
 \section{Data Processing}\label{sec:data_processing}
-All of the corpora listed above (see~\ref{sec:data_source}) are created at different times, for different purposes, and by different authors.
+All of the corpora listed above (see Section \ref{sec:data_source}) are created at different times, for different purposes, and by different authors.
 As a result, the data in these corpora all have different formats and need to be processed into a common format to be used together, with the goal
@@
 human error that introduced those errors in the first place. Instead, the frequency of occurrence of tokens in the corpora combined with manual review
-thereafter, was used to detect and correct the most common errors. Tokens with the lowest frequency across corpora were manually reviewed and
-corrected where necessary. An \(A \rightarrow B\) mapping system was used to correct the errors, where \(A\) is the erroneous token and \(B\) is the
-corrected or replaced token. A total of 22 unique errors were detected and subsequently corrected or removed using this method, although multiple
+thereafter, was used to detect and correct the most common errors. The frequency of occurrence of a token is calculated using equation
+(\ref{eq:token_frequency_occurrence}).
+
+\begin{equation}\label{eq:token_frequency_occurrence}
+    f_w = \frac{\phi_w(w)}{D_{tokens}}
+\end{equation}
+
+\noindent
+Where $f_w$ is the frequency at which a token occurs in a corpus, $\phi_w(w)$ is the number of times a token, $w$, occurs, and $D_{tokens}$ is the
+total number of tokens in the corpus. Tokens with the lowest frequency across corpora (and lowest number of occurrences) were manually reviewed and
+corrected where necessary. An $A \rightarrow B$ mapping system was used to correct the errors, where \(A\) is the erroneous token and \(B\) is the
+corrected or replaced token. As the text is parsed from the corpora, each sample is analyzed, and if an erroneous token, $A$, is found, it is replaced
+with the corrected version, $B$. A total of 22 unique errors were detected and subsequently corrected or removed using this method, although multiple
 instances of those errors typically occurred.
@@
 
-\begin{table}[h!]
+\begin{table}[!t]
     \centering
@@
 \subsection{Outlier Detection and Analysis}\label{sec:outliers}
-The Local Outlier Factor (LOF) algorithm~\cite{breunig_lof_2000} was used to detect and remove outliers in the corpora for further analysis. The
-sequence lengths of the data samples were used to determine outliers. Since all four corpora are aggregated, LOF analysis was performed on the
-aggregated data instead of on the individual corpora.
-
-Sci-Kit Learn's implementation of the LOF
-algorithm was used with a value of \(N = 35\) for \(N\) neighbors in the K-Nearest Neighbors algorithm and Minkowski distance for distance
-metrics/calculations (in this case, since \(p=2\), Euclidean distance is effectively used for distance calculations). A total of 47 samples were
-detected as outliers and removed for analysis and are summarized in table~\ref{tab:outlier_stats}.
-
-\begin{table}[h!]
+The Local Outlier Factor (LOF) algorithm \cite{breunig_lof_2000} was used to detect and analyze outliers in the corpora. The sequence lengths of the
+data samples were used to determine outliers. Since all four corpora are aggregated, LOF analysis was performed on the aggregated data instead of on
+the individual corpora.
+
+Sci-Kit Learn's \cite{pedregosa_scikit-learn_2011} implementation of the LOF algorithm was used with a value of \(N = 35\) for \(N\) neighbors in the
+K-Nearest Neighbors algorithm and Minkowski distance for distance metrics/calculations (in this case, since \(p=2\), Euclidean distance is effectively
+used for distance calculations). A total of 47 samples were detected as outliers and removed for analysis and are summarized in table
+\ref{tab:outlier_stats}.
+
+\begin{table}[!t]
     \centering
@@
     the airspace) is in a format that enables us to extract the turn taking phenomenon that is present in most air-traffic communications.} to improve
-its pretraining results. MLM and NSP pretraining are explained more in-depth in section \ref{sec:bert}, below.
+its pretraining results. MLM and NSP pretraining are explained more in-depth in Section \ref{sec:bert}, below.
 
@@
 \begin{equation}\label{eq:ngram_prob_wj}
-    P(w) = \frac{\textit{occurrences}(w)}{D_{tokens}}
+    P(w) = \frac{\phi_w(w)}{D_{tokens}}
 \end{equation}
@@
 \noindent
-where $P(w)$ is the probability of a token, $w$, occurring anywhere in the corpus, $\textit{occurrences}(w)$ is the number of times of a specific
+where $P(w)$ is the probability of a token, $w$, occurring anywhere in the corpus, $\phi_w(w)$ is the number of times of a specific
 token appears in the corpus, and $D_{tokens}$ is total number of tokens in the corpus, $D$. Computationally, the process of determining the
@@
 
-\textbf{Next Sentence Prediction}. NSP is meant to support downstream tasks, while MLM conditions the model to represent tokens within sentences, NSP
-is meant to condition the model to understand relationships between sentences/sequences. Since BERTs training set consists primarily of literary works
-and excerpts from Wikipedia, the model simply predicts whether two sentences appear together (one after the other) in the original text. Instead of a
-single input sequence, the input consists of two sequences separated by a ``separation'' token and beginning with a ``classification'' token, these
-are represented by \lstinline|[SEP]| and \lstinline|[CLS]|, respectively in Devlin et al.~\cite{devlin_bert_2019}.
+\textbf{Next Sentence Prediction}. While MLM conditions the model to represent tokens within sentences, NSP is meant to condition the model to
+understand relationships between sentences/sequences to further support downstream tasks. Since BERTs training set consists primarily of literary
+works and excerpts from Wikipedia, the  model simply predicts whether two sentences appear together (one after the other) in the original text.
+Instead of a single input sequence, the input consists of two sequences separated by a ``separation'' token and beginning with a ``classification''
+token, these are represented by \lstinline|[SEP]| and \lstinline|[CLS]|, respectively in Devlin et al.~\cite{devlin_bert_2019}.
 
 For symmetry between pretraining tasks, the special tokens specifying the beginning and end of input sequences are shared between tasks. For both MLM
-and NSP objectives, input sequences always start with a classification token and end with a separation token (see figure \ref{fig:bert_mlm_input_example}).
-
-\begin{figure}[h!]
+and NSP objectives, input sequences always start with a classification token and end with a separation token (see figure
+\ref{fig:bert_mlm_input_example}). Likewise, NSP inputs always begin with a classification token and each input sequence is concatenated with a
+separation token terminating each sequence (see figure \ref{fig:bert_nsp_input_example}).
+
+\begin{figure}[!t]
     \includegraphics[width=\linewidth]{figures/BERT_MLM_input.png}
@@
 
-\noindent
-Likewise, NSP inputs always begin with a classification token and each input sequence is concatenated with a separation token terminating each
-sequence (see figure \ref{fig:bert_nsp_input_example}).
-
-\begin{figure}[h!]
+\begin{figure}[!t]
     \includegraphics[width=\linewidth]{figures/BERT_NSP_input.png}
@@
 \subsection{RoBERTa}\label{sec:roberta}
-As the name implies, RoBERTa, is an iteration on BERT (described in section \ref{sec:bert}, above). The results from Liu et
-al.~\cite{liu_roberta_2019} indicate that BERT was initially significantly undertrained, so by training longer, with larger batch sizes, on
-longer sequences, and dynamically changing the masking pattern applied to the MLM pretraining task, the performance of BERT was improved. The new
-checkpoint of this version of BERT was therefore called RoBERTa (Robustly Optimized BERT Pretraining Approach).
+As the name implies, Robustly Optimized BERT Pretraining Approach (RoBERTa), is an iteration on BERT (described in Section \ref{sec:bert}, above).
+The results from Liu et al.~\cite{liu_roberta_2019} indicate that BERT was initially significantly undertrained, so by training longer, with larger
+batch sizes, on longer sequences, and dynamically changing the masking pattern applied to the MLM pretraining task, the performance of BERT was
+improved. The new checkpoint of this version of BERT was therefore called RoBERTa.
 
@@
 
-% \begin{figure}[h!]
+% \begin{figure}[!t]
 %     \includegraphics[width=\linewidth]{figures/jasper_vertical.png}
@@
 
-% \begin{figure}[h!]
+% \begin{figure}[!t]
 %     \includegraphics[width=\linewidth]{figures/quartz_vertical.png}
@@
 
-% \begin{figure}[h!]
+% \begin{figure}[!t]
 %     \includegraphics[width=\linewidth]{figures/citrinet_vertical.png}
@@
 \subsection{WordLevel}\label{sec:wordlevel}
-The WordLevel tokenization algorithm, mentioned briefly as an introductory example above, is the most technically simple algorithm among those
-introduced in this section. Tokens are extrapolated from words in a sentence based on the whitespace separating each word. This means that
-contractions, hyphenations, etc.~are treated as their own words i.e.~ ``you'', ``you're'', and ``you've'' are all seen as unique, distinct words that
-are tokenized and mapped to their own integral representations.
+The WordLevel \cite{wolf_transformers_2020} tokenization algorithm, mentioned briefly as an introductory example above, is the most technically simple
+algorithm among those introduced in this section. Tokens are extrapolated from words in a sentence based on the whitespace separating each word.
+This means that contractions, hyphenations, etc.~are treated as their own words i.e.~ ``you'', ``you're'', and ``you've'' are all seen as unique,
+distinct words that are tokenized and mapped to their own integral representations.
 
 In order to map individual tokens to numerical values or IDs, an index of the words that appear in a corpus must be created. The numerical values for
-each token are derived from this index. For the WordLevel algorithm, this process is referred to as the training the tokenizer model.
+each token are derived from this index. For the WordLevel algorithm, this process is referred to as the training the tokenizer model (see algorithm
+\ref{alg:wordlevel_training}).
 
@@
 
-\begin{algorithm}
+\begin{algorithm}[!t]
     \caption{Training procedure for the WordLevel tokenizer.}
@@
 
-\noindent
 The simplicity of the algorithm speaks for itself and serves as a good introduction to tokenization as well as an illustrative example of the OOV
@@
 
-The tokenization of words with shared roots but different suffixes is shown in table \ref{tab:wordlevel_tokenization_shared_roots_example}.
-
-\begin{table}[h!]
+The tokenization of words with shared roots but different suffixes is shown in table \ref{tab:wordlevel_tokenization_shared_roots_example}. Clearly
+all three words are seen as independent and unique words by the tokenizer, since they are mapped to their own integer representations.
+
+\begin{table}[!t]
     \centering
@@
 
-\noindent
-Clearly all three words are seen as independent and unique words by the tokenizer, since they are mapped to their own integer representations.
-
 The tokenizer used to generate the examples in tables \ref{tab:wordlevel_tokenization_example} and
 \ref{tab:wordlevel_tokenization_shared_roots_example} was trained on two sentences, notably lacking the word ``you'll''. Running the tokenizer on the
-word ``you'll'' alone results in the following output:
-
-\begin{table}[h!]
+word ``you'll'' alone results in the output in table \ref{tab:wordlevel_unk_word}. Although the word shares a root with some of the words in the
+training data (``you'', ``you're'', and ``you've'') and is semantically very similar, the tokenizer does not recognize the word at all (represented
+by the ``[UNK]'' or unknown token). This word would be considered out-of-vocabulary since it did not appear in the training data and thus reveals one
+of the major downsides of this algorithm, as mentioned in section \ref{sec:tokenizers}. This problem can be alleviated to some extent by modifying the
+preprocessing strategy of the tokenizer to partition contracted words into their component parts e.g.~the word ``you're'' would be partitioned to
+\lstinline|[``you'', ``''', ``re'']|, however, it eventually reappears, for example, when using the past, present, and future tenses of a verb
+(e.g.~``go'' and ``goes'' in the training data, then encountering the future tense: ``going'').
+
+\begin{table}[!t]
     \centering
@@
 
-\noindent
-Although the word shares a root with some of the words in the training data (``you'', ``you're'', and ``you've'') and is semantically very similar,
-the tokenizer does not recognize the word at all (represented by the ``[UNK]'' or unknown token). This word would be considered out-of-vocabulary
-since it did not appear in the training data and thus reveals one of the major downsides of this algorithm, as mentioned in section
-\ref{sec:tokenizers}. This problem can be alleviated to some extent by modifying the preprocessing strategy of the tokenizer to partition contracted
-words into their component parts e.g.~the word ``you're'' would be partitioned to \lstinline|[``you'', ``''', ``re'']|, however, it eventually
-reappears, for example, when using the past, present, and future tenses of a verb (e.g.~``go'' and ``goes'' in the training data, then encountering
-the future tense: ``going'').
-
 \subsection{WordPiece}\label{sec:wordpiece}
 % Wu et al. 2016; Schuster and Nakajima 2012
-The WordPiece algorithm was developed by Schuster and Nakajima in 2012~\cite{schuster_japanese_2012} as a word segmentation algorithm for language
-modeling in Japanese and Korean voice search systems. It is further used by Wu et al.~\cite{wu_googles_2016} for neural machine translation tasks. The
-algorithm demonstrated proficiency and increased performance for the models used in both tasks and effectively addressed the OOV problem.
+The WordPiece algorithm was developed by Schuster and Nakajima \cite{schuster_japanese_2012} as a word segmentation algorithm for language modeling in
+Japanese and Korean voice search systems. It is further used by Wu et al.~\cite{wu_googles_2016} for neural machine translation tasks. The algorithm
+demonstrated proficiency and increased performance for the models used in both tasks and effectively addressed the OOV problem.
 
@@
 
-\noindent
 The training procedure is simple, but computationally expensive. Schuster and Nakajima calculated the time complexity of the algorithm at $O(K^2)$
@@
 Table \ref{tab:wordpiece_tokenization_example}, below, shows an example of a WordPiece tokenizer model output on a sentence that was seen during the
-training procedure (trained on the same data as the WordLevel tokenizer in section \ref{sec:wordlevel}).
-
-\begin{table}[h!]
+training procedure (trained on the same data as the WordLevel tokenizer in Section \ref{sec:wordlevel}). Since this sentence was in the training data,
+the segmentation is identical to that of the WordLevel tokenizer in table \ref{tab:wordlevel_tokenization_example}, however, tokenization of similar
+words e.g.~words with shared roots or similar contractions are handled differently as in table \ref{tab:wordpiece_tokenization_shared_roots_example}.
+Since each word has a shared root word in the contractions (as well as an apostrophe separating the contraction), the tokenizer segments each word and
+represents them as groups of small, more common (thus with higher probabilities of occurring) tokens. Additionally, the WordPiece algorithm
+effectively handles OOV words. Take the OOV in table \ref{tab:wordpiece_unk_word_similar} with a shared root and similar contraction as those in table
+\ref{tab:wordpiece_tokenization_shared_roots_example}, which appear in the training data. Since the base of the contraction, ``you'', occurs in the
+training data as well as the apostrophe in the contraction, so the tokenizer segments those parts of the word as tokens. The last part of the word,
+``ll'', does not appear at all in the training data, so it is broken down into its component characters which are represented as ``l'' (beginning of
+a word) and ``\#\#l'' (segmented part of a word that should be concatenated to the previous token when they are recombined). Comparing this to the
+results of the WordLevel tokenizer output on the same word in table \ref{tab:wordlevel_unk_word}, the WordPiece algorithm sidesteps the OOV problem
+for words similar to those in the trining corpus and, according to the output of the WordPiece tokenizer in table \ref{tab:wordpiece_unk_word}, for
+words unlike those in the training corpus.
+
+\begin{table}[!t]
     \centering
@@
 
-\noindent
-Since this sentence was in the training data, the segmentation is identical to that of the WordLevel tokenizer in table
-\ref{tab:wordlevel_tokenization_example}, however, tokenization of similar words e.g.~words with shared roots or similar contractions are handled
-differently as in table \ref{tab:wordpiece_tokenization_shared_roots_example}.
-
-\begin{table}[h!]
+\begin{table}[!t]
     \centering
@@
 
-\noindent
-Since each word has a shared root word in the contractions (as well as an apostrophe separating the contraction), the tokenizer segments each word and
-represents them as groups of small, more common (thus with higher probabilities of occurring) tokens. Additionally, the WordPiece algorithm
-effectively handles OOV words. Take the OOV in table \ref{tab:wordpiece_unk_word_similar} with a shared root and similar contraction as those in table
-\ref{tab:wordpiece_tokenization_shared_roots_example}, which appear in the training data.
-
-\begin{table}[h!]
+\begin{table}[!t]
     \centering
@@
 
-\noindent
-Since the base of the contraction, ``you'', occurs in the training data as well as the apostrophe in the contraction, so the tokenizer segments those
-parts of the word as tokens. The last part of the word, ``ll'', does not appear at all in the training data, so it is broken down into its component
-characters which are represented as ``l'' (beginning of a word) and ``\#\#l'' (segmented part of a word that should be concatenated to the previous
-token when they are recombined). Comparing this to the results of the WordLevel tokenizer output on the same word in table
-\ref{tab:wordlevel_unk_word}, the WordPiece algorithm sidesteps the OOV problem for words similar to those in the trining corpus and, according to
-the output of the WordPiece tokenizer in table \ref{tab:wordpiece_unk_word}, for words unlike those in the training corpus.
-
-\begin{table}[h!]
+\begin{table}[!t]
     \centering
@@
 
-\begin{algorithm}[h!]
+\begin{algorithm}[!t]
     \caption{BPE training algorithm implementation in Python. Modified from Sennrich et al.~\cite{sennrich_neural_2016}.}
@@
 
-Finally, comparing the tokenization process to the WordLevel and WordPiece algorithms, we can see that the outputs of the algorithm are nearly
+Finally, comparing the BPE tokenization process with the WordLevel and WordPiece algorithms, we can see that the outputs of the algorithm are nearly
 identical when the sentences being tokenized have words that appear in the training data (i.e.~the BPE outputs in tables
@@
 
-\noindent
 Now, testing the BPE algorithm on words that did not appear in the training data, it is clear that the BPE algorithm effectively deals with and
@@
 \lstinline|tokenizers| library implements it, this is a preprocessing step for the tokenizer model that maps all bytes in a string/sentence to their
-own unique and visible character. Functionally, byte-level BPE is nearly identical to standard BPE models with the main exception being that the
-beginning of word symbol is actually rendered and represented by a visible character in the tokenized sequence (see table
-\ref{tab:byte_level_bpe_example}).
-
-\begin{table}[h!]
+own unique and visible character. Functionally, byte-level BPE is nearly identical to standard BPE models except that the beginning of word symbol is
+actually rendered and represented by a visible character in the tokenized sequence (see table \ref{tab:byte_level_bpe_example}).
+
+\begin{table}[!t]
     \centering
@@
 \subsection{Language Model Pretraining}\label{sec:lm_pretraining}
-A standard masked language modeling training procedure is used to pretrain the language models, as described in section \ref{sec:bert}. Tokens are
-masked at random with at a uniform probability of 0.2. Input sequences are padded to the maximum length allowed by the model and truncated to the
-maximum length, if greater. The maximum sequence length for BERT and RoBERTa models is 512 tokens \cite{devlin_bert_2019,liu_roberta_2019}. Based on
-the data analysis in section \ref{sec:data_analysis}, there are few, if any, samples that meet or exceed the maximum sequence length for BERT and
-RoBERTa models.
+A standard masked language modeling training procedure is used to pretrain the language models, as described in Section \ref{sec:bert}. Tokens are
+masked at random with a uniform probability of 0.2. Input sequences are padded to the maximum length allowed by the model and truncated to the maximum
+length, if greater. The maximum sequence length for BERT and RoBERTa models is 512 tokens \cite{devlin_bert_2019,liu_roberta_2019}. Based on the data
+analysis in Section \ref{sec:data_analysis}, there are few, if any, samples that meet or exceed the maximum sequence length for BERT and RoBERTa
+models.
 
@@
     \begin{gathered}
-        \mathcal{L}(x, y) = \{\mbox{$l_1$, $l_2$, ..., $l_N$}\}^T\\
-        l_n = -\log \left[\frac{\exp(x_{n,y_n})}{\sum_{c=1}^{C}\exp(x_{n,c})}\right]
+        \mathcal{L}(x, y) = L = \{\mbox{$l_1$, $l_2$, ..., $l_N$}\}^T\\
+        l_n = -\log \left[\frac{\exp(x_{n,y_n})}{\sum_{c=1}^{C}\exp(x_{n,c})}\right] \quad \forall n=1..N
     \end{gathered}
@@
 \noindent
-Where $x$ and $y$ are $N \times 1$ vectors representing the inputs and labels, respectively, for each sample in the batch. $N$ is the number of
-samples in the batch, $l_n$ is the loss for an individual sample, $n$, in the batch, and $C$ is the number of class labels. For the purpose of
-language modeling, the set of class labels is equivalent to the vocabulary of the tokenizer (and thus the language model). Numerical labels will
-correspond to token IDs.
+where $x$ is a $512 \times N$ vector representing model inputs (a sequence of tokens) and $y$ is a $1 \times N$ vector representing the target/label
+for each sample in the batch. $N$ is the number of samples in the batch, $l_n$ is the loss for an individual sample, $n$, in the batch, and $C$ is the
+number of class labels. For the purpose of language modeling, the set of class labels is equivalent to the vocabulary of the tokenizer (and thus the
+language model). Numerical labels will correspond to token IDs. The overall loss is then found by taking the mean of the loss over the batch, likewise
+for an epoch, the loss is found by taking the average of each batch in the epoch.
 
@@
 Graphs of the training loss (average loss over the samples/batches in the training partition of the corpus) are presented in figure
-\ref{fig:training_loss}.
-
-\begin{figure}[h!]
+\ref{fig:training_loss}. Both classes of models (scratch and pretrained) quickly reach a steady state in terms of loss although the initial loss
+values are significantly different. It is notable that the pretrained models start with a significantly lower loss, about 1.3, compared to models
+trained from scratch, which started with a loss value of just over 5.0. This initially indicates that the models are training well and beginning to
+converge on either local or global minima.
+
+\begin{figure}[!t]
     \includegraphics[width=\linewidth]{figures/training_loss.png}
@@
 
-\noindent
-Both classes of models (scratch and pretrained) quickly reach a steady state in terms of loss although the initial loss values are significantly
-different. It is notable that the pretrained models start with a significantly lower loss, about 1.3, compared to models trained from scratch, which
-started with a loss value of just over 5.0. This initially indicates that the models are training well and beginning to converge on either local or
-global minima.
-
 The perplexity of the models on the training data yields interesting results (figure \ref{fig:training_ppl}). Although the pretrained models have a
@@
 y-axis in figure \ref{fig:training_ppl}; the final perplexity value of each language model trained from scratch is at least two orders of magnitude
-lower than that of the pretrained models.
-
-\begin{figure}[h!]
+lower than that of the pretrained models. At the end of training, the BERT and RoBERTa models trained from scratch had perplexity values of
+$0.52\times 10^7$ and $0.16\times 10^7$, respectively, whereas the pretrained models reached perplexity values of $0.29\times 10^9$ and
+$0.17\times 10^9$, for BERT and RoBERTa, respectively. This indicates that best language model performance may be achieved by pretraining models on
+in-domain data, then continuing to fine-tune on specific downstream tasks. This is a direct contradiction of the current recommended practice of using
+one of the pretrained checkpoints from Devlin et al.~\cite{devlin_bert_2019} and Liu et al.~\cite{liu_roberta_2019} then fine-tuning on specific
+downstream tasks. This is further upheld by the results of the model performance on the test partition of the corpus in table
+\ref{tab:pretraining_results}, below. As with the training data, the perplexity of the language models trained from scratch over the test partition of
+the corpus is significantly lower than that of the pretrained language models. The loss of each model over the test set is somewhat high, but they are
+around the same values for each model, which is consistent with the loss statistics during training. A higher loss value on unseen (test) data is
+normal and expected. The BERT model trained from scratch had the best performance among the models tested despite both RoBERTa models training better
+within their model classes. Within the class of pretrained models, however, the RoBERTa model still technically performed better than the BERT model
+(values shown in table \ref{tab:pretraining_results} were rounded to three significant figures) with a perplexity value of $0.007\times 10^9$ compared
+to BERT's $0.014\times 10^9$.
+
+\begin{figure}[!t]
     \includegraphics[width=\linewidth]{figures/training_ppl.png}
@@
 
-\noindent
-At the end of training, the BERT and RoBERTa models trained from scratch had perplexity values of $0.52\times 10^7$ and $0.16\times 10^7$,
-respectively, whereas the pretrained models reached perplexity values of $0.29\times 10^9$ and $0.17\times 10^9$, for BERT and RoBERTa, respectively.
-This indicates that best language model performance may be achieved by pretraining models on in-domain data, then continuing to fine-tune on specific
-downstream tasks. This is a direct contradiction of the current recommended practice of using one of the pretrained checkpoints from Devlin et
-al.~\cite{devlin_bert_2019} and Liu et al.~\cite{liu_roberta_2019} then fine-tuning on specific downstream tasks. This is further upheld by the
-results of the model performance on the test partition of the corpus in table \ref{tab:pretraining_results}, below.
-
-\begin{table}[h!]
+\begin{table}[!t]
     \centering
@@
 \end{table}
-
-\noindent
-As with the training data, the perplexity of the language models trained from scratch over the test partition of the corpus is significantly lower
-than that of the pretrained language models. The loss of each model over the test set is somewhat high, but they are around the same values for each
-model, which is consistent with the loss statistics during training. A higher loss value on unseen (test) data is normal and expected. The BERT model
-trained from scratch had the best performance among the models tested despite both RoBERTa models training better within their model classes. Within
-the class of pretrained models, however, the RoBERTa model still technically performed better than the BERT model (values shown in table
-\ref{tab:pretraining_results} were rounded to three significant figures) with a perplexity value of $0.007\times 10^9$ compared to BERT's
-$0.014\times 10^9$.
 
@@
 English. It is difficult to determine what exactly is causing this phenomenon, so several theories were developed in an attempt to explain it (some
-of which will be elaborated upon in section \ref{sec:future_work}):
-
-\noindent
+of which will be elaborated upon in Section \ref{sec:future_work}):
+
 \textbf{Small volume of training data}. Especially when compared to \textit{Attention is All You Need}, BERT, and RoBERTa,
@@
 
-\noindent
 \textbf{Different forms of English}. The BERT and RoBERTa pretraining checkpoints were created by training the models on literary works and typed
@@
 
-\noindent
-\textbf{Mixed regional data sources}. As described in section \ref{sec:data_source}, the data aggregated for use here is comes from several different
+\textbf{Mixed regional data sources}. As described in Section \ref{sec:data_source}, the data aggregated for use here is comes from several different
 regions, namely international airports in the US \cite{godfrey_air_1994} and international and domestic airports in Europe
@@
 \textbf{Next Sentence Prediction (NSP)}. This training procedure was introduced by Devlin et al.~\cite{devlin_bert_2019} as a pretraining procedure
-for BERT and is described in more detail in section \ref{sec:bert}. This can be used on top of masked language modeling for language model
+for BERT and is described in more detail in Section \ref{sec:bert}. This can be used on top of masked language modeling for language model
 pretraining, however, it requires a knowledge of which sequences in the corpus appear next to each other. For pretraining transformer-based neural
@@
 recognition models in the aviation domain, the pretraining procedures introduced here could be used to further reduce the error rates of these models.
-This will be elaborated further in appendix A in section \ref{sec:appendix_a}.
+This will be elaborated further in Appendix B in Section \ref{sec:appendix_a}.
 
@@
 \newpage
-\section{Appendix A: Neural Language Modeling Application to Speech Recognition}\label{sec:appendix_a}
+\section{Appendix A: Examples of Samples Detected as Outliers}
+
+\newpage
+\section{Appendix B: Neural Language Modeling Application to Speech Recognition}\label{sec:appendix_a}
 NVIDIA's NeMo (Neural Modules) framework \cite{kuchaiev_nemo_2019} is a state-of-the-art deep learning framework with implementations for dozens of
@@
 language models knowledge of sequence structures. The beam search algorithm is then used to select tokens such that the overall likelihood of the
-sequence occurring is maximized. As mentioned above in section \ref{sec:ngram}, N-grams are left-associative language models (assuming the model is
+sequence occurring is maximized. As mentioned above in Section \ref{sec:ngram}, N-grams are left-associative language models (assuming the model is
 being applied to the English language in this case), which is an inherently limiting factor. A neural language model capable of representing

### 2023-12-04 thesis/MSECE_thesis.tex
more thesis edits.

@@
 of frequencies monitored vary by airport, but the presence of departure, approach, and ground frequencies is relatively consistent across subcorpora
-\cite{godfrey_air_1994}.
-% TODO: data length/size; samples, hours, tokens
+\cite{godfrey_air_1994}. See Table \ref{tab:atcc_stats} for corpus-specific statistics.
+
+\begin{table}
+    \centering
+    \begin{tabular}{l r}
+        \toprule
+        \multicolumn{2}{c}{ATCC Corpus Properties} \\
+        \midrule
+        Samples              & 9,556               \\
+        Audio (hours)        & 72.48               \\
+        Mean Sequence Length & 11.90               \\
+        Std.~Sequence Length & 7.22                \\
+
+        Total Tokens         & 355,485             \\
+        Unique Tokens        & 2,209               \\
+        \bottomrule
+    \end{tabular}
+    \caption{Data statistics for the Air Traffic Control Complete (ATCC) corpus.}
+    \label{tab:atcc_stats}
+\end{table}
 
@@
 development in addition to language detection (although in this work it is only used for ASR and LM development)~\cite{szoke_detecting_2021}.
-% TODO: data length/size; samples, hours, tokens
+See Table \ref{tab:atco2_stats} for corpus-specific statistics.
+
+\begin{table}
+    \centering
+    \begin{tabular}{l r}
+        \toprule
+        \multicolumn{2}{c}{ATCO2 Corpus Properties} \\
+        \midrule
+        Samples              & 874                  \\
+        Audio (hours)        & 1.10                 \\
+        Mean Sequence Length & 12.28                \\
+        Std.~Sequence Length & 5.78                 \\
+        Total Tokens         & 10,733               \\
+        Unique Tokens        & 786                  \\
+        \bottomrule
+    \end{tabular}
+    \caption{Data statistics for the ATCO2 corpus.}
+    \label{tab:atco2_stats}
+\end{table}
 
@@
 by one person, trained according to the guidelines and formatting requirements of the corpus. After all data was transcribed, it was reviewed and
-corrected and any remaining problems were reviewed by an operational air traffic controller and resolved \cite{hofbauer_atcosim_2008}.
-% TODO: data length/size; samples, hours, tokens
+corrected and any remaining problems were reviewed by an operational air traffic controller and resolved \cite{hofbauer_atcosim_2008}. See Table
+\ref{tab:atcosim_stats} for corpus-specific statistics.
+
+\begin{table}
+    \centering
+    \begin{tabular}{l r}
+        \toprule
+        \multicolumn{2}{c}{ATCOSIM Corpus Properties} \\
+        \midrule
+        Samples              & 9,556                  \\
+        Audio (hours)        & 10.69                  \\
+        Mean Sequence Length & 11.29                  \\
+        Std.~Sequence Length & 4.17                   \\
+        Total Tokens         & 107,881                \\
+        Unique Tokens        & 827                    \\
+        \bottomrule
+    \end{tabular}
+    \caption{Data statistics for the ATCOSIM corpus.}
+    \label{tab:atcosim_stats}
+\end{table}
 
@@
 for review during the annotation process. After the dataset was completely labeled, all samples were checked, corrected, and
-unified\cite{smidl_air_2019}.
-% TODO: data length/size; samples, hours, tokens
+unified\cite{smidl_air_2019}. See Table \ref{tab:zcu_cz_atc_stats} for corpus-specific statistics.
+
+\begin{table}
+    \centering
+    \begin{tabular}{l r}
+        \toprule
+        \multicolumn{2}{c}{ZCU CZ ATC Corpus Properties} \\
+        \midrule
+        Samples              & 14,435                    \\
+        Audio (hours)        & 20.58                     \\
+        Mean Sequence Length & 10.05                     \\
+        Std.~Sequence Length & 5.50                      \\
+        Total Tokens         & 145,107                   \\
+        Unique Tokens        & 3,100                     \\
+        \bottomrule
+    \end{tabular}
+    \caption{Data statistics for ZCU CZ ATC corpus.}
+    \label{tab:zcu_cz_atc_stats}
+\end{table}
 
@@
 A special token is used to represent hesitations when no other context regarding the realization of hesitation is present. If other corpora use a
-special token to represent hesitations, it is translated to match the convention in table~\ref{tab:phenomena_conventions}. Partial/cut-off words
+special token to represent hesitations, it is translated to match the convention in Table \ref{tab:phenomena_conventions}. Partial/cut-off words
 are represented with a hyphen (-) at the point where the pronounced word is cut off. For example, if the last syllable of ``approaching'' is cut
@@
 ``-roaching''. Lastly, if there is enough information in the original transcript to indicate the realization of a hesitation, it is represented by
-the most appropriate token from ``Filled pauses'' in table~\ref{tab:phenomena_conventions}.
+the most appropriate token from ``Filled pauses'' in Table \ref{tab:phenomena_conventions}.
 
@@
 % sequence lengths, unique tokens, total number of tokens, total number of samples, breakdown by each dataset
-Relevant corpora statistics were observed or calculated and are presented in Table~\ref{tab:corpora_stats}. This includes the number of samples in
+Relevant corpora statistics were observed or calculated and are presented in Table \ref{tab:corpora_stats}. This includes the number of samples in
 the corpus, mean sequence length per sample in the corpus, the number of unique tokens in the corpus, and the total number of tokens in the corpus.
@@
 K-Nearest Neighbors algorithm and Minkowski distance for distance metrics/calculations (in this case, since \(p=2\), Euclidean distance is effectively
-used for distance calculations). A total of 47 samples were detected as outliers and removed for analysis and are summarized in table
+used for distance calculations). A total of 47 samples were detected as outliers and removed for analysis and are summarized in Table
 \ref{tab:outlier_stats}.
@@
 the outlying samples manually and reviewing each reveals that, for the most part, the samples are typical albeit somewhat complex communications.
-With the exception of four outliers, all of the samples labeled as outlying by the LOF algorithm were kept in the data.
+With the exception of four outliers, all of the samples labeled as outlying by the LOF algorithm were kept in the data. Examples of the outliers are
+listed in Appendix A in Section \ref{sec:appendix_a}, including the sample removed.
 
@@
 \begin{equation}\label{eq:ngram_sequence}
-    X = \mbox{$w_1$, $w_2$, $w_3$, ..., $w_T$}
+    X = \mbox{$w_1$, $w_2$, $w_3$, ..., $w_T$}\\
 \end{equation}
@@
 \noindent
-the probability of a token, $w_i$, occurring in the sequence can be represented as
-
-\begin{equation}\label{eq:ngram_preceding}
+the probability of tokens occurring in order, in a sequence can be represented using the chain-rule of probability:
+
+\begin{equation}
     \begin{gathered}
-        P(w_i) = P(w_i|w_{i-1},w_{i-2},w_{i-3},...,w_{i-N}) \\
-        P(w_i) = \prod_{j=i-1}^{i-N} P(w_j)
+        P(X) = P(w_1) P(w_2|w_1) P(w_3|w_2, w_1) ... P(w_T|w_{T-1}, w_{T-2}, ..., w_1)\\
+        P(X) = P(w_i)\prod_{i=2}^{T} P(w_i|w_{i-1}, w_{i-2}, ..., w_1)
     \end{gathered}
@@
 \noindent
-or for the case of succeeding tokens (looking ahead $N$ tokens)
-
-\begin{equation}\label{eq:ngram_succeeding}
-    \begin{gathered}
-        P(w_i) = P(w_i|w_{i+1},w_{i+2},w_{i+3},...,w_{i+N}) \\
-        P(w_i) = \prod_{j=i+1}^{i+N} P(w_j)
-    \end{gathered}
-\end{equation}
-
-\noindent
-where $N$ is the number of preceding (or succeeding) tokens. Since the $N$ preceding tokens\footnote{The most common use case for an N-gram model is
-    the preceding tokens.} are needed prior to the estimation of the probability of $w_i$, N-grams are sometimes referred to as left-associative
-models (in a left-to-right language).
-
-The probability of a token occurring anywhere in a sequence (referred to as $w_j$ in equations \ref{eq:ngram_preceding} and \ref{eq:ngram_succeeding})
-is determined statistically as the number occurrences of a token in a corpus divided by the total number of tokens in the corpus.
+Except for the first token in the sequence, the probability of a token occurring at a specific point in the sequence depends on the tokens preceding it
+\footnote{The most common use case for an N-gram model is the preceding tokens.}. Due to this fact, N-grams are sometimes referred to as
+left-associative models (in a left-to-right language).
+
+The probability of a token occurring at all is determined statistically as the number occurrences of a token in a corpus divided by the total number
+of tokens in the corpus.
 
@@
 understand relationships between sentences/sequences to further support downstream tasks. Since BERTs training set consists primarily of literary
-works and excerpts from Wikipedia, the  model simply predicts whether two sentences appear together (one after the other) in the original text.
-Instead of a single input sequence, the input consists of two sequences separated by a ``separation'' token and beginning with a ``classification''
-token, these are represented by \lstinline|[SEP]| and \lstinline|[CLS]|, respectively in Devlin et al.~\cite{devlin_bert_2019}.
+works and excerpts from Wikipedia, the  model simply predicts whether two sentences appear together (one after the other) in the original text,
+making this a binary classification task as opposed to MLM which is a multi-class classification task. Instead of a single input sequence, the input
+consists of two sequences separated by a ``separation'' token and beginning with a ``classification'' token, these are represented by
+\lstinline|[SEP]| and \lstinline|[CLS]|, respectively in Devlin et al.~\cite{devlin_bert_2019}.
 
 For symmetry between pretraining tasks, the special tokens specifying the beginning and end of input sequences are shared between tasks. For both MLM
-and NSP objectives, input sequences always start with a classification token and end with a separation token (see figure
+and NSP objectives, input sequences always start with a classification token and end with a separation token (see Figure
 \ref{fig:bert_mlm_input_example}). Likewise, NSP inputs always begin with a classification token and each input sequence is concatenated with a
-separation token terminating each sequence (see figure \ref{fig:bert_nsp_input_example}).
+separation token terminating each sequence (see Figure \ref{fig:bert_nsp_input_example} and equations (\ref{eq:bert_mlm_input_example}) and
+(\ref{eq:bert_nsp_input_example})).
+
+\begin{equation}\label{eq:bert_mlm_input_example}
+    \mbox{[CLS], the, quick, brown, fox, [SEP]}
+\end{equation}
 
@@
 \end{figure}
+
+\begin{equation}\label{eq:bert_nsp_input_example}
+    \mbox{[CLS], the, quick, brown, fox, [SEP], jumps, over, the, [SEP]}
+\end{equation}
 
@@
 
-The tokenization of words with shared roots but different suffixes is shown in table \ref{tab:wordlevel_tokenization_shared_roots_example}. Clearly
+The tokenization of words with shared roots but different suffixes is shown in Table \ref{tab:wordlevel_tokenization_shared_roots_example}. Clearly
 all three words are seen as independent and unique words by the tokenizer, since they are mapped to their own integer representations.
@@
 \ref{tab:wordlevel_tokenization_shared_roots_example} was trained on two sentences, notably lacking the word ``you'll''. Running the tokenizer on the
-word ``you'll'' alone results in the output in table \ref{tab:wordlevel_unk_word}. Although the word shares a root with some of the words in the
+word ``you'll'' alone results in the output in Table \ref{tab:wordlevel_unk_word}. Although the word shares a root with some of the words in the
 training data (``you'', ``you're'', and ``you've'') and is semantically very similar, the tokenizer does not recognize the word at all (represented
@@
 training procedure (trained on the same data as the WordLevel tokenizer in Section \ref{sec:wordlevel}). Since this sentence was in the training data,
-the segmentation is identical to that of the WordLevel tokenizer in table \ref{tab:wordlevel_tokenization_example}, however, tokenization of similar
-words e.g.~words with shared roots or similar contractions are handled differently as in table \ref{tab:wordpiece_tokenization_shared_roots_example}.
+the segmentation is identical to that of the WordLevel tokenizer in Table \ref{tab:wordlevel_tokenization_example}, however, tokenization of similar
+words e.g.~words with shared roots or similar contractions are handled differently as in Table \ref{tab:wordpiece_tokenization_shared_roots_example}.
 Since each word has a shared root word in the contractions (as well as an apostrophe separating the contraction), the tokenizer segments each word and
 represents them as groups of small, more common (thus with higher probabilities of occurring) tokens. Additionally, the WordPiece algorithm
-effectively handles OOV words. Take the OOV in table \ref{tab:wordpiece_unk_word_similar} with a shared root and similar contraction as those in table
+effectively handles OOV words. Take the OOV in Table \ref{tab:wordpiece_unk_word_similar} with a shared root and similar contraction as those in Table
 \ref{tab:wordpiece_tokenization_shared_roots_example}, which appear in the training data. Since the base of the contraction, ``you'', occurs in the
@@
 a word) and ``\#\#l'' (segmented part of a word that should be concatenated to the previous token when they are recombined). Comparing this to the
-results of the WordLevel tokenizer output on the same word in table \ref{tab:wordlevel_unk_word}, the WordPiece algorithm sidesteps the OOV problem
-for words similar to those in the trining corpus and, according to the output of the WordPiece tokenizer in table \ref{tab:wordpiece_unk_word}, for
+results of the WordLevel tokenizer output on the same word in Table \ref{tab:wordlevel_unk_word}, the WordPiece algorithm sidesteps the OOV problem
+for words similar to those in the trining corpus and, according to the output of the WordPiece tokenizer in Table \ref{tab:wordpiece_unk_word}, for
 words unlike those in the training corpus.
@@
 Now, testing the BPE algorithm on words that did not appear in the training data, it is clear that the BPE algorithm effectively deals with and
-processes OOV words (tables \ref{tab:bpe_unk_similar} and \ref{tab:bpe_unk}), even when no subunits of the word appear in the training data (table
+processes OOV words (tables \ref{tab:bpe_unk_similar} and \ref{tab:bpe_unk}), even when no subunits of the word appear in the training data (Table
 \ref{tab:bpe_unk}).
@@
 own unique and visible character. Functionally, byte-level BPE is nearly identical to standard BPE models except that the beginning of word symbol is
-actually rendered and represented by a visible character in the tokenized sequence (see table \ref{tab:byte_level_bpe_example}).
+actually rendered and represented by a visible character in the tokenized sequence (see Table \ref{tab:byte_level_bpe_example}).
 
@@
 
-The models are trained with the \textit{AdamW} optimizer \cite{loshchilov_decoupled_2019} using the hyperparameters in table \ref{tab:optim_params}
+The models are trained with the \textit{AdamW} optimizer \cite{loshchilov_decoupled_2019} using the hyperparameters in Table \ref{tab:optim_params}
 based on the most effective pretraining configuration presented in \cite{devlin_bert_2019} and \cite{liu_roberta_2019}.
@@
 \subsection{Performance Metrics}\label{sec:performance_metrics}
-As mentioned previously, perplexity and cross-entropy loss (often shortened to loss) are used to track the performance of the model on the training
-and validation sets, per epoch, while training. Cross-entropy loss is defined by the following equation:
+\textbf{Cross-Entropy Loss}. As mentioned previously, perplexity and cross-entropy loss (often shortened to loss) are used to track the performance of
+the model on the training and validation sets, per epoch, while training. Cross-entropy loss is defined by the following equation:
 
@@
 where $x$ is a $512 \times N$ vector representing model inputs (a sequence of tokens) and $y$ is a $1 \times N$ vector representing the target/label
-for each sample in the batch. $N$ is the number of samples in the batch, $l_n$ is the loss for an individual sample, $n$, in the batch, and $C$ is the
-number of class labels. For the purpose of language modeling, the set of class labels is equivalent to the vocabulary of the tokenizer (and thus the
-language model). Numerical labels will correspond to token IDs. The overall loss is then found by taking the mean of the loss over the batch, likewise
-for an epoch, the loss is found by taking the average of each batch in the epoch.
-
-Perplexity is related to the probability of a token appearing in a sequence. It is used as a measure of how well a language model can predict tokens
-in a sequence and is defined by the equation below:
+for each sample in the batch. $L$ is an $N \times 1$ vector containing the losses for each sample in the batch, $N$ is the number of samples in the
+batch, $l_n$ is the loss for an individual sample, $n$, in the batch, and $C$ is the number of class labels. For the purpose of language modeling, the
+set of class labels is equivalent to the vocabulary of the tokenizer (and thus the language model). Numerical labels will correspond to token IDs. The
+overall loss is then found by taking the mean of the loss over the batch, likewise for an epoch, the loss is found by taking the average of each batch
+in the epoch.
+
+\textbf{Perplexity}. Perplexity is related to the probability of a token appearing in a sequence. It is used as a measure of how well a language model
+can predict tokens in a sequence and is defined by the equation below:
 
@@
 
-The perplexity of the models on the training data yields interesting results (figure \ref{fig:training_ppl}). Although the pretrained models have a
+The perplexity of the models on the training data yields interesting results (Figure \ref{fig:training_ppl}). Although the pretrained models have a
 more stable curve over time, the models trained from scratch result in a significantly lower perplexity. Note the difference in the scales on the
-y-axis in figure \ref{fig:training_ppl}; the final perplexity value of each language model trained from scratch is at least two orders of magnitude
+y-axis in Figure \ref{fig:training_ppl}; the final perplexity value of each language model trained from scratch is at least two orders of magnitude
 lower than that of the pretrained models. At the end of training, the BERT and RoBERTa models trained from scratch had perplexity values of
@@
 one of the pretrained checkpoints from Devlin et al.~\cite{devlin_bert_2019} and Liu et al.~\cite{liu_roberta_2019} then fine-tuning on specific
-downstream tasks. This is further upheld by the results of the model performance on the test partition of the corpus in table
+downstream tasks. This is further upheld by the results of the model performance on the test partition of the corpus in Table
 \ref{tab:pretraining_results}, below. As with the training data, the perplexity of the language models trained from scratch over the test partition of
@@
 within their model classes. Within the class of pretrained models, however, the RoBERTa model still technically performed better than the BERT model
-(values shown in table \ref{tab:pretraining_results} were rounded to three significant figures) with a perplexity value of $0.007\times 10^9$ compared
+(values shown in Table \ref{tab:pretraining_results} were rounded to three significant figures) with a perplexity value of $0.007\times 10^9$ compared
 to BERT's $0.014\times 10^9$.
@@
 
-\textbf{Mixed regional data sources}. As described in Section \ref{sec:data_source}, the data aggregated for use here is comes from several different
+\textbf{Mixed regional data sources}. As described in Section \ref{sec:data_source}, the data aggregated for use here comes from several different
 regions, namely international airports in the US \cite{godfrey_air_1994} and international and domestic airports in Europe
@@
 In the aviation domain and for this particular data, that would mean training for tasks such as callsign identification, speaker identification,
-flight phase classification, clearance recognition, etc., however, labels for these tasks are not present in the corpora used here. Corpora with
+flight phase classification, clearance recognition, etc. However, labels for these tasks are not present in the corpora used here. Corpora with
 relevant labels for downstream tasks could not be obtained within an appropriate amount of time to run and include the experiments in this work.
@@
 \section{Conclusion}\label{sec:conclusion}
-Four aviation English corpora are aggregated and unified for use in neural language modeling. A masked language modeling objective is used pretrain
-two classes of BERT and RoBERTa models (models trained from scratch and pretrained models transfer learned to aviation English) on the aggregated
-data.
+Four aviation English corpora are aggregated and unified for use in neural language modeling. Masked language modeling is then used pretrain two
+classes of BERT and RoBERTa models (models trained from scratch and pretrained models transfer learned to aviation English).
 
@@
 The results from the experiments in this work indicate that, for natural language processing tasks in the aviation domain, models pretrained from
-scratch are more effective than transfer learning pretrained models. Thus, models trained from scratch should fine-tuned and used for downstream tasks
-over pretrained models. The most significant limiting factor to the performance of the models is the small volume of aviation English data available
-for pretraining and downstream task use.
+scratch are more effective than transfer learning pretrained models. Thus, models trained from scratch should be fine-tuned and used for downstream
+tasks over pretrained models. The most significant limiting factor to the performance of the models is the small volume of aviation English data
+available for pretraining and downstream task use.
 
@@
 \section{Appendix A: Examples of Samples Detected as Outliers}
+The outlying sample that was removed from the corpus is below\footnote{The preprocessing effects on the string, capitalization, punctuation, etc.~are left
+    unmodified for display purposes}:
+
+\begin{quote}
+    uh uh have a nice evening thank you bye bye ciao yes to the irish pub you know it uh you have to ask i have no idea where it is i'm just walking
+    behind the others uh no uh but you come in front and you ask and he will tell you but i really don't know i don't even know the address okay it's
+    in paris definitely okay bye
+\end{quote}
+
+This sample has a sequence length of 71 tokens, well above the mean of the aggregated corpora. The decision was made to remove this sample, because
+it has little to no relevance to aviation communications, especially the ones being examined by the model in this thesis.
+
+A few other samples, detected as outliers, but not removed are shown in Table \ref{tab:outlier_examples}.
+
+\begin{table}
+    \centering
+    \begin{tabular}{p{0.75\linewidth} r}
+        \toprule
+        Outlying Sample                                                                                                                                                                                                                                                                                                             & Sequence Length \\
+        \midrule
+        ah they're probably putting well i don't know if they are are not the weather is niner hundred scattered measured ceiling one thousand five hundred broken two thousand five hundred overcast visibility one zero temperature five three dew point five zero wind ah zero six zero at three altimeter two niner niner seven & 53              \\
+        \midrule
+        american eight forty six descend and maintain one thousand six hundred turn left heading zero three zero join the localizer you're four from oxonn maintain one thousand six hundred till established cleared i l s three six approach                                                                                      & 38              \\
+        \midrule
+        t w a seven hundred traffic's a helicopter below you at the pentagon he's no factor there's a cessna at one o'clock three miles three thousand five hundred v f r circling he'll stay east of the river                                                                                                                     & 38              \\
+        \midrule
+        yeah we're supposed to ah the thing is now we're going to get you a way out of line for your profile ah in your turn you can continue right around the left to ah zero four zero and then we're going to have to bring you all the way back around to the right                                                             & 55              \\
+        \midrule
+        actually continental ten seventy two he's just ahead a little bit to your right now ten seventy two you're five from ripit maintain two thousand til established cleared i l s d m e approach runway two seven                                                                                                              & 38              \\
+        \midrule
+        continental three eighty two from the i loner cross loner at three or above cleared i l s d m e two seven approach maintain one seven zero knots to loner correction one seven zero knots until ripit                                                                                                                       & 38              \\
+        \midrule
+        u s r sixteen fifteen turn left heading two niner zero three from loner cross loner at three or above cleared i l s d m e runway two seven approach maintain one seven zero knots until ripit                                                                                                                               & 38              \\
+        \midrule
+        clipper sixty five forty seven miles from ripit heading three zero zero maintain three thousand til established on the localizer cleared i l s d m e runway two seven approach speed one ninety or greater to ripit                                                                                                         & 38              \\
+        \midrule
+        cessna three ten bravo whiskey boston tower ah move up and ah the heavy jet's going to runway two two left you'll ah depart prior to both heavies you can move up to the next avlable taxiway it's just it's about a thousand feet ahead turn right and hold short of runway two two right acknowledge hold                 & 57              \\
+        \bottomrule
+    \end{tabular}
+    \caption{A few examples of samples detected as outliers that were kept in the aggregated corpus.}
+    \label{tab:outlier_examples}
+\end{table}
 

### 2023-12-04 source/data/atccomplete.py

@@ -2,10 +2,12 @@ import os
 import glob
 import re
 import json
+import subprocess
 from pprint import pprint
 from typing import *
 from typing import List
 from data import Data, atccutils
+from pathlib import Path
 
 
 class TransmissionTurn(object):
@@ -121,6 +123,21 @@ class ATCCompleteData(Data):
         )
         transcript_glob_string = os.path.join(data_root, "**/data/transcripts/*.txt")
         self._transcript_glob = glob.glob(transcript_glob_string, recursive=True)
+
+        sphere_glob_string = os.path.join(data_root, "**/data/audio/*.sph")
+        wav_glob_string = os.path.join(data_root, "**/data/audio/*.wav")
+
+        # collect audio and transcript file, convert if needed
+        sphere_files = glob.glob(sphere_glob_string, recursive=True)
+
+        #: list of paths to audio files in the dataset.
+        self._audio_glob = glob.glob(wav_glob_string, recursive=True)
+
+        if len(self._audio_glob) != len(self._transcript_glob):
+            self.sphere_to_wav(sphere_files)
+            self._audio_glob = glob.glob(wav_glob_string, recursive=True)
+
+        self.audio_data_paths = self._audio_glob
 
     def parse_transcripts(self) -> List[Dict[str, Union[str, float]]]:
         """
@@ -276,3 +293,34 @@ class ATCCompleteData(Data):
                 )
 
         pprint(transmission_groups)
+
+    def sphere_to_wav(self, sphere_glob: List[str]) -> List[str]:
+        """
+        Runs ffmpeg on each sphere file in `sphere_glob` to convert the
+        files from the NIST Sphere format to the MS Wav format. Also resamples
+        the audio to 16k.
+
+        :param sphere_glob: List of paths (strings) to sphere files to convert
+        :returns: List of converted files
+        """
+        assert len(sphere_glob) > 0
+        converted_paths = []
+
+        for sphere_file in sphere_glob:
+            sphere_file = Path(sphere_file)
+            converted_file = str(sphere_file).replace(".sph", ".wav")
+            if sphere_file.exists():
+                # program and arguments to run to reformat/resample the sphere file
+                process = [
+                    "sox",
+                    f"{sphere_file}",
+                    "-r",
+                    "16000",
+                    f"{converted_file}",
+                ]
+                # for debugging
+                # print(" ".join(ffmpeg_options))
+                subprocess.run(process)
+                converted_paths.append(converted_file)
+
+        return converted_paths

### 2023-12-04 source/data/atco2sim.py

@@ -29,6 +29,7 @@ class ATCO2SimData(Data):
         super(ATCO2SimData, self).__init__(dataset_name="ATCO2", **kwargs)
 
         self.transcripts = glob.glob(os.path.join(data_root, "DATA/*.xml"))
+        self.audio_data_paths = glob.glob(os.path.join(data_root, "DATA/*.wav"))
         assert len(self.transcripts) != 0
 
     def parse_transcripts(self) -> List[str]:

### 2023-12-04 source/data/atcosim.py

@@ -37,6 +37,9 @@ class ATCOSimData(Data):
         self.text_glob = glob.glob(
             os.path.join(data_root, "txtdata/**/*.txt"), recursive=True
         )
+        self.audio_data_paths = glob.glob(
+            os.path.join(data_root, "wavdata/**/*.wav"), recursive=True
+        )
 
         # at the moment this is easier than updating the regex to exclude this specific file
         # TODO: update regex

### 2023-12-04 source/data/czechdataset.py

@@ -14,13 +14,12 @@ class ZCUCZATCDataset(Data):
     Dataset can be obtained here: https://lindat.mff.cuni.cz/repository/xmlui/handle/11858/00-097C-0000-0001-CCA1-0
     """
 
-    _transcript_corrections = [
-        ("+", "-")
-    ]
+    _transcript_corrections = [("+", "-")]
 
     def __init__(self, data_root: str, **kwargs):
         super(ZCUCZATCDataset, self).__init__(dataset_name="ZCU CZ ATC", **kwargs)
         self.transcript_paths = glob.glob(os.path.join(data_root, "*.trs"))
+        self.audio_data_paths = glob.glob(os.path.join(data_root, "*.wav"))
 
         assert (
             len(self.transcript_paths) != 0

### 2023-12-04 source/data/data.py

@@ -6,6 +6,7 @@ from copy import deepcopy
 import matplotlib.pyplot as plt
 import numpy as np
 import pandas as pd
+import librosa
 from numpy.random import default_rng
 from pytorch_lightning.utilities.types import EVAL_DATALOADERS
 from torch.utils.data import IterableDataset, DataLoader
@@ -40,6 +41,7 @@ class Data(IterableDataset):
         """
         self.dataset_name: str = dataset_name
         self.data: List[str] = []
+        self.audio_data_paths: List[str] = []
         self.seq_lens = []
         # create random number generator sequence with specified seed, if applicable
         Data._random: np.random.Generator = np.random.default_rng(random_seed)
@@ -215,6 +217,7 @@ class Data(IterableDataset):
         dataset_summary = f"""
 Name: {self.name}
 Samples: {self.num_samples}
+Hours of audio: {self.audio_hours}
 Mean Sequence Length: {self.average_sequence_length}
 Std of Sequence Length: {self.std_sequence_length}
 Number of Tokens: {self.total_tokens}
@@ -322,6 +325,22 @@ Ratio of unique tokens to the total number of tokens: {self.token_ratio()}, {sel
         Number of samples in the dataset
         """
         return len(self.data)
+
+    @property
+    def audio_hours(self) -> float:
+        """
+        Hours of audio in the dataset
+        """
+        assert len(self.audio_data_paths) != 0
+
+        duration_seconds = 0
+
+        for path in self.audio_data_paths:
+            # duration in seconds
+            duration_seconds += librosa.get_duration(path=path)
+
+        # convert seconds to hours (second -> minutes -> hours; 3600)
+        return duration_seconds / 3600
 
     @property
     def total_tokens(self) -> int:

### 2023-12-04 source/main.py

@@ -84,33 +84,33 @@ def parse_datasets():
     data_objects[0].dataset_name = "All"
     data_objects[0].summary()
 
-    # outliers = data_objects[0].detect_outliers()
-    # print(f"Mean outlier sequence length: {outliers['seq_len'].mean()}")
-    # print(f"Min outlier sequence length: {outliers['seq_len'].min()}")
-    # print(f"Max outlier sequence length: {outliers['seq_len'].max()}")
-    # print(f"Std outlier sequence length: {outliers['seq_len'].std()}")
-    # print(
-    #     f"Most significant outlier:\n{outliers.where(outliers['seq_len'] == outliers['seq_len'].max()).dropna()}"
-    # )
+    outliers = data_objects[0].detect_outliers()
+    print(f"Mean outlier sequence length: {outliers['seq_len'].mean()}")
+    print(f"Min outlier sequence length: {outliers['seq_len'].min()}")
+    print(f"Max outlier sequence length: {outliers['seq_len'].max()}")
+    print(f"Std outlier sequence length: {outliers['seq_len'].std()}")
+    print(
+        f"Most significant outlier:\n{outliers.where(outliers['seq_len'] == outliers['seq_len'].max()).dropna()}"
+    )
 
-    # print(f"Max LOF score: {outliers['negative_outlier_factor_'].min()}")
-    # print(
-    #     outliers["text"]
-    #     .where(
-    #         outliers["negative_outlier_factor_"]
-    #         == outliers["negative_outlier_factor_"].min()
-    #     )
-    #     .dropna()
-    # )
+    print(f"Max LOF score: {outliers['negative_outlier_factor_'].min()}")
+    print(
+        outliers["text"]
+        .where(
+            outliers["negative_outlier_factor_"]
+            == outliers["negative_outlier_factor_"].min()
+        )
+        .dropna()
+    )
 
-    # data_objects[0].summary()
+    data_objects[0].summary()
 
-    # with open("outliers.txt", "a") as f:
-    #     for item, length in zip(
-    #         outliers["text"].tolist(), outliers["seq_len"].tolist()
-    #     ):
-    #         f.write(f"{int(length)}\n")
-    #         f.write(item + "\n")
+    with open("outliers.txt", "w") as f:
+        for item, length in zip(
+            outliers["text"].tolist(), outliers["seq_len"].tolist()
+        ):
+            f.write(f"{int(length)}\n")
[…]

### 2023-12-05 thesis/MSECE_thesis.tex
Last round of thesis edits.

@@
 domain of aviation communications, linguistics research in aviation English, or both. All data in all four corpora include both recordings and
-transcriptions, audio and text, respectively. The transcriptions are used to pretrain and train the language models and tokenizers (described in
-sections \ref{sec:language_modeling} and \ref{sec:tokenizers}).
+transcriptions, which are audio and text, respectively. The transcriptions are used to pretrain and train the language models and tokenizers
+(described in sections \ref{sec:language_modeling} and \ref{sec:tokenizers}).
 
@@
 are removed where necessary\footnote{A copy of the original data is made, then the annotations are removed from the copy.}, and the resulting text
-is written to a file in the standard corpus format (an ASCII text file with one line in the file corresponding to one sample from the corpus) to a
-file.
+is written to a file in the standard corpus format (an ASCII text file with one line in the file corresponding to one sample from the corpus).
 
 \subsection{Handling Spelling Errors \& Inconsistencies}\label{sec:spelling_errors}
-By the nature of the data labeling process, there are bound to be spelling errors introduced by the annotators during corpora creation. This gets
-exacerbated by the aggregation of several corpora with varying controls for correcting spelling errors and inconsistencies. Nearly all of the errors
-are recoverable/correctable given context. For example, ``possibility of \textbf{tornado's}'' should clearly have been ``possibility of
+By the nature of the data labeling process, there is a high likelihood to be spelling errors introduced by the annotators during corpora creation.
+This gets exacerbated by the aggregation of several corpora with varying controls for correcting spelling errors and inconsistencies. Nearly all of
+the errors are recoverable/correctable given context. For example, ``possibility of \textbf{tornado's}'' should clearly have been ``possibility of
 \textbf{tornadoes}'', changing the possessive form of ``tornado'' to the plural form. There are also instances of spelling conventions that are
@@
 instances that were cut-off mid pronunciation). In order to facilitate the unification of the transcripts for all four corpora, the conventions for
-transcribing these phenomena are redefined below:
+transcribing these phenomena are redefined in Table \ref{tab:phenomena_definitions}.
 
@@
 data, and ATCC and ZCU CZ ATC together make up approximately 81\% of the data. The sequence lengths of the samples are fairly consistent across
-corpora with at most about an 18\% percent difference between sequence lengths. The number of unique tokens does not scale additively to the
-combined corpora, which is as expected. Considering that aviation English generally conforms to and follows ICAO phraseology standards we would expect
-that many terms would be shared across corpora with the exception of geographic/location specific terms, such as city, landmark, airport, and airline
-names, etc.~that typically vary across regions.
+corpora with at most about an 18\% percent difference between sequence lengths (calculated using equation (\ref{eq:percent_difference}), below). The
+number of unique tokens does not scale additively to the combined corpora, which is as expected since there will be an overlap of tokens between
+corpora. Aviation English generally conforms to and follows ICAO phraseology standards, so we would expect that many terms would be
+shared across corpora with the exception of geographic/location specific terms, such as city, landmark, airport, and airline names, etc.~that
+typically vary across regions.
+
+\begin{equation}\label{eq:percent_difference}
+    \mbox{PD} = \frac{|V_1 - V_2|}{\frac{V_1 + V_2}{2}} \times 100
+\end{equation}
 
@@
 
-Sci-Kit Learn's \cite{pedregosa_scikit-learn_2011} implementation of the LOF algorithm was used with a value of \(N = 35\) for \(N\) neighbors in the
-K-Nearest Neighbors algorithm and Minkowski distance for distance metrics/calculations (in this case, since \(p=2\), Euclidean distance is effectively
-used for distance calculations). A total of 47 samples were detected as outliers and removed for analysis and are summarized in Table
-\ref{tab:outlier_stats}.
+Sci-Kit Learn's \cite{pedregosa_scikit-learn_2011} implementation of the LOF algorithm was used with a value of \(K = 35\) for the K-Nearest Neighbors
+algorithm and Minkowski distance for distance metrics/calculations (in this case, since \(p=2\), Euclidean distance is effectively used for distance
+calculations). A total of 47 samples were detected as outliers and removed for analysis and are summarized in Table \ref{tab:outlier_stats}.
 
@@
 
-Although all of the outliers have higher sequence lengths than the typical sequence lengths of the other samples in the data, going the through
-the outlying samples manually and reviewing each reveals that, for the most part, the samples are typical albeit somewhat complex communications.
-With the exception of four outliers, all of the samples labeled as outlying by the LOF algorithm were kept in the data. Examples of the outliers are
+Although all of the outliers have longer sequence lengths than the typical sequence lengths of the other samples in the data, going the through the
+outlying samples manually and reviewing each reveals that, for the most part, the samples are typical albeit somewhat complex communications.
+With the exception of one outlier, all of the samples labeled as outlying by the LOF algorithm were kept in the data. Examples of the outliers are
 listed in Appendix A in Section \ref{sec:appendix_a}, including the sample removed.
@@
 % n-gram
-This section introduces the chosen language model architectures as well as their pretrained checkpoints, if one is used. Two transformer-based
-architectures are used, BERT and RoBERTa, and statistical/machine learning-based approaches such as N-gram and word lattices.
+This section introduces the chosen language model architectures as well as their pretrained checkpoints. Two transformer-based architectures are used,
+BERT and RoBERTa, as well as one statistical/machine learning-based approach: N-gram language modeling.
 
@@
 BERT for bidirectional encoding representation and likewise in Liu et al.~\cite{liu_roberta_2019} for training RoBERTa. Given the results of the
-RoBERTa pretraining and downstream task training thereafter, it was initially decided to eliminate the next sentence prediction (NSP) task as it was
-used in BERT, however, their results suggest that the success of the MLM-only approach was due to the increase in batch and overall data
-size~\cite{liu_roberta_2019}. Taking this into account, along with the very limited amount of data available in the domain of aviation an NSP
-pretraining task was implemented where possible\footnote{Only about half of the data (only the data from ATCC, since it is continuous recordings of
-    the airspace) is in a format that enables us to extract the turn taking phenomenon that is present in most air-traffic communications.} to improve
-its pretraining results. MLM and NSP pretraining are explained more in-depth in Section \ref{sec:bert}, below.
+RoBERTa pretraining and downstream task training thereafter, it was decided to eliminate the next sentence prediction (NSP) task that was used to
+pretrain BERT. MLM and NSP pretraining are explained more in-depth in Section \ref{sec:bert}, below.
 
@@
 \textbf{Masked Language Modeling}. The ability of BERT to represent text in a bidirectional context is achieved through the Masked Language Modeling
-(MLM) pretraining task~\cite{devlin_bert_2019}. MLM pretraining choses a certain percentage of token positions at random for prediction, then tokens
-are replaced with either a ``mask'' token, a random token, or the original token according to preconfigured probabilities (BERT used 80\% for the mask
-token, 10\% for random tokens, and 10\% for the original/unchanged token). The original token is then predicted with cross-entropy loss.
-
-\textbf{Next Sentence Prediction}. While MLM conditions the model to represent tokens within sentences, NSP is meant to condition the model to
-understand relationships between sentences/sequences to further support downstream tasks. Since BERTs training set consists primarily of literary
-works and excerpts from Wikipedia, the  model simply predicts whether two sentences appear together (one after the other) in the original text,
-making this a binary classification task as opposed to MLM which is a multi-class classification task. Instead of a single input sequence, the input
-consists of two sequences separated by a ``separation'' token and beginning with a ``classification'' token, these are represented by
-\lstinline|[SEP]| and \lstinline|[CLS]|, respectively in Devlin et al.~\cite{devlin_bert_2019}.
+(MLM) pretraining task~\cite{devlin_bert_2019}. MLM pretraining chooses a certain percentage of token positions at random as labels for prediction,
+those tokens are then replaced with either a ``mask'' token, a random token, or the original token according to preconfigured probabilities (BERT used
+80\% for the mask token, 10\% for random tokens, and 10\% for the original/unchanged token). During training, the masked sequences are used as input
+to the model and the original is used as the prediction target. The model prediction is scored against the target using cross-entropy loss.
+
+\textbf{Next Sentence Prediction}. While MLM conditions the model to represent tokens within sequences and predict the most probable token based on
+its position in the sequence, NSP is meant to condition the model to understand relationships between sentences/sequences to further support
+downstream tasks. Since BERTs training set consists primarily of literary works and excerpts from Wikipedia, the model simply predicts whether two
+sentences appear together (one after the other) in the original text, making this a binary classification task as opposed to MLM which is a
+multi-class classification task. Instead of a single sequence, the input consists of two sequences separated by a ``separation'' token and
+beginning with a ``classification'' token, these are represented by \lstinline|[SEP]| and \lstinline|[CLS]|, respectively in Devlin et
+al.~\cite{devlin_bert_2019}.
 
@@
 \ref{fig:bert_mlm_input_example}). Likewise, NSP inputs always begin with a classification token and each input sequence is concatenated with a
-separation token terminating each sequence (see Figure \ref{fig:bert_nsp_input_example} and equations (\ref{eq:bert_mlm_input_example}) and
-(\ref{eq:bert_nsp_input_example})).
-
-\begin{equation}\label{eq:bert_mlm_input_example}
-    \mbox{[CLS], the, quick, brown, fox, [SEP]}
+separation token terminating each sequence (see Figure \ref{fig:bert_nsp_input_example} and equation (\ref{eq:bert_input_example})).
+
+\begin{equation}\label{eq:bert_input_example}
+    \begin{gathered}
+        \mbox{[CLS], the, quick, brown, fox, [SEP]}\\
+        \mbox{[CLS], the, quick, brown, fox, [SEP], jumps, over, the, [SEP]}
+    \end{gathered}
 \end{equation}
@@
 
-\begin{equation}\label{eq:bert_nsp_input_example}
-    \mbox{[CLS], the, quick, brown, fox, [SEP], jumps, over, the, [SEP]}
-\end{equation}
-
 \begin{figure}[!t]
@@
 segments are separated by a separation token (\lstinline|[SEP]|). Rather than having sequences end with a separation token as well, they are
-terminated with an end of sequence token (\lstinline|[EOS]|). See equation \ref{eq:roberta_input_example} modified from Liu et
+terminated with an end of sequence token (\lstinline|[EOS]|). See equation (\ref{eq:roberta_input_example}) modified from Liu et
 al.~\cite{liu_roberta_2019}, below.
@@
 referred to as a sequence. The rules by which the tokens are created or segmented depend entirely on the algorithm used. The simplest and most
-intuitive way to do this by creating tokens based on whitespace i.e., each ``word'' in the corpus is  treated as a token\footnote{The Huggingface
+intuitive way to do this is by creating tokens based on whitespace i.e., each word in the corpus is treated as a token\footnote{The Huggingface
     Tokenizer library implements this as the \lstinline|WordLevel| tokenizer.}. While this is simple and works in theory, it does not take into
@@
 the original training corpus is known as the out-of-vocabulary (OOV) problem~\cite{wu_googles_2016}. The OOV problem has been addressed by several
-different algorithms. For brevity's sake, only the tokenization algorithms used by the models in this work are defined and explained.
+different algorithms. For brevity's sake, only the tokenization algorithms used by the models in this thesis are defined and explained.
 
@@
 In order to map individual tokens to numerical values or IDs, an index of the words that appear in a corpus must be created. The numerical values for
-each token are derived from this index. For the WordLevel algorithm, this process is referred to as the training the tokenizer model (see algorithm
+each token are derived from this index. For the WordLevel algorithm, this process is referred to as the training the tokenizer model (see Algorithm
 \ref{alg:wordlevel_training}).
@@
 
-The tokenization of words with shared roots but different suffixes is shown in Table \ref{tab:wordlevel_tokenization_shared_roots_example}. Clearly
-all three words are seen as independent and unique words by the tokenizer, since they are mapped to their own integer representations.
+The tokenization of words with shared roots but different suffixes is shown in Table \ref{tab:wordlevel_tokenization_shared_roots_example}. All three
+words are seen as independent and unique words by the tokenizer, since they are mapped to their own integer representations.
 
@@
 Japanese and Korean voice search systems. It is further used by Wu et al.~\cite{wu_googles_2016} for neural machine translation tasks. The algorithm
-demonstrated proficiency and increased performance for the models used in both tasks and effectively addressed the OOV problem.
+demonstrated proficiency and increased performance for the models used in both tasks and effectively addresses the OOV problem.
 
@@
 operation. To achieve this, during training, the vocabulary of the tokenizer is initialized to a basic set of characters; since WordPiece is designed
-to be language agnostic, the initial vocabulary is specified as all basic unicode characters (for the chosen language) in addition to all ASCII
-characters.
+to be language agnostic, the initial vocabulary is specified as all basic unicode characters in addition to all ASCII characters.
 
@@
 the segmentation is identical to that of the WordLevel tokenizer in Table \ref{tab:wordlevel_tokenization_example}, however, tokenization of similar
-words e.g.~words with shared roots or similar contractions are handled differently as in Table \ref{tab:wordpiece_tokenization_shared_roots_example}.
+words (e.g.~words with shared roots or similar contractions) are handled differently, as in Table \ref{tab:wordpiece_tokenization_shared_roots_example}.
 Since each word has a shared root word in the contractions (as well as an apostrophe separating the contraction), the tokenizer segments each word and
-represents them as groups of small, more common (thus with higher probabilities of occurring) tokens. Additionally, the WordPiece algorithm
-effectively handles OOV words. Take the OOV in Table \ref{tab:wordpiece_unk_word_similar} with a shared root and similar contraction as those in Table
-\ref{tab:wordpiece_tokenization_shared_roots_example}, which appear in the training data. Since the base of the contraction, ``you'', occurs in the
-training data as well as the apostrophe in the contraction, so the tokenizer segments those parts of the word as tokens. The last part of the word,
-``ll'', does not appear at all in the training data, so it is broken down into its component characters which are represented as ``l'' (beginning of
-a word) and ``\#\#l'' (segmented part of a word that should be concatenated to the previous token when they are recombined). Comparing this to the
-results of the WordLevel tokenizer output on the same word in Table \ref{tab:wordlevel_unk_word}, the WordPiece algorithm sidesteps the OOV problem
-for words similar to those in the trining corpus and, according to the output of the WordPiece tokenizer in Table \ref{tab:wordpiece_unk_word}, for
-words unlike those in the training corpus.
+represents them as groups of small, more common tokens (thus with higher probabilities of occurring). Additionally, the WordPiece algorithm
+effectively handles OOV words. For example, the OOV tokens in Table \ref{tab:wordpiece_unk_word_similar}, with a shared root and similar contractions
+as those in Table \ref{tab:wordpiece_tokenization_shared_roots_example}, which appear in the training data. The base of the contraction, ``you'',
+occurs in the training data as well as the apostrophe in the contraction, so the tokenizer segments those parts of the word as tokens. The last part
+of the word, ``ll'', does not appear at all in the training data, so it is broken down into its component characters which are represented as ``l''
+(beginning of a word) and ``\#\#l'' (segmented part of a word that should be concatenated to the previous token when they are recombined). Comparing
+this to the output of the WordLevel tokenizer on the same word in Table \ref{tab:wordlevel_unk_word}, the WordPiece algorithm effectively addresses
+the OOV problem for words similar to those in the trining corpus and, according to the output of the WordPiece tokenizer in Table
+\ref{tab:wordpiece_unk_word}, for words unlike those in the training corpus.
 
@@
 % Sennrich et al. 2016; Gage 1994
-The Byte-Pair Encoding (BPE) tokenizer algorithm was adapted from a data compression algorithm by Gage 1994~\cite{gage_feb94_1994}. The data compression
-algorithm works by replacing common pairs of bytes in data with an unused byte to reduce the overall size of the data. The tokenization algorithm
-works under the same principle, merging frequent pairs of characters or tokens into one token~\cite{sennrich_neural_2016}.
+The Byte-Pair Encoding (BPE) tokenizer algorithm was adapted from a data compression algorithm by Phillip Gage in 1994~\cite{gage_feb94_1994}. The
+data compression algorithm works by replacing common pairs of bytes in data with an unused byte to reduce the overall size of the data. The
+tokenization algorithm works under the same principle, merging frequent pairs of characters or tokens into one token~\cite{sennrich_neural_2016}.
 
@@
 
-Now, testing the BPE algorithm on words that did not appear in the training data, it is clear that the BPE algorithm effectively deals with and
+When testing the BPE algorithm on words that did not appear in the training data, it is clear that the BPE algorithm effectively deals with and
 processes OOV words (tables \ref{tab:bpe_unk_similar} and \ref{tab:bpe_unk}), even when no subunits of the word appear in the training data (Table
@@
 \lstinline|tokenizers| library implements it, this is a preprocessing step for the tokenizer model that maps all bytes in a string/sentence to their
-own unique and visible character. Functionally, byte-level BPE is nearly identical to standard BPE models except that the beginning of word symbol is
-actually rendered and represented by a visible character in the tokenized sequence (see Table \ref{tab:byte_level_bpe_example}).
+own unique and visible character. Functionally, byte-level BPE is nearly identical to the standard BPE algorithm, described above, except that the
+beginning of word symbol is actually rendered and represented by a visible character in the tokenized sequence (see Table
+\ref{tab:byte_level_bpe_example}).
 
@@
 tasks, more generally, in the conversational and literary English domains, it follows that transformer architectures should see widespread use in the
-domain of aviation English for domain-specific NLP tasks, however, according to the literature, transformer-based language models have seen sparse
-use. The most notable example of the use of a transformer deep neural network is the application of BERT for knowledge extraction from notice to
-airmen (NOTAM) messages \cite{arnold_knowledge_2022}.
+domain of aviation English for domain-specific NLP tasks. However, according to the literature, transformer-based language models have seen sparse
+use. The most notable example of the use of a transformer neural network is the application of BERT for knowledge extraction from notice to airmen
+(NOTAM) messages \cite{arnold_knowledge_2022}.
+
+The expectation is that the language models trained from scratch will outperform the pretrained language models due to the nature of the differences
+between literary English (the language the pretrained models were trained on) and aviation English. Additionally, the data being modeled are
+orthographic transcriptions, which further extends the difference between the training corpora, thus leading to the hypothesis of the differences
+in model performance. To test this hypothesis, the models will be divided into two sets or classes: (1) models trained from scratch and (2) pretrained
+models that are fine-tuned (or transfer learned) using the data aggregated for this thesis (as described in Section \ref{sec:data_source}), these are
+denoted as the sets of \textit{scratch} and \textit{pretrained} models, respectively. Each class includes a version of the BERT and RoBERTa models,
+starting from scratch or the pretrained checkpoint, depending on the class of the model. In order to perform a fair comparison between models, the
+following parameters (and/or hyperparameters) are kept the same across all four models:
+
+\begin{itemize}
+    \item Optimization algorithm
+    \item Optimizer hyperparameters (see Table \ref{tab:optim_params})
+    \item Number of hidden layers
+    \item Number of neurons per layer
+    \item Number of attention heads
+    \item Activation functions
+    \item Type of tokenizer algorithm
+    \item Training, testing, and validation data
+\end{itemize}
 
@@
 A standard masked language modeling training procedure is used to pretrain the language models, as described in Section \ref{sec:bert}. Tokens are
-masked at random with a uniform probability of 0.2. Input sequences are padded to the maximum length allowed by the model and truncated to the maximum
-length, if greater. The maximum sequence length for BERT and RoBERTa models is 512 tokens \cite{devlin_bert_2019,liu_roberta_2019}. Based on the data
-analysis in Section \ref{sec:data_analysis}, there are few, if any, samples that meet or exceed the maximum sequence length for BERT and RoBERTa
-models.
+masked at random as described in the MLM process in Section \ref{sec:bert}. Input sequences are padded to the maximum length allowed by the model with
+a special \lstinline|[PAD]| token and truncated to the maximum length, if greater. The maximum sequence length for BERT and RoBERTa models is 512
+tokens \cite{devlin_bert_2019,liu_roberta_2019}. Based on the data analysis in Section \ref{sec:data_analysis}, there are few, if any, samples that
+meet or exceed the maximum sequence length for BERT and RoBERTa models.
 
@@
 
-A train/test split of 70\% and 30\% and a validation split of 10\% of the training data was used for each model. The blank models (the models that are
-trained from randomly initialized weights; also referred to as being trained from scratch) were trained for 200 epochs and the pretrained models were
-trained for 100 epochs. These values were chosen through trial and error, monitoring the perplexity and loss curves of the
-validation set to ensure that the pretraining does not result in net negative effects on the loss and perplexity of the model (i.e.~to ensure that the
-models are not overfitting to the training data). Batch sizes of 16 were used for training all four models; this was the maximum batch size that
-worked reliably with the hardware available to train the models\footnote{Values larger than 16 would sometimes cause CUDA Out of Memory errors, so 16
-    was used to ensure all models trained continuously without error and/or interruption.}.
-
-\begin{table}
+A train/test split of 70\% and 30\% and a validation split of 10\% of the training data was used for each model. The models trained from scratch were
+trained for 200 epochs and the pretrained models were trained for 100 epochs. These values were chosen through trial and error, monitoring the
+perplexity and loss curves of the validation set to ensure that the pretraining does not result in net negative effects on the loss and perplexity of
+the model (i.e.~to ensure that the models are not overfitting to the training data). Batch sizes of 16 were used for training all four models; this
+was the maximum batch size that worked reliably with the hardware available to train the models\footnote{Values larger than 16 would sometimes cause
+    Out of Memory errors on the GPU device used for training, so 16 was used to ensure all models trained continuously without error and/or
+    interruption. Models were trained on a workstation 64 GB of RAM and an NVIDIA RTX 3090 GPU with 24 GB of VRAM.}.
+
+\begin{table}[!t]
     \centering
@@
 \textbf{Cross-Entropy Loss}. As mentioned previously, perplexity and cross-entropy loss (often shortened to loss) are used to track the performance of
-the model on the training and validation sets, per epoch, while training. Cross-entropy loss is defined by the following equation:
+the model on the training and validation sets, per epoch, while training. Cross-entropy loss is defined below.
+
+Let $\mathcal{L}(x, y)$ be the cross-entropy loss function that outputs a $N \times 1$ vector, $L$, that contains the loss values for each sample in a
+batch of $N$ samples.
+
+\begin{equation}\label{eq:cross_entropy_defs}
+    \mathcal{L}(x, y) = L = \{\mbox{$l_1$, $l_2$, ..., $l_N$}\}^T
+\end{equation}
+
+\noindent
+Where $x$ is a $512 \times N$ vector representing model inputs (a sequence of tokens) and $y$ is a $1 \times N$ vector representing the target/label
+for each sample in the batch. The cross-entropy loss for a single sample is then calculated as
 
 \begin{equation}\label{eq:cross_entropy_loss}
+    l_n = -\log \left[\frac{\exp(x_{n,y_n})}{\sum_{c=1}^{C}\exp(x_{n,c})}\right] \quad \forall n=1..N
+\end{equation}
+
+\noindent
+where $l_n$ is the loss for an individual sample in the batch, and $C$ is the number of class labels. For the purpose of language modeling, the
+set of class labels is equivalent to the vocabulary of the tokenizer (and thus the language model). Numerical labels will correspond to token IDs. The
+loss for a batch is calculated as the mean of the loss for each sample in the batch.
+
+\begin{equation}\label{eq:cross_entropy_mean}
+    \mathcal{L}(x, y) = \frac{1}{N}\sum_{i=1}^{N}l_i
+\end{equation}
+
+\noindent
+Equation (\ref{eq:cross_entropy_mean}) can then be extended to find the mean cross-entropy loss over an epoch by swapping $l_i$ for batch means and
+$\mathcal{L}(x, y)$ for the epoch mean. $N$ would then represent the number of batch in an epoch rather than the number of samples in a batch.
+
+\textbf{Perplexity}. This metric is used to measure a models ability to predict the most probable token in a sequence. It will be used to measure and
+track how well the language models can predict the most probable token for a specific point in a sequence.
+
+Let $P$ and $\hat{Y}$ be vectors representing the perplexity for samples in a batch, and non-normalized output neurons in a neural network,
+respectively.
+\begin{equation}\label{eq:perplexity_defs}
     \begin{gathered}
-        \mathcal{L}(x, y) = L = \{\mbox{$l_1$, $l_2$, ..., $l_N$}\}^T\\
-        l_n = -\log \left[\frac{\exp(x_{n,y_n})}{\sum_{c=1}^{C}\exp(x_{n,c})}\right] \quad \forall n=1..N
+        P = \{p_1, p_2, ..., p_N\}^T\\
+        \hat{Y} = \{\hat{y}_1, \hat{y}_2, ..., \hat{y}_C\}^T
     \end{gathered}
@@
 \noindent
-where $x$ is a $512 \times N$ vector representing model inputs (a sequence of tokens) and $y$ is a $1 \times N$ vector representing the target/label
-for each sample in the batch. $L$ is an $N \times 1$ vector containing the losses for each sample in the batch, $N$ is the number of samples in the
-batch, $l_n$ is the loss for an individual sample, $n$, in the batch, and $C$ is the number of class labels. For the purpose of language modeling, the
-set of class labels is equivalent to the vocabulary of the tokenizer (and thus the language model). Numerical labels will correspond to token IDs. The
-overall loss is then found by taking the mean of the loss over the batch, likewise for an epoch, the loss is found by taking the average of each batch
-in the epoch.
-
-\textbf{Perplexity}. Perplexity is related to the probability of a token appearing in a sequence. It is used as a measure of how well a language model
-can predict tokens in a sequence and is defined by the equation below:
+Where $N$ is the number of samples in a batch and $C$ is the number of class labels. Then the perplexity for each sample in a batch can be calculated
+as
 
 \begin{equation}\label{eq:perplexity}
-    \begin{gathered}
-        X = \{x_1, x_2, ..., x_N\}\\
-        \mbox{PPL} = \exp\left[\frac{\sum_{i=1}^N - \log_2 x_i}{N}\right]
-    \end{gathered}
+    p_n = \exp \left[ \frac{1}{T}\sum_{i=1}^C - \log(\hat{y}_i) \right] \quad \forall n = 1..N
 \end{equation}
@@
 \noindent
-Where $X$ is an input sequence, $N$ is the length of the sequence, and $x_i$ is an individual token in the sequence, $X$.
+where $p_n$ is the perplexity for an individual sample in the batch, and $T$ is the number of tokens in the sample. The perplexity for the batch is
+then calculated as the mean of the perplexities in the batch.
+
+\begin{equation}\label{eq:perplexity_mean}
+    \mbox{PPL} = \frac{1}{N}\sum_{i=1}^{N}p_i
+\end{equation}
+
+\noindent
+Where $\mbox{PPL}$ is the mean perplexity over a batch with $N$ samples. This equation can then be extended to find the mean perplexity over each
+batch in the epoch, swapping $p_i$ for batch means and $\mbox{PPL}$ for the epoch mean. $N$ would then represent the number of batches in an epoch
+rather than the number of samples in the batch.
 
@@
 With the metrics above defined, an analysis of the results of the model pretraining can be performed. The objective of the pretraining procedure is to
-build the models' understanding of the structure of the text, in this case transcripts of air-traffic communications. By tracking the average
+build the models' understanding of the structure of the text, transcripts of air-traffic communications in this case. By tracking the average
 cross-entropy loss (loss) and perplexity (PPL) during training, we can estimate how well (and how correctly) the models are predicting tokens and if
@@
 
-Graphs of the training loss (average loss over the samples/batches in the training partition of the corpus) are presented in figure
+Graphs of the training loss (average loss over the samples/batches in the training partition of the corpus) are presented in Figure
 \ref{fig:training_loss}. Both classes of models (scratch and pretrained) quickly reach a steady state in terms of loss although the initial loss
@@
 The perplexity of the models on the training data yields interesting results (Figure \ref{fig:training_ppl}). Although the pretrained models have a
-more stable curve over time, the models trained from scratch result in a significantly lower perplexity. Note the difference in the scales on the
-y-axis in Figure \ref{fig:training_ppl}; the final perplexity value of each language model trained from scratch is at least two orders of magnitude
-lower than that of the pretrained models. At the end of training, the BERT and RoBERTa models trained from scratch had perplexity values of
-$0.52\times 10^7$ and $0.16\times 10^7$, respectively, whereas the pretrained models reached perplexity values of $0.29\times 10^9$ and
-$0.17\times 10^9$, for BERT and RoBERTa, respectively. This indicates that best language model performance may be achieved by pretraining models on
-in-domain data, then continuing to fine-tune on specific downstream tasks. This is a direct contradiction of the current recommended practice of using
-one of the pretrained checkpoints from Devlin et al.~\cite{devlin_bert_2019} and Liu et al.~\cite{liu_roberta_2019} then fine-tuning on specific
-downstream tasks. This is further upheld by the results of the model performance on the test partition of the corpus in Table
-\ref{tab:pretraining_results}, below. As with the training data, the perplexity of the language models trained from scratch over the test partition of
-the corpus is significantly lower than that of the pretrained language models. The loss of each model over the test set is somewhat high, but they are
-around the same values for each model, which is consistent with the loss statistics during training. A higher loss value on unseen (test) data is
-normal and expected. The BERT model trained from scratch had the best performance among the models tested despite both RoBERTa models training better
-within their model classes. Within the class of pretrained models, however, the RoBERTa model still technically performed better than the BERT model
-(values shown in Table \ref{tab:pretraining_results} were rounded to three significant figures) with a perplexity value of $0.007\times 10^9$ compared
-to BERT's $0.014\times 10^9$.
+more stable curve over time, the models trained from scratch result in a significantly lower perplexity. The final perplexity values (over the
+training set) of the models trained from scratch were significantly better than that of the pretrained models. At the end of training, the BERT and
+RoBERTa models trained from scratch had perplexity values of $0.52\times 10^7$ and $0.16\times 10^7$, respectively, whereas the pretrained models
+reached perplexity values of $0.29\times 10^9$ and $0.17\times 10^9$, for BERT and RoBERTa, respectively. Using the following equation,
+
+\begin{equation}\label{eq:relative_difference}
+    \mbox{RD} = \frac{v - v_{ref}}{v_{ref}}
+\end{equation}
+
+\noindent
+the relative difference in perplexity between the scratch and pretrained models on the final training epoch is 55.24, for the BERT models, and 104.57,
+for the RoBERTa models, which equates to an absolute difference (equation (\ref{eq:absolute_difference})) of $28.74 \times 10^7$ and
+$16.43 \times 10^7$ for the BERT and RoBERTa model, respectively.
+
+\begin{equation}\label{eq:absolute_difference}
+    \mbox{AD} = |a - b|
+\end{equation}
+
+The results from the training procedure show that the models trained from scratch outperformed their pretrained counterparts. This initially
+indicates that best language model performance may be achieved by pretraining models on in-domain data, then continuing to fine-tune on specific
+downstream tasks. This seems to be a direct contradiction of the current recommended practice of using one of the pretrained checkpoints from Devlin
+et al.~\cite{devlin_bert_2019} and Liu et al.~\cite{liu_roberta_2019} then fine-tuning on specific downstream tasks. However, this is only partially
+upheld by the results of the model performance on the test partition of the corpus (Table \ref{tab:pretraining_results}). For the BERT models, the
+pattern exhibited during training is consistent with the results on the test partition, resulting in a relative difference of 8.62 and an absolute
+difference of $1.24 \times 10^7$. For the RoBERTa models, however, the relative difference in perplexity is -0.34 with an absolute difference of
+$0.36 \times 10^7$. These results show that the pretrained RoBERTa model performed better during pretraining than the corresponding model
+that was trained from scratch, unlike the BERT models (theories as to why this occurs will be explored in the next paragraph). The loss of each
+model over the test set is somewhat high, but they are around the same values for each model, which is consistent with the loss statistics during
+training. A higher loss value on unseen (test) data is normal and expected.
+
+The BERT model trained from scratch had the best performance among the models tested followed closely by the pretrained RoBERTa model, which partially
+confirms the hypothesis that this thesis is based upon. The pretrained BERT model used both MLM and NSP pretraining objectives to create the
+checkpoint used in this thesis. The pretrained RoBERTa model, on the other hand, used only the MLM pretraining objective. It is readily apparent from
+the results of \cite{liu_roberta_2019} that RoBERTas modified pretraining methodology resulted in better model performance for in-domain problems.
+However, it seems that the pretraining method also allows RoBERTa to generalize better than BERT when transfer learned to other NLP domains. It is
+conceivable that, given additional training data in the future, the pretrained RoBERTa model could surpass the performance of the BERT model trained
+from scratch. Although, with the current volume of data available in this domain, it is clear from the results of the models on the test partition of
+the corpus that a BERT model trained from scratch will still be the best option for aviation English NLP tasks.
 
@@
         \toprule
-        Model                & Loss  & Perplexity        \\
-        \midrule
-        BERT (scratch)       & 10.54 & $0.14\times 10^7$ \\ % 1435064.875
-        RoBERTa (scratch)    & 11.41 & $1.04\times 10^7$ \\ % 10376598.000
-        BERT (fine-tuned)    & 11.00 & $0.01\times 10^9$ \\ % 13805354.000
-        RoBERTa (fine-tuned) & 11.80 & $0.01\times 10^9$ \\ % 6816011.500
+        Model                & Loss  & Perplexity ($\times 10^7$) \\
+        \midrule
+        BERT (scratch)       & 10.54 & 0.14                       \\ % 1435064.875
+        RoBERTa (scratch)    & 11.41 & 1.04                       \\ % 10376598.000
+        BERT (pretrained)    & 11.00 & 1.38                       \\ % 13805354.000
+        RoBERTa (pretrained) & 11.80 & 0.68                       \\ % 6816011.500
         \bottomrule
@@
 articles such as those from \textit{Wikipedia}. The data used in this work are orthographic transcriptions of aviation English and while the two data
-sources share the common root of English, transfer learning from literary English to aviation English yields limited results, as evidenced above.
-Especially with relatively small amounts of data.
+sources share the common root of English, transfer learning from literary English to aviation English yields limited results, as evidenced above,
+especially with relatively small amounts of data.
 
@@
 relevant labels for downstream tasks could not be obtained within an appropriate amount of time to run and include the experiments in this work.
-Some potential algorithms were developed in an attempt to mine labels from the corpora aggregated for this work which will be expanded upon in section
-\ref{sec:future_work}. For these reasons, testing model performance and effectiveness on downstream tasks must be left for future work.
+Some potential algorithms were developed in an attempt to mine labels from the corpora aggregated for this work, which will be expanded upon in
+section \ref{sec:future_work}. For these reasons, testing model performance and effectiveness on downstream tasks must be left for future work.
 
 \section{Conclusion}\label{sec:conclusion}
-Four aviation English corpora are aggregated and unified for use in neural language modeling. Masked language modeling is then used pretrain two
-classes of BERT and RoBERTa models (models trained from scratch and pretrained models transfer learned to aviation English).
+Four aviation English corpora are aggregated and unified for use in neural language modeling. A masked language modeling pretraining methodology is
+then used to pretrain two classes of BERT and RoBERTa models (models trained from scratch and pretrained models transfer learned to aviation English).
 
@@
 English, models achieve better performance when pretrained from scratch than model checkpoints transfer learned from the available BERT and RoBERTa
-pretrained checkpoints. The BERT and RoBERTa models trained from scratch, with parameters identical to the pretrained models, reached perplexity
-values nearly two orders of magnitude lower than their corresponding pretrained checkpoints on the training partition of the corpus. Additionally,
-the BERT and RoBERTa models trained from scratch consistently outperformed the corresponding pretrained models on the test partition of the corpus,
-with BERT performing the best overall.
-
-The results from the experiments in this work indicate that, for natural language processing tasks in the aviation domain, models pretrained from
-scratch are more effective than transfer learning pretrained models. Thus, models trained from scratch should be fine-tuned and used for downstream
-tasks over pretrained models. The most significant limiting factor to the performance of the models is the small volume of aviation English data
-available for pretraining and downstream task use.
+pretrained checkpoints. Although the pretrained RoBERTa model achieved better performance on the test partition of the corpus than expected, the
+pretraining methods laid out in this thesis resulted in the BERT model trained from scratch achieving the lowest loss and perplexity overall.
+
+The results from the experiments in this work indicate that, for natural language processing tasks with limited data availability in the aviation
+domain, models trained from scratch are more effective than transfer learning pretrained models. Thus, models trained from scratch should be
+fine-tuned and used for downstream tasks over pretrained models. The most significant limiting factor to the performance of the models is the small
+volume of aviation English data available for pretraining and downstream task use. Investment in the creation of expansive, high quality, openly
+available aviation English corpora with appropriate formatting for the development machine learning applications is also highly recommended.
 

### 2023-12-05 thesis/MSECE_thesis.tex
Grammar revisions.

@@
 \section{Abstract}\label{sec:abstract}
-With recent advances in machine learning and deep learning technologies along with the creation of larger aviation-specific corpora, the application
-of natural language processing technologies to aviation communications, especially those based on transformer neural networks, is becoming
-increasingly feasible. Previous work has been done focusing on machine learning applications to natural language processing such as N-grams and word
-lattices. This thesis will layout a process for pretraining transformer-based language models on aviation English corpora and compare the
-effectiveness and performance of language models transfer learned from pretrained checkpoints and those trained from their base weight initializations
-(trained from scratch). The results show that transformer language models trained from scratch outperform models fine-tuned from pretrained
-checkpoints, achieving perplexity values on the test partition of the corpus nearly two orders of magnitude lower than that of the pretrained models.
-The work concludes by making recommendations for future work to improve pretraining performance as well as suggestions for downstream, in-domain tasks
-such as semantic extraction, named entity recognition (callsign identification), speaker role identification and speech recognition.
+With recent advances in machine learning and deep learning technologies and the creation of larger aviation-specific corpora, applying
+natural language processing technologies to aviation communications, especially those based on transformer neural networks, is becoming
+increasingly feasible. Previous work has focused on machine learning applications to natural language processing, such as N-grams and
+word lattices. This thesis will lay out a process for pretraining transformer-based language models on aviation English corpora and
+compare the effectiveness and performance of language models transfer learned from pretrained checkpoints and those trained from their
+base weight initializations (trained from scratch). The results show that transformer language models trained from scratch outperform
+models fine-tuned from pretrained checkpoints. The work concludes by recommending future work to improve pretraining performance and
+suggestions for downstream, in-domain tasks such as semantic extraction, named entity recognition (callsign identification), speaker
+role identification, and speech recognition.
 
 \section{Introduction}\label{sec:introduction}
-There have been noticeable and significant advancements in the general domain of artificial intelligence (AI) recently. Natural language processing
-and language modeling technologies, in particular, have made tremendous advances over the past two decades achieving performance levels good enough to
-be brought to consumer facing products such as personal assistants, search engines, phones, televisions, etc. Due to the success in the general
-domain, applications have begun to spin-off into problem-specific domains, such as (and most notably for this work) aviation. For example, several
-speech recognition systems with rule- and machine learning-based natural language processing, inference, and semantic extraction algorithms have been
-implemented at EUROCONTROL simulation centers in Europe to train air traffic controllers and study the potential effects of AI augmented air traffic
-management (ATM) systems on controller workloads. Preliminary studies have found that AI augmented ATM systems not only reduce controller workloads
-but also reduce fuel consumption of aircraft by optimizing aircraft routes in the airspace and reducing departure/arrival times thereby reducing fuel
-consumption~\cite{helmke_quantifying_2017}.
-
-At the time of this writing, the current focus of general machine learning and deep learning algorithms in the language modeling domains has been on a
-variety of deep learning methods/architectures, namely; recurrent, convolutional, and transformer neural networks due to the availability of large
-natural language corpora such as WikiText~\cite{merity_pointer_2016} and IMDB~\cite{maas_learning_2011}. In contrast to the general domain, the
-aviation domain still prefers, and sees much success with, traditional machine learning models for language modeling applications. Typically, the most
-prevalent of the traditional models are N-grams, word lattices, and/or rule-based approaches depending on the application and operating environment of
-the models. The most likely reason for this divergence between the domains is the difference and relative lack of labeled data in the
-aviation domain. By comparison, the largest corpora in the general domain is either WikiText or the BookCorpus which contain trillions of tokens
-whereas, for aviation, the largest corpus is Air Traffic Control Complete with approximately 26 hours of labeled data amounting to just over 300,000
-tokens\footnote{Air Traffic Control Complete has approximately 70 hours of audio data, however, about 26 hours of the 70 are labeled.}. This disparity
-in data availability is exacerbated by the fact that aviation English uses a specialized vocabulary in addition to specialized pronunciation for some
-words to enhance understandability in a radiotelephonic (R/T) medium~\cite{paltridge_handbook_2013}, making it significantly more difficult to
-transcribe/label data as it requires transcribers to have domain knowledge and experience in order to create effective transcriptions. This, in
-addition to the following facets, further contribute to the difficulty of transcription: (1) the medium of communication, (2) the environment in which
-the communication occurs\footnote{For example, a noisy and possibly low fidelity medium, depending on the equipment used to transmit and receive
-    communications}, (3) a noisy and active communication environment i.e.~an active flight deck under the influence of wind, engine noise, etc., (4)
-an active Air Traffic Control (ATC) tower and/or channels, and (5) a rapid rate of speech \cite{paltridge_handbook_2013}.
+There have been noticeable and significant advancements in the general domain of artificial intelligence (AI) recently. Natural
+language processing and language modeling technologies, in particular, have made tremendous advances over the past two decades,
+achieving performance levels good enough to be brought to consumer-facing products such as personal assistants, search engines, phones,
+televisions, etc. Due to the success in the general domain, applications have begun to spin off into problem-specific domains, such as
+(and most notably for this work) aviation. For example, several speech recognition systems with rule- and machine learning-based
+natural language processing, inference, and semantic extraction algorithms have been implemented at EUROCONTROL simulation centers in
+Europe to train air traffic controllers and study the potential effects of AI-augmented air traffic management (ATM) systems on
+controller workloads. Preliminary studies have found that AI-augmented ATM systems reduce controller workloads and fuel consumption of
+aircraft by optimizing aircraft routes in the airspace and reducing departure/arrival times, thereby reducing fuel consumption
+\cite{helmke_quantifying_2017}.
+
+At the time of this writing, the current focus of general machine learning and deep learning algorithms in the language modeling domains has been on a variety of deep learning methods/architectures, namely, recurrent, convolutional, and transformer neural networks, due to the availability of large natural language corpora such as WikiText~\cite{merity_pointer_2016} and IMDB~\cite{maas_learning_2011}.
+In contrast to the general domain, the aviation domain still prefers and sees much success with traditional machine learning models for language modeling applications.
+Typically, the most prevalent of the traditional models are N-grams, word lattices, and rule-based approaches, depending on the application and operating environment of the models.
+The most likely reason for this divergence between the domains is the difference and relative lack of labeled data in the aviation domain.
+By comparison, the largest corpora in the general domain are either WikiText or the BookCorpus, which contain trillions of tokens, whereas, for aviation, the largest corpus is Air Traffic Control Complete with approximately 26 hours of labeled data amounting to just over 300,000 tokens\footnote{Air Traffic Control Complete has approximately 70 hours of audio data, however, about 26 hours of the 70 are labeled.}.
+This disparity in data availability is exacerbated by the fact that aviation English uses a specialized vocabulary in addition to specialized pronunciation for some words to enhance understandability in a radiotelephonic (R/T) medium~\cite{paltridge_handbook_2013}, making it significantly more difficult to transcribe/label data as it requires transcribers to have domain knowledge and experience to create effective transcriptions.
+This, in addition to the following facets, further contributes to the difficulty of transcription: (1) the medium of communication, (2) the environment in which the communication occurs\footnote{For example, a noisy and possibly low fidelity medium, depending on the equipment used to transmit and receive communications}, (3) a noisy and active communication environment i.e.~an active flight deck under the influence of wind, engine noise, etc., (4) an active Air Traffic Control (ATC) tower and channels, and (5) a rapid rate of speech \cite{paltridge_handbook_2013}.
 
 % connect complexity of aviation english to utility/necessity of NLP models (ASR and LM) in the aviation domain
-The potential for speech recognition and language modeling algorithms to reach better than human performance (for example, the Stanford Question
-Answering Dataset (SQuAD) benchmark has recorded results consistently better than human since
-2019\footnote{https://rajpurkar.github.io/SQuAD-explorer/})~\cite{zhang_ai_2022} suggests that these NLP algorithms could have high utility in the
-aviation domain. If they can achieve better-than-human performance in the aviation domain, and be used in conjunction with pilot and air traffic
-controller systems, they could easily boost efficiency, reduce workloads, and reduce and/or mitigate errors during normal operations.
-
-The methods and results laid out in this thesis were done with the intention of creating a streamlined process for training and testing deep learning
-based NLP methods, specifically language modeling. The results from the language models suggest that the volume of text used to train the language
-models is not yet sufficient to produce results akin to those in the conversational and/or literary English domains. Additionally, the results show
-that language models trained ``from scratch'' on in-domain data perform better than model trained on out-of-domain data and transfer learned on
-in-domain data i.e.~pretrained BERT and RoBERTa models transfer learned to aviation English.
+The potential for speech recognition and language modeling algorithms to reach better than human performance (for example, the Stanford Question Answering Dataset (SQuAD) benchmark has recorded results consistently better than humans since 2019\footnote{https://rajpurkar.github.io/SQuAD-explorer/})~\cite{zhang_ai_2022} suggests that these NLP algorithms could have high utility in the aviation domain.
+If they can achieve better-than-human performance in the aviation domain and be used with pilot and air traffic controller systems, they could easily boost efficiency, reduce workloads, and reduce and mitigate errors during normal operations.
+
+The methods and results laid out in this thesis were done with the intention of creating a streamlined process for training and testing deep learning-based NLP methods, specifically language modeling.
+The results from the language models suggest that the volume of text used to train the language models is not yet sufficient to produce results akin to those in the conversational and/or literary English domains.
+Additionally, the results show that language models trained ``from scratch'' on in-domain data perform better than models trained on out-of-domain data and transfer learned on in-domain data, i.e., constrained BERT and RoBERTa models transfer learned to aviation English.
 
 \section{Notation and Terminology}
-This section is included for the sake of clarity throughout this thesis. In pursuit of that, there are a few terms and notations that are defined
-below.
-
-Set notation (or set-builder notation)\footnote{That is, the notation used to denote sets and associated operations therein as it relates to the field
-    of set theory, in mathematics.} as defined in \cite{jech_chapter_1978} is used in parts of this thesis, particularly in equations and algorithms,
-to convey mathematical concepts. With that being said, the shorthand below is defined
+This section is included for the sake of clarity throughout this thesis. In pursuit of that, a few terms and notations are defined below.
+
+Set notation (or set-builder notation)\footnote{That is, the notation used to denote sets and associated operations therein as it relates to the field of set theory, in mathematics.} as defined in \cite{jech_chapter_1978} is used in parts of this thesis, particularly in equations and algorithms, to convey mathematical concepts.
+The shorthand below is defined
 \begin{equation*}
@@
 % transformer history (background)
-The transformer neural network architecture was proposed in 2017 for Neural Machine Translation tasks and immediately achieved
-state-of-the-art \cite{vaswani_attention_2017}. Transformer architectures have been found to be extremely effective at learning representations and
-understandings of languages to predict token probabilities as opposed to transforming one language into another
-\cite{devlin_bert_2019,liu_roberta_2019}. Transformers have also been found to be very effective at other NLP-related tasks such as prompt completion
-and sentiment analysis among others (i.e.~auto-regressive and sequence classification tasks, respectively)
-\cite{lewis_bart_2019,radford_improving_2018}. Devlin et al.~\cite{devlin_bert_2019}proposed pretraining methods such as masked language modeling and
-next sentence prediction to improve natural language understanding for downstream tasks, which was further expanded upon by Liu et
-al.~\cite{liu_roberta_2019} by training longer, with larger batch sizes, and removing the next sentence prediction task.
+The transformer neural network architecture was proposed in 2017 for Neural Machine Translation tasks and immediately achieved state-of-the-art \cite{vaswani_attention_2017}.
+Transformer architectures are extremely effective at learning representations and understandings of languages to predict token probabilities instead of transforming one language into another \cite{devlin_bert_2019,liu_roberta_2019}.
+Transformers have also been found to be very effective at other NLP-related tasks such as prompt completion and sentiment analysis among others (i.e.~auto-regressive and sequence classification tasks, respectively) \cite{lewis_bart_2019,radford_improving_2018}.
+Devlin et al.~\cite{devlin_bert_2019}proposed pretraining methods such as masked language modeling and next sentence prediction to improve natural language understanding for downstream tasks, which was further expanded upon by Liu et al.~\cite{liu_roberta_2019} by training longer, with larger batch sizes, and removing the next sentence prediction task.
 
@@
 % LMs and NLP in aero. domain
-Various natural language processing methods have been applied to the aviation domain to help deal with miscommunications
-and try to mitigate safety incidents\cite{ragnarsdottir_language_2003,tanguy_natural_2016,madeira_machine_2021}.
-Some machine learning approaches have been implemented to analyze the text in aviation incident and safety reports to predict
-contributing factors and topic models\cite{tanguy_natural_2016,madeira_machine_2021}. An ASR system with NLP post-processing
-has also been proposed to analyze Air-Traffic communications and condense significant features (e.g.~weather, runway, and
-heading info) into an XML language structure\cite{ragnarsdottir_language_2003}. Transformer language models such as BERT and RoBERTa have been applied
-to notice to airmen (NOTAM) messages to perform named entity recognition (NER) tasks, translation tasks (between notations; e.g.~NOTAM notation to
-Airlang to make parsing tasks easier) and reduce the workload for pilots during long-haul flights\cite{arnold_knowledge_2022}. Transformer models
-have also been used for speaker role identification tasks in the aviation domain (e.g.~identifying pilot versus controller in communications),
-specifically, a pretrained BERT model was used and fine-tuned on problem specific data and compared to other models that performed well at speaker
-and role identification tasks in general\cite{guo_comparative_2022}.
+Various natural language processing methods have been applied to the aviation domain to help deal with miscommunications and try to mitigate safety incidents\cite{ragnarsdottir_language_2003,tanguy_natural_2016,madeira_machine_2021}.
+Some machine learning approaches have been implemented to analyze the text in aviation incident and safety reports to predict contributing factors and topic models\cite{tanguy_natural_2016,madeira_machine_2021}.
+An ASR system with NLP post-processing has also been proposed to analyze Air-Traffic communications and condense significant features (e.g.~weather, runway, and heading info) into an XML language structure\cite{ragnarsdottir_language_2003}.
+Transformer language models such as BERT and RoBERTa have been applied to notice to airmen (NOTAM) messages to perform named entity recognition (NER) tasks, translation tasks (between notations; e.g.~NOTAM notation to Airlang to make parsing tasks easier) and reduce the workload for pilots during long-haul flights\cite{arnold_knowledge_2022}.
+Transformer models have also been used for speaker role identification tasks in the aviation domain (e.g.~identifying pilot versus controller in communications); specifically, a pretrained BERT model was used and fine-tuned on problem-specific data and compared to other models that performed well at speaker and role identification tasks in general\cite{guo_comparative_2022}.
 
 % ASR context (suggest some need for semi-supervised approaches and therefore LMs)
-Automatic speech recognition is seeing increased use in aviation for tasks such as transcription, callsign detection, speaker identification,
-etc. A majority of these approaches use machine learning approaches (as opposed to deep learning) where acoustic, pronunciation, and language
-modeling is used to transcribe speech
-\cite{guo_comparative_2022,smidl_air_2019,zuluaga-gomez_automatic_2020,badrinath_automatic_2022,hofbauer_atcosim_2008,helmke_quantifying_2017}.
-Language models have been used in the general domain (both neural and machine learning-based approaches) to boost the performance of speech
-recognition models \cite{han_contextnet_2020,kriman_quartznet_2020,majumdar_citrinet_2021}. Machine learning-based language models, usually N-grams
-and word lattices have also been used in the aviation domain alongside semi-supervised approaches to automatic speech recognition to leverage
-unlabeled data. However, none of the developed methodologies use neural language models for their applications
-\cite{zuluaga-gomez_contextual_2021,srinivasamurthy_semi-supervised_2017,badrinath_automatic_2022}.
+Automatic speech recognition is seeing increased use in aviation for tasks such as transcription, callsign detection, speaker identification, etc.
+A majority of these approaches use machine learning approaches (as opposed to deep learning) where acoustic, pronunciation, and language modeling is used to transcribe speech \cite{guo_comparative_2022,smidl_air_2019,zuluaga gomez_automatic_2020,badrinath_automatic_2022,hofbauer_atcosim_2008,helmke_quantifying_2017}.
+Language models have been used in the general domain (both neural and machine learning-based approaches) to boost the performance of speech recognition models \cite{han_contextnet_2020,kriman_quartznet_2020,majumdar_citrinet_2021}.
+Machine learning-based language models, usually N-grams and word lattices, have also been used in aviation alongside semi-supervised approaches to automatic speech recognition to leverage unlabeled data.
+However, none of the developed methodologies use neural language models for their applications \cite{zuluaga-gomez_contextual_2021,srinivasamurthy_semi supervised_2017,badrinath_automatic_2022}.
 
 % differences in language (transcripts vs literary english), narrow model selection
-The use of the English language in aviation has been shown to be sufficiently different enough to be classified under its own category (aviation
-English) \cite{paltridge_handbook_2013}. The issue with using pretrained checkpoints of transformer-based language models is characterizing the
-similarity (or, conversely, the difference) between literary English and aviation English. If the two domains are sufficiently different, then
-training the pretrained models on aviation English data would amount to transfer learning between specific domains. Transfer learning between text
-domains has been shown to be possible even with data volumes significantly lower than the original data used to train the models
-\cite{raffel_exploring_2020}. Yadlowsky et al.~\cite{yadlowsky_pretraining_2023} recently showed that if problem domains are sufficiently different,
-model pretraining on data considered out-of-domain will have negative effects on model performance.
+The use of the English language in aviation has been shown to be sufficiently different enough to be classified under its own category (aviation English) \cite{paltridge_handbook_2013}.
+The issue with using pretrained checkpoints of transformer-based language models is characterizing the similarity (or, conversely, the difference) between literary English and aviation English.
+If the two domains are sufficiently different, then training the pretrained models on aviation English data would amount to transfer learning between specific domains.
+Transfer learning between text domains has been shown to be possible even with data volumes significantly lower than the original data used to train the models \cite{raffel_exploring_2020}.
+Yadlowsky et al.~\cite{yadlowsky_pretraining_2023} recently showed that if problem domains are sufficiently different, model pretraining on out-of-domain data will negatively affect model performance.
 
 \section{Data Sources}\label{sec:data_source}
-There are four corpora used in this thesis. The datasets are primarily speech corpora intended either for automatic speech recognition research in the
-domain of aviation communications, linguistics research in aviation English, or both. All data in all four corpora include both recordings and
-transcriptions, which are audio and text, respectively. The transcriptions are used to pretrain and train the language models and tokenizers
-(described in sections \ref{sec:language_modeling} and \ref{sec:tokenizers}).
-
-The four corpora that were selected and used for this work are briefly described below. Further analysis is performed in Section
+Four corpora are used in this thesis.
+The datasets are primarily speech corpora intended for automatic speech recognition research in aviation communications, linguistics research in aviation English, or both.
+All data in all four corpora include both recordings and transcriptions, which are audio and text, respectively.
+The transcriptions are used to pretrain and train the language models and tokenizers (described in sections \ref{sec:language_modeling} and \ref{sec:tokenizers}).
+
+The four corpora selected and used for this work are briefly described below. Further analysis is performed in Section
 \ref{sec:data_analysis}.
@@
 The Air Traffic Control Complete (ATCC) corpus is a speech corpus consisting of audio recordings with corresponding transcriptions.
-ATCC is a collection of three smaller corpora recorded at Dallas-Fort Worth, Logan International, and Washington National airports and
-transcribed by current or former air traffic controllers familiar with the respective areas. Audio data was recorded through the placement of Very
-High Frequency (VHF) antennae configured to monitor several frequencies at each airport, such as arrival, departure, approach, ground, etc. The types
-of frequencies monitored vary by airport, but the presence of departure, approach, and ground frequencies is relatively consistent across subcorpora
-\cite{godfrey_air_1994}. See Table \ref{tab:atcc_stats} for corpus-specific statistics.
+ATCC is a collection of three smaller corpora recorded at Dallas-Fort Worth, Logan International, and Washington National airports and transcribed by current or former air traffic controllers familiar with the respective areas.
+Audio data was recorded by placing Very High Frequency (VHF) antennae configured to monitor several frequencies at each airport, such as arrival, departure, approach, ground, etc.
+The types of frequencies observed vary by airport, but the presence of departure, approach, and ground frequencies is relatively consistent across subcorpora \cite{godfrey_air_1994}.
+See Table \ref{tab:atcc_stats} for corpus-specific statistics.
 
@@
 \subsection{ATCO2}\label{sec:atco2}
-The ATCO2 dataset is a speech corpus that consists of audio communications at Prague and Brno airports, in Czechia, along with corresponding
-transcriptions. The speech recordings and transcripts are crowd sourced from volunteers. The dataset was created and labeled, in part, as an English
-language detection corpus, however, the labeling also includes transcripts of speech segments which makes it practical for ASR and language model
-development in addition to language detection (although in this work it is only used for ASR and LM development)~\cite{szoke_detecting_2021}.
+The ATCO2 dataset is a speech corpus comprising audio communications at Prague and Brno airports in Czechia and corresponding transcriptions.
+The speech recordings and transcripts are crowd-sourced from volunteers.
+The dataset was created and labeled, in part, as an English language detection corpus.
+However, the labeling also includes transcripts of speech segments, which makes it practical for ASR and language model development in addition to language detection (although in this work, it is only used for ASR and LM development)~\cite{szoke_detecting_2021}.
 See Table \ref{tab:atco2_stats} for corpus-specific statistics.
@@
 The Air Traffic Control Simulation (ATCOSIM) corpus is a speech corpus consisting of audio recordings and corresponding transcriptions.
-The data was recorded at the EUROCONTROL Experimental Centre for air traffic control simulation in Bretigny-sur-Orge, France. In this corpus, only the
-controllers' voices are included and therefore the transcripts only include the controller's side of each interaction. The audio data was transcribed
-by one person, trained according to the guidelines and formatting requirements of the corpus. After all data was transcribed, it was reviewed and
-corrected and any remaining problems were reviewed by an operational air traffic controller and resolved \cite{hofbauer_atcosim_2008}. See Table
-\ref{tab:atcosim_stats} for corpus-specific statistics.
+The data was recorded at the EUROCONTROL Experimental Centre for air traffic control simulation in Bretigny-sur-Orge, France.
+In this corpus, only the controllers' voices are included, and therefore, the transcripts only include the controller's side of each interaction.
+The audio data was transcribed by one person, trained according to the guidelines and formatting requirements of the corpus.
+After all data was transcribed, it was reviewed and corrected, and any remaining problems were reviewed by an operational air traffic controller and resolved \cite{hofbauer_atcosim_2008}.
+See Table \ref{tab:atcosim_stats} for corpus-specific statistics.
 
@@
 \subsection{ZCU CZ ATC Corpus}\label{sec:zcu_atc}
-The ZCU CZ ATC corpus consists of audio recordings and corresponding transcripts in the Czech airspace. Both controller and pilot sides of the
-communications are included in this data. Annotations were created by experienced transcribers/annotators and labeled samples were randomly selected
-for review during the annotation process. After the dataset was completely labeled, all samples were checked, corrected, and
-unified\cite{smidl_air_2019}. See Table \ref{tab:zcu_cz_atc_stats} for corpus-specific statistics.
+The ZCU CZ ATC corpus consists of audio recordings and corresponding transcripts in the Czech airspace.
+Both the controller and pilot sides of the communications are included in this data.
+Experienced transcribers/annotators created annotations, and labeled samples were randomly selected for review during the annotation process.
+After the dataset was completely labeled, all samples were checked, corrected, and unified\cite{smidl_air_2019}.
+See Table \ref{tab:zcu_cz_atc_stats} for corpus-specific statistics.
 
@@
 \section{Data Processing}\label{sec:data_processing}
-All of the corpora listed above (see Section \ref{sec:data_source}) are created at different times, for different purposes, and by different authors.
-As a result, the data in these corpora all have different formats and need to be processed into a common format to be used together, with the goal
-being to essentially create a ``super-corpus'' that includes all the corpora listed above (i.e., a superset of the other corpora).
-
-All data was processed in Python using primarily built-in functions/modules\footnote{The \lstinline|re| and \lstinline|xml| modules were also used
-    although these are built into most Python distributions by default.}. The text data is extracted from the corpus transcripts by parsing from the
-corpora-specific formats where necessary i.e.~Lisp in ATCC, XML in ATCO2 and ZCU CZ ATC, and plain text in ATCOSIM, the transcriber annotations
-are removed where necessary\footnote{A copy of the original data is made, then the annotations are removed from the copy.}, and the resulting text
-is written to a file in the standard corpus format (an ASCII text file with one line in the file corresponding to one sample from the corpus).
+All corpora listed above (see Section \ref{sec:data_source}) are created at different times, for various purposes, and by multiple authors.
+As a result, the data in these corpora all have different formats and need to be processed into a common format to be used together, with the goal being to essentially create a ``super-corpus'' that includes all the corpora listed above (i.e., a superset of the other corpora).
+
+All data was processed in Python using primarily built-in functions/modules\footnote{The \lstinline|re| and \lstinline|xml| modules were also used, although these are built into most Python distributions by default.}.
+The text data is extracted from the corpus transcripts by parsing from the corpora-specific formats where necessary, e.g.~Lisp in ATCC, XML in ATCO2 and ZCU CZ ATC, and plain text in ATCOSIM, the transcriber annotations are removed where necessary\footnote{A copy of the original data is made, then the annotations are removed from the copy.}, and the resulting text is written to a file in the standard corpus format (an ASCII text file with one line in the file corresponding to one sample from the corpus).
 
 \subsection{Handling Spelling Errors \& Inconsistencies}\label{sec:spelling_errors}
-By the nature of the data labeling process, there is a high likelihood to be spelling errors introduced by the annotators during corpora creation.
-This gets exacerbated by the aggregation of several corpora with varying controls for correcting spelling errors and inconsistencies. Nearly all of
-the errors are recoverable/correctable given context. For example, ``possibility of \textbf{tornado's}'' should clearly have been ``possibility of
-\textbf{tornadoes}'', changing the possessive form of ``tornado'' to the plural form. There are also instances of spelling conventions that are
-consistent within corpora, but become inconsistent when the corpora are combined.
+By the nature of the data labeling process, there is a high likelihood of spelling errors introduced by the annotators during corpora creation.
+This gets exacerbated by aggregating several corpora with varying controls for correcting spelling errors and inconsistencies.
+Nearly all of the mistakes are recoverable/correctable, given proper context.
+For example, ``possibility of \textbf{tornado's}'' should have been ``possibility of \textbf{tornadoes}'', changing the possessive form of ``tornado'' to the plural form.
+There are also instances of spelling conventions that are consistent within corpora but become inconsistent when combined.
 
 % 22 spelling errors
-Manually reviewing every sample in the corpus is unrealistic considering that there are over 50,000 data samples and would be subject the same
-human error that introduced those errors in the first place. Instead, the frequency of occurrence of tokens in the corpora combined with manual review
-thereafter, was used to detect and correct the most common errors. The frequency of occurrence of a token is calculated using equation
-(\ref{eq:token_frequency_occurrence}).
+Manually reviewing every sample in the corpus is unrealistic, considering that there are over 50,000 data samples and would be subject to the same human error that introduced those errors in the first place.
+Instead, the frequency of token occurrence in the corpora, combined with manual review thereafter, was used to detect and correct the most common errors.
+The frequency of occurrence of a token is calculated using equation (\ref{eq:token_frequency_occurrence}).
 
@@
 \noindent
-Where $f_w$ is the frequency at which a token occurs in a corpus, $\phi_w(w)$ is the number of times a token, $w$, occurs, and $D_{tokens}$ is the
-total number of tokens in the corpus. Tokens with the lowest frequency across corpora (and lowest number of occurrences) were manually reviewed and
-corrected where necessary. An $A \rightarrow B$ mapping system was used to correct the errors, where \(A\) is the erroneous token and \(B\) is the
-corrected or replaced token. As the text is parsed from the corpora, each sample is analyzed, and if an erroneous token, $A$, is found, it is replaced
-with the corrected version, $B$. A total of 22 unique errors were detected and subsequently corrected or removed using this method, although multiple
-instances of those errors typically occurred.
+Where $f_w$ is the frequency at which a token occurs in a corpus, $\phi_w(w)$ is the number of times a token, $w$, occurs, and $D_{tokens}$ is the total number of tokens in the corpus.
+Tokens with the lowest frequency across corpora (and lowest number of occurrences) were manually reviewed and corrected where necessary.
+An $A \rightarrow B$ mapping system was used to correct the errors, where \(A\) is the erroneous token and \(B\) is the corrected or replaced token.
+As the text is parsed from the corpora, each sample is analyzed.
+If an erroneous token, $A$, is found, it is replaced with the corrected version, $B$.
+22 unique errors were detected and subsequently corrected or removed using this method.
+However, multiple instances of those errors typically occurred.
 
 % hesitations, cut off words, different kinds of filled pauses
-The mapping system described above is also used to modify within corpora conventions to be consistent across corpora. Three areas are addressed here:
-hesitations, cut-off or partial words, and filled pauses within speech. These areas are defined differently between the four corpora, which makes
-unifying the transcripts difficult. Hesitations, for example, are defined by one corpus as any incomplete or non-english speech sounds and left
-completely undefined by another (despite annotators making notes about hesitations). The way in which hesitations, pauses, and partial words are
-transcribed differs between corpora as well. Two corpora prescribe special tokens to these phenomena such as ``[hes]'', ``[hesitation]'', ``<pause>'',
-or similar, while the others transcribe the data as it sounds (i.e., ``uh'', ``eh'', or ``er'' for filled pauses and ``cir-'' or ``cir+'' for
-instances that were cut-off mid pronunciation). In order to facilitate the unification of the transcripts for all four corpora, the conventions for
-transcribing these phenomena are redefined in Table \ref{tab:phenomena_definitions}.
+The mapping system described above is also used to modify corpora conventions to be consistent across corpora.
+Three areas are addressed here: hesitations, cut-off, or partial words, and filled pauses within speech.
+These areas are defined differently between the four corpora, which makes unifying the transcripts difficult.
+Hesitations, for example, are defined by one corpus as incomplete or non-English speech sounds and left completely undefined by another (despite annotators making notes about hesitations).
+The way in which hesitations, pauses, and partial words are transcribed differs between corpora as well.
+Two corpora prescribe special tokens to these phenomena, such as ``[hes]'', ``[hesitation]'', ``<pause>'', or similar.
+In contrast, the others transcribe the data as it sounds (i.e., ``uh'', ``eh'', or ``er'' for filled pauses and ``cir-'' or ``cir+'' for instances that were cut-off mid pronunciation).
+To facilitate the unification of the transcripts for all four corpora, the conventions for transcribing these phenomena are redefined in Table \ref{tab:phenomena_definitions}.
 
@@
 
-A special token is used to represent hesitations when no other context regarding the realization of hesitation is present. If other corpora use a
-special token to represent hesitations, it is translated to match the convention in Table \ref{tab:phenomena_conventions}. Partial/cut-off words
-are represented with a hyphen (-) at the point where the pronounced word is cut off. For example, if the last syllable of ``approaching'' is cut
-mid transmission, it would be represented in the transcript as ``approa-'' or, likewise, if the first syllable is cut off, it would be written as
-``-roaching''. Lastly, if there is enough information in the original transcript to indicate the realization of a hesitation, it is represented by
-the most appropriate token from ``Filled pauses'' in Table \ref{tab:phenomena_conventions}.
-
-One of the intended downstream use cases of the language models developed from this work is decoding speech recognition predictions into transcripts,
-so including as much data as possible regarding the realization of spoken words is useful given the context. The \lstinline|[HES]| token is used as a
-catch-all, fallback token in the case that no knowledge can be extracted about the utterance from the transcript.
-
+A special token represents hesitation when no other context is present regarding the realization of hesitation.
+If other corpora use a special token to express hesitations, it is translated to match the convention in Table \ref{tab:phenomena_conventions}.
+Partial/cut-off words are represented with a hyphen (-) at the point where the pronounced word is cut off.
+For example, if the last syllable of ``approaching'' is cut mid-transmission, it would be represented in the transcript as ``approa-'' or, likewise, if the first syllable is cut off, it would be written as ``-roaching''.
+Lastly, if there is enough information in the original transcript to indicate the realization of a hesitation, it is represented by the most appropriate token from ``Filled pauses'' in Table \ref{tab:phenomena_conventions}.
+
+One of the intended downstream use cases of the language models developed from this work is decoding speech recognition predictions into transcripts, so including as much data as possible regarding spoken word realization is useful given the context.
+The \lstinline|[HES]| token is used as a catch-all fallback token if no knowledge about the utterance from the transcript can be extracted.
 \section{Data Analysis}\label{sec:data_analysis}
 % sequence lengths, unique tokens, total number of tokens, total number of samples, breakdown by each dataset
-Relevant corpora statistics were observed or calculated and are presented in Table \ref{tab:corpora_stats}. This includes the number of samples in
-the corpus, mean sequence length per sample in the corpus, the number of unique tokens in the corpus, and the total number of tokens in the corpus.
-The same statistics were also calculated across corpora, since all four corpora will be combined for training and testing the language and speech
-recognition models. Note that this is based on the raw data after loading and unifying the corpora, prior to preprocessing for model training/testing
-and outlier removal.
+Relevant corpora statistics were observed or calculated and are presented in Table \ref{tab:corpora_stats}.
+This includes the number of samples in the corpus, the mean sequence length per sample, the number of unique tokens, and the total number of tokens in the corpus.
+The same statistics were also calculated across corpora since all four corpora will be combined for training and testing the language and speech recognition models.
+Note that this is based on the raw data after loading and unifying the corpora before preprocessing for model training/testing and outlier removal.
 
@@
 
-The ATCC and ZCU CZ ATC corpora make up a significant portion of the data, ATCC alone makes up more than half (approximately 54\%) of the combined
-data, and ATCC and ZCU CZ ATC together make up approximately 81\% of the data. The sequence lengths of the samples are fairly consistent across
-corpora with at most about an 18\% percent difference between sequence lengths (calculated using equation (\ref{eq:percent_difference}), below). The
-number of unique tokens does not scale additively to the combined corpora, which is as expected since there will be an overlap of tokens between
-corpora. Aviation English generally conforms to and follows ICAO phraseology standards, so we would expect that many terms would be
-shared across corpora with the exception of geographic/location specific terms, such as city, landmark, airport, and airline names, etc.~that
-typically vary across regions.
+The ATCC and ZCU CZ ATC corpora make up a significant portion of the data, ATCC alone makes up more than half (approximately 54\%) of the combined data, and ATCC and ZCU CZ ATC together make up approximately 81\% of the data.
+The sequence lengths of the samples are fairly consistent across corpora with at most about an 18\% percent difference between sequence lengths (calculated using equation (\ref{eq:percent_difference}), below).
+The number of unique tokens does not scale additively to the combined corpora, which is expected since there will be an overlap of tokens.
+Aviation English generally conforms to and follows ICAO phraseology standards, so we expect many terms to be shared across corpora except geographic and location-specific terms, such as city, landmark, airport, and airline names, which vary across regions.
 
@@
 \subsection{Outlier Detection and Analysis}\label{sec:outliers}
+% TODO: should this be described in more detail? component equations, algs, etc.?
 The Local Outlier Factor (LOF) algorithm \cite{breunig_lof_2000} was used to detect and analyze outliers in the corpora. The sequence lengths of the
@@
 
-Sci-Kit Learn's \cite{pedregosa_scikit-learn_2011} implementation of the LOF algorithm was used with a value of \(K = 35\) for the K-Nearest Neighbors
-algorithm and Minkowski distance for distance metrics/calculations (in this case, since \(p=2\), Euclidean distance is effectively used for distance
-calculations). A total of 47 samples were detected as outliers and removed for analysis and are summarized in Table \ref{tab:outlier_stats}.
+Sci-Kit Learn's \cite{pedregosa_scikit-learn_2011} implementation of the LOF algorithm was used with a value of \(K = 35\) for the K-Nearest Neighbors algorithm and Minkowski distance for distance metrics/calculations (in this case, since \(p=2\), Euclidean distance is effectively used for distance calculations).
+A total of 47 samples were detected as outliers and removed for analysis.
+They are summarized in Table \ref{tab:outlier_stats}.
 
@@
 
-Although all of the outliers have longer sequence lengths than the typical sequence lengths of the other samples in the data, going the through the
-outlying samples manually and reviewing each reveals that, for the most part, the samples are typical albeit somewhat complex communications.
-With the exception of one outlier, all of the samples labeled as outlying by the LOF algorithm were kept in the data. Examples of the outliers are
-listed in Appendix A in Section \ref{sec:appendix_a}, including the sample removed.
+Although all of the outliers have longer sequence lengths than the typical sequence lengths of the other samples in the data, going through the outlying samples manually and reviewing each reveals that, for the most part, the samples are typical albeit somewhat complex communications.
+Except for one outlier, all of the samples labeled outlying by the LOF algorithm were kept in the data.
+Examples of the outliers are listed in Appendix A in Section \ref{sec:appendix_a}, including the sample removed.
 
@@
 % n-gram
-This section introduces the chosen language model architectures as well as their pretrained checkpoints. Two transformer-based architectures are used,
-BERT and RoBERTa, as well as one statistical/machine learning-based approach: N-gram language modeling.
-
-The transformers are pre-trained using the same masked language modeling (MLM) objective presented in Devlin et al.~\cite{devlin_bert_2019} to train
-BERT for bidirectional encoding representation and likewise in Liu et al.~\cite{liu_roberta_2019} for training RoBERTa. Given the results of the
-RoBERTa pretraining and downstream task training thereafter, it was decided to eliminate the next sentence prediction (NSP) task that was used to
-pretrain BERT. MLM and NSP pretraining are explained more in-depth in Section \ref{sec:bert}, below.
+This section introduces the chosen language model architectures and their pretrained checkpoints.
+Two transformer-based architectures, BERT and RoBERTa, and one statistical/machine learning-based approach: N-gram language modeling.
+
+The transformers are pre-trained using the same masked language modeling (MLM) objective presented in Devlin et al.~\cite{devlin_bert_2019} to train BERT for bidirectional encoding representation and likewise in Liu et al.~\cite{liu_roberta_2019} for training RoBERTa.
+Given the results of the RoBERTa pretraining and downstream task training thereafter, it was decided to eliminate the next sentence prediction (NSP) task that was used to pretrain BERT.
+MLM and NSP pretraining are explained more in-depth in Section \ref{sec:bert} below.
 
@@
 \noindent
-Except for the first token in the sequence, the probability of a token occurring at a specific point in the sequence depends on the tokens preceding it
-\footnote{The most common use case for an N-gram model is the preceding tokens.}. Due to this fact, N-grams are sometimes referred to as
-left-associative models (in a left-to-right language).
-
-The probability of a token occurring at all is determined statistically as the number occurrences of a token in a corpus divided by the total number
-of tokens in the corpus.
+Except for the first token in the sequence, the probability of a token occurring at a specific point depends on the tokens preceding it\footnote{The most common use case for an N-gram model is the preceding tokens.}.
+Due to this fact, N-grams are sometimes called left-associative models (in a left-to-right language).
+
+The probability of a token occurring at all is determined statistically as the number of occurrences in a corpus divided by the total number of tokens.
 
@@
 \noindent
-where $P(w)$ is the probability of a token, $w$, occurring anywhere in the corpus, $\phi_w(w)$ is the number of times of a specific
-token appears in the corpus, and $D_{tokens}$ is total number of tokens in the corpus, $D$. Computationally, the process of determining the
-probabilities of the individual tokens in the corpus is typically referred to as training or fitting the N-gram model.
-
-$N$ is a predetermined value based on model or application requirements and while it can technically take any value from one to infinity, its
-usability becomes very limited at the minimum sequence length in the corpus and unusable at any value at or above the maximum sequence length.
-Depending on implementation, the time and resources required to train a model with a large value of $N$ can be very expensive. The most common values
-for $N$ in the literature are one, two, and three, sometimes called unigrams, bigrams, and trigrams, respectively.
+where $P(w)$ is the probability of a token, $w$, occurring anywhere in the corpus, $\phi_w(w)$ is the number of times a specific token appears in the corpus, and $D_{tokens}$ is the total number of tokens in the corpus, $D$.
+Computationally, determining the probabilities of the individual tokens in the corpus is typically referred to as training or fitting the N-gram model.
+
+$N$ is a predetermined value based on model or application requirements.
+While it can technically take any value from one to infinity, its usability becomes very limited at the minimum sequence length in the corpus and unusable at any value at or above the maximum sequence length.
+Depending on implementation, the time and resources required to train a model with a large value of $N$ can be very expensive.
+Common values for $N$ are one, two, and three, sometimes called unigrams, bigrams, and trigrams, respectively.
 
 \subsection{BERT}\label{sec:bert}
-The purpose of BERT (Bidirectional Encoder Representations from Transformers) is to represent tokens in text sequences in relation to the tokens
-preceding and following the token in question. Whereas an N-gram model represents token probabilities based on the $N$ tokens preceding a specific
-token or, conversely, the $N$ tokens succeeding said token, BERT models are able to predict the probability of a token based on both preceding and
-succeeding tokens, thus the bidirectionality novelty.
-
-\textbf{Masked Language Modeling}. The ability of BERT to represent text in a bidirectional context is achieved through the Masked Language Modeling
-(MLM) pretraining task~\cite{devlin_bert_2019}. MLM pretraining chooses a certain percentage of token positions at random as labels for prediction,
-those tokens are then replaced with either a ``mask'' token, a random token, or the original token according to preconfigured probabilities (BERT used
-80\% for the mask token, 10\% for random tokens, and 10\% for the original/unchanged token). During training, the masked sequences are used as input
-to the model and the original is used as the prediction target. The model prediction is scored against the target using cross-entropy loss.
-
-\textbf{Next Sentence Prediction}. While MLM conditions the model to represent tokens within sequences and predict the most probable token based on
-its position in the sequence, NSP is meant to condition the model to understand relationships between sentences/sequences to further support
-downstream tasks. Since BERTs training set consists primarily of literary works and excerpts from Wikipedia, the model simply predicts whether two
-sentences appear together (one after the other) in the original text, making this a binary classification task as opposed to MLM which is a
-multi-class classification task. Instead of a single sequence, the input consists of two sequences separated by a ``separation'' token and
-beginning with a ``classification'' token, these are represented by \lstinline|[SEP]| and \lstinline|[CLS]|, respectively in Devlin et
-al.~\cite{devlin_bert_2019}.
-
-For symmetry between pretraining tasks, the special tokens specifying the beginning and end of input sequences are shared between tasks. For both MLM
-and NSP objectives, input sequences always start with a classification token and end with a separation token (see Figure
-\ref{fig:bert_mlm_input_example}). Likewise, NSP inputs always begin with a classification token and each input sequence is concatenated with a
-separation token terminating each sequence (see Figure \ref{fig:bert_nsp_input_example} and equation (\ref{eq:bert_input_example})).
+The purpose of BERT (Bidirectional Encoder Representations from Transformers) is to represent tokens in text sequences based on the tokens preceding and following the token in question.
+Whereas an N-gram model represents token probabilities based on the $N$ tokens preceding a specific token or, conversely, the $N$ tokens succeeding said token, BERT models can predict the probability of a token based on both preceding and succeeding tokens; the bidirectionality novelty.
+
+\textbf{Masked Language Modeling}.
+The ability of BERT to represent text in a bidirectional context is achieved through the Masked Language Modeling (MLM) pretraining task~\cite{devlin_bert_2019}.
+MLM pretraining chooses a certain percentage of token positions at random as labels for prediction; those tokens are then replaced with either a ``mask'' token, a random token, or the original token according to preconfigured probabilities (BERT used 80\% for the mask token, 10\% for random tokens, and 10\% for the original/unchanged token).
+During training, the masked sequences are used as input to the model, and the original is used as the prediction target.
+The model prediction is scored against the target using cross-entropy loss.
+
+\textbf{Next Sentence Prediction}.
+While MLM conditions the model to represent tokens within sequences and predict the most probable token based on its position in the sequence, NSP is meant to condition the model to understand relationships between sentences/sequences to support downstream tasks.
+Since the BERT training set consists primarily of literary works and excerpts from Wikipedia, the model simply predicts whether two sentences appear together (one after the other) in the original text, making this a binary classification task as opposed to MLM, which is a multi-class classification task.
+Instead of a single sequence, the input consists of two sequences separated by a ``separation'' token and beginning with a ``classification'' token; these are represented by \lstinline|[SEP]| and \lstinline|[CLS]|, respectively in Devlin et al.~\cite{devlin_bert_2019}.
+
+The special tokens specifying the beginning and end of input sequences are shared between tasks for symmetry between pretraining tasks.
+For both MLM and NSP objectives, input sequences always start with a classification token and end with a separation token (see Figure \ref{fig:bert_mlm_input_example}).
+Likewise, NSP inputs always begin with a classification token, and each input sequence is concatenated with a separation token terminating each sequence (see Figure \ref{fig:bert_nsp_input_example} and equation (\ref{eq:bert_input_example})).
 
@@
 \subsection{RoBERTa}\label{sec:roberta}
-As the name implies, Robustly Optimized BERT Pretraining Approach (RoBERTa), is an iteration on BERT (described in Section \ref{sec:bert}, above).
-The results from Liu et al.~\cite{liu_roberta_2019} indicate that BERT was initially significantly undertrained, so by training longer, with larger
-batch sizes, on longer sequences, and dynamically changing the masking pattern applied to the MLM pretraining task, the performance of BERT was
-improved. The new checkpoint of this version of BERT was therefore called RoBERTa.
-
-The pretraining methodology introduced by RoBERTa eliminated the NSP pretraining task that was originally used in BERT and opted instead to increase
-the quantity of training data along with the batch sizes in addition to training the model longer. Lastly, there is a small difference in the
-construction of input sequences as compared to BERT; input sequences always begin with a classification token (\lstinline|[CLS]|) and individual
-segments are separated by a separation token (\lstinline|[SEP]|). Rather than having sequences end with a separation token as well, they are
-terminated with an end of sequence token (\lstinline|[EOS]|). See equation (\ref{eq:roberta_input_example}) modified from Liu et
-al.~\cite{liu_roberta_2019}, below.
+As the name implies, the Robustly Optimized BERT Pretraining Approach (RoBERTa) is an iteration on BERT (described in Section \ref{sec:bert}, above).
+The results from Liu et al.~\cite{liu_roberta_2019} indicate that BERT was initially significantly undertrained, so by training longer, with larger batch sizes, on longer sequences, and dynamically changing the masking pattern applied to the MLM pretraining task, the performance of BERT was improved.
+The new checkpoint of this version of BERT was, therefore, called RoBERTa.
+
+The pretraining methodology introduced by RoBERTa eliminated the NSP pretraining task originally used in BERT and opted to increase the quantity of training data along with the batch sizes and train the model longer.
+Lastly, there is a small difference in the construction of input sequences as compared to BERT; input sequences always begin with a classification token (\lstinline|[CLS]|), and individual segments are separated by a separation token (\lstinline|[SEP]|).
+Rather than having sequences end with a separation token, they are terminated with an end-of-sequence token (\lstinline|[EOS]|).
+See equation (\ref{eq:roberta_input_example}) modified from Liu et al.~\cite{liu_roberta_2019}, below.
 
@@
 
-% \section{Speech Recognition}\label{sec:speech_recognition}
-% Automatic speech recognition (ASR) or simply speech recognition is the process of translating spoken language to text i.e.~transcribing spoken
-% language. There have been a myriad of approaches to ASR problems, including, but not limited to rule-based, machine learning, and deep learning
-% methods. For the purposes of this work, we will focus primarily on deep learning approaches to ASR tasks with a specific focus on transcribing
-% spoken language in aviation (following from the data accumulated, processed, and analyzed in sections \ref{sec:data_source},
-% \ref{sec:data_processing}, and \ref{sec:data_analysis}).
-
-% Three ASR architectures were chosen due to their success in the conversational English domain and ease of reproducing results. Nvidia maintains a
-% Python package called Neural Modules (shortened to NeMo) with PyTorch implementations of many well-performing ASR models\footnote{Performance of these
-%     models can be seen in their respective papers (referenced hereafter).}, including the three models used in this work and described in more detail
-% below \cite{kuchaiev_nemo_2019}.
-
-% Although the internal processing of data varies between architectures, the format of the models input and output is generally the same. ASR models
-% have seen much success and performance gains through the use of Mel-frequency Cepstral Coefficients (MFCCs), so the audio files provided by the
-% corpora used in this work were processed and resampled into the same format, then the MFCCs are extracted according to the process
-% in~\cite{sahidullah_design_2012}. The MFCCs are then used as input to the ASR models. Each model outputs a probability distribution of characters or
-% tokens (depending on the vocabulary the ASR model is initialized with) over a segment of audio. All the ASR models used are trained with a
-% Connectionist Temporal Classification (CTC) loss function~\cite{graves_connectionist_2006}.
-
-% \subsection{Jasper}\label{sec:jasper}
-% Jasper is a family of end-to-end convolutional neural networks (CNNs) developed by NVIDIA to replace traditional ASR models that use separately
-% learned components for each stage of the pipeline (acoustic, pronunciation, language modeling, etc.) \cite{li_jasper_2019}.
-
-% Jasper models are designed with $B$ blocks and $R$ sub-blocks and are named accordingly as \textit{Jasper $B\mbox{x}R$} models. The Jasper
-% architecture also contains four extra convolutional blocks, one for preprocessing and three for postprocessing (see figure \ref{fig:jasper}). Each
-% convolutional sub-block contains one dimensional convolutions. Li et al.~trained and tested a Jasper 10x5 model that produced state-of-the-art
-% results on the LibriSpeech, Wall Street Journal, and Fisher+Switchboard speech corpora \cite{li_jasper_2019}.
-
-% \begin{figure}[!t]
-%     \includegraphics[width=\linewidth]{figures/jasper_vertical.png}
-%     \caption{Block diagram of the Jasper architecture. Copied from Li et al.~\cite{li_jasper_2019}.}
-%     \label{fig:jasper}
-% \end{figure}
-
-% \subsection{QuartzNet}\label{sec:quartznet}
-% The QuartzNet architecture is based on Jasper with some key modifications, namely, that the one dimensional convolutions are replaced with ``one
-% dimensional time-channel separable convolutions''. As with Jasper, QuartzNet models are denoted as \textit{QuartzNet $B\mbox{x}R$}, however, blocks
-% are repeated $S$ times and have a certain number of input and output channels, $c_{in}$ and $c_{out}$, respectively.
-
-% In Kriman et al.~experiments with QuartzNet were conducted on the LibriSpeech and Wall Street Journal speech corpora and achieved near state-of-the-art
-% performance on both. A transfer learning experiment was also performed (transferring between corpora). The model was initially trained on LibriSpeech
-% and Mozilla Common Voice, then fine-tuned on the Wall Street Journal corpus and also achieved near state-of-the-art results \cite{kriman_quartznet_2020}.
-% These results suggest that the QuartzNet architecture not only performs well on in-domain tasks, but can generalize and fine-tune well on related
-% data.
-
-% \begin{figure}[!t]
-%     \includegraphics[width=\linewidth]{figures/quartz_vertical.png}
-%     \caption{Block diagram of the QuartzNet architecture. Copied from Kriman et al.~\cite{kriman_quartznet_2020}.}
-%     \label{fig:quartznet}
-% \end{figure}
-
-% \subsection{Citrinet}\label{sec:citrinet}
-% Citrinet is an end-to-end CNN based on QuartzNet (and thus Jasper) and augmented with one dimensional Squeeze and Excitation modules
-% \cite{hu_squeeze-and-excitation_2018}. Citrinet models are denoted as \textit{Citrinet $B$x$R$x$C$} with $B$ representing the number of blocks, $R$
-% representing the number of repeated sub-blocks, and $C$ representing the number of channels. Similarly to Jasper and QuartzNet, Citrinet has a set of
-% prolog and epilog blocks, in this case one of each, the first and last-most blocks, respectively. The core of the model architecture consists of three
-% ``mega blocks'', which are themselves made up of seven QuartzNet blocks that are repeated $R$ times along with a Squeeze and Excitation module at the
-% end of the block. The value of $C$ follows from that of QuartzNet (section \ref{sec:quartznet}, above; $c_{in}\mbox{x}c_{out}$) \cite{majumdar_citrinet_2021}.
-
-% \begin{figure}[!t]
-%     \includegraphics[width=\linewidth]{figures/citrinet_vertical.png}
-%     \caption{Block diagram of the Citrinet architecture. Copied from Majumdar et al.~\cite{majumdar_citrinet_2021}.}
-% \end{figure}
-
-% Citrinet also uses a WordPiece tokenizer (described in more detail in section \ref{sec:wordpiece}), typically with a vocabulary size of 1024 tokens
-% (or word pieces) in the experiments performed by Majumdar et al. The results from the experiments showed model performance on ASR transcription tasks
-% near or better than the other state-of-the-art models that the \textit{Citrinet-1024} was compared against \cite{majumdar_citrinet_2021}.
-
 \section{Tokenization Algorithms}\label{sec:tokenizers}
 % present and explain the tokenization algorithms used (possibly which models they correspond to)
-Tokenization is the process of separating or segmenting words in sentences into their component ``tokens''. The tokenized version of a sentence is
-referred to as a sequence. The rules by which the tokens are created or segmented depend entirely on the algorithm used. The simplest and most
-intuitive way to do this is by creating tokens based on whitespace i.e., each word in the corpus is treated as a token\footnote{The Huggingface
-    Tokenizer library implements this as the \lstinline|WordLevel| tokenizer.}. While this is simple and works in theory, it does not take into
-account the fact that not every word in a language will occur in a corpus. The inevitable possibility for a tokenizer to see a word that was not in
-the original training corpus is known as the out-of-vocabulary (OOV) problem~\cite{wu_googles_2016}. The OOV problem has been addressed by several
-different algorithms. For brevity's sake, only the tokenization algorithms used by the models in this thesis are defined and explained.
+Tokenization is the process of separating or segmenting words in sentences into their component ``tokens''.
+The tokenized version of a sentence is referred to as a sequence.
+The rules by which the tokens are created or segmented depend entirely on the algorithm used.
+The simplest and most intuitive way to do this is by creating tokens based on whitespace, i.e., each word in the corpus is treated as a token\footnote{The Huggingface Tokenizer library implements this as the \lstinline|WordLevel| tokenizer.}.
+While this is simple and works in theory, it does not consider that not every word in a language will occur in a corpus.
+The inevitable possibility of a tokenizer seeing a word not in the original training corpus is known as the out-of-vocabulary (OOV) problem~\cite{wu_googles_2016}.
+Several different algorithms have addressed the OOV problem.
+For brevity's sake, only the tokenization algorithms used by the models in this thesis are defined and explained.
 
 \subsection{WordLevel}\label{sec:wordlevel}
-The WordLevel \cite{wolf_transformers_2020} tokenization algorithm, mentioned briefly as an introductory example above, is the most technically simple
-algorithm among those introduced in this section. Tokens are extrapolated from words in a sentence based on the whitespace separating each word.
-This means that contractions, hyphenations, etc.~are treated as their own words i.e.~ ``you'', ``you're'', and ``you've'' are all seen as unique,
-distinct words that are tokenized and mapped to their own integral representations.
-
-In order to map individual tokens to numerical values or IDs, an index of the words that appear in a corpus must be created. The numerical values for
-each token are derived from this index. For the WordLevel algorithm, this process is referred to as the training the tokenizer model (see Algorithm
-\ref{alg:wordlevel_training}).
-
-The first step of the training procedure is to index any specified special tokens, such as mask, classification, or separation tokens (this is why
-special tokens often have low-valued token IDs). The second step is to split the sentences in the training data into individual words. This is done
-by splitting each string based on whitespace, in other words the tokenizer treats sentences as lists of words delimited by space characters. Lastly,
-the tokenizer iterates through each word in each sentence, indexing new words as it finds them, until all sentences in the training corpus have been
-analyzed. Token IDs are then derived based on the position of the word in the tokenizer index or vocabulary.
+The WordLevel \cite{wolf_transformers_2020} tokenization algorithm, mentioned briefly as an introductory example above, is the most technically simple algorithm among those introduced in this section.
+Tokens are extrapolated from words in a sentence based on the whitespace separating each word.
+This means that contractions, hyphenations, etc.
+are treated as their own words, i.e.~ ``you'', ``you're'', and ``you've'' are all seen as unique, distinct words tokenized and mapped to their integral representations.
+
+An index of the words that appear in a corpus must be created to map individual tokens to numerical values or IDs.
+The numerical values for each token are derived from this index.
+For the WordLevel algorithm, this process is referred to as the tokenizer model training (see Algorithm \ref{alg:wordlevel_training}).
+
+The first step of the training procedure is to index any specified special tokens, such as mask, classification, or separation tokens (this is why special tokens often have low-valued token IDs).
+The second step is to split the sentences in the training data into individual words.
+This is done by splitting each string based on whitespace; in other words, the tokenizer treats sentences as lists of words delimited by space characters.
+Lastly, the tokenizer iterates through each word in each sentence, indexing new words as it finds them until all sentences in the training corpus have been analyzed.
+Token IDs are then derived based on the word's position in the tokenizer index or vocabulary.
 
@@
 
-The simplicity of the algorithm speaks for itself and serves as a good introduction to tokenization as well as an illustrative example of the OOV
-problem. Note that the values of the integers representing the tokens are determined based on the order in which the tokens are seen during the
-training process (special tokens are defined and therefore seen first), so these results are completely reproducible as long as the data appears in
-the same order.
-
-The tokenization of words with shared roots but different suffixes is shown in Table \ref{tab:wordlevel_tokenization_shared_roots_example}. All three
-words are seen as independent and unique words by the tokenizer, since they are mapped to their own integer representations.
+The algorithm's simplicity speaks for itself and serves as a good introduction to tokenization and an illustrative example of the OOV problem.
+Note that the values of the integers representing the tokens are determined based on the order in which the tokens are seen during the training process (special tokens are defined and, therefore, seen first), so these results are completely reproducible as long as the data appears in the same order.
+
+The tokenization of words with shared roots but different suffixes is shown in Table \ref{tab:wordlevel_tokenization_shared_roots_example}.
+The tokenizer sees all three words as independent and unique words since they are mapped to their own integer representations.
 
@@
 
-The tokenizer used to generate the examples in tables \ref{tab:wordlevel_tokenization_example} and
-\ref{tab:wordlevel_tokenization_shared_roots_example} was trained on two sentences, notably lacking the word ``you'll''. Running the tokenizer on the
-word ``you'll'' alone results in the output in Table \ref{tab:wordlevel_unk_word}. Although the word shares a root with some of the words in the
-training data (``you'', ``you're'', and ``you've'') and is semantically very similar, the tokenizer does not recognize the word at all (represented
-by the ``[UNK]'' or unknown token). This word would be considered out-of-vocabulary since it did not appear in the training data and thus reveals one
-of the major downsides of this algorithm, as mentioned in section \ref{sec:tokenizers}. This problem can be alleviated to some extent by modifying the
-preprocessing strategy of the tokenizer to partition contracted words into their component parts e.g.~the word ``you're'' would be partitioned to
-\lstinline|[``you'', ``''', ``re'']|, however, it eventually reappears, for example, when using the past, present, and future tenses of a verb
-(e.g.~``go'' and ``goes'' in the training data, then encountering the future tense: ``going'').
+The tokenizer used to generate the examples in tables \ref{tab:wordlevel_tokenization_example} and \ref{tab:wordlevel_tokenization_shared_roots_example} was trained on two sentences, notably lacking the word ``you'll''.
+Running the tokenizer on the word ``you'll'' alone results in the output in Table \ref{tab:wordlevel_unk_word}.
+Although the word shares a root with some of the words in the training data (``you'', ``you're'', and ``you've'') and is semantically very similar, the tokenizer does not recognize the word at all (represented by the ``[UNK]'' or unknown token).
+This word would be considered out-of-vocabulary since it did not appear in the training data and thus reveals one of the major downsides of this algorithm, as mentioned in section \ref{sec:tokenizers}.
+This problem can be alleviated to some extent by modifying the preprocessing strategy of the tokenizer to partition contracted words into their component parts, e.g.~the word ``you're'' would be partitioned to \lstinline|[``you'', ``''', ``re'']|, however, it eventually reappears, for example, when using the past, present, and future tenses of a verb (e.g.~``go'' and ``goes'' in the training data, then encountering the future tense: ``going'').
 
@@
 % Wu et al. 2016; Schuster and Nakajima 2012
-The WordPiece algorithm was developed by Schuster and Nakajima \cite{schuster_japanese_2012} as a word segmentation algorithm for language modeling in
-Japanese and Korean voice search systems. It is further used by Wu et al.~\cite{wu_googles_2016} for neural machine translation tasks. The algorithm
-demonstrated proficiency and increased performance for the models used in both tasks and effectively addresses the OOV problem.
+Schuster and Nakajima \cite{schuster_japanese_2012} developed the WordPiece algorithm as a word segmentation algorithm for language modeling in Japanese and Korean voice search systems.
+It is further used by Wu et al.~\cite{wu_googles_2016} for neural machine translation tasks.
+The algorithm demonstrated proficiency and increased performance for the models used in both tasks and effectively addresses the OOV problem.
 
@@
 
-The training procedure is simple, but computationally expensive. Schuster and Nakajima calculated the time complexity of the algorithm at $O(K^2)$
-where $K$ is the current size of the vocabulary~\cite{schuster_japanese_2012}. Due to the high time complexity of the algorithm, the authors suggest
-the following considerations to reduce the computational complexity:
+The training procedure is simple but computationally expensive.
+Schuster and Nakajima calculated the algorithm's time complexity at $O(K^2)$ where $K$ is the current size of the vocabulary~\cite{schuster_japanese_2012}.
+Due to the high time complexity of the algorithm, the authors suggest the following considerations to reduce the computational complexity:
 
@@
 
-Table \ref{tab:wordpiece_tokenization_example}, below, shows an example of a WordPiece tokenizer model output on a sentence that was seen during the
-training procedure (trained on the same data as the WordLevel tokenizer in Section \ref{sec:wordlevel}). Since this sentence was in the training data,
-the segmentation is identical to that of the WordLevel tokenizer in Table \ref{tab:wordlevel_tokenization_example}, however, tokenization of similar
-words (e.g.~words with shared roots or similar contractions) are handled differently, as in Table \ref{tab:wordpiece_tokenization_shared_roots_example}.
-Since each word has a shared root word in the contractions (as well as an apostrophe separating the contraction), the tokenizer segments each word and
-represents them as groups of small, more common tokens (thus with higher probabilities of occurring). Additionally, the WordPiece algorithm
-effectively handles OOV words. For example, the OOV tokens in Table \ref{tab:wordpiece_unk_word_similar}, with a shared root and similar contractions
-as those in Table \ref{tab:wordpiece_tokenization_shared_roots_example}, which appear in the training data. The base of the contraction, ``you'',
-occurs in the training data as well as the apostrophe in the contraction, so the tokenizer segments those parts of the word as tokens. The last part
-of the word, ``ll'', does not appear at all in the training data, so it is broken down into its component characters which are represented as ``l''
-(beginning of a word) and ``\#\#l'' (segmented part of a word that should be concatenated to the previous token when they are recombined). Comparing
-this to the output of the WordLevel tokenizer on the same word in Table \ref{tab:wordlevel_unk_word}, the WordPiece algorithm effectively addresses
-the OOV problem for words similar to those in the trining corpus and, according to the output of the WordPiece tokenizer in Table
-\ref{tab:wordpiece_unk_word}, for words unlike those in the training corpus.
+Table \ref{tab:wordpiece_tokenization_example}, below, shows an example of a WordPiece tokenizer model output on a sentence that was seen during the training procedure (trained on the same data as the WordLevel tokenizer in Section \ref{sec:wordlevel}).
+Since this sentence was in the training data, the segmentation is identical to that of the WordLevel tokenizer in Table \ref{tab:wordlevel_tokenization_example}, however, tokenization of similar words (e.g.~words with shared roots or similar contractions) are handled differently, as in Table \ref{tab:wordpiece_tokenization_shared_roots_example}.
+Since each word has a shared root word in the contractions (as well as an apostrophe separating the contraction), the tokenizer segments each word and represents them as groups of small, more common tokens (thus with higher probabilities of occurring).
+Additionally, the WordPiece algorithm effectively handles OOV words.
+For example, the OOV tokens in Table \ref{tab:wordpiece_unk_word_similar}, with a shared root and similar contractions as those in Table \ref{tab:wordpiece_tokenization_shared_roots_example}, which appear in the training data.
+The base of the contraction, ``you'', occurs in the training data and the apostrophe in the contraction, so the tokenizer segments those parts of the word as tokens.
+The last part of the word, ``ll'', does not appear at all in the training data, so it is broken down into its component characters, which are represented as ``l'' (beginning of a word) and ``\#\#l'' (segmented part of a word that should be concatenated to the previous token when they are recombined).
+Comparing this to the output of the WordLevel tokenizer on the same word in Table \ref{tab:wordlevel_unk_word}, the WordPiece algorithm effectively addresses the OOV problem for words similar to those in the training corpus and, according to the output of the WordPiece tokenizer in Table \ref{tab:wordpiece_unk_word}, for words unlike those in the training corpus.
 
@@
 % Sennrich et al. 2016; Gage 1994
-The Byte-Pair Encoding (BPE) tokenizer algorithm was adapted from a data compression algorithm by Phillip Gage in 1994~\cite{gage_feb94_1994}. The
-data compression algorithm works by replacing common pairs of bytes in data with an unused byte to reduce the overall size of the data. The
-tokenization algorithm works under the same principle, merging frequent pairs of characters or tokens into one token~\cite{sennrich_neural_2016}.
-
-The vocabulary of the tokenizer is initialized to the base set of characters (letters, numbers, punctuation, etc.~). The training procedure begins by
-counting all pairs (``A'', ``B'') of characters that appear in the training data and combines the most frequently occurring pair into one token
-(``A'', ``B'' $\rightarrow$ ``AB''). This process repeats until a specified vocabulary size has been reached (or a specified number of merge
-operations have occurred; the final vocabulary size is equal to the base character set plus the number of merge operations). Algorithm
-\ref{alg:bep_training} shows a minimal Python implementation for the BPE training procedure.
+The Byte-Pair Encoding (BPE) tokenizer algorithm was adapted from a data compression algorithm by Phillip Gage in 1994~\cite{gage_feb94_1994}.
+The data compression algorithm works by replacing common pairs of bytes in data with an unused byte to reduce the overall size of the data.
+The tokenization algorithm works under the same principle, merging frequent pairs of characters or tokens into one token~\cite{sennrich_neural_2016}.
+
+The vocabulary of the tokenizer is initialized to the base set of characters (letters, numbers, punctuation, etc.~).
+The training procedure begins by counting all pairs (``A'', ``B'') of characters that appear in the training data and combines the most frequently occurring pair into one token (``A'', ``B'' $\rightarrow$ ``AB'').
+This process repeats until a specified vocabulary size has been reached (or a specified number of merge operations have occurred; the final vocabulary size is equal to the base character set plus the number of merge operations).
+Algorithm \ref{alg:bep_training} shows a minimal Python implementation for the BPE training procedure.
 
@@
 
-Finally, comparing the BPE tokenization process with the WordLevel and WordPiece algorithms, we can see that the outputs of the algorithm are nearly
-identical when the sentences being tokenized have words that appear in the training data (i.e.~the BPE outputs in tables
-\ref{tab:bpe_tokenization_example} and \ref{tab:bpe_shared_roots_example} are almost identical to the WordPiece output in tables
-\ref{tab:wordpiece_tokenization_example} and \ref{tab:wordpiece_tokenization_shared_roots_example} with the exception of the numerical representation
-of the tokens) and very similar to that of the WordLevel tokenizer.
+Finally, comparing the BPE tokenization process with the WordLevel and WordPiece algorithms, we can see that the outputs of the algorithm are nearly identical when the sentences being tokenized have words that appear in the training data (i.e.~the BPE outputs in tables \ref{tab:bpe_tokenization_example} and \ref{tab:bpe_shared_roots_example} are almost identical to the WordPiece output in tables \ref{tab:wordpiece_tokenization_example} and \ref{tab:wordpiece_tokenization_shared_roots_example} except the numerical representation of the tokens) and very similar to that of the WordLevel tokenizer.
 
@@
 
-\textbf{Byte-Level Byte-Pair Encoding}. This is a subset of the BPE algorithm that is also commonly used for NLP tasks such as neural machine
-translation, language modeling, and generative tasks (GPT-2 uses a byte-level BPE tokenizer~\cite{radford_language_2019}). As Huggingface's
-\lstinline|tokenizers| library implements it, this is a preprocessing step for the tokenizer model that maps all bytes in a string/sentence to their
-own unique and visible character. Functionally, byte-level BPE is nearly identical to the standard BPE algorithm, described above, except that the
-beginning of word symbol is actually rendered and represented by a visible character in the tokenized sequence (see Table
-\ref{tab:byte_level_bpe_example}).
+\textbf{Byte-Level Byte-Pair Encoding}.
+This is a subset of the BPE algorithm that is also commonly used for NLP tasks such as neural machine translation, language modeling, and generative tasks (GPT-2 uses a byte-level BPE tokenizer~\cite{radford_language_2019}).
+As Huggingface's \lstinline|tokenizers| library implements it, this is a preprocessing step for the tokenizer model that maps all bytes in a string/sentence to their own unique and visible character.
+Functionally, byte-level BPE is nearly identical to the standard BPE algorithm, described above, except that the beginning of word symbol is rendered and represented by a visible character in the tokenized sequence (see Table \ref{tab:byte_level_bpe_example}).
 
@@
 \section{Discussion}\label{sec:discussion}
-Considering the success of deep learning models with transformer-based architectures for language modeling and natural language processing (NLP)
-tasks, more generally, in the conversational and literary English domains, it follows that transformer architectures should see widespread use in the
-domain of aviation English for domain-specific NLP tasks. However, according to the literature, transformer-based language models have seen sparse
-use. The most notable example of the use of a transformer neural network is the application of BERT for knowledge extraction from notice to airmen
-(NOTAM) messages \cite{arnold_knowledge_2022}.
-
-The expectation is that the language models trained from scratch will outperform the pretrained language models due to the nature of the differences
-between literary English (the language the pretrained models were trained on) and aviation English. Additionally, the data being modeled are
-orthographic transcriptions, which further extends the difference between the training corpora, thus leading to the hypothesis of the differences
-in model performance. To test this hypothesis, the models will be divided into two sets or classes: (1) models trained from scratch and (2) pretrained
-models that are fine-tuned (or transfer learned) using the data aggregated for this thesis (as described in Section \ref{sec:data_source}), these are
-denoted as the sets of \textit{scratch} and \textit{pretrained} models, respectively. Each class includes a version of the BERT and RoBERTa models,
-starting from scratch or the pretrained checkpoint, depending on the class of the model. In order to perform a fair comparison between models, the
-following parameters (and/or hyperparameters) are kept the same across all four models:
+Considering the success of deep learning models with transformer-based architectures for language modeling and natural language processing (NLP) tasks, more generally, in the conversational and literary English domains, it follows that transformer architectures should see widespread use in aviation English for domain-specific NLP tasks.
+However, the literature shows transformer-based language models have seen sparse use.
+The most notable example of using a transformer neural network is the application of BERT for knowledge extraction from notice to airmen (NOTAM) messages \cite{arnold_knowledge_2022}.
+
+The expectation is that the language models trained from scratch will outperform the pretrained language models due to the nature of the differences between literary English (the language the pretrained models were trained on) and aviation English.
+Additionally, the modeled data are orthographic transcriptions, extending the difference between the training corpora, thus leading to the hypothesis of the differences in model performance.
+To test this hypothesis, the models will be divided into two sets or classes: (1) models trained from scratch and (2) pretrained models that are fine-tuned (or transfer learned) using the data aggregated for this thesis (as described in Section \ref{sec:data_source}), these are denoted as the sets of \textit{scratch} and \textit{pretrained} models, respectively.
+Each class includes a version of the BERT and RoBERTa models, starting from scratch or the pretrained checkpoint, depending on the class of the model.
+To perform a fair comparison between models, the following parameters (and hyperparameters) are kept the same across all four models:
 
@@
 \subsection{Language Model Pretraining}\label{sec:lm_pretraining}
-A standard masked language modeling training procedure is used to pretrain the language models, as described in Section \ref{sec:bert}. Tokens are
-masked at random as described in the MLM process in Section \ref{sec:bert}. Input sequences are padded to the maximum length allowed by the model with
-a special \lstinline|[PAD]| token and truncated to the maximum length, if greater. The maximum sequence length for BERT and RoBERTa models is 512
-tokens \cite{devlin_bert_2019,liu_roberta_2019}. Based on the data analysis in Section \ref{sec:data_analysis}, there are few, if any, samples that
-meet or exceed the maximum sequence length for BERT and RoBERTa models.
+A standard masked language modeling training procedure is used to pretrain the language models, as described in Section \ref{sec:bert}.
+Tokens are masked at random as described in the MLM process in Section \ref{sec:bert}.
+Input sequences are padded to the maximum length the models accept with a special \lstinline|[PAD]| token and truncated to the maximum length, if greater.
+The maximum sequence length for BERT and RoBERTa models is 512 tokens \cite{devlin_bert_2019,liu_roberta_2019}.
+Based on the data analysis in Section \ref{sec:data_analysis}, there are few, if any, samples that meet or exceed the maximum sequence length for BERT and RoBERTa models.
 
@@
 
-A train/test split of 70\% and 30\% and a validation split of 10\% of the training data was used for each model. The models trained from scratch were
-trained for 200 epochs and the pretrained models were trained for 100 epochs. These values were chosen through trial and error, monitoring the
-perplexity and loss curves of the validation set to ensure that the pretraining does not result in net negative effects on the loss and perplexity of
-the model (i.e.~to ensure that the models are not overfitting to the training data). Batch sizes of 16 were used for training all four models; this
-was the maximum batch size that worked reliably with the hardware available to train the models\footnote{Values larger than 16 would sometimes cause
-    Out of Memory errors on the GPU device used for training, so 16 was used to ensure all models trained continuously without error and/or
-    interruption. Models were trained on a workstation 64 GB of RAM and an NVIDIA RTX 3090 GPU with 24 GB of VRAM.}.
+A train/test split of 70\% and 30\% and a validation split of 10\% of the training data was used for each model.
+The models trained from scratch were trained for 200 epochs and the pretrained models were trained for 100 epochs.
+These values were chosen through trial and error, monitoring the perplexity and loss curves of the validation set to ensure that the pretraining does not result in net negative effects on the loss and perplexity of the model (i.e., ensure that the models are not overfitting to the training data).
+Batch sizes of 16 were used for training all four models; this was the maximum batch size that worked reliably with the hardware available to train the models\footnote{Values larger than 16 would sometimes cause Out of Memory errors on the GPU device used for training, so 16 was used to ensure all models trained continuously without error or interruption.
+Models were trained on a workstation with 64 GB of RAM and an NVIDIA RTX 3090 GPU with 24 GB of VRAM.}.
 
@@
 \subsection{Performance Metrics}\label{sec:performance_metrics}
-\textbf{Cross-Entropy Loss}. As mentioned previously, perplexity and cross-entropy loss (often shortened to loss) are used to track the performance of
-the model on the training and validation sets, per epoch, while training. Cross-entropy loss is defined below.
-
-Let $\mathcal{L}(x, y)$ be the cross-entropy loss function that outputs a $N \times 1$ vector, $L$, that contains the loss values for each sample in a
-batch of $N$ samples.
+\textbf{Cross-Entropy Loss}.
+As mentioned previously, perplexity and cross-entropy loss (often shortened to loss) are used to track the model's performance on the training and validation sets per epoch while training.
+Cross-entropy loss is defined below.
+
+Let $\mathcal{L}(x, y)$ be the cross-entropy loss function that outputs a $N \times 1$ vector, $L$, that contains the loss values for each sample in a batch of $N$ samples.
 
@@
 \noindent
-where $l_n$ is the loss for an individual sample in the batch, and $C$ is the number of class labels. For the purpose of language modeling, the
-set of class labels is equivalent to the vocabulary of the tokenizer (and thus the language model). Numerical labels will correspond to token IDs. The
-loss for a batch is calculated as the mean of the loss for each sample in the batch.
+where $l_n$ is the loss for an individual sample in the batch, and $C$ is the number of class labels.
+For language modeling, the set of class labels is equivalent to the vocabulary of the tokenizer (and thus the language model).
+Numerical labels will correspond to token IDs.
+The loss for a batch is calculated as the mean of the loss for each sample in the batch.
 
@@
 \noindent
-Equation (\ref{eq:cross_entropy_mean}) can then be extended to find the mean cross-entropy loss over an epoch by swapping $l_i$ for batch means and
-$\mathcal{L}(x, y)$ for the epoch mean. $N$ would then represent the number of batch in an epoch rather than the number of samples in a batch.
-
-\textbf{Perplexity}. This metric is used to measure a models ability to predict the most probable token in a sequence. It will be used to measure and
-track how well the language models can predict the most probable token for a specific point in a sequence.
-
-Let $P$ and $\hat{Y}$ be vectors representing the perplexity for samples in a batch, and non-normalized output neurons in a neural network,
-respectively.
+Equation (\ref{eq:cross_entropy_mean}) can then be extended to find the mean cross-entropy loss over an epoch by swapping $l_i$ for batch means and $\mathcal{L}(x, y)$ for the epoch mean.
+$N$ would then represent the number of batches in an epoch rather than the number of samples in a batch.
+
+\textbf{Perplexity}.
+This metric measures a model's ability to predict the most probable token in a sequence.
+It will measure and track how well the language models can predict the most probable token for a specific point in a sequence.
+
+Let $P$ and $\hat{Y}$ be vectors representing the perplexity for samples in a batch and non-normalized output neurons in a neural network, respectively.
+
 \begin{equation}\label{eq:perplexity_defs}
@@
 \noindent
-Where $N$ is the number of samples in a batch and $C$ is the number of class labels. Then the perplexity for each sample in a batch can be calculated
-as
+Where $N$ is the number of samples in a batch and $C$ is the number of class labels. Then, the perplexity for each sample in a batch can be calculated as
 
@@
 \subsection{Pretraining Results}\label{sec:pretraining_results}
-With the metrics above defined, an analysis of the results of the model pretraining can be performed. The objective of the pretraining procedure is to
-build the models' understanding of the structure of the text, transcripts of air-traffic communications in this case. By tracking the average
-cross-entropy loss (loss) and perplexity (PPL) during training, we can estimate how well (and how correctly) the models are predicting tokens and if
-the models are effectively ``understanding'' token placements in sequences.
-
-Graphs of the training loss (average loss over the samples/batches in the training partition of the corpus) are presented in Figure
-\ref{fig:training_loss}. Both classes of models (scratch and pretrained) quickly reach a steady state in terms of loss although the initial loss
-values are significantly different. It is notable that the pretrained models start with a significantly lower loss, about 1.3, compared to models
-trained from scratch, which started with a loss value of just over 5.0. This initially indicates that the models are training well and beginning to
-converge on either local or global minima.
+With the metrics above defined, an analysis of the results of the model pretraining can be performed.
+The pretraining procedure aims to build the models' understanding of the text structure and transcripts of air-traffic communications in this case.
+By tracking the average cross-entropy loss (loss) and perplexity (PPL) during training, we can estimate how well (and how correctly) the models are predicting tokens and if the models are effectively ``understanding'' token placements in sequences.
+
+Graphs of the training loss (average loss over the samples/batches in the training partition of the corpus) are presented in Figure \ref{fig:training_loss}.
+Both classes of models (scratch and pretrained) quickly reach a steady state in terms of loss, although the initial loss values are significantly different.
+Notably, the pretrained models start with a considerably lower loss, about 1.3, compared to models trained from scratch, which began with a loss value of just over 5.0.
+This indicates that the models are training well and starting to converge on local or global minima.
 
@@
 
-The perplexity of the models on the training data yields interesting results (Figure \ref{fig:training_ppl}). Although the pretrained models have a
-more stable curve over time, the models trained from scratch result in a significantly lower perplexity. The final perplexity values (over the
-training set) of the models trained from scratch were significantly better than that of the pretrained models. At the end of training, the BERT and
-RoBERTa models trained from scratch had perplexity values of $0.52\times 10^7$ and $0.16\times 10^7$, respectively, whereas the pretrained models
-reached perplexity values of $0.29\times 10^9$ and $0.17\times 10^9$, for BERT and RoBERTa, respectively. Using the following equation,
+The perplexity of the models on the training data yields interesting results (Figure \ref{fig:training_ppl}).
+Although the pretrained models have a more stable curve over time, the models trained from scratch result in a significantly lower perplexity.
+The final perplexity values (over the training set) of the models trained from scratch were markedly better than that of the pretrained models.
+At the end of training, the BERT and RoBERTa models trained from scratch had perplexity values of $0.52\times 10^7$ and $0.16\times 10^7$, respectively, whereas the pretrained models reached perplexity values of $0.29\times 10^9$ and $0.17\times 10^9$, for BERT and RoBERTa, respectively.
+Using the following equation,
 
@@
 \noindent
-the relative difference in perplexity between the scratch and pretrained models on the final training epoch is 55.24, for the BERT models, and 104.57,
-for the RoBERTa models, which equates to an absolute difference (equation (\ref{eq:absolute_difference})) of $28.74 \times 10^7$ and
-$16.43 \times 10^7$ for the BERT and RoBERTa model, respectively.
+the relative difference in perplexity between the scratch and pretrained models on the final training epoch is 55.24 for the BERT models, and 104.57, for the RoBERTa models, which equates to an absolute difference (equation (\ref{eq:absolute_difference})) of $28.74 \times 10^7$ and $16.43 \times 10^7$ for the BERT and RoBERTa model, respectively.
 
@@
 
-The results from the training procedure show that the models trained from scratch outperformed their pretrained counterparts. This initially
-indicates that best language model performance may be achieved by pretraining models on in-domain data, then continuing to fine-tune on specific
-downstream tasks. This seems to be a direct contradiction of the current recommended practice of using one of the pretrained checkpoints from Devlin
-et al.~\cite{devlin_bert_2019} and Liu et al.~\cite{liu_roberta_2019} then fine-tuning on specific downstream tasks. However, this is only partially
-upheld by the results of the model performance on the test partition of the corpus (Table \ref{tab:pretraining_results}). For the BERT models, the
-pattern exhibited during training is consistent with the results on the test partition, resulting in a relative difference of 8.62 and an absolute
-difference of $1.24 \times 10^7$. For the RoBERTa models, however, the relative difference in perplexity is -0.34 with an absolute difference of
-$0.36 \times 10^7$. These results show that the pretrained RoBERTa model performed better during pretraining than the corresponding model
-that was trained from scratch, unlike the BERT models (theories as to why this occurs will be explored in the next paragraph). The loss of each
-model over the test set is somewhat high, but they are around the same values for each model, which is consistent with the loss statistics during
-training. A higher loss value on unseen (test) data is normal and expected.
-
-The BERT model trained from scratch had the best performance among the models tested followed closely by the pretrained RoBERTa model, which partially
-confirms the hypothesis that this thesis is based upon. The pretrained BERT model used both MLM and NSP pretraining objectives to create the
-checkpoint used in this thesis. The pretrained RoBERTa model, on the other hand, used only the MLM pretraining objective. It is readily apparent from
-the results of \cite{liu_roberta_2019} that RoBERTas modified pretraining methodology resulted in better model performance for in-domain problems.
-However, it seems that the pretraining method also allows RoBERTa to generalize better than BERT when transfer learned to other NLP domains. It is
-conceivable that, given additional training data in the future, the pretrained RoBERTa model could surpass the performance of the BERT model trained
-from scratch. Although, with the current volume of data available in this domain, it is clear from the results of the models on the test partition of
-the corpus that a BERT model trained from scratch will still be the best option for aviation English NLP tasks.
+The results from the training procedure show that the models trained from scratch outperformed their pretrained counterparts.
+This initially indicates that pretraining models on in-domain data may achieve the best language model performance, then continuing to fine-tune on specific downstream tasks.
+This seems to be a direct contradiction of the current recommended practice of using one of the pretrained checkpoints from Devlin et al.~\cite{devlin_bert_2019} and Liu et al.~\cite{liu_roberta_2019} then fine-tuning on specific downstream tasks.
+However, this is only partially upheld by the results of the model performance on the test partition of the corpus (Table \ref{tab:pretraining_results}).
+For the BERT models, the pattern exhibited during training is consistent with the results on the test partition, resulting in a relative difference of 8.62 and an absolute difference of $1.24 \times 10^7$.
+For the RoBERTa models, however, the relative difference in perplexity is -0.34 with an absolute difference of $0.36 \times 10^7$.
+These results show that the pretrained RoBERTa model performed better during pretraining than the corresponding model that was trained from scratch, unlike the BERT models (theories of why this occurs will be explored in the next paragraph).
+The loss of each model over the test set is somewhat high, but they are around the same values for each model, which is consistent with the loss statistics during training.
+A higher loss value on unseen (test) data is expected.
+
+The BERT model trained from scratch had the best performance among the tested models followed closely by the pretrained RoBERTa model, which partially confirms the hypothesis this thesis is based upon.
+The pretrained BERT model used both MLM and NSP pretraining objectives to create the checkpoint used in this thesis.
+The pretrained RoBERTa model, on the other hand, used only the MLM pretraining objective.
+It is readily apparent from the results of \cite{liu_roberta_2019} that RoBERTas modified pretraining methodology resulted in better model performance for in-domain problems.
+However, it seems that the pretraining method also allows RoBERTa to generalize better than BERT when transfer learned to other NLP domains.
+It is conceivable that, given additional training data in the future, the pretrained RoBERTa model could surpass the performance of the BERT model trained from scratch.
+However, with the current volume of data available in this domain, it is clear from the models' results on the corpus test partition that a BERT model trained from scratch will still be the best option for aviation English NLP tasks.
 
@@
 
-With the results and performance of the models established relative to each other, the high perplexity of each model must be addressed. While the loss
-values are converging nicely during training, the perplexity is somewhat erratic and consistently high, rarely converging to a consistent curve,
-although it is exhibiting a clear trend as training progresses. Adjusting the hyperparameters of the optimizer did not solve this issue and changing
-the hyperparameters of the model (e.g.~sequence length, number of hidden layers, attention heads, etc.~) was not desirable for doing a direct
-comparison between the pretrained checkpoints and the base models to study the effectiveness of the models on orthographic transcriptions of aviation
-English. It is difficult to determine what exactly is causing this phenomenon, so several theories were developed in an attempt to explain it (some
-of which will be elaborated upon in Section \ref{sec:future_work}):
-
-\textbf{Small volume of training data}. Especially when compared to \textit{Attention is All You Need}, BERT, and RoBERTa,
-\cite{vaswani_attention_2017,devlin_bert_2019,liu_roberta_2019} the number of samples used in the training data dwarfs that of the state-of-the-art.
-It has been exhaustively demonstrated in the past that neural networks perform better with large volumes of data. The inception of RoBERTa itself is
-based on the idea that BERT was not trained on enough data \cite{liu_roberta_2019}. Increasing the volume of in-domain data used for pretraining
-(and training) would be the most effective way to immediately improve the performance of the models.
-
-\textbf{Different forms of English}. The BERT and RoBERTa pretraining checkpoints were created by training the models on literary works and typed
-articles such as those from \textit{Wikipedia}. The data used in this work are orthographic transcriptions of aviation English and while the two data
-sources share the common root of English, transfer learning from literary English to aviation English yields limited results, as evidenced above,
-especially with relatively small amounts of data.
-
-\textbf{Mixed regional data sources}. As described in Section \ref{sec:data_source}, the data aggregated for use here comes from several different
-regions, namely international airports in the US \cite{godfrey_air_1994} and international and domestic airports in Europe
-\cite{smidl_air_2019,hofbauer_atcosim_2008,szoke_detecting_2021}. During data processing it was ensured that only English transcriptions were included
-in the corpus, however, proper nouns from each region will be present in the data regardless. Additionally, dialectical differences and varying levels
-of English proficiency will affect the structure of token sequences across regions. All of these facets combined may be contributing to the high
-perplexity of the language models relative to other works. It should be noted that the varied regional sources of data are not necessarily a bad
-thing. Depending on the application and desired robustness of the model, the ability to interpret international sources and contexts can be a
-desirable trait for model selection, however, it will require significantly more data and proportionally balanced regional contexts within the corpus
-in order for the pretraining process to be effective for downstream tasks.
-
-Up to this point we have followed a procedure for studying the effectiveness of the models that is somewhat symmetrical to that of BERT and RoBERTa,
-so at this point the models should be trained for downstream tasks in order to further measure the effectiveness of the pretraining procedures.
-In the aviation domain and for this particular data, that would mean training for tasks such as callsign identification, speaker identification,
-flight phase classification, clearance recognition, etc. However, labels for these tasks are not present in the corpora used here. Corpora with
-relevant labels for downstream tasks could not be obtained within an appropriate amount of time to run and include the experiments in this work.
-Some potential algorithms were developed in an attempt to mine labels from the corpora aggregated for this work, which will be expanded upon in
-section \ref{sec:future_work}. For these reasons, testing model performance and effectiveness on downstream tasks must be left for future work.
+With the results and performance of the models established relative to each other, the high perplexity of each model must be addressed.
+While the loss values converge nicely during training, the perplexity is somewhat erratic and consistently high, rarely converging to a consistent curve.
+However, it exhibits a clear trend as training progresses.
+Adjusting the hyperparameters of the optimizer did not solve this issue, and changing the hyperparameters of the model (e.g.~sequence length, number of hidden layers, attention heads, etc.~) was not desirable for performing a direct comparison between the pretrained checkpoints and the base models to study the effectiveness of the models on orthographic transcriptions of aviation English.
+It is difficult to determine what exactly is causing this phenomenon, so several theories were developed in an attempt to explain it (some of which will be elaborated upon in Section \ref{sec:future_work}):
+
+\textbf{Small volume of training data}.
+Especially when compared to \textit{Attention is All You Need}, BERT, and RoBERTa, \cite{vaswani_attention_2017,devlin_bert_2019,liu_roberta_2019} the number of samples used in the training data dwarfs that of the state-of-the-art.
+It has been exhaustively demonstrated that neural networks perform better with large volumes of data.
+The inception of RoBERTa itself is based on the idea that BERT was not trained on enough data \cite{liu_roberta_2019}.
+Increasing the volume of in-domain data used for pretraining (and training) would be the most effective way to immediately improve the performance of the models.
+
+\textbf{Different forms of English}.
+The BERT and RoBERTa pretraining checkpoints were created by training the models on literary works and typed articles such as those from \textit{Wikipedia}.
+The data used in this work are orthographic transcriptions of aviation English, and while the two data sources share the common root of English, transfer learning from literary English to aviation English yields limited results, as evidenced above, especially with relatively small amounts of data.
+
+\textbf{Mixed regional data sources}.
+As described in Section \ref{sec:data_source}, the data aggregated for use here comes from several different regions, namely international airports in the US \cite{godfrey_air_1994} and international and domestic airports in Europe \cite{smidl_air_2019,hofbauer_atcosim_2008,szoke_detecting_2021}.
+During data processing, it was ensured that only English transcriptions were included in the corpus.
+However, proper nouns from each region will be present in the data.
+Additionally, dialectical differences and varying levels of English proficiency will affect the structure of token sequences across regions.
+All of these combined facets may contribute to the high perplexity of the language models relative to other works.
+It should be noted that the varied regional data sources are not necessarily bad.
+Depending on the application and desired robustness of the model, the ability to interpret international sources and contexts can be desirable for model selection; however, it will require significantly more data and proportionally balanced regional contexts within the corpus for the pretraining process to be effective for downstream tasks.
+
+Up to this point, we have followed a procedure for studying the effectiveness of the models that is somewhat symmetrical to that of BERT and RoBERTa, so at this point, the models should be trained for downstream tasks to measure the effectiveness of the pretraining procedures further.
+In the aviation domain and for this particular data, that would mean training for tasks such as callsign identification, speaker identification, flight phase classification, clearance recognition, etc.
+However, labels for these tasks are not present in the corpora used here.
+Corpora with relevant labels for downstream tasks could not be obtained within an appropriate time to run and include the experiments in this work.
+Some potential algorithms were developed to mine labels from the corpora aggregated for this work, which will be expanded upon in section \ref{sec:future_work}.
+For these reasons, testing model performance and effectiveness on downstream tasks must be left for future work.
 
@@
 
-Contrary to the recommended best practice of using pretrained checkpoints and training for downstream tasks, we find that, in the domain of aviation
-English, models achieve better performance when pretrained from scratch than model checkpoints transfer learned from the available BERT and RoBERTa
-pretrained checkpoints. Although the pretrained RoBERTa model achieved better performance on the test partition of the corpus than expected, the
-pretraining methods laid out in this thesis resulted in the BERT model trained from scratch achieving the lowest loss and perplexity overall.
-
-The results from the experiments in this work indicate that, for natural language processing tasks with limited data availability in the aviation
-domain, models trained from scratch are more effective than transfer learning pretrained models. Thus, models trained from scratch should be
-fine-tuned and used for downstream tasks over pretrained models. The most significant limiting factor to the performance of the models is the small
-volume of aviation English data available for pretraining and downstream task use. Investment in the creation of expansive, high quality, openly
-available aviation English corpora with appropriate formatting for the development machine learning applications is also highly recommended.
+Contrary to the recommended best practice of using pretrained checkpoints and training for downstream tasks, we find that, in the domain of aviation English, models achieve better performance when pretrained from scratch than model checkpoints transfer learned from the available BERT and RoBERTa pretrained checkpoints.
+Although the pretrained RoBERTa model achieved better performance on the test partition of the corpus than expected, the pretraining methods laid out in this thesis resulted in the BERT model trained from scratch achieving the lowest loss and perplexity overall.
+
+The results from the experiments in this work indicate that models trained from scratch for natural language processing tasks with limited data availability in the aviation domain are more effective than transfer learning pretrained models.
+Thus, models trained from scratch should be fine-tuned and used for downstream tasks over pretrained models.
+The most significant limiting factor to the performance of the models is the small volume of aviation English data available for pretraining and downstream task use.
+Investment in creating expansive, high-quality, openly available aviation English corpora with appropriate formatting for developing machine learning applications is also highly recommended.
 
 \section{Future Work}\label{sec:future_work}
-This section is intended to serve as a brief proposal or set of suggestions for future work based on the results of this work. The main focus of these
-suggestions are corpus specific algorithms and data augmentation techniques to enable testing and experiments to determine the effectiveness of the
-pretraining procedures covered earlier in this thesis.
-
-\textbf{Next Sentence Prediction (NSP)}. This training procedure was introduced by Devlin et al.~\cite{devlin_bert_2019} as a pretraining procedure
-for BERT and is described in more detail in Section \ref{sec:bert}. This can be used on top of masked language modeling for language model
-pretraining, however, it requires a knowledge of which sequences in the corpus appear next to each other. For pretraining transformer-based neural
-language models for the aviation domain (with orthographic transcriptions), this same procedure can be used to classify whether two transmissions
-occur in the correct order. In other words; classifying a pilots response to an air-traffic controller directive or an air-traffic controllers
-response to a pilot query. Of the four corpora used in this work, ATCC is the only corpus that can be mined for an NSP-like pretraining procedure
-since the raw transcripts have indicators for time (start and end indicators), transmitting party, and intended receiver. With this information we can
-estimate the correct order of transmissions between two parties to a reasonable degree of certainty (either manually or programmatically).
-
-\textbf{Callsign Identification}. This is the process of identifying an aircraft callsign from the transcription of the spoken callsign. This is a
-popular subtask for automatic speech recognition applications in aviation, but can also be accomplished by postprocessing speech recognition outputs
-through a language model. ATCC is a good candidate corpus for this task. There are two potential methods for implementing this as a downstream task:
-(1) In transmissions from air-traffic controllers to specific aircraft, use the receiver (``TO'' label) as the label for each sequence or (2) Identify
-the sequence of tokens, in each transmission, that correspond to the callsign of the aircraft and use those tokens as the labels for each sequence.
-Each of these methods bring their own problems. The first method effectively creates an ever expanding classification problem with infinitely many
-labels, so some kind of tokenization or decoding scheme would need to be created to address this. The second method requires manually labeling every
-sequence in the corpus, which may not be feasible in all cases. Lastly, both methods may lack sufficient context to correctly predict aircraft
-callsigns. For example, general aviation (GA) callsigns in North America typically start with the letter \textit{N} followed by a series of numbers
-and/or letters, however, the spoken callsign is based on aircraft manufacturer (e.g.~Cessna, Cirrus, Gulfstream, etc.~), so the spoken callsign will
-begin with the manufacturer name instead of ``november''.
-
-\textbf{Speaker Identification}. Generally, speaker identification is a task to classify specific speakers in a series of communications. This is
-usually a subtask that falls under automatic speech recognition, but it is also feasible for language models under some circumstances as well. In
-aviation this could be a task as simple as identifying the side of the communication i.e.~an air-traffic controller or a pilot or it can be complex
-enough to be grouped under callsign identification tasks. In the simplest form, this can be accomplished easily using ATCC since controller positions
-are labeled along with transmitting/receiving aircraft.
-
-\textbf{Speech Recognition}. Finally, speech recognition system performance can be improved using a sufficiently trained language model. This approach
-has been used by several state-of-the-art automatic speech recognition models to improve sequence decoding and predictions. NVIDIA's Citrinet and
-QuartzNet models, for example, both use this approach to reduce their relative word error rates compared to the performance of other models on the
-same corpora \cite{majumdar_citrinet_2021,kriman_quartznet_2020}. Considering the relatively high word error rates of most automatic speech
-recognition models in the aviation domain, the pretraining procedures introduced here could be used to further reduce the error rates of these models.
+This section is intended to serve as a brief proposal or set of suggestions for future work based on the results of this work.
+The main focus of these suggestions is corpus-specific algorithms and data augmentation techniques to enable testing and experiments to determine the effectiveness of the pretraining procedures covered earlier in this thesis.
+
+\textbf{Next Sentence Prediction (NSP)}.
+This training procedure was introduced by Devlin et al.~\cite{devlin_bert_2019} as a pretraining procedure for BERT and is described in more detail in Section \ref{sec:bert}.
+This can be used on top of masked language modeling for language model pretraining.
+However, it requires a knowledge of which sequences in the corpus appear next to each other.
+For pretraining transformer-based neural language models for the aviation domain (with orthographic transcriptions), this same procedure can be used to classify whether two transmissions occur in the correct order.
+In other words, it is classifying a pilot's response to an air traffic controller directive or an air traffic controller's response to a pilot query.
+Of the four corpora used in this work, ATCC is the only corpus that can be mined for an NSP-like pretraining procedure since the raw transcripts have indicators for time (start and end indicators), transmitting party, and intended receiver.
+With this information, we can estimate the correct order of transmissions between two parties to a reasonable degree of certainty (manually or programmatically).
+
+\textbf{Callsign Identification}.
+This is the process of identifying an aircraft callsign from the transcription of the spoken callsign.
+This is a popular subtask for automatic speech recognition applications in aviation, but postprocessing speech recognition outputs can also be accomplished through a language model.
+ATCC is a good candidate corpus for this task.
+There are two potential methods for implementing this as a downstream task: (1) In transmissions from air-traffic controllers to specific aircraft, use the receiver (``TO'' label) as the label for each sequence or (2) Identify the sequence of tokens, in each transmission, that correspond to the callsign of the aircraft and use those tokens as the labels for each sequence.
+Each of these methods brings its problems.
+The first method effectively creates an ever-expanding classification problem with infinitely many labels, so a tokenization or decoding scheme must be created to address this.
+The second method requires manually labeling every sequence in the corpus, which may not be feasible in all cases.
+Lastly, both methods may lack sufficient context to predict aircraft callsigns correctly.
+For example, general aviation (GA) callsigns in North America typically start with the letter \textit{N} followed by a series of numbers and letters.
+However, the spoken callsign is based on the aircraft manufacturer (e.g.~Cessna, Cirrus, Gulfstream, etc.~), so the spoken callsign will begin with the manufacturer name instead of ``November''.
+
+\textbf{Speaker Identification}.
+Generally, speaker identification is a task to classify specific speakers in a series of communications.
+This subtask usually falls under automatic speech recognition, but it is also feasible for language models under some circumstances.
+In aviation, this could be as simple as identifying the side of the communication, i.e., an air-traffic controller or a pilot, or it can be complex enough to be grouped under callsign identification tasks.
+In the simplest form, this can be accomplished easily using ATCC since controller positions are labeled along with transmitting/receiving aircraft.
+
+\textbf{Speech Recognition}.
+Finally, a sufficiently trained language model can improve speech recognition system performance.
+This approach has been used by several state-of-the-art automatic speech recognition models to improve sequence decoding and predictions.
+NVIDIA's Citrinet and QuartzNet models, for example, both use this approach to reduce their relative word error rates compared to the performance of other models on the same corpora \cite{majumdar_citrinet_2021,kriman_quartznet_2020}.
+Considering the relatively high word error rates of most automatic speech recognition models in the aviation domain, the pretraining procedures introduced here could further reduce these models' error rates.
 This will be elaborated further in Appendix B in Section \ref{sec:appendix_a}.
@@
 \section{Appendix B: Neural Language Modeling Application to Speech Recognition}\label{sec:appendix_a}
-NVIDIA's NeMo (Neural Modules) framework \cite{kuchaiev_nemo_2019} is a state-of-the-art deep learning framework with implementations for dozens of
-natural language processing models and utilities. Notably among those are the automatic speech recognition (ASR) model implementations and tools for
-decoding/augmenting predictions from model outputs.
-
-For ASR models, there are generally two methods of decoding predictions: greedy decoding and decoding with beam search. Greedy decoding methods simply
-use the neuron, in the last layer of the network (the output layer), with the highest value to determine the prediction for that time step. Beam
-search-based decoding methods use a language model along with a beam search algorithm to determine predictions at each time step. The most commonly
-used language model for these types of decoding methods is an N-gram and, in fact, are what PyTorch's \lstinline|CTCDecoder|\footnote{More
-    specifically, the \lstinline|torchaudio| package of PyTorch} \cite{paszke_pytorch_2019} and NeMo's \lstinline|BeamSearchDecoderWithLM|
-\cite{kuchaiev_nemo_2019} use to decode predictions.
-
-The purpose of language models in beam search decoding algorithms is to estimate the likelihood of the decoded sequence occurring based on the
-language models knowledge of sequence structures. The beam search algorithm is then used to select tokens such that the overall likelihood of the
-sequence occurring is maximized. As mentioned above in Section \ref{sec:ngram}, N-grams are left-associative language models (assuming the model is
-being applied to the English language in this case), which is an inherently limiting factor. A neural language model capable of representing
-sequences bidirectionally, such as those covered in this work, is a significantly more robust method of language modeling in speech recognition and
-has been used in the past to improve word error rates \cite{kriman_quartznet_2020,majumdar_citrinet_2021}. There is at least one instance of a deep
-learning-based language model being used for beam search decoding in aviation applications, however, it did not perform well, likely due to small
-size of the dataset that it was trained on \cite{pellegrini_airbus_2019}. The models trained in this work used considerably more data and, therefore,
-will likely have better performance in beam search decoding algorithms.
+NVIDIA's NeMo (Neural Modules) framework \cite{kuchaiev_nemo_2019} is a state-of-the-art deep learning framework with implementations for dozens of natural language processing models and utilities.
+Among those are the automatic speech recognition (ASR) model implementations and tools for decoding/augmenting predictions from model outputs.
+
+For ASR models, there are generally two methods of decoding predictions: greedy decoding and decoding with beam search.
+Greedy decoding methods simply use the neuron in the network's last layer (the output layer) with the highest value to determine the prediction for that time step.
+Beam search-based decoding methods use a language model and a beam search algorithm to determine predictions at each time step.
+The most commonly used language model for these types of decoding methods is an N-gram and is what PyTorch's \lstinline|CTCDecoder|\footnote{More specifically, the \lstinline|torchaudio| package of PyTorch} \cite{paszke_pytorch_2019} and NeMo's \lstinline|BeamSearchDecoderWithLM| \cite{kuchaiev_nemo_2019} use to decode predictions.
+
+The purpose of language models in beam search decoding algorithms is to estimate the likelihood of the decoded sequence occurring based on the language models' knowledge of sequence structures.
+The beam search algorithm is then used to select tokens to maximize the likelihood of the sequence occurring.
+As mentioned above in Section \ref{sec:ngram}, N-grams are left-associative language models (assuming the model is being applied to the English language in this case), which is an inherently limiting factor.
+A neural language model capable of representing sequences bidirectionally, such as those covered in this work, is a significantly more robust method of language modeling in speech recognition and has been used in the past to improve word error rates \cite{kriman_quartznet_2020,majumdar_citrinet_2021}.
+At least one instance of a deep learning-based language model being used for beam search decoding in aviation applications.
+However, it did not perform well, likely due to the small size of the dataset that it was trained on \cite{pellegrini_airbus_2019}.
+The models trained in this work used considerably more data and will likely perform better in beam search decoding algorithms.
 

### 2023-12-05 thesis/MSECE_thesis.tex
Fix broken citations.

@@
 
-At the time of this writing, the current focus of general machine learning and deep learning algorithms in the language modeling domains has been on a variety of deep learning methods/architectures, namely, recurrent, convolutional, and transformer neural networks, due to the availability of large natural language corpora such as WikiText~\cite{merity_pointer_2016} and IMDB~\cite{maas_learning_2011}.
-In contrast to the general domain, the aviation domain still prefers and sees much success with traditional machine learning models for language modeling applications.
-Typically, the most prevalent of the traditional models are N-grams, word lattices, and rule-based approaches, depending on the application and operating environment of the models.
-The most likely reason for this divergence between the domains is the difference and relative lack of labeled data in the aviation domain.
-By comparison, the largest corpora in the general domain are either WikiText or the BookCorpus, which contain trillions of tokens, whereas, for aviation, the largest corpus is Air Traffic Control Complete with approximately 26 hours of labeled data amounting to just over 300,000 tokens\footnote{Air Traffic Control Complete has approximately 70 hours of audio data, however, about 26 hours of the 70 are labeled.}.
-This disparity in data availability is exacerbated by the fact that aviation English uses a specialized vocabulary in addition to specialized pronunciation for some words to enhance understandability in a radiotelephonic (R/T) medium~\cite{paltridge_handbook_2013}, making it significantly more difficult to transcribe/label data as it requires transcribers to have domain knowledge and experience to create effective transcriptions.
-This, in addition to the following facets, further contributes to the difficulty of transcription: (1) the medium of communication, (2) the environment in which the communication occurs\footnote{For example, a noisy and possibly low fidelity medium, depending on the equipment used to transmit and receive communications}, (3) a noisy and active communication environment i.e.~an active flight deck under the influence of wind, engine noise, etc., (4) an active Air Traffic Control (ATC) tower and channels, and (5) a rapid rate of speech \cite{paltridge_handbook_2013}.
+At the time of this writing, the current focus of general machine learning and deep learning algorithms in the language modeling domains has been on a
+variety of deep learning methods/architectures, namely, recurrent, convolutional, and transformer neural networks, due to the availability of large
+natural language corpora such as WikiText \cite{merity_pointer_2016} and IMDB \cite{maas_learning_2011}. In contrast to the general domain, the
+aviation domain still prefers and sees much success with traditional machine learning models for language modeling applications. Typically, the most
+prevalent of the traditional models are N-grams, word lattices, and rule-based approaches, depending on the application and operating environment of
+the models. The most likely reason for this divergence between the domains is the difference and relative lack of labeled data in the aviation domain.
+By comparison, the largest corpora in the general domain are either WikiText or the BookCorpus, which contain trillions of tokens, whereas, for
+aviation, the largest corpus is Air Traffic Control Complete with approximately 26 hours of labeled data amounting to just over 300,000
+tokens\footnote{Air Traffic Control Complete has approximately 70 hours of audio data, however, about 26 hours of the 70 are labeled.}. This disparity
+in data availability is exacerbated by the fact that aviation English uses a specialized vocabulary in addition to specialized pronunciation for some
+words to enhance understandability in a radiotelephonic (R/T) medium \cite{paltridge_handbook_2013}, making it significantly more difficult to
+transcribe/label data as it requires transcribers to have domain knowledge and experience to create effective transcriptions. This, in addition to the
+following facets, further contributes to the difficulty of transcription: (1) the medium of communication, (2) the environment in which the
+communication occurs\footnote{For example, a noisy and possibly low fidelity medium, depending on the equipment used to transmit and receive
+    communications}, (3) a noisy and active communication environment i.e.~an active flight deck under the influence of wind, engine noise, etc., (4)
+an active Air Traffic Control (ATC) tower and channels, and (5) a rapid rate of speech \cite{paltridge_handbook_2013}.
 
 % connect complexity of aviation english to utility/necessity of NLP models (ASR and LM) in the aviation domain
-The potential for speech recognition and language modeling algorithms to reach better than human performance (for example, the Stanford Question Answering Dataset (SQuAD) benchmark has recorded results consistently better than humans since 2019\footnote{https://rajpurkar.github.io/SQuAD-explorer/})~\cite{zhang_ai_2022} suggests that these NLP algorithms could have high utility in the aviation domain.
-If they can achieve better-than-human performance in the aviation domain and be used with pilot and air traffic controller systems, they could easily boost efficiency, reduce workloads, and reduce and mitigate errors during normal operations.
-
-The methods and results laid out in this thesis were done with the intention of creating a streamlined process for training and testing deep learning-based NLP methods, specifically language modeling.
-The results from the language models suggest that the volume of text used to train the language models is not yet sufficient to produce results akin to those in the conversational and/or literary English domains.
-Additionally, the results show that language models trained ``from scratch'' on in-domain data perform better than models trained on out-of-domain data and transfer learned on in-domain data, i.e., constrained BERT and RoBERTa models transfer learned to aviation English.
+The potential for speech recognition and language modeling algorithms to reach better than human performance (for example, the Stanford Question
+Answering Dataset (SQuAD) benchmark has recorded results consistently better than humans since
+2019\footnote{https://rajpurkar.github.io/SQuAD-explorer/}) \cite{zhang_ai_2022} suggests that these NLP algorithms could have high utility in the
+aviation domain. If they can achieve better-than-human performance in the aviation domain and be used with pilot and air traffic controller systems,
+they could easily boost efficiency, reduce workloads, and reduce and mitigate errors during normal operations.
+
+The methods and results laid out in this thesis were done with the intention of creating a streamlined process for training and testing deep
+learning-based NLP methods, specifically language modeling. The results from the language models suggest that the volume of text used to train the
+language models is not yet sufficient to produce results akin to those in the conversational and/or literary English domains. Additionally, the
+results show that language models trained ``from scratch'' on in-domain data perform better than models trained on out-of-domain data and transfer
+learned on in-domain data, i.e., constrained BERT and RoBERTa models transfer learned to aviation English.
 
@@
 
-Set notation (or set-builder notation)\footnote{That is, the notation used to denote sets and associated operations therein as it relates to the field of set theory, in mathematics.} as defined in \cite{jech_chapter_1978} is used in parts of this thesis, particularly in equations and algorithms, to convey mathematical concepts.
-The shorthand below is defined
+Set notation (or set-builder notation)\footnote{That is, the notation used to denote sets and associated operations therein as it relates to the field
+    of set theory, in mathematics.} as defined in \cite{jech_chapter_1978} is used in parts of this thesis, particularly in equations and algorithms, to
+convey mathematical concepts. The shorthand below is defined
 \begin{equation*}
@@
 % transformer history (background)
-The transformer neural network architecture was proposed in 2017 for Neural Machine Translation tasks and immediately achieved state-of-the-art \cite{vaswani_attention_2017}.
-Transformer architectures are extremely effective at learning representations and understandings of languages to predict token probabilities instead of transforming one language into another \cite{devlin_bert_2019,liu_roberta_2019}.
-Transformers have also been found to be very effective at other NLP-related tasks such as prompt completion and sentiment analysis among others (i.e.~auto-regressive and sequence classification tasks, respectively) \cite{lewis_bart_2019,radford_improving_2018}.
-Devlin et al.~\cite{devlin_bert_2019}proposed pretraining methods such as masked language modeling and next sentence prediction to improve natural language understanding for downstream tasks, which was further expanded upon by Liu et al.~\cite{liu_roberta_2019} by training longer, with larger batch sizes, and removing the next sentence prediction task.
+The transformer neural network architecture was proposed in 2017 for Neural Machine Translation tasks and immediately achieved state-of-the-art
+\cite{vaswani_attention_2017}. Transformer architectures are extremely effective at learning representations and understandings of languages to
+predict token probabilities instead of transforming one language into another \cite{devlin_bert_2019,liu_roberta_2019}. Transformers have also been
+found to be very effective at other NLP-related tasks such as prompt completion and sentiment analysis among others (i.e.~auto-regressive and sequence
+classification tasks, respectively) \cite{lewis_bart_2019,radford_improving_2018}. Devlin et al.~\cite{devlin_bert_2019} proposed pretraining methods
+such as masked language modeling and next sentence prediction to improve natural language understanding for downstream tasks, which was further
+expanded upon by Liu et al.~\cite{liu_roberta_2019} by training longer, with larger batch sizes, and removing the next sentence prediction task.
 
@@
 % LMs and NLP in aero. domain
-Various natural language processing methods have been applied to the aviation domain to help deal with miscommunications and try to mitigate safety incidents\cite{ragnarsdottir_language_2003,tanguy_natural_2016,madeira_machine_2021}.
-Some machine learning approaches have been implemented to analyze the text in aviation incident and safety reports to predict contributing factors and topic models\cite{tanguy_natural_2016,madeira_machine_2021}.
-An ASR system with NLP post-processing has also been proposed to analyze Air-Traffic communications and condense significant features (e.g.~weather, runway, and heading info) into an XML language structure\cite{ragnarsdottir_language_2003}.
-Transformer language models such as BERT and RoBERTa have been applied to notice to airmen (NOTAM) messages to perform named entity recognition (NER) tasks, translation tasks (between notations; e.g.~NOTAM notation to Airlang to make parsing tasks easier) and reduce the workload for pilots during long-haul flights\cite{arnold_knowledge_2022}.
-Transformer models have also been used for speaker role identification tasks in the aviation domain (e.g.~identifying pilot versus controller in communications); specifically, a pretrained BERT model was used and fine-tuned on problem-specific data and compared to other models that performed well at speaker and role identification tasks in general\cite{guo_comparative_2022}.
+Various natural language processing methods have been applied to the aviation domain to help deal with miscommunications and try to mitigate safety
+incidents\cite{ragnarsdottir_language_2003,tanguy_natural_2016,madeira_machine_2021}. Some machine learning approaches have been implemented to
+analyze the text in aviation incident and safety reports to predict contributing factors and topic models
+\cite{tanguy_natural_2016,madeira_machine_2021}. An ASR system with NLP post-processing has also been proposed to analyze Air-Traffic communications
+and condense significant features (e.g.~weather, runway, and heading info) into an XML language structure\cite{ragnarsdottir_language_2003}.
+Transformer language models such as BERT and RoBERTa have been applied to notice to airmen (NOTAM) messages to perform named entity recognition (NER)
+tasks, translation tasks (between notations; e.g.~NOTAM notation to Airlang to make parsing tasks easier) and reduce the workload for pilots during
+long-haul flights\cite{arnold_knowledge_2022}. Transformer models have also been used for speaker role identification tasks in the aviation domain
+(e.g.~identifying pilot versus controller in communications); specifically, a pretrained BERT model was used and fine-tuned on problem-specific data
+and compared to other models that performed well at speaker and role identification tasks in general\cite{guo_comparative_2022}.
 
 % ASR context (suggest some need for semi-supervised approaches and therefore LMs)
-Automatic speech recognition is seeing increased use in aviation for tasks such as transcription, callsign detection, speaker identification, etc.
-A majority of these approaches use machine learning approaches (as opposed to deep learning) where acoustic, pronunciation, and language modeling is used to transcribe speech \cite{guo_comparative_2022,smidl_air_2019,zuluaga gomez_automatic_2020,badrinath_automatic_2022,hofbauer_atcosim_2008,helmke_quantifying_2017}.
-Language models have been used in the general domain (both neural and machine learning-based approaches) to boost the performance of speech recognition models \cite{han_contextnet_2020,kriman_quartznet_2020,majumdar_citrinet_2021}.
-Machine learning-based language models, usually N-grams and word lattices, have also been used in aviation alongside semi-supervised approaches to automatic speech recognition to leverage unlabeled data.
-However, none of the developed methodologies use neural language models for their applications \cite{zuluaga-gomez_contextual_2021,srinivasamurthy_semi supervised_2017,badrinath_automatic_2022}.
+Automatic speech recognition is seeing increased use in aviation for tasks such as transcription, callsign detection, speaker identification, etc. A
+majority of these approaches use machine learning approaches (as opposed to deep learning) where acoustic, pronunciation, and language modeling is
+used to transcribe speech
+\cite{guo_comparative_2022,smidl_air_2019,zuluaga-gomez_automatic_2020,badrinath_automatic_2022,hofbauer_atcosim_2008,helmke_quantifying_2017}.
+Language models have been used in the general domain (both neural and machine learning-based approaches) to boost the performance of speech
+recognition models \cite{han_contextnet_2020,kriman_quartznet_2020,majumdar_citrinet_2021}. Machine learning-based language models, usually N-grams
+and word lattices, have also been used in aviation alongside semi-supervised approaches to automatic speech recognition to leverage unlabeled data.
+However, none of the developed methodologies use neural language models for their applications
+\cite{zuluaga-gomez_contextual_2021,srinivasamurthy_semi-supervised_2017,badrinath_automatic_2022}.
 

### 2023-12-22 thesis/MSECE_thesis.tex
Final edits from committee.

@@
 \usepackage{xcolor}
+\usepackage{url}
 \usepackage[T1]{fontenc}
@@
 There have been noticeable and significant advancements in the general domain of artificial intelligence (AI) recently. Natural
-language processing and language modeling technologies, in particular, have made tremendous advances over the past two decades,
-achieving performance levels good enough to be brought to consumer-facing products such as personal assistants, search engines, phones,
-televisions, etc. Due to the success in the general domain, applications have begun to spin off into problem-specific domains, such as
-(and most notably for this work) aviation. For example, several speech recognition systems with rule- and machine learning-based
-natural language processing, inference, and semantic extraction algorithms have been implemented at EUROCONTROL simulation centers in
-Europe to train air traffic controllers and study the potential effects of AI-augmented air traffic management (ATM) systems on
-controller workloads. Preliminary studies have found that AI-augmented ATM systems reduce controller workloads and fuel consumption of
+language processing and language modeling technologies, in particular, have made tremendous advances over the past two decades
+\cite{hannun_deep_2014,lee_unsupervised_2009,li_jasper_2019,kriman_quartznet_2020}, achieving performance levels good enough to be brought to
+consumer-facing products such as personal assistants, search engines, phones, televisions, etc. Due to the success in the general domain, applications
+have begun to spin off into problem-specific domains, such as (and most notably for this work) aviation. For example, several speech recognition
+systems with rule- and machine learning-based natural language processing, inference, and semantic extraction algorithms have been implemented at
+EUROCONTROL simulation centers in Europe to train air traffic controllers and study the potential effects of AI-augmented air traffic management (ATM)
+systems on controller workloads. Preliminary studies have found that AI-augmented ATM systems reduce controller workloads and fuel consumption of
 aircraft by optimizing aircraft routes in the airspace and reducing departure/arrival times, thereby reducing fuel consumption
@@
 in data availability is exacerbated by the fact that aviation English uses a specialized vocabulary in addition to specialized pronunciation for some
-words to enhance understandability in a radiotelephonic (R/T) medium \cite{paltridge_handbook_2013}, making it significantly more difficult to
-transcribe/label data as it requires transcribers to have domain knowledge and experience to create effective transcriptions. This, in addition to the
-following facets, further contributes to the difficulty of transcription: (1) the medium of communication, (2) the environment in which the
-communication occurs\footnote{For example, a noisy and possibly low fidelity medium, depending on the equipment used to transmit and receive
-    communications}, (3) a noisy and active communication environment i.e.~an active flight deck under the influence of wind, engine noise, etc., (4)
-an active Air Traffic Control (ATC) tower and channels, and (5) a rapid rate of speech \cite{paltridge_handbook_2013}.
+words to enhance understandability in a radiotelephonic (R/T) medium \cite{paltridge_handbook_2013}. This makes the data labeling process
+significantly more difficult because it not only requires two types of data (audio and text) to create labels, but also requires labelers to have
+domain knowledge and experience to create effective transcriptions. This, in addition to the following facets, further contributes to the difficulty
+of transcription: (1) the medium of communication, (2) the environment in which the communication occurs\footnote{For example, a noisy and possibly
+    low fidelity medium, depending on the equipment used to transmit and receive communications}, (3) a noisy and active communication environment
+i.e.~an active flight deck under the influence of wind, engine noise, etc., (4) an active Air Traffic Control (ATC) tower and channels, and (5) a
+rapid rate of speech. Pilots are also operating sophisticated equipment during while communicating, requiring significant attention and mental
+resources further contributing to the complexity of the communications \cite{paltridge_handbook_2013}.
 
@@
 Answering Dataset (SQuAD) benchmark has recorded results consistently better than humans since
-2019\footnote{https://rajpurkar.github.io/SQuAD-explorer/}) \cite{zhang_ai_2022} suggests that these NLP algorithms could have high utility in the
+2019\footnote{\url{https://rajpurkar.github.io/SQuAD-explorer/}}) \cite{zhang_ai_2022} suggests that these NLP algorithms could have high utility in the
 aviation domain. If they can achieve better-than-human performance in the aviation domain and be used with pilot and air traffic controller systems,
@@
 language models is not yet sufficient to produce results akin to those in the conversational and/or literary English domains. Additionally, the
-results show that language models trained ``from scratch'' on in-domain data perform better than models trained on out-of-domain data and transfer
-learned on in-domain data, i.e., constrained BERT and RoBERTa models transfer learned to aviation English.
-
-\section{Notation and Terminology}
+results in Section \ref{sec:pretraining_results} show that language models trained ``from scratch'' on in-domain data perform better than models
+trained on out-of-domain data and transfer learned on in-domain data.
+
+\subsection{Problem Statement}\label{sec:problem_statement}
+Other work has shown that the word error rates of automatic speech recognition models can and have been significantly reduced by using language models
+to determine the most likely token at a time step in a sequence \cite{li_jasper_2019,kriman_quartznet_2020}. Language models have also been shown to
+be effective for natural language processing tasks such as named entity recognition \cite{devlin_bert_2019,liu_roberta_2019}. Speech recognition and
+callsign detection (CSD) have become popular tasks for machine learning applications in aviation
+\cite{pellegrini_airbus_2019,delpech_real-life_2018,helmke_quantifying_2017}; callsign detection is a task analogous to named entity recognition since
+the elements of a callsign in a transcription can be treated as named entities in a sentence. For example, in the following transcription
+
+\begin{quote}
+    skyshuttle one one four zero now descend flight level three three zero
+\end{quote}
+
+\noindent
+the sequence ``skyshuttle one one four zero'' should be detected as the callsign. This is similar to detecting ``EU'', ``German'', and ``British'' as
+named entities in the sequence below.
+
+\begin{quote}
+    EU rejects German call to boycott British lamb
+\end{quote}
+
+The key component missing for both of these applications is an effectively pretrained language model that can be used for these downstream tasks, so
+the goal of this thesis is to run several experiments using transformer-based language models to determine the most effective strategy for pretraining
+the model for downstream tasks. Recommended best practice is to use a pre-existing language model checkpoint and fine-tune on additional data
+\cite{devlin_bert_2019,liu_roberta_2019,lewis_bart_2019}, although recent work suggests that if domains are sufficiently different, the performance
+of the fine-tuned model will be limited \cite{yadlowsky_pretraining_2023}. The main point of comparison in this thesis will be the performance of
+pretrained models fine-tuned on in-domain (aviation) data and models trained from scratch on in-domain data.
+
+\subsection{Notation and Terminology}
 This section is included for the sake of clarity throughout this thesis. In pursuit of that, a few terms and notations are defined below.
@@
 Set notation (or set-builder notation)\footnote{That is, the notation used to denote sets and associated operations therein as it relates to the field
-    of set theory, in mathematics.} as defined in \cite{jech_chapter_1978} is used in parts of this thesis, particularly in equations and algorithms, to
-convey mathematical concepts. The shorthand below is defined
+    of set theory, in mathematics.} as defined in \cite{jech_chapter_1978} is used in parts of this thesis, particularly in equations and algorithms,
+to convey mathematical concepts. The shorthand below is defined
 \begin{equation*}
@@
 as the union of the set $A$ and a set containing the element $x$. The equation below can be read as adding the element $x$ to the set $A$:
+
 \begin{equation*}
@@
 The following words (in bold) are used throughout the thesis as they relate to language modeling and are defined concretely as follows:
+
+The definition of words and sentences stem from their definitions in the English language, succinctly, as defined in \cite{noauthor_american_2011}:
+
 \begin{itemize}
+    \item A \textbf{word} is the representation in writing of a sound or combination of sounds that symbolizes and communicates a meaning
+    \item A \textbf{sentence} is a grammatical unit that is syntactically independent that is expressed or understood and contains at least one finite
+          verb
+\end{itemize}
+
+\noindent
+In terms of the structure of individual samples in a corpus, we will consider each sample in the corpus to be one sentence that is made up of words
+that are separated by whitespace (one or more spaces) and/or punctuation. Structurally, we consider a word to be made up of one or more characters
+as defined by the unicode standard \cite{allen_unicode_2012} and a sentence to be made up of one or more words.
+
+\begin{itemize}
+    \item A \textbf{token} shall be defined as the smallest, fundamental component of a sequence; these can be single characters, groups of
+          characters, punctuation, etc. \cite{devlin_bert_2019,wu_googles_2016,sennrich_neural_2016}
     \item A \textbf{sequence} shall be defined as a set of tokens (distinct from a sentence)
-    \item A \textbf{token} shall be defined as the smallest, fundamental component of a sequence (distinct from characters and words)
-    \item The rules for the formation of tokens from a sentence (and into sequences) shall be defined by tokenization algorithms (Section
-          \ref{sec:tokenizers})
-    \item The definition of \textbf{words} and \textbf{sentences} follow from that of the English language
-    \item The \textbf{vocabulary}, \textbf{inventory}, \textbf{index}, and \textbf{word list} of a model shall be semantically identical
 \end{itemize}
 
-\section{Literature Review}\label{sec:lit_review}
+\noindent
+As mentioned above, we consider each sample in a corpus to be a sentence, then as the corpus is collated, processed, and tokenized, each sample will
+be processed into sequences of tokens. These sequences are what will be analyzed and modeled by the language models.
+
+The rules by which words in sentences are broken up into tokens are defined by tokenization algorithms, otherwise known as tokenizers (elaborated
+further in Section \ref{sec:tokenizers}). While tokens can be individual characters or words as they appear in the original sample, this is not always
+the case. The intention of most tokenization algorithms is the flexibility to break up unknown words into smaller tokens known to the tokenizer
+\cite{wu_googles_2016,sennrich_neural_2016}.
+
+\section{Background \& Literature Review}\label{sec:lit_review}
+
+\subsection{Transformer Neural Networks}
 % from thesis proposal; revise and add to this as more references are used
@@
 The transformer neural network architecture was proposed in 2017 for Neural Machine Translation tasks and immediately achieved state-of-the-art
-\cite{vaswani_attention_2017}. Transformer architectures are extremely effective at learning representations and understandings of languages to
-predict token probabilities instead of transforming one language into another \cite{devlin_bert_2019,liu_roberta_2019}. Transformers have also been
-found to be very effective at other NLP-related tasks such as prompt completion and sentiment analysis among others (i.e.~auto-regressive and sequence
-classification tasks, respectively) \cite{lewis_bart_2019,radford_improving_2018}. Devlin et al.~\cite{devlin_bert_2019} proposed pretraining methods
-such as masked language modeling and next sentence prediction to improve natural language understanding for downstream tasks, which was further
-expanded upon by Liu et al.~\cite{liu_roberta_2019} by training longer, with larger batch sizes, and removing the next sentence prediction task.
-
+performance on language translation tasks \cite{vaswani_attention_2017}. Transformer architectures are extremely effective at learning representations
+and understandings of languages to predict token probabilities instead of transforming one language into another
+\cite{devlin_bert_2019,liu_roberta_2019}. Transformers have also been found to be very effective at other NLP tasks such as prompt completion
+and sentiment analysis among others (i.e.~auto-regressive and sequence classification tasks, respectively)
+\cite{lewis_bart_2019,radford_improving_2018}. Devlin et al.~\cite{devlin_bert_2019} proposed pretraining methods such as masked language modeling
+and next sentence prediction to improve natural language understanding for downstream tasks, which was further expanded upon by Liu et
+al.~\cite{liu_roberta_2019} by training longer, with larger batch sizes, and removing the next sentence prediction task.
+
+\subsection{Tokenizer Algorithms}
 % tokenizer algorithms
@@
 
+\subsection{Language Modeling \& Natural Language Processing in Aviation}
 % LMs and NLP in aero. domain
@@
 
+\subsection{Automatic Speech Recognition in Aviation}
 % ASR context (suggest some need for semi-supervised approaches and therefore LMs)
-Automatic speech recognition is seeing increased use in aviation for tasks such as transcription, callsign detection, speaker identification, etc. A
+Automatic speech recognition (ASR) is seeing increased use in aviation for tasks such as transcription, callsign detection, speaker identification, etc. A
 majority of these approaches use machine learning approaches (as opposed to deep learning) where acoustic, pronunciation, and language modeling is
@@
 
-% differences in language (transcripts vs literary english), narrow model selection
-The use of the English language in aviation has been shown to be sufficiently different enough to be classified under its own category (aviation English) \cite{paltridge_handbook_2013}.
-The issue with using pretrained checkpoints of transformer-based language models is characterizing the similarity (or, conversely, the difference) between literary English and aviation English.
-If the two domains are sufficiently different, then training the pretrained models on aviation English data would amount to transfer learning between specific domains.
-Transfer learning between text domains has been shown to be possible even with data volumes significantly lower than the original data used to train the models \cite{raffel_exploring_2020}.
-Yadlowsky et al.~\cite{yadlowsky_pretraining_2023} recently showed that if problem domains are sufficiently different, model pretraining on out-of-domain data will negatively affect model performance.
+ASR and speech synthesis technologies have seen use for improving communications and reducing pilot workload as early as 1981
+\cite{money_aural_1981,north_application_1984,north_systems_1984,mayer_electronic_1981}. Early approaches focused on developing and improving speech
+synthesis technologies to shift pilot focus in military and general aviation cockpits to urgent systems or alarms
+\cite{mayer_electronic_1981,money_aural_1981,mosko_clear_1981}. ASR system concepts and technologies were also initially developed to reduce pilot
+workloads and automate simple tasks; these systems were somewhat rigid with fixed vocabularies and keywords, but were able to successfully reduce
+workload and received positive feedback from pilots \cite{north_application_1984,cotton_development_1983}. Similar ASR technologies were later applied
+to Air Traffic Control (ATC) contexts to reduce controller workloads and increase the efficiency of Air Traffic Management (ATM) solutions
+\cite{schafer_context-sensitive_2000,lechner_voice_2002,ragnarsdottir_language_2003,cordero_automated_2012}. As ASR technologies have become more
+sophisticated and more aviation-specific speech corpora have become available
+\cite{godfrey_air_1994,smidl_air_2019,hofbauer_atcosim_2008,szoke_detecting_2021,segura_hiwire_2007,graglia_vocalise_2005,delpech_real-life_2018,zuluaga-gomez_atco2_2023}
+there have been more applications of ASR to aviation
+\cite{moere_implementing_2009,gurluk_assistant_2015,helmke_assistant-based_2015,helmke_machine_2020,helmke_readback_2021,kleinert_automated_2021,badrinath_automatic_2022,shi_end--end_2022,xiao_speech_2022,zuluaga-gomez_atco2_2023}.
+
+\subsection{Differences in Language}
+% differences in language (aviation vs casual english), narrow model selection
+The use of the English language in aviation has been shown to be sufficiently different enough to be classified under its own category (aviation
+English) \cite{paltridge_handbook_2013}. The issue with using pretrained checkpoints of transformer-based language models is characterizing the
+similarity (or, conversely, the difference) between casual and technical English i.e., in this case, the differences between written English, as in
+books and articles, and the spoken, transcribed English used in aviation. If the two domains are sufficiently different, then training the pretrained
+models on aviation English data would amount to transfer learning between specific domains. Transfer learning between text domains has been shown to
+be possible even with data volumes significantly lower than the original data used to train the models \cite{raffel_exploring_2020}. Yadlowsky et
+al.~\cite{yadlowsky_pretraining_2023} recently showed that if problem domains are sufficiently different, model pretraining on out-of-domain data will
+negatively affect model performance.
 
 \section{Data Sources}\label{sec:data_source}
-Four corpora are used in this thesis.
-The datasets are primarily speech corpora intended for automatic speech recognition research in aviation communications, linguistics research in aviation English, or both.
-All data in all four corpora include both recordings and transcriptions, which are audio and text, respectively.
-The transcriptions are used to pretrain and train the language models and tokenizers (described in sections \ref{sec:language_modeling} and \ref{sec:tokenizers}).
-
-The four corpora selected and used for this work are briefly described below. Further analysis is performed in Section
-\ref{sec:data_analysis}.
+Four corpora are used in this thesis. The datasets are primarily speech corpora intended for automatic speech recognition research in aviation
+communications, linguistics research in aviation English, or both. All data in all four corpora include both recordings and transcriptions, which are
+audio and text, respectively. The transcriptions are used to pretrain and train the language models and tokenizers (described in sections
+\ref{sec:language_models} and \ref{sec:tokenizers}).
+
+The four corpora selected and used for this work are briefly described below. Further analysis is performed in Section \ref{sec:data_analysis}.
 
 \subsection{Air Traffic Control Complete}\label{sec:atcc}
-The Air Traffic Control Complete (ATCC) corpus is a speech corpus consisting of audio recordings with corresponding transcriptions.
-ATCC is a collection of three smaller corpora recorded at Dallas-Fort Worth, Logan International, and Washington National airports and transcribed by current or former air traffic controllers familiar with the respective areas.
-Audio data was recorded by placing Very High Frequency (VHF) antennae configured to monitor several frequencies at each airport, such as arrival, departure, approach, ground, etc.
-The types of frequencies observed vary by airport, but the presence of departure, approach, and ground frequencies is relatively consistent across subcorpora \cite{godfrey_air_1994}.
-See Table \ref{tab:atcc_stats} for corpus-specific statistics.
-
-\begin{table}
+The Air Traffic Control Complete (ATCC) corpus is a speech corpus consisting of audio recordings with corresponding transcriptions. ATCC is a
+collection of three smaller corpora recorded at Dallas-Fort Worth, Logan International, and Washington National airports and transcribed by current or
+former air traffic controllers familiar with the respective areas. Audio data was recorded by placing Very High Frequency (VHF) antennae configured to
+monitor several frequencies at each airport, such as arrival, departure, approach, ground, etc. The types of frequencies observed vary by airport, but
+the presence of departure, approach, and ground frequencies is relatively consistent across subcorpora \cite{godfrey_air_1994}. See Table
+\ref{tab:atcc_stats} for corpus-specific statistics.
+
+\begin{table}[!t]
     \centering
@@
 \subsection{ATCO2}\label{sec:atco2}
-The ATCO2 dataset is a speech corpus comprising audio communications at Prague and Brno airports in Czechia and corresponding transcriptions.
-The speech recordings and transcripts are crowd-sourced from volunteers.
-The dataset was created and labeled, in part, as an English language detection corpus.
-However, the labeling also includes transcripts of speech segments, which makes it practical for ASR and language model development in addition to language detection (although in this work, it is only used for ASR and LM development)~\cite{szoke_detecting_2021}.
+The ATCO2 dataset is a speech corpus comprising audio communications at Prague and Brno airports in Czechia and corresponding transcriptions. The
+speech recordings and transcripts are crowd-sourced from volunteers. The dataset was created and labeled, in part, as an English language detection
+corpus. However, the labeling also includes transcripts of speech segments, which makes it practical for ASR and language model development in
+addition to language detection (although in this work, it is only used for ASR and LM development)~\cite{szoke_detecting_2021}.
 See Table \ref{tab:atco2_stats} for corpus-specific statistics.
 
-\begin{table}
+\begin{table}[!t]
     \centering
@@
 \subsection{Air Traffic Control Simulation}\label{sec:atcosim}
-The Air Traffic Control Simulation (ATCOSIM) corpus is a speech corpus consisting of audio recordings and corresponding transcriptions.
-The data was recorded at the EUROCONTROL Experimental Centre for air traffic control simulation in Bretigny-sur-Orge, France.
-In this corpus, only the controllers' voices are included, and therefore, the transcripts only include the controller's side of each interaction.
-The audio data was transcribed by one person, trained according to the guidelines and formatting requirements of the corpus.
-After all data was transcribed, it was reviewed and corrected, and any remaining problems were reviewed by an operational air traffic controller and resolved \cite{hofbauer_atcosim_2008}.
-See Table \ref{tab:atcosim_stats} for corpus-specific statistics.
-
-\begin{table}
+The Air Traffic Control Simulation (ATCOSIM) corpus is a speech corpus consisting of audio recordings and corresponding transcriptions. The data was
+recorded at the EUROCONTROL Experimental Centre for air traffic control simulation in Bretigny-sur-Orge, France. In this corpus, only the controllers'
+voices are included, and therefore, the transcripts only include the controller's side of each interaction. The audio data was transcribed by one
+person, trained according to the guidelines and formatting requirements of the corpus. After all data was transcribed, it was reviewed and corrected,
+and any remaining problems were reviewed by an operational air traffic controller and resolved \cite{hofbauer_atcosim_2008}. See Table
+\ref{tab:atcosim_stats} for corpus-specific statistics.
+
+\begin{table}[!t]
     \centering
@@
 \subsection{ZCU CZ ATC Corpus}\label{sec:zcu_atc}
-The ZCU CZ ATC corpus consists of audio recordings and corresponding transcripts in the Czech airspace.
-Both the controller and pilot sides of the communications are included in this data.
-Experienced transcribers/annotators created annotations, and labeled samples were randomly selected for review during the annotation process.
-After the dataset was completely labeled, all samples were checked, corrected, and unified\cite{smidl_air_2019}.
+The ZCU CZ ATC corpus consists of audio recordings and corresponding transcripts in the Czech airspace. Both the controller and pilot sides of the
+communications are included in this data. Experienced transcribers/annotators created annotations, and labeled samples were randomly selected for
+review during the annotation process. After the dataset was completely labeled, all samples were checked, corrected, and unified\cite{smidl_air_2019}.
 See Table \ref{tab:zcu_cz_atc_stats} for corpus-specific statistics.
 
-\begin{table}
+\begin{table}[!t]
     \centering
@@
 
-\section{Data Processing}\label{sec:data_processing}
-All corpora listed above (see Section \ref{sec:data_source}) are created at different times, for various purposes, and by multiple authors.
-As a result, the data in these corpora all have different formats and need to be processed into a common format to be used together, with the goal being to essentially create a ``super-corpus'' that includes all the corpora listed above (i.e., a superset of the other corpora).
-
-All data was processed in Python using primarily built-in functions/modules\footnote{The \lstinline|re| and \lstinline|xml| modules were also used, although these are built into most Python distributions by default.}.
-The text data is extracted from the corpus transcripts by parsing from the corpora-specific formats where necessary, e.g.~Lisp in ATCC, XML in ATCO2 and ZCU CZ ATC, and plain text in ATCOSIM, the transcriber annotations are removed where necessary\footnote{A copy of the original data is made, then the annotations are removed from the copy.}, and the resulting text is written to a file in the standard corpus format (an ASCII text file with one line in the file corresponding to one sample from the corpus).
+\section{Data Preparation}\label{sec:data_preparation}
+All corpora listed above (see Section \ref{sec:data_source}) are created at different times, for various purposes, and by multiple authors. As a
+result, the data in these corpora all have different formats. ATCC, for example, uses Lisp-style lists as below:
+
+\begin{lstlisting}
+((FROM NERA3788) (NUM L02F1-0001)
+(TO F1-1)
+(TEXT THOUSAND ONE NINETY WE (QUOTE LL) GIVE YOU THAT ON THE SPEED AND WE (QUOTE RE) CLEARED FOR THE APPROACH AH NERA THIRTY SEVEN EIGHTY EIGHT WE (QUOTE LL) HOLD SHORT OF TWO SEVEN)
+(TIMES 1.49 6.57))
+\end{lstlisting}
+
+ATCO2 and ZCU CZ ATC use XML to isolate transcribed segments of speech in the audio recordings, although the format used between the two differs.
+ATCO2 uses a form like below:
+
+\clearpage
+\begin{lstlisting}
+<data>
+<segment>
+<start>3.79</start>
+<end>6.85</end>
+<speaker>B</speaker>
+<speaker_label>OK-PMB</speaker_label>
+<text>level one hundred Oscar Kilo Papa Mike Bravo</text>
+<tags>
+<correct>0</correct>
+<correct_transcript>1</correct_transcript>
+<correct_tagging>0</correct_tagging>
+<non_english>0</non_english>
+</tags>
+</segment>
+</data>
+\end{lstlisting}
+
+\clearpage
+ZCU CZ ATC uses a different nesting and labeling style, like below\footnote{Some information is excluded for readability.}:
+
+\begin{lstlisting}
+<Trans audio_filename="e2_ACCU-0agmXf.wav">
+<Episode>
+<Section type="report" startTime="0" endTime="61.47">
+<Turn startTime="0" endTime="61.47">
+<Sync time="0.000"/>
+[ground]Skyshuttle 1 1 4 0 now descend FL 3 3 0[speaker]
+<Sync time="4.790"/>
+[air]descending FL 3
+<Sync time="6.650"/>
+[air]K O Z
+<Sync time="9.640"/>
+[ground_|]Austrian 3 2 3 G climb FL 3 4 0 [|_ground][air_|]level 3 4 0 Austrian 3 2 3 G[|_air]
+</Turn>
+</Section>
+</Episode>
+</Trans>
+\end{lstlisting}
+
+Lastly, ATCOSIM uses a text-only format with transcriptions occurring on their own lines in text files:
+
+\begin{lstlisting}
+[FRAGMENT] contact geneva one two eight decimal one five good bye
+\end{lstlisting}
+
+\noindent
+Each text file corresponds to the audio file from which the speech was transcribed.
+
+Due to the differing formats above, and the intention to aggregate and use all these transcripts together, all transcripts need to be processed into
+a common format to be used together.
+
+% TODO: fix; lost some text somehow
+All data was processed in Python using primarily built-in functions/modules\footnote{The \lstinline|re| and \lstinline|xml| modules were also used,
+    although these are built into most Python distributions by default.}. The text data is extracted from the corpus transcripts by reading each
+transcript into the main memory of the system according to the format prescribed by the corpus. The text corresponding to the transcript is isolated,
+copied, and stored to a separate region of memory (a Python list of strings) that corresponds to the original corpus. The resulting isolated
+transcripts are written to files that correspond to the original corpus as well as combined, shuffled, and written to a file in the standard corpus
+format (an ASCII text file with one line in the file corresponding to one sample from the corpus). The resulting output is four corpus files that
+correspond to the original corpora in addition to a fifth corpus file with the samples from all four corpora combined and shuffled for later use.
 
 \subsection{Handling Spelling Errors \& Inconsistencies}\label{sec:spelling_errors}
-By the nature of the data labeling process, there is a high likelihood of spelling errors introduced by the annotators during corpora creation.
-This gets exacerbated by aggregating several corpora with varying controls for correcting spelling errors and inconsistencies.
-Nearly all of the mistakes are recoverable/correctable, given proper context.
-For example, ``possibility of \textbf{tornado's}'' should have been ``possibility of \textbf{tornadoes}'', changing the possessive form of ``tornado'' to the plural form.
-There are also instances of spelling conventions that are consistent within corpora but become inconsistent when combined.
+By the nature of the data labeling process, there is a high likelihood of spelling errors introduced by the annotators during corpora creation. This
+gets exacerbated by aggregating several corpora with varying controls for correcting spelling errors and inconsistencies. Nearly all of the mistakes
+are recoverable/correctable, given proper context. For example, ``possibility of \textbf{tornado's}'' should have been ``possibility of
+\textbf{tornadoes}'', changing the possessive form of ``tornado'' to the plural form. There are also instances of spelling conventions that are
+consistent within corpora but become inconsistent when combined.
 
 % 22 spelling errors
-Manually reviewing every sample in the corpus is unrealistic, considering that there are over 50,000 data samples and would be subject to the same human error that introduced those errors in the first place.
-Instead, the frequency of token occurrence in the corpora, combined with manual review thereafter, was used to detect and correct the most common errors.
-The frequency of occurrence of a token is calculated using equation (\ref{eq:token_frequency_occurrence}).
+Manually reviewing every sample in the corpus is unrealistic, considering that there are over 50,000 data samples and would be subject to the same
+human error that introduced those errors in the first place. Instead, the frequency of token occurrence in the corpora, combined with manual review
+thereafter, was used to detect and correct the most common errors. The frequency of occurrence of a token is calculated using equation
+(\ref{eq:token_frequency_occurrence}).
 
@@
 \noindent
-Where $f_w$ is the frequency at which a token occurs in a corpus, $\phi_w(w)$ is the number of times a token, $w$, occurs, and $D_{tokens}$ is the total number of tokens in the corpus.
-Tokens with the lowest frequency across corpora (and lowest number of occurrences) were manually reviewed and corrected where necessary.
-An $A \rightarrow B$ mapping system was used to correct the errors, where \(A\) is the erroneous token and \(B\) is the corrected or replaced token.
-As the text is parsed from the corpora, each sample is analyzed.
-If an erroneous token, $A$, is found, it is replaced with the corrected version, $B$.
-22 unique errors were detected and subsequently corrected or removed using this method.
-However, multiple instances of those errors typically occurred.
+Where $f_w$ is the frequency at which a token occurs in a corpus, $\phi_w(w)$ is the number of times a token, $w$, occurs, and $D_{tokens}$ is the
+total number of tokens in the corpus. Tokens with the lowest frequency across corpora (and lowest number of occurrences) were manually reviewed and
+corrected where necessary. An $A \rightarrow B$ mapping system was used to correct the errors, where \(A\) is the erroneous token and \(B\) is the
+corrected or replaced token. As the text is parsed from the corpora, each sample is analyzed. If an erroneous token, $A$, is found, it is replaced
+with the corrected version, $B$. 22 unique errors were detected and subsequently corrected or removed using this method. However, multiple instances
+of those errors typically occurred.
 
 % hesitations, cut off words, different kinds of filled pauses
-The mapping system described above is also used to modify corpora conventions to be consistent across corpora.
-Three areas are addressed here: hesitations, cut-off, or partial words, and filled pauses within speech.
-These areas are defined differently between the four corpora, which makes unifying the transcripts difficult.
-Hesitations, for example, are defined by one corpus as incomplete or non-English speech sounds and left completely undefined by another (despite annotators making notes about hesitations).
-The way in which hesitations, pauses, and partial words are transcribed differs between corpora as well.
-Two corpora prescribe special tokens to these phenomena, such as ``[hes]'', ``[hesitation]'', ``<pause>'', or similar.
-In contrast, the others transcribe the data as it sounds (i.e., ``uh'', ``eh'', or ``er'' for filled pauses and ``cir-'' or ``cir+'' for instances that were cut-off mid pronunciation).
-To facilitate the unification of the transcripts for all four corpora, the conventions for transcribing these phenomena are redefined in Table \ref{tab:phenomena_definitions}.
+The mapping system described above is also used to modify corpora conventions to be consistent across corpora. Three areas are addressed here:
+hesitations, cut-off, or partial words, and filled pauses within speech. These areas are defined differently between the four corpora, which makes
+unifying the transcripts difficult. Hesitations, for example, are defined by one corpus as incomplete or non-English speech sounds and left completely
+undefined by another (despite annotators making notes about hesitations). The way in which hesitations, pauses, and partial words are transcribed
+differs between corpora as well. Two corpora prescribe special tokens to these phenomena, such as ``[hes]'', ``[hesitation]'', ``<pause>'', or
+similar. In contrast, the others transcribe the data as it sounds (i.e., ``uh'', ``eh'', or ``er'' for filled pauses and ``cir-'' or ``cir+'' for
+instances that were cut-off mid pronunciation). To facilitate the unification of the transcripts for all four corpora, the conventions for
+transcribing these phenomena are redefined in Table \ref{tab:phenomena_definitions}.
 
@@
 
-A special token represents hesitation when no other context is present regarding the realization of hesitation.
-If other corpora use a special token to express hesitations, it is translated to match the convention in Table \ref{tab:phenomena_conventions}.
-Partial/cut-off words are represented with a hyphen (-) at the point where the pronounced word is cut off.
-For example, if the last syllable of ``approaching'' is cut mid-transmission, it would be represented in the transcript as ``approa-'' or, likewise, if the first syllable is cut off, it would be written as ``-roaching''.
-Lastly, if there is enough information in the original transcript to indicate the realization of a hesitation, it is represented by the most appropriate token from ``Filled pauses'' in Table \ref{tab:phenomena_conventions}.
-
-One of the intended downstream use cases of the language models developed from this work is decoding speech recognition predictions into transcripts, so including as much data as possible regarding spoken word realization is useful given the context.
-The \lstinline|[HES]| token is used as a catch-all fallback token if no knowledge about the utterance from the transcript can be extracted.
-\section{Data Analysis}\label{sec:data_analysis}
+A special token represents hesitation when no other context is present regarding the realization of hesitation. If other corpora use a special token
+to express hesitations, it is translated to match the convention in Table \ref{tab:phenomena_conventions}. Partial/cut-off words are represented with
+a hyphen (-) at the point where the pronounced word is cut off. For example, if the last syllable of ``approaching'' is cut mid-transmission, it would
+be represented in the transcript as ``approa-'' or, likewise, if the first syllable is cut off, it would be written as ``-roaching''. Lastly, if there
+is enough information in the original transcript to indicate the realization of a hesitation, it is represented by the most appropriate token from
+``Filled pauses'' in Table \ref{tab:phenomena_conventions}.
+
+One of the intended downstream use cases of the language models developed from this work is decoding speech recognition predictions into transcripts,
+so including as much data as possible regarding spoken word realization is useful given the context. The \lstinline|[HES]| token is used as a
+fallback token if no knowledge about the utterance can be extracted from the transcript. This usually occurs when the original transcript did not
+include any detail regarding the special case of an utterance, for example, an \lstinline|[UNINTELLIGIBLE]| token to mark unintelligible speech, or
+\lstinline|[UNK]| to mark an unknown word or utterance that could not be transcribed. The \lstinline|[HES]| token is therefore used as a way of
+unifying the different approaches to mark hesitations, unintelligible speech, etc. when a special token was used with no indication of the realized
+speech.
+
+\subsection{Data Analysis}\label{sec:data_analysis}
 % sequence lengths, unique tokens, total number of tokens, total number of samples, breakdown by each dataset
-Relevant corpora statistics were observed or calculated and are presented in Table \ref{tab:corpora_stats}.
-This includes the number of samples in the corpus, the mean sequence length per sample, the number of unique tokens, and the total number of tokens in the corpus.
-The same statistics were also calculated across corpora since all four corpora will be combined for training and testing the language and speech recognition models.
-Note that this is based on the raw data after loading and unifying the corpora before preprocessing for model training/testing and outlier removal.
+Relevant corpora statistics were observed or calculated and are presented in Table \ref{tab:corpora_stats}. This includes the number of samples in the
+corpus, the mean sequence length per sample, the number of unique tokens, and the total number of tokens in the corpus. The same statistics were also
+calculated across corpora since all four corpora will be combined for training and testing the language and speech recognition models. Note that this
+is based on the raw data after loading and unifying the corpora before preprocessing for model training/testing and outlier removal.
 
@@
 
-The ATCC and ZCU CZ ATC corpora make up a significant portion of the data, ATCC alone makes up more than half (approximately 54\%) of the combined data, and ATCC and ZCU CZ ATC together make up approximately 81\% of the data.
-The sequence lengths of the samples are fairly consistent across corpora with at most about an 18\% percent difference between sequence lengths (calculated using equation (\ref{eq:percent_difference}), below).
-The number of unique tokens does not scale additively to the combined corpora, which is expected since there will be an overlap of tokens.
-Aviation English generally conforms to and follows ICAO phraseology standards, so we expect many terms to be shared across corpora except geographic and location-specific terms, such as city, landmark, airport, and airline names, which vary across regions.
+The ATCC and ZCU CZ ATC corpora make up a significant portion of the data, ATCC alone makes up more than half (approximately 54\%) of the combined
+data, and ATCC and ZCU CZ ATC together make up approximately 81\% of the data. The sequence lengths of the samples are fairly consistent across
+corpora with at most about an 18\% percent difference between sequence lengths (calculated using equation (\ref{eq:percent_difference}), below). The
+number of unique tokens does not scale additively to the combined corpora, which is expected since there will be an overlap of tokens. Aviation
+English generally conforms to and follows International Civil Aviation Organization (ICAO) or Federal Aviation Administration (FAA) phraseology
+standards\footnote{The applicable standard depends on the airspace and country. The ICAO and FAA standards are similar but different, so both are
+    included.}, so we expect many terms to be shared across corpora except geographic and location-specific terms, such as city, landmark, airport,
+and airline names, which vary across regions.
 
@@
 
-Sci-Kit Learn's \cite{pedregosa_scikit-learn_2011} implementation of the LOF algorithm was used with a value of \(K = 35\) for the K-Nearest Neighbors algorithm and Minkowski distance for distance metrics/calculations (in this case, since \(p=2\), Euclidean distance is effectively used for distance calculations).
-A total of 47 samples were detected as outliers and removed for analysis.
-They are summarized in Table \ref{tab:outlier_stats}.
+Sci-Kit Learn's \cite{pedregosa_scikit-learn_2011} implementation of the LOF algorithm was used with a value of \(K = 35\) for the K-Nearest Neighbors
+algorithm and Minkowski distance for distance metrics/calculations (in this case, since \(p=2\), Euclidean distance is effectively used for distance
+calculations). A total of 47 samples were detected as outliers and removed for analysis. They are summarized in Table \ref{tab:outlier_stats}.
 
@@
 
-Although all of the outliers have longer sequence lengths than the typical sequence lengths of the other samples in the data, going through the outlying samples manually and reviewing each reveals that, for the most part, the samples are typical albeit somewhat complex communications.
-Except for one outlier, all of the samples labeled outlying by the LOF algorithm were kept in the data.
-Examples of the outliers are listed in Appendix A in Section \ref{sec:appendix_a}, including the sample removed.
-
-\section{Language Modeling}\label{sec:language_modeling}
+Although all of the outliers have longer sequence lengths than the typical sequence lengths of the other samples in the data, going through the
+outlying samples manually and reviewing each reveals that, for the most part, the samples are typical albeit somewhat complex communications. Except
+for one outlier, all of the samples labeled outlying by the LOF algorithm were kept in the data. Examples of the outliers are listed in Appendix A in
+Section \ref{sec:appendix_a}, including the sample removed.
+
+\section{Model Architectures}
+
+\subsection{Automatic Speech Recognition}\label{sec:automatic_speech_recognition}
+Automatic speech recognition (ASR) or simply speech recognition is the process of translating spoken language to text i.e.~transcribing spoken
+language. There have been a myriad of approaches to ASR problems, including, but not limited to rule-based, machine learning, and deep learning
+methods. For the purposes of this work, we will focus primarily on deep learning approaches to ASR tasks with a specific focus on transcribing
+spoken language in aviation (following from the data accumulated, processed, and analyzed in sections \ref{sec:data_source},
+\ref{sec:data_preparation}, and \ref{sec:data_analysis}).
+
+Although the internal processing of data varies between architectures, the format of the models input and output is the same for the models described
+below. ASR models have seen much success and performance gains for speech-related tasks such as speech and speaker recognition through the use of
+Mel-frequency Cepstral Coefficients (MFCCs) \cite{sahidullah_design_2012}. Since ASR is not the main focus of this thesis, the process for computing
+the MFCCs will only be described from a high level as it is described in \cite{sahidullah_design_2012}. Readers should refer to
+\cite{sahidullah_design_2012} for a more detailed, matrix-based approach for computing MFCCs. Given an audio signal:
+
+\begin{enumerate}
+    \item Multiply the audio signal by a tapered window such as a Hamming or Hanning window to obtain windowed speech frames
+    \item Pad the windowed audio signal with zeroes to facilitate the fast Fourier transform of the signal
+    \item Multiply the zero-padded windowed speech frames by the twiddle factor matrix to obtain the discrete Fourier transform (DFT) coefficients
+    \item Multiply each DFT coefficient with its conjugate to obtain the power spectrum of the signal
+    \item Pass the signal through a triangular filter bank with linearly spaced filters in the Mel-scale. Obtain the log-energy output by taking the
+          log of the output --- the filter bank log energy
+    \item Finally, multiply the filter bank log energy from the previous step by the discrete cosine transform matrix to obtain the MFCCs
+\end{enumerate}
+
+\noindent
+The MFCCs are then used as input to the ASR models. Each model outputs a probability distribution of characters or tokens (depending on the vocabulary
+the ASR model is initialized with) over a segment of audio. All the ASR models used are trained with a Connectionist Temporal Classification (CTC)
+loss function~\cite{graves_connectionist_2006}.
+
+\subsubsection{Jasper}\label{sec:jasper}
+Jasper is a family of end-to-end convolutional neural networks (CNNs) developed by NVIDIA to replace traditional ASR models that use separately
+learned components for each stage of the pipeline (acoustic, pronunciation, language modeling, etc.) \cite{li_jasper_2019}.
+
+Jasper models are designed with $B$ blocks and $R$ sub-blocks and are named accordingly as \textit{Jasper $B\mbox{x}R$} models. The Jasper
+architecture also contains four extra convolutional blocks, one for preprocessing and three for postprocessing (see figure \ref{fig:jasper}). Each
+convolutional sub-block contains one dimensional convolutions. Li et al.~\cite{li_jasper_2019} trained and tested a Jasper 10x5 model that produced
+state-of-the-art results on the LibriSpeech, Wall Street Journal, and Fisher+Switchboard speech corpora.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=0.5\linewidth]{figures/jasper_vertical.png}
+    \caption{Block diagram of the Jasper architecture. Copied from Li et al.~\cite{li_jasper_2019}.}
+    \label{fig:jasper}
+\end{figure}
+
+\subsubsection{QuartzNet}\label{sec:quartznet}
+The QuartzNet architecture is based on Jasper with some key modifications, namely, that the one dimensional convolutions are replaced with ``one
+dimensional time-channel separable convolutions''. As with Jasper, QuartzNet models are denoted as \textit{QuartzNet $B\mbox{x}R$}, however, blocks
+are repeated $S$ times and have a certain number of input and output channels, $c_{in}$ and $c_{out}$, respectively.
+
+In Kriman et al.~\cite{kriman_quartznet_2020} experiments with QuartzNet were conducted on the LibriSpeech and Wall Street Journal speech corpora and
+achieved near state-of-the-art performance on both. A transfer learning experiment was also performed (transferring between corpora). The model was
+initially trained on LibriSpeech and Mozilla Common Voice, then fine-tuned on the Wall Street Journal corpus and also achieved near state-of-the-art
+results \cite{kriman_quartznet_2020}. These results suggest that the QuartzNet architecture not only performs well on in-domain tasks, but can
+generalize and fine-tune well on related data.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=0.5\linewidth]{figures/quartz_vertical.png}
+    \caption{Block diagram of the QuartzNet architecture. Copied from Kriman et al.~\cite{kriman_quartznet_2020}.}
+    \label{fig:quartznet}
+\end{figure}
+
+\subsection{Language Models}\label{sec:language_models}
 % bert, roberta
 % n-gram
-This section introduces the chosen language model architectures and their pretrained checkpoints.
-Two transformer-based architectures, BERT and RoBERTa, and one statistical/machine learning-based approach: N-gram language modeling.
-
-The transformers are pre-trained using the same masked language modeling (MLM) objective presented in Devlin et al.~\cite{devlin_bert_2019} to train BERT for bidirectional encoding representation and likewise in Liu et al.~\cite{liu_roberta_2019} for training RoBERTa.
-Given the results of the RoBERTa pretraining and downstream task training thereafter, it was decided to eliminate the next sentence prediction (NSP) task that was used to pretrain BERT.
-MLM and NSP pretraining are explained more in-depth in Section \ref{sec:bert} below.
-
-\subsection{N-Gram}\label{sec:ngram}
-An N-gram language model estimates the probability of a token occurring in a sequence based on the \(N\) tokens preceding or succeeding that token.
+This section introduces the chosen language model architectures and their pretrained checkpoints. Two transformer-based architectures, BERT and
+RoBERTa, and one statistical/machine learning-based approach: N-gram language modeling.
+
+The transformers are pre-trained using the same masked language modeling (MLM) objective presented in Devlin et al.~\cite{devlin_bert_2019} to train
+BERT for bidirectional encoding representation and likewise in Liu et al.~\cite{liu_roberta_2019} for training RoBERTa. Given the results of the
+RoBERTa pretraining and downstream task training thereafter, it was decided to eliminate the next sentence prediction (NSP) task that was used to
+pretrain BERT. MLM and NSP pretraining are explained more in-depth in Section \ref{sec:bert} below.
+
+\subsubsection{N-Gram}\label{sec:ngram}
+An N-gram language model estimates the probability of a token occurring in a sequence based on the $N$ tokens preceding or succeeding that token.
 In other words, for a sequence, $X$, of $T$ tokens
@@
 \noindent
-Except for the first token in the sequence, the probability of a token occurring at a specific point depends on the tokens preceding it\footnote{The most common use case for an N-gram model is the preceding tokens.}.
-Due to this fact, N-grams are sometimes called left-associative models (in a left-to-right language).
+Except for the first token in the sequence, the probability of a token occurring at a specific point depends on the tokens preceding
+it\footnote{The most common use case for an N-gram model is the preceding tokens.}. Due to this fact, N-grams are sometimes called left-associative
+models (in a left-to-right language).
 
@@
 \noindent
-where $P(w)$ is the probability of a token, $w$, occurring anywhere in the corpus, $\phi_w(w)$ is the number of times a specific token appears in the corpus, and $D_{tokens}$ is the total number of tokens in the corpus, $D$.
-Computationally, determining the probabilities of the individual tokens in the corpus is typically referred to as training or fitting the N-gram model.
-
-$N$ is a predetermined value based on model or application requirements.
-While it can technically take any value from one to infinity, its usability becomes very limited at the minimum sequence length in the corpus and unusable at any value at or above the maximum sequence length.
-Depending on implementation, the time and resources required to train a model with a large value of $N$ can be very expensive.
-Common values for $N$ are one, two, and three, sometimes called unigrams, bigrams, and trigrams, respectively.
-
-\subsection{BERT}\label{sec:bert}
-The purpose of BERT (Bidirectional Encoder Representations from Transformers) is to represent tokens in text sequences based on the tokens preceding and following the token in question.
-Whereas an N-gram model represents token probabilities based on the $N$ tokens preceding a specific token or, conversely, the $N$ tokens succeeding said token, BERT models can predict the probability of a token based on both preceding and succeeding tokens; the bidirectionality novelty.
-
-\textbf{Masked Language Modeling}.
-The ability of BERT to represent text in a bidirectional context is achieved through the Masked Language Modeling (MLM) pretraining task~\cite{devlin_bert_2019}.
-MLM pretraining chooses a certain percentage of token positions at random as labels for prediction; those tokens are then replaced with either a ``mask'' token, a random token, or the original token according to preconfigured probabilities (BERT used 80\% for the mask token, 10\% for random tokens, and 10\% for the original/unchanged token).
-During training, the masked sequences are used as input to the model, and the original is used as the prediction target.
-The model prediction is scored against the target using cross-entropy loss.
-
-\textbf{Next Sentence Prediction}.
-While MLM conditions the model to represent tokens within sequences and predict the most probable token based on its position in the sequence, NSP is meant to condition the model to understand relationships between sentences/sequences to support downstream tasks.
-Since the BERT training set consists primarily of literary works and excerpts from Wikipedia, the model simply predicts whether two sentences appear together (one after the other) in the original text, making this a binary classification task as opposed to MLM, which is a multi-class classification task.
-Instead of a single sequence, the input consists of two sequences separated by a ``separation'' token and beginning with a ``classification'' token; these are represented by \lstinline|[SEP]| and \lstinline|[CLS]|, respectively in Devlin et al.~\cite{devlin_bert_2019}.
-
-The special tokens specifying the beginning and end of input sequences are shared between tasks for symmetry between pretraining tasks.
-For both MLM and NSP objectives, input sequences always start with a classification token and end with a separation token (see Figure \ref{fig:bert_mlm_input_example}).
-Likewise, NSP inputs always begin with a classification token, and each input sequence is concatenated with a separation token terminating each sequence (see Figure \ref{fig:bert_nsp_input_example} and equation (\ref{eq:bert_input_example})).
+where $P(w)$ is the probability of a token, $w$, occurring anywhere in the corpus, $\phi_w(w)$ is the number of times a specific token appears in the
+corpus, and $D_{tokens}$ is the total number of tokens in the corpus, $D$. Computationally, determining the probabilities of the individual tokens in
+the corpus is typically referred to as training or fitting the N-gram model.
+
+$N$ is a predetermined value based on model or application requirements. While it can technically take any value from one to infinity, its usability
+becomes very limited at the minimum sequence length in the corpus and unusable at any value at or above the maximum sequence length. Depending on
+implementation, the time and resources required to train a model with a large value of $N$ can be very expensive. Common values for $N$ are one, two,
+and three, sometimes called unigrams, bigrams, and trigrams, respectively.
+
+\subsubsection{Transformer}\label{sec:transformer}
+The transformer neural network uses an Encoder-Decoder architecture with multi-head self-attention layers throughout the network. Details of the
+general design of the network will be laid out here in more detail, including the construction of the layers, sub-layers, and operation of the
+attention mechanism. The design discussed here is identical to the network described in \textit{Attention Is All You Need}
+\cite{vaswani_attention_2017} and is the base of the design of BERT and RoBERTa models discussed in Sections \ref{sec:bert} and \ref{sec:roberta},
+respectively.
+
+\textbf{The Encoder Stack}. The encoding layers (referred to as the ``Encoder Stack''; see Figure \ref{fig:transformer_encoder_stack}) are composed
+of $N$ identical layers, each containing two sub-layers: (1) a multi-head self-attention mechanism and (2) a position-wise fully connected
+feed-forward network. Each sub-layer includes a residual connection that is added to the output of the sub-layer and normalized. All sub-layers
+produce outputs of dimension $d_{model}$. Vaswani et al.~\cite{vaswani_attention_2017} used $N=6$ encoding layers and $d_{model} = 512$ output
+dimensions. The feed-forward network uses an inner dimension of $d_{ff} = 2048$
+
+\begin{figure}
+    \centering
+    \includegraphics[width=0.5\linewidth]{figures/transformer_encoder_stack.png}
+    \caption{The Encoder Stack of the transformer architecture. Modified from \cite{vaswani_attention_2017}.}
+    \label{fig:transformer_encoder_stack}
+\end{figure}
+
+\textbf{The Decoder Stack}. The decoding layers are also comprised of $N$ layers (the ``Decoding Stack''; see Figure
+\ref{fig:transformer_decoder_stack}) with three sub-layers for each layer: (1) a masked multi-head attention sub-layer, (2) a multi-head attention
+sub-layer over the outputs of the Encoder Stack, and (3) a position-wise fully connected feed-forward network. The last two sub-layers are identical
+to those in the Encoder Stack. Each sub-layer also includes a residual connection that is added to the output of the layer and normalized. The first
+multi-head attention layer is modified to a masked multi-head attention layer, masking off tokens greater than the $i$th position. This, in
+combination with the output embeddings being shifted right by one position, ensure that predictions for a token position, $i$, are only dependent on
+outputs for positions less than $i$ \cite{vaswani_attention_2017}.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=0.6\linewidth]{figures/transformer_decoder_stack.png}
+    \caption{The Decoder Stack of the transformer architecture. Modified from \cite{vaswani_attention_2017}.}
+    \label{fig:transformer_decoder_stack}
+\end{figure}
+
+\textbf{The Multi-Head Attention Mechanism}. By the simplest of terms, attention mechanisms can be described as mapping a query and a set of keys and
+values to an output. In the case of the transformer described by Vaswani et al.~\cite{vaswani_attention_2017}, a version of attention called ``Scaled
+Dot-Product Attention'' is used. The query and keys are represented as vectors $Q$ and $K$, respectively, with dimensions of $d_k$. The values are
+represented by a vector, $V$, with a dimension of $d_v$. A dot-product of the query and keys is computed, scaled by $\sqrt{d_k}$. A softmax function
+is then applied to the output, given by equation (\ref{eq:softmax}).
+
+\begin{equation}\label{eq:softmax}
+    \sigma(x) = \frac{\exp(x_i)}{\sum_{j=1}^{N}\exp(x_j)} \quad \forall i=1..N
+\end{equation}
+
+\noindent
+Where $x$ is a vector of dimension $N$. Scaled-dot product attention can then be given by equation (\ref{eq:scaled_dot_attention}), below.
+
+\begin{equation}\label{eq:scaled_dot_attention}
+    \mbox{Attention}(Q,K,V) = \sigma\left(\frac{QK^T}{\sqrt{d_k}}\right)V
+\end{equation}
+
+Multi-Head attention is defined as multiple scaled-dot product attention layers, running in parallel. More specifically, the queries, keys, and values
+are linearly projected $h$ times with different learned linear projections to $d_k$, $d_k$, and $d_v$ dimensions, respectively. The attention function
+is applied to each projection resulting in output vectors with dimensions of $d_v$. These are concatenated and linearly projected once more to produce
+the output for the multi-head attention function \cite{vaswani_attention_2017}.
+
+\begin{equation}\label{eq:multi_head_attention}
+    \begin{gathered}
+        \mbox{MultiHeadAttention}(Q,K,V) = \mbox{Concat}(\mbox{head$_1$, ..., head$_h$})W^O\\
+        \mbox{where head$_i$ = Attention}(QW_i^Q, KW_i^K, VW_i^V)
+    \end{gathered}
+\end{equation}
+
+\noindent
+Where $W_i^Q$ and $W_i^K$ are weight matrices for the projections of dimensions $d_{model} \times d_k$, $W_i^V$ is a weight matrix for the projection
+of dimension $d_{model} \times d_v$ and $W_i^O$ is a weight matrix for the projections of dimension $h d_v \times d_{model}$.
+
+Vaswani et al.~\cite{vaswani_attention_2017} used $h = 8$ attention heads and $d_k = d_v = d_{model} / h = 64$.
+
+\textbf{Token Embeddings}. Input and output tokens are converted to learned representations of dimension $d_{model}$ by the model
+\cite{vaswani_attention_2017}.
+
+\subsubsection{BERT}\label{sec:bert}
+The purpose of BERT (Bidirectional Encoder Representations from Transformers) is to represent tokens in text sequences based on the tokens preceding
+and following the token in question. Whereas an N-gram model represents token probabilities based on the $N$ tokens preceding a specific token or,
+conversely, the $N$ tokens succeeding said token, BERT models can predict the probability of a token based on both preceding and succeeding tokens;
+the bidirectionality novelty.
+
+BERT is based on the transformer neural network described in Section \ref{sec:transformer} and uses an implementation that is nearly identical to that
+of the original model \cite{devlin_bert_2019,vaswani_attention_2017}. The notable differences between the original implementation and the
+implementation of the base and large BERT models are shown in Table \ref{tab:bert_params}.
+
+\begin{table}
+    \centering
+    \begin{tabular}{l r r r}
+        \toprule
+        Model          & Layers ($N$) & Hidden Layer Size ($d_{ff}$) & Attention Heads ($h$) \\
+        \midrule
+        AIAYN          & 6            & 2048                         & 8                     \\
+        BERT$_{base}$  & 12           & 768                          & 12                    \\
+        BERT$_{large}$ & 24           & 1024                         & 16                    \\
+        \bottomrule
+    \end{tabular}
+    \caption{Number of layers, hidden layer sizes, and number of attention heads for the base and large versions of the BERT model
+        compared to the model used in \textit{Attention Is All You Need} (AIAYN) \cite{devlin_bert_2019,vaswani_attention_2017}.}
+    \label{tab:bert_params}
+\end{table}
+
+\textbf{Representation of Inputs}. BERT uses a summation of the token embeddings (token IDs), segment embeddings (zero or one to represent the first
+or second sentence in a sentence pair), and position embeddings (the index of the token in the sequence) to form model inputs. Additionally, the
+WordPiece tokenizer algorithm (detailed in Section \ref{sec:wordpiece}) is used to segment tokens. The maximum sequence length of the model is 512
+tokens and the size of the learned embeddings corresponds to $d_{model}$ (768 for BERT$_{base}$ and 1024 for BERT$_{large}$) \cite{devlin_bert_2019}.
+
+\textbf{Masked Language Modeling}. The ability of BERT to represent text in a bidirectional context is achieved through the Masked Language Modeling
+(MLM) pretraining task~\cite{devlin_bert_2019}. MLM pretraining chooses a certain percentage of token positions at random as labels for prediction;
+those tokens are then replaced with either a ``mask'' token, a random token, or the original token according to preconfigured probabilities (BERT used
+80\% for the mask token, 10\% for random tokens, and 10\% for the original/unchanged token). During training, the masked sequences are used as input
+to the model, and the original is used as the prediction target. The model prediction is scored against the target using cross-entropy loss.
+
+\textbf{Next Sentence Prediction}. While MLM conditions the model to represent tokens within sequences and predict the most probable token based on
+its position in the sequence, NSP is meant to condition the model to understand relationships between sentences/sequences to support downstream tasks.
+Since the BERT training set consists primarily of literary works and excerpts from Wikipedia, the model simply predicts whether two sentences appear
+together (one after the other) in the original text, making this a binary classification task as opposed to MLM, which is a multi-class classification
+task. Instead of a single sequence, the input consists of two sequences separated by a ``separation'' token and beginning with a ``classification''
+token; these are represented by \lstinline|[SEP]| and \lstinline|[CLS]|, respectively in Devlin et al.~\cite{devlin_bert_2019}.
+
+The special tokens specifying the beginning and end of input sequences are shared between tasks for symmetry between pretraining tasks. For both MLM
+and NSP objectives, input sequences always start with a classification token and end with a separation token (see Figure
+\ref{fig:bert_mlm_input_example}). Likewise, NSP inputs always begin with a classification token, and each input sequence is concatenated with a
+separation token terminating each sequence (see Figure \ref{fig:bert_nsp_input_example} and equation (\ref{eq:bert_input_example})).
 
@@
 
-\begin{figure}[!t]
+\begin{figure}
     \includegraphics[width=\linewidth]{figures/BERT_MLM_input.png}
@@
 
-\begin{figure}[!t]
+\begin{figure}
     \includegraphics[width=\linewidth]{figures/BERT_NSP_input.png}
@@
 
-\subsection{RoBERTa}\label{sec:roberta}
+\subsubsection{RoBERTa}\label{sec:roberta}
 As the name implies, the Robustly Optimized BERT Pretraining Approach (RoBERTa) is an iteration on BERT (described in Section \ref{sec:bert}, above).
-The results from Liu et al.~\cite{liu_roberta_2019} indicate that BERT was initially significantly undertrained, so by training longer, with larger batch sizes, on longer sequences, and dynamically changing the masking pattern applied to the MLM pretraining task, the performance of BERT was improved.
-The new checkpoint of this version of BERT was, therefore, called RoBERTa.
-
-The pretraining methodology introduced by RoBERTa eliminated the NSP pretraining task originally used in BERT and opted to increase the quantity of training data along with the batch sizes and train the model longer.
-Lastly, there is a small difference in the construction of input sequences as compared to BERT; input sequences always begin with a classification token (\lstinline|[CLS]|), and individual segments are separated by a separation token (\lstinline|[SEP]|).
-Rather than having sequences end with a separation token, they are terminated with an end-of-sequence token (\lstinline|[EOS]|).
-See equation (\ref{eq:roberta_input_example}) modified from Liu et al.~\cite{liu_roberta_2019}, below.
+The results from Liu et al.~\cite{liu_roberta_2019} indicate that BERT was initially significantly undertrained, so by training longer, with larger
+batch sizes, on longer sequences, and dynamically changing the masking pattern applied to the MLM pretraining task, the performance of BERT was
+improved. The new checkpoint of this version of BERT was, therefore, called RoBERTa. The architecture and parameters of the RoBERTa models are
+identical to that of BERT.
+
+The pretraining methodology introduced by RoBERTa eliminated the NSP pretraining task originally used in BERT and opted to increase the quantity of
+training data along with the batch sizes and train the model longer. Lastly, there is a small difference in the construction of input sequences as
+compared to BERT; input sequences always begin with a classification token (\lstinline|[CLS]|), and individual segments are separated by a separation
+token (\lstinline|[SEP]|). Rather than having sequences end with a separation token, they are terminated with an end-of-sequence token
+(\lstinline|[EOS]|). See equation (\ref{eq:roberta_input_example}) modified from Liu et al.~\cite{liu_roberta_2019}, below.
 
@@
 
-\section{Tokenization Algorithms}\label{sec:tokenizers}
+\subsection{Tokenization Algorithms}\label{sec:tokenizers}
 % present and explain the tokenization algorithms used (possibly which models they correspond to)
-Tokenization is the process of separating or segmenting words in sentences into their component ``tokens''.
-The tokenized version of a sentence is referred to as a sequence.
-The rules by which the tokens are created or segmented depend entirely on the algorithm used.
-The simplest and most intuitive way to do this is by creating tokens based on whitespace, i.e., each word in the corpus is treated as a token\footnote{The Huggingface Tokenizer library implements this as the \lstinline|WordLevel| tokenizer.}.
-While this is simple and works in theory, it does not consider that not every word in a language will occur in a corpus.
-The inevitable possibility of a tokenizer seeing a word not in the original training corpus is known as the out-of-vocabulary (OOV) problem~\cite{wu_googles_2016}.
-Several different algorithms have addressed the OOV problem.
-For brevity's sake, only the tokenization algorithms used by the models in this thesis are defined and explained.
-
-\subsection{WordLevel}\label{sec:wordlevel}
-The WordLevel \cite{wolf_transformers_2020} tokenization algorithm, mentioned briefly as an introductory example above, is the most technically simple algorithm among those introduced in this section.
-Tokens are extrapolated from words in a sentence based on the whitespace separating each word.
-This means that contractions, hyphenations, etc.
-are treated as their own words, i.e.~ ``you'', ``you're'', and ``you've'' are all seen as unique, distinct words tokenized and mapped to their integral representations.
-
-An index of the words that appear in a corpus must be created to map individual tokens to numerical values or IDs.
-The numerical values for each token are derived from this index.
-For the WordLevel algorithm, this process is referred to as the tokenizer model training (see Algorithm \ref{alg:wordlevel_training}).
-
-The first step of the training procedure is to index any specified special tokens, such as mask, classification, or separation tokens (this is why special tokens often have low-valued token IDs).
-The second step is to split the sentences in the training data into individual words.
-This is done by splitting each string based on whitespace; in other words, the tokenizer treats sentences as lists of words delimited by space characters.
-Lastly, the tokenizer iterates through each word in each sentence, indexing new words as it finds them until all sentences in the training corpus have been analyzed.
-Token IDs are then derived based on the word's position in the tokenizer index or vocabulary.
+Tokenization is the process of separating or segmenting words in sentences into their component ``tokens''. The tokenized version of a sentence is
+referred to as a sequence. The rules by which the tokens are created or segmented depend entirely on the algorithm used. The simplest and most
+intuitive way to do this is by creating tokens based on whitespace, i.e., each word in the corpus is treated as a token\footnote{The Huggingface
+    Tokenizer library implements this as the \lstinline|WordLevel| tokenizer.}. While this is simple and works in theory, it does not consider that
+not every word in a language will occur in a corpus. The inevitable possibility of a tokenizer seeing a word not in the original training corpus is
+known as the out-of-vocabulary (OOV) problem~\cite{wu_googles_2016}. Several different algorithms have addressed the OOV problem. For brevity's sake,
+only the tokenization algorithms used by the models in this thesis are defined and explained.
+
+\subsubsection{WordLevel}\label{sec:wordlevel}
+The WordLevel \cite{wolf_transformers_2020} tokenization algorithm, mentioned briefly as an introductory example above, is the most technically simple
+algorithm among those introduced in this section. Tokens are extrapolated from words in a sentence based on the whitespace separating each word. This
+means that contractions, hyphenations, etc. are treated as their own words, i.e.~ ``you'', ``you're'', and ``you've'' are all seen as unique, distinct
+words tokenized and mapped to their integral representations.
+
+An index of the words that appear in a corpus must be created to map individual tokens to numerical values or IDs. The numerical values for each token
+are derived from this index. For the WordLevel algorithm, this process is referred to as the tokenizer model training (see Algorithm
+\ref{alg:wordlevel_training}).
+
+The first step of the training procedure is to index any specified special tokens, such as mask, classification, or separation tokens (this is why
+special tokens often have low-valued token IDs). The second step is to split the sentences in the training data into individual words. This is done by
+splitting each string based on whitespace; in other words, the tokenizer treats sentences as lists of words delimited by space characters. Lastly, the
+tokenizer iterates through each word in each sentence, indexing new words as it finds them until all sentences in the training corpus have been
+analyzed. Token IDs are then derived based on the word's position in the tokenizer index or vocabulary.
 
@@
 
-The algorithm's simplicity speaks for itself and serves as a good introduction to tokenization and an illustrative example of the OOV problem.
-Note that the values of the integers representing the tokens are determined based on the order in which the tokens are seen during the training process (special tokens are defined and, therefore, seen first), so these results are completely reproducible as long as the data appears in the same order.
-
-The tokenization of words with shared roots but different suffixes is shown in Table \ref{tab:wordlevel_tokenization_shared_roots_example}.
-The tokenizer sees all three words as independent and unique words since they are mapped to their own integer representations.
+The algorithm's simplicity speaks for itself and serves as a good introduction to tokenization and an illustrative example of the OOV problem. Note
+that the values of the integers representing the tokens are determined based on the order in which the tokens are seen during the training process
+(special tokens are defined and, therefore, seen first), so these results are completely reproducible as long as the data appears in the same order.
+
+The tokenization of words with shared roots but different suffixes is shown in Table \ref{tab:wordlevel_tokenization_shared_roots_example}. The
+tokenizer sees all three words as independent and unique words since they are mapped to their own integer representations.
 
@@
 
-The tokenizer used to generate the examples in tables \ref{tab:wordlevel_tokenization_example} and \ref{tab:wordlevel_tokenization_shared_roots_example} was trained on two sentences, notably lacking the word ``you'll''.
-Running the tokenizer on the word ``you'll'' alone results in the output in Table \ref{tab:wordlevel_unk_word}.
-Although the word shares a root with some of the words in the training data (``you'', ``you're'', and ``you've'') and is semantically very similar, the tokenizer does not recognize the word at all (represented by the ``[UNK]'' or unknown token).
-This word would be considered out-of-vocabulary since it did not appear in the training data and thus reveals one of the major downsides of this algorithm, as mentioned in section \ref{sec:tokenizers}.
-This problem can be alleviated to some extent by modifying the preprocessing strategy of the tokenizer to partition contracted words into their component parts, e.g.~the word ``you're'' would be partitioned to \lstinline|[``you'', ``''', ``re'']|, however, it eventually reappears, for example, when using the past, present, and future tenses of a verb (e.g.~``go'' and ``goes'' in the training data, then encountering the future tense: ``going'').
+The tokenizer used to generate the examples in tables \ref{tab:wordlevel_tokenization_example} and
+\ref{tab:wordlevel_tokenization_shared_roots_example} was trained on two sentences, notably lacking the word ``you'll''. Running the tokenizer on the
+word ``you'll'' alone results in the output in Table \ref{tab:wordlevel_unk_word}. Although the word shares a root with some of the words in the
+training data (``you'', ``you're'', and ``you've'') and is syntactically very similar, the tokenizer does not recognize the word at all (represented
+by the ``[UNK]'' or unknown token). This word would be considered out-of-vocabulary since it did not appear in the training data and thus reveals one
+of the major downsides of this algorithm, as mentioned in section \ref{sec:tokenizers}. This problem can be alleviated to some extent by modifying the
+preprocessing strategy of the tokenizer to partition contracted words into their component parts, e.g.~the word ``you're'' would be partitioned to
+\lstinline|[``you'', ``''', ``re'']|, however, it eventually reappears, for example, when using the past, present, and future tenses of a verb
+(e.g.~``go'' and ``goes'' in the training data, then encountering the future tense: ``going'').
 
@@
 
-\subsection{WordPiece}\label{sec:wordpiece}
+\subsubsection{WordPiece}\label{sec:wordpiece}
 % Wu et al. 2016; Schuster and Nakajima 2012
-Schuster and Nakajima \cite{schuster_japanese_2012} developed the WordPiece algorithm as a word segmentation algorithm for language modeling in Japanese and Korean voice search systems.
-It is further used by Wu et al.~\cite{wu_googles_2016} for neural machine translation tasks.
-The algorithm demonstrated proficiency and increased performance for the models used in both tasks and effectively addresses the OOV problem.
+Schuster and Nakajima \cite{schuster_japanese_2012} developed the WordPiece algorithm as a word segmentation algorithm for language modeling in
+Japanese and Korean voice search systems. It is further used by Wu et al.~\cite{wu_googles_2016} for neural machine translation tasks. The algorithm
+demonstrated proficiency and increased performance for the models used in both tasks and effectively addresses the OOV problem.
 
@@
 
-The training procedure is simple but computationally expensive.
-Schuster and Nakajima calculated the algorithm's time complexity at $O(K^2)$ where $K$ is the current size of the vocabulary~\cite{schuster_japanese_2012}.
-Due to the high time complexity of the algorithm, the authors suggest the following considerations to reduce the computational complexity:
+The training procedure is simple but computationally expensive. Schuster and Nakajima calculated the algorithm's time complexity at $O(K^2)$ where $K$
+is the current size of the vocabulary~\cite{schuster_japanese_2012}. Due to the high time complexity of the algorithm, the authors suggest the
+following considerations to reduce the computational complexity:
 
@@
 
-\subsection{Byte-Pair Encoding}\label{sec:bpe}
+\subsubsection{Byte-Pair Encoding}\label{sec:bpe}
 % Sennrich et al. 2016; Gage 1994
@@
 
-Finally, comparing the BPE tokenization process with the WordLevel and WordPiece algorithms, we can see that the outputs of the algorithm are nearly identical when the sentences being tokenized have words that appear in the training data (i.e.~the BPE outputs in tables \ref{tab:bpe_tokenization_example} and \ref{tab:bpe_shared_roots_example} are almost identical to the WordPiece output in tables \ref{tab:wordpiece_tokenization_example} and \ref{tab:wordpiece_tokenization_shared_roots_example} except the numerical representation of the tokens) and very similar to that of the WordLevel tokenizer.
+Finally, comparing the BPE tokenization process with the WordLevel and WordPiece algorithms, we can see that the outputs of the algorithm are nearly
+identical when the sentences being tokenized have words that appear in the training data (i.e.~the BPE outputs in tables
+\ref{tab:bpe_tokenization_example} and \ref{tab:bpe_shared_roots_example} are almost identical to the WordPiece output in tables
+\ref{tab:wordpiece_tokenization_example} and \ref{tab:wordpiece_tokenization_shared_roots_example} except the numerical representation of the tokens)
+and very similar to that of the WordLevel tokenizer.
 
@@
     \end{tabular}
-    \caption{Byte-Pair Encoding tokenizer output for semantically similar words with shared roots.}
+    \caption{Byte-Pair Encoding tokenizer output for syntactically similar words with shared roots.}
     \label{tab:bpe_shared_roots_example}
@@
 
-\textbf{Byte-Level Byte-Pair Encoding}.
-This is a subset of the BPE algorithm that is also commonly used for NLP tasks such as neural machine translation, language modeling, and generative tasks (GPT-2 uses a byte-level BPE tokenizer~\cite{radford_language_2019}).
-As Huggingface's \lstinline|tokenizers| library implements it, this is a preprocessing step for the tokenizer model that maps all bytes in a string/sentence to their own unique and visible character.
-Functionally, byte-level BPE is nearly identical to the standard BPE algorithm, described above, except that the beginning of word symbol is rendered and represented by a visible character in the tokenized sequence (see Table \ref{tab:byte_level_bpe_example}).
+\textbf{Byte-Level Byte-Pair Encoding}. This is a subset of the BPE algorithm that is also commonly used for NLP tasks such as neural machine
+translation, language modeling, and generative tasks (GPT-2 uses a byte-level BPE tokenizer~\cite{radford_language_2019}). As Huggingface's
+\lstinline|tokenizers| library implements it, this is a preprocessing step for the tokenizer model that maps all bytes in a string/sentence to
+their own unique and visible character. Functionally, byte-level BPE is nearly identical to the standard BPE algorithm, described above, except that
+the beginning of word symbol is rendered and represented by a visible character in the tokenized sequence (see Table
+\ref{tab:byte_level_bpe_example}).
 
@@
 
-\section{Discussion}\label{sec:discussion}
-Considering the success of deep learning models with transformer-based architectures for language modeling and natural language processing (NLP) tasks, more generally, in the conversational and literary English domains, it follows that transformer architectures should see widespread use in aviation English for domain-specific NLP tasks.
-However, the literature shows transformer-based language models have seen sparse use.
-The most notable example of using a transformer neural network is the application of BERT for knowledge extraction from notice to airmen (NOTAM) messages \cite{arnold_knowledge_2022}.
-
-The expectation is that the language models trained from scratch will outperform the pretrained language models due to the nature of the differences between literary English (the language the pretrained models were trained on) and aviation English.
-Additionally, the modeled data are orthographic transcriptions, extending the difference between the training corpora, thus leading to the hypothesis of the differences in model performance.
-To test this hypothesis, the models will be divided into two sets or classes: (1) models trained from scratch and (2) pretrained models that are fine-tuned (or transfer learned) using the data aggregated for this thesis (as described in Section \ref{sec:data_source}), these are denoted as the sets of \textit{scratch} and \textit{pretrained} models, respectively.
-Each class includes a version of the BERT and RoBERTa models, starting from scratch or the pretrained checkpoint, depending on the class of the model.
-To perform a fair comparison between models, the following parameters (and hyperparameters) are kept the same across all four models:
-
-\begin{itemize}
-    \item Optimization algorithm
-    \item Optimizer hyperparameters (see Table \ref{tab:optim_params})
-    \item Number of hidden layers
-    \item Number of neurons per layer
-    \item Number of attention heads
-    \item Activation functions
-    \item Type of tokenizer algorithm
-    \item Training, testing, and validation data
-\end{itemize}
-
-\subsection{Language Model Pretraining}\label{sec:lm_pretraining}
-A standard masked language modeling training procedure is used to pretrain the language models, as described in Section \ref{sec:bert}.
-Tokens are masked at random as described in the MLM process in Section \ref{sec:bert}.
-Input sequences are padded to the maximum length the models accept with a special \lstinline|[PAD]| token and truncated to the maximum length, if greater.
-The maximum sequence length for BERT and RoBERTa models is 512 tokens \cite{devlin_bert_2019,liu_roberta_2019}.
-Based on the data analysis in Section \ref{sec:data_analysis}, there are few, if any, samples that meet or exceed the maximum sequence length for BERT and RoBERTa models.
+\section{Experiments}\label{sec:experiments}
+This section lays out the details of the setup and results of each set of experiments. The experiments are divided into two sets: ASR model
+experiments and language model experiments. Each section defines the performance metrics used to evaluate the models during the training and testing
+procedures for each experiment as well as the results of the experiments, individually and compared to each other where possible.
+
+\subsection{Automatic Speech Recognition Experiments}\label{sec:asr_experiments}
+This section will summarize the experiment setup, evaluation metrics, and final results of the two best performing ASR models. Two Jasper and
+QuartzNet models were trained from scratch and pretrained checkpoints (four models evaluated in total), similar to how the language models in Section
+\ref{sec:lm_pretraining} were setup.
+
+\subsubsection{Performance Metrics}\label{sec:asr_performance_metrics}
+% word error rate
+This section defines the metrics used to track the performance of the ASR models during training and the performance of the models on the test
+partitions of their datasets.
+
+\textbf{Word Error Rate (WER)}. This is, simply, the number of words predicted incorrectly by the model. WER is based on the edit distance between the
+models prediction and the prediction target \cite{hunt_figures_1990}. It is defined formally as follows:
+
+Let $N$, $S$, $I$, and $D$ be the number of words, substitutions, insertions, and deletions, respectively. Then the WER is calculated as
+
+\begin{equation}\label{eq:word_error_rate}
+    \mbox{WER} = \frac{S + D + I}{N} \times 100
+\end{equation}
+
+\noindent
+In this case, we calculate WER as a percentage by multiplying by 100. Note that it is possible for the WER to be greater than 100\% if
+$S + D + I > N$. This can occur, for example, if the model predicts more words than are actually present in the audio signal.
+
+\subsubsection{Optimizer}\label{sec:asr_optimizer}
+The NovoGrad optimizer \cite{ginsburg_stochastic_2019} is used with the parameters shown in Table \ref{tab:novograd_params}.
+
+\begin{table}[!t]
+    \centering
+    \begin{tabular}{l r}
+        \toprule
+        Parameter     & Value \\
+        \midrule
+        Learning rate & 0.01  \\
+        Weight Decay  & 0.001 \\
+        $\beta_1$     & 0.8   \\
+        $\beta_2$     & 0.5   \\
+        \bottomrule
+    \end{tabular}
+    \caption{NovoGrad optimizer parameters for the Jasper and QuartzNet ASR model training procedures.}
+    \label{tab:novograd_params}
+\end{table}
+
+\subsubsection{Results}\label{sec:asr_results}
+The Jasper model trained from scratch performed best overall with about a 15\% WER on the test data. The pretrained QuartzNet model that was
+fine-tuned on the aggregated data performed second best among the evaluated models reaching a WER of about 30\%. These tests were run using a greedy
+decoding method, meaning that at each time step, the most active neuron or that with the highest value after softmax was chosen as the prediction.
+These results can likely be improved further by using a beam search decoding algorithm with a language model. The beam search decoding algorithm is
+explained further in Appendix B in Section \ref{sec:appendix_b} The best-performing ASR models are summarized in Table \ref{tab:asr_performance}.
+
+\begin{table}[!t]
+    \centering
+    \begin{tabular}{l r}
+        \toprule
+        Model                  & WER (\%) \\
+        \midrule
+        Jasper (scratch)       & 15.3     \\
+        QuartzNet (pretrained) & 30.2     \\
+        \bottomrule
+    \end{tabular}
+    \caption{WERs of the best-performing ASR models on the test partition of the corpus.}
+    \label{tab:asr_performance}
+\end{table}
+
+The other two models that were trained and tested are left out of Table \ref{tab:asr_performance} due to their unusually poor performance on average
+over the test samples in the corpus. The QuartzNet model trained from scratch consistently reached WERs above 100\% and the pretrained Jasper model
+did not perform better than 80\% WER on the test set.
+
+\subsection{Language Model Pretraining Experiments}\label{sec:lm_pretraining}
+This section will layout the experiment setup and results for the language modeling pretraining experiments. The language models are divided into two
+categories, with two models in each category. The first category of models are those trained from scratch meaning that the models weights are
+initialized according to a truncated normal distribution \cite{burkardt_truncated_2014}. The second category of models are based on pretrained
+checkpoints from BERT and RoBERTa \cite{devlin_bert_2019,liu_roberta_2019}. Each category contains a BERT and RoBERTa model.
+
+Considering the success of deep learning models with transformer-based architectures for language modeling and natural language processing (NLP)
+tasks, more generally, in the conversational and literary English domains, it follows that transformer architectures should see widespread use in
+aviation English for domain-specific NLP tasks. However, the literature shows transformer-based language models have seen sparse use. The most notable
+example of using a transformer neural network is the application of BERT for knowledge extraction from notice to airmen (NOTAM) messages
+\cite{arnold_knowledge_2022}.
+
+A standard masked language modeling training procedure is used to pretrain the language models, as described in Section \ref{sec:bert}. Tokens are
+masked at random as described in the MLM process in Section \ref{sec:bert}. Input sequences are padded to the maximum length the models accept with a
+special \lstinline|[PAD]| token and truncated to the maximum length, if greater. The maximum sequence length for BERT and RoBERTa models is 512 tokens
+\cite{devlin_bert_2019,liu_roberta_2019}. Based on the data analysis in Section \ref{sec:data_analysis}, there are few, if any, samples that meet or
+exceed the maximum sequence length for BERT and RoBERTa models.
+
+\subsubsection{Performance Metrics}\label{sec:performance_metrics}
+The following metrics are defined and used to track the performance of the language models over the course of the training procedures. They also serve
+as a means of measuring and comparing the performance of the models over the course of training and on the test partition of the corpus.
+
+\textbf{Cross-Entropy Loss}. As mentioned previously, perplexity and cross-entropy loss (often shortened to loss) are used to track the model's
+performance on the training and validation sets per epoch while training. Cross-entropy loss is defined below.
+
+The cross-entropy loss between two probability distributions, $p$ and $q$, is typically defined as
+
+\begin{equation}\label{eq:cross_entropy}
+    H(p, q) = -\sum_{x \in X} p(x) \log q(x)
+\end{equation}
+
+\noindent
+where $x$ is a single sample in the discrete distribution $X$. However, the way in which cross-entropy loss is measured can differ depending on
+application. PyTorch \cite{paszke_pytorch_2019}, for example, defines and measures cross-entropy loss as follows:
+
+Let $\mathcal{L}(x, y)$ be the cross-entropy loss function that outputs a $N \times 1$ vector, $L$, that contains the loss values for each sample in
+a batch of $N$ samples.
+
+\begin{equation}\label{eq:cross_entropy_defs}
+    \mathcal{L}(x, y) = L = \{\mbox{$l_1$, $l_2$, ..., $l_N$}\}^T
+\end{equation}
+
+\noindent
+Where $x$ is a $512 \times N$ vector representing model inputs (a sequence of tokens) and $y$ is a $1 \times N$ vector representing the target/label
+for each sample in the batch. The cross-entropy loss for a single sample is then calculated as
+
+\begin{equation}\label{eq:cross_entropy_loss}
+    l_n = -\log \left[\frac{\exp(x_{n,y_n})}{\sum_{c=1}^{C}\exp(x_{n,c})}\right] \quad \forall n=1..N
+\end{equation}
+
+\noindent
+where $l_n$ is the loss for an individual sample in the batch, and $C$ is the number of class labels. For language modeling, the set of class labels
+is equivalent to the vocabulary of the tokenizer (and thus the language model). Numerical labels will correspond to token IDs. The loss for a batch is
+calculated as the mean of the loss for each sample in the batch.
+
+\begin{equation}\label{eq:cross_entropy_mean}
+    \mathcal{L}(x, y) = \frac{1}{N}\sum_{i=1}^{N}l_i
+\end{equation}
+
+\noindent
+Equation (\ref{eq:cross_entropy_mean}) can then be extended to find the mean cross-entropy loss over an epoch by swapping $l_i$ for batch means and
+$\mathcal{L}(x, y)$ for the epoch mean. $N$ would then represent the number of batches in an epoch rather than the number of samples in a batch.
+
+In this case, it is important to note that cross-entropy loss is, basically, a measure of the distance between a models prediction, $\hat{y}$, and the
+intended target for the prediction, $y$.
+
+\textbf{Perplexity}. This metric measures a model's ability to predict the most probable token in a sequence. It will measure and track how well the
+language models can predict the most probable token for a specific point in a sequence. Perplexity is typically defined as
+
+\begin{equation}\label{eq:ppl}
+    \mbox{PPL} = 2^{-\frac{1}{N}\sum_{i=1}^{N} \log_2 q(x_i)}
+\end{equation}
+
+\noindent
+where $N$ is the number of samples in a sequence, $x$ is an individual sample, and $q$ is the probability model (a language model, for the purposes of
+this thesis). This is, essentially, the inverse of the probability of a sample appearing. A helpful way to view this is as how ``surprised'' a
+probability model is to see a sample, thus the lower the probability of a sample, the higher the perplexity will be.
+
+As with cross-entropy loss, the measure of perplexity can differ between implementations. PyTorch's \cite{paszke_pytorch_2019} implementation for
+perplexity is defined as follows:
+
+Let $P$ and $\hat{Y}$ be vectors representing the perplexity for samples in a batch and non-normalized output neurons in a neural network,
+respectively.
+
+\begin{equation}\label{eq:perplexity_defs}
+    \begin{gathered}
+        P = \{p_1, p_2, ..., p_N\}^T\\
+        \hat{Y} = \{\hat{y}_1, \hat{y}_2, ..., \hat{y}_C\}^T
+    \end{gathered}
+\end{equation}
+
+\noindent
+Where $N$ is the number of samples in a batch and $C$ is the number of class labels. Then, the perplexity for each sample in a batch can be
+calculated as
+
+\begin{equation}\label{eq:perplexity}
+    p_n = \exp \left[ \frac{1}{T}\sum_{i=1}^C - \log(\hat{y}_i) \right] \quad \forall n = 1..N
+\end{equation}
+
+\noindent
+where $p_n$ is the perplexity for an individual sample in the batch, and $T$ is the number of tokens in the sample. The perplexity for the batch is
+then calculated as the mean of the perplexities in the batch.
+
+\begin{equation}\label{eq:perplexity_mean}
+    \mbox{PPL} = \frac{1}{N}\sum_{i=1}^{N}p_i
+\end{equation}
+
+\noindent
+Where $\mbox{PPL}$ is the mean perplexity over a batch with $N$ samples. This equation can then be extended to find the mean perplexity over each
+batch in the epoch, swapping $p_i$ for batch means and $\mbox{PPL}$ for the epoch mean. $N$ would then represent the number of batches in an epoch
+rather than the number of samples in the batch.
+
+\subsubsection{Optimizer Settings}
 
 The models are trained with the \textit{AdamW} optimizer \cite{loshchilov_decoupled_2019} using the hyperparameters in Table \ref{tab:optim_params}
-based on the most effective pretraining configuration presented in \cite{devlin_bert_2019} and \cite{liu_roberta_2019}.
+based on the most effective pretraining configuration presented in \cite{devlin_bert_2019} and \cite{liu_roberta_2019}. These settings are kept
+constant across language model pretraining experiments.
 
@@
 
-A train/test split of 70\% and 30\% and a validation split of 10\% of the training data was used for each model.
-The models trained from scratch were trained for 200 epochs and the pretrained models were trained for 100 epochs.
-These values were chosen through trial and error, monitoring the perplexity and loss curves of the validation set to ensure that the pretraining does not result in net negative effects on the loss and perplexity of the model (i.e., ensure that the models are not overfitting to the training data).
-Batch sizes of 16 were used for training all four models; this was the maximum batch size that worked reliably with the hardware available to train the models\footnote{Values larger than 16 would sometimes cause Out of Memory errors on the GPU device used for training, so 16 was used to ensure all models trained continuously without error or interruption.
-Models were trained on a workstation with 64 GB of RAM and an NVIDIA RTX 3090 GPU with 24 GB of VRAM.}.
+\subsubsection{Data Splits}
+A train/test split of 70\% and 30\% and a validation split of 10\% of the training data was used for each model. A breakdown of the number of samples
+in each split is shown in Table \ref{tab:corpus_splits}. The samples present in each split are kept consistent across language model experiments so
+that the model performances are being compared based on identical data.
 
@@
 
-\subsection{Performance Metrics}\label{sec:performance_metrics}
-\textbf{Cross-Entropy Loss}.
-As mentioned previously, perplexity and cross-entropy loss (often shortened to loss) are used to track the model's performance on the training and validation sets per epoch while training.
-Cross-entropy loss is defined below.
-
-Let $\mathcal{L}(x, y)$ be the cross-entropy loss function that outputs a $N \times 1$ vector, $L$, that contains the loss values for each sample in a batch of $N$ samples.
-
-\begin{equation}\label{eq:cross_entropy_defs}
-    \mathcal{L}(x, y) = L = \{\mbox{$l_1$, $l_2$, ..., $l_N$}\}^T
-\end{equation}
-
-\noindent
-Where $x$ is a $512 \times N$ vector representing model inputs (a sequence of tokens) and $y$ is a $1 \times N$ vector representing the target/label
-for each sample in the batch. The cross-entropy loss for a single sample is then calculated as
-
-\begin{equation}\label{eq:cross_entropy_loss}
-    l_n = -\log \left[\frac{\exp(x_{n,y_n})}{\sum_{c=1}^{C}\exp(x_{n,c})}\right] \quad \forall n=1..N
-\end{equation}
-
-\noindent
-where $l_n$ is the loss for an individual sample in the batch, and $C$ is the number of class labels.
-For language modeling, the set of class labels is equivalent to the vocabulary of the tokenizer (and thus the language model).
-Numerical labels will correspond to token IDs.
-The loss for a batch is calculated as the mean of the loss for each sample in the batch.
-
-\begin{equation}\label{eq:cross_entropy_mean}
-    \mathcal{L}(x, y) = \frac{1}{N}\sum_{i=1}^{N}l_i
-\end{equation}
-
-\noindent
-Equation (\ref{eq:cross_entropy_mean}) can then be extended to find the mean cross-entropy loss over an epoch by swapping $l_i$ for batch means and $\mathcal{L}(x, y)$ for the epoch mean.
-$N$ would then represent the number of batches in an epoch rather than the number of samples in a batch.
-
-\textbf{Perplexity}.
-This metric measures a model's ability to predict the most probable token in a sequence.
-It will measure and track how well the language models can predict the most probable token for a specific point in a sequence.
-
-Let $P$ and $\hat{Y}$ be vectors representing the perplexity for samples in a batch and non-normalized output neurons in a neural network, respectively.
-
-\begin{equation}\label{eq:perplexity_defs}
-    \begin{gathered}
-        P = \{p_1, p_2, ..., p_N\}^T\\
-        \hat{Y} = \{\hat{y}_1, \hat{y}_2, ..., \hat{y}_C\}^T
-    \end{gathered}
-\end{equation}
-
-\noindent
-Where $N$ is the number of samples in a batch and $C$ is the number of class labels. Then, the perplexity for each sample in a batch can be calculated as
-
-\begin{equation}\label{eq:perplexity}
-    p_n = \exp \left[ \frac{1}{T}\sum_{i=1}^C - \log(\hat{y}_i) \right] \quad \forall n = 1..N
-\end{equation}
-
-\noindent
-where $p_n$ is the perplexity for an individual sample in the batch, and $T$ is the number of tokens in the sample. The perplexity for the batch is
-then calculated as the mean of the perplexities in the batch.
-
-\begin{equation}\label{eq:perplexity_mean}
-    \mbox{PPL} = \frac{1}{N}\sum_{i=1}^{N}p_i
-\end{equation}
-
-\noindent
-Where $\mbox{PPL}$ is the mean perplexity over a batch with $N$ samples. This equation can then be extended to find the mean perplexity over each
-batch in the epoch, swapping $p_i$ for batch means and $\mbox{PPL}$ for the epoch mean. $N$ would then represent the number of batches in an epoch
-rather than the number of samples in the batch.
-
-\subsection{Pretraining Results}\label{sec:pretraining_results}
-With the metrics above defined, an analysis of the results of the model pretraining can be performed.
-The pretraining procedure aims to build the models' understanding of the text structure and transcripts of air-traffic communications in this case.
-By tracking the average cross-entropy loss (loss) and perplexity (PPL) during training, we can estimate how well (and how correctly) the models are predicting tokens and if the models are effectively ``understanding'' token placements in sequences.
-
-Graphs of the training loss (average loss over the samples/batches in the training partition of the corpus) are presented in Figure \ref{fig:training_loss}.
-Both classes of models (scratch and pretrained) quickly reach a steady state in terms of loss, although the initial loss values are significantly different.
-Notably, the pretrained models start with a considerably lower loss, about 1.3, compared to models trained from scratch, which began with a loss value of just over 5.0.
-This indicates that the models are training well and starting to converge on local or global minima.
+\subsubsection{Results for the Models Trained from Scratch}\label{sec:scratch_pretraining_results}
+The models trained from scratch were trained for 200 epochs. This value was found through two rounds of trial and error by monitoring the curves of
+the loss and perplexity on the training and validation partitions of the corpus. This value was chosen to ensure that the pretraining procedure did
+not have an overall negative effect on the model performance. Batch sizes of 16 were used over the course of training; this was the maximum batch size
+that could be used with the hardware available\footnote{Values larger than 16 would sometimes cause Out of Memory errors on the GPU device used for
+    training, so 16 was used to ensure all models trained continuously without error or interruption. Models were trained on a workstation with 64 GB
+    of RAM and an NVIDIA RTX 3090 GPU with 24 GB of VRAM.}.
+
+A graph of the average loss per step of scratch models on the training partition are shown in Figure \ref{fig:scratch_training_loss}. Both models
+quickly converged to low loss values and maintained a relatively stable loss curve for the duration of the training procedure. This suggests that the
+models are converging on either local or global minima.
 
 \begin{figure}[!t]
-    \includegraphics[width=\linewidth]{figures/training_loss.png}
-    \caption{Loss on the training partition of the corpus over time (steps).}
-    \label{fig:training_loss}
+    \includegraphics[width=\linewidth]{figures/scratch_training_loss.png}
+    \caption{Cross-entropy loss of the models trained from scratch on the training partition of the corpus over time (steps).}
+    \label{fig:scratch_training_loss}
 \end{figure}
 
-The perplexity of the models on the training data yields interesting results (Figure \ref{fig:training_ppl}).
-Although the pretrained models have a more stable curve over time, the models trained from scratch result in a significantly lower perplexity.
-The final perplexity values (over the training set) of the models trained from scratch were markedly better than that of the pretrained models.
-At the end of training, the BERT and RoBERTa models trained from scratch had perplexity values of $0.52\times 10^7$ and $0.16\times 10^7$, respectively, whereas the pretrained models reached perplexity values of $0.29\times 10^9$ and $0.17\times 10^9$, for BERT and RoBERTa, respectively.
-Using the following equation,
+The graph of the perplexity of the models on the training partition of the corpus is shown in Figure \ref{fig:scratch_training_ppl}. The perplexity
+of the models over time is erratic and rarely settles to a consistent trend over the duration of the training procedure. Also note that the perplexity
+of the models is quite high throughout the training process. This is likely due to an issue with the models rapidly overfitting to the training data
+as evidenced by graphs of the loss and perplexity of the models on the validation partition of the corpus in Figures \ref{fig:scratch_validation_loss}
+and \ref{fig:scratch_validation_ppl}. This problem of rapid overfitting and consistently high perplexity is present in the results of the pretrained
+models as well (see Section \ref{sec:pretrained_pretraining_results}). Theories as to why the models' perplexities are so consistently high and why
+the models are overfitting so quickly will be explored in Section \ref{sec:results_comparison}.
+
+\begin{figure}[!t]
+    \includegraphics[width=\linewidth]{figures/scratch_training_ppl.png}
+    \caption{Perplexity of the models trained from scratch on the training partition of the corpus over time (steps).}
+    \label{fig:scratch_training_ppl}
+\end{figure}
+
+\begin{figure}[!t]
+    \includegraphics[width=\linewidth]{figures/scratch_validation_loss.png}
+    \caption{Cross-entropy loss of the models trained from scratch on the validation partition of the corpus over time (steps).}
+    \label{fig:scratch_validation_loss}
+\end{figure}
+
+\begin{figure}[!t]
+    \includegraphics[width=\linewidth]{figures/scratch_validation_ppl.png}
+    \caption{Perplexity of the models trained from scratch on the validation partition of the corpus over time (steps).}
+    \label{fig:scratch_validation_ppl}
+\end{figure}
+
+\subsubsection{Results for the Pretrained Models}\label{sec:pretrained_pretraining_results}
+The pretrained models were each trained for 100 epochs. As with the models trained from scratch, this value was found through two rounds of trial and
+error by monitoring the loss and perplexity values of the models on the training and validation partitions of the corpus to find a value that would
+ensure (or at least minimize) the net negative effects of overfitting. The batch size used for during training for the pretrained models is also 16
+(for the same reason as with the models trained from scratch).
+
+\begin{figure}[!t]
+    \includegraphics[width=\linewidth]{figures/pretrained_training_loss.png}
+    \caption{Cross-entropy loss of the pretrained models on the training partition of the corpus over time (steps).}
+    \label{fig:pretrained_training_loss}
+\end{figure}
+
+The graph of the training loss over time for the pretrained models can be seen in Figure \ref{fig:pretrained_training_loss}. Similarly to the models
+trained from scratch, the pretrained models quickly converge to a consistent low loss value over the duration of the training procedure with little
+to no deviation. The perplexity of the pretrained models on the training partition over the course of training, however, is much more stable and
+exhibits a somewhat consistent trend. There are still large spikes in the value of the perplexity throughout the training procedure and the overall
+values of the perplexity per step are still extremely high (note the scale of $10^9$ on the y-axis). As mentioned in Section
+\ref{sec:scratch_pretraining_results}, this is likely due to overfitting of the models to the data, evidenced by the rapid overfitting shown in
+Figures \ref{fig:pretrained_validation_loss} and \ref{fig:pretrained_validation_ppl}. A potential explanation as to why this phenomenon occurs is
+proposed and explored in Section \ref{sec:results_comparison}.
+
+\begin{figure}[!t]
+    \includegraphics[width=\linewidth]{figures/pretrained_training_ppl.png}
+    \caption{Perplexity of the pretrained models on the training partition of the corpus over time (steps).}
+    \label{fig:pretrained_training_ppl}
+\end{figure}
+
+\begin{figure}[!t]
+    \includegraphics[width=\linewidth]{figures/pretrained_validation_loss.png}
+    \caption{Cross-entropy loss of the pretrained models on the validation partition of the corpus over time (steps).}
+    \label{fig:pretrained_validation_loss}
+\end{figure}
+
+\begin{figure}[!t]
+    \includegraphics[width=\linewidth]{figures/pretrained_validation_ppl.png}
+    \caption{Perplexity of the pretrained models on the validation partition of the corpus over time (steps).}
+    \label{fig:pretrained_validation_ppl}
+\end{figure}
+
+\subsubsection{Comparison of Results}\label{sec:results_comparison}
+Both classes of models (scratch and pretrained) quickly reach a steady state in terms of loss on the training partition of the corpus, although the
+initial loss values are significantly different. Notably, the pretrained models start with a considerably lower loss, about 1.3, compared to the
+models trained from scratch, which began with a loss value of just over 5.0. This indicates that the models are training well and starting to converge
+on local or global minima. Although, as the graphs of the loss on the validation partition show, the models are quickly overfitting to the training
+data, causing the models to begin performing poorly the validation data, which is not seen by the model during training.
+
+Based on the trends exhibited by the cross-entropy loss of the models on the training and validation data, we would expect to see similar trends with
+the perplexity on the training and validation partitions. However, the perplexity of the models on the training and validation data yields interesting
+and somewhat unexpected results. Although the pretrained models have a more stable curve over time, the models trained from scratch result in a
+significantly lower perplexity. The final perplexity values (over the training set) of the models trained from scratch were markedly better than that
+of the pretrained models. At the end of training, the BERT and RoBERTa models trained from scratch had perplexity values of $0.52\times 10^7$ and
+$0.16\times 10^7$, respectively, whereas the pretrained models reached perplexity values of $0.29\times 10^9$ and $0.17\times 10^9$, for BERT and
+RoBERTa, respectively. Using the following equation
 
@@
 \noindent
-the relative difference in perplexity between the scratch and pretrained models on the final training epoch is 55.24 for the BERT models, and 104.57, for the RoBERTa models, which equates to an absolute difference (equation (\ref{eq:absolute_difference})) of $28.74 \times 10^7$ and $16.43 \times 10^7$ for the BERT and RoBERTa model, respectively.
+where $v$ is a tested value and $v_{ref}$ is a reference value. The relative difference in perplexity between the scratch and pretrained models on the
+final training epoch is 55.24 for the BERT models, and 104.57, for the RoBERTa models. Using the following equation for absolute difference,
 
@@
 
-The results from the training procedure show that the models trained from scratch outperformed their pretrained counterparts.
-This initially indicates that pretraining models on in-domain data may achieve the best language model performance, then continuing to fine-tune on specific downstream tasks.
-This seems to be a direct contradiction of the current recommended practice of using one of the pretrained checkpoints from Devlin et al.~\cite{devlin_bert_2019} and Liu et al.~\cite{liu_roberta_2019} then fine-tuning on specific downstream tasks.
-However, this is only partially upheld by the results of the model performance on the test partition of the corpus (Table \ref{tab:pretraining_results}).
-For the BERT models, the pattern exhibited during training is consistent with the results on the test partition, resulting in a relative difference of 8.62 and an absolute difference of $1.24 \times 10^7$.
-For the RoBERTa models, however, the relative difference in perplexity is -0.34 with an absolute difference of $0.36 \times 10^7$.
-These results show that the pretrained RoBERTa model performed better during pretraining than the corresponding model that was trained from scratch, unlike the BERT models (theories of why this occurs will be explored in the next paragraph).
-The loss of each model over the test set is somewhat high, but they are around the same values for each model, which is consistent with the loss statistics during training.
-A higher loss value on unseen (test) data is expected.
-
-The BERT model trained from scratch had the best performance among the tested models followed closely by the pretrained RoBERTa model, which partially confirms the hypothesis this thesis is based upon.
-The pretrained BERT model used both MLM and NSP pretraining objectives to create the checkpoint used in this thesis.
-The pretrained RoBERTa model, on the other hand, used only the MLM pretraining objective.
-It is readily apparent from the results of \cite{liu_roberta_2019} that RoBERTas modified pretraining methodology resulted in better model performance for in-domain problems.
-However, it seems that the pretraining method also allows RoBERTa to generalize better than BERT when transfer learned to other NLP domains.
-It is conceivable that, given additional training data in the future, the pretrained RoBERTa model could surpass the performance of the BERT model trained from scratch.
-However, with the current volume of data available in this domain, it is clear from the models' results on the corpus test partition that a BERT model trained from scratch will still be the best option for aviation English NLP tasks.
-
-\begin{figure}[!t]
-    \includegraphics[width=\linewidth]{figures/training_ppl.png}
-    \caption{Language model perplexity on the training partition of the corpus over time (steps). Note the scales on the y axes.}
-    \label{fig:training_ppl}
-\end{figure}
+\noindent
+this equates to an absolute difference of $28.74 \times 10^7$ and $16.43 \times 10^7$ for the BERT and RoBERTa models, respectively.
+
+The results from the training procedure show that the models trained from scratch outperformed their pretrained counterparts. This initially indicates
+that pretraining models on in-domain data may achieve the best language model performance, then continuing to fine-tune on specific downstream tasks.
+This seems to be a direct contradiction of the current recommended practice of using one of the pretrained checkpoints from Devlin et
+al.~\cite{devlin_bert_2019} and Liu et al.~\cite{liu_roberta_2019} then fine-tuning on specific downstream tasks. However, this is only partially
+upheld by the results of the model performance on the test partition of the corpus (Table \ref{tab:pretraining_results}). For the BERT models, the
+pattern exhibited during training is consistent with the results on the test partition, resulting in a relative difference of 8.62 and an absolute
+difference of $1.24 \times 10^7$ (using equations (\ref{eq:absolute_difference}) and (\ref{eq:relative_difference}), respectively). For the RoBERTa
+models, however, the relative difference in perplexity is -0.34 with an absolute difference of $0.36 \times 10^7$. These results show that the
+pretrained RoBERTa model performed better during pretraining than the corresponding model that was trained from scratch, unlike the BERT models. The
+loss of each model over the test set is somewhat high, but they are around the same values for each model, which is consistent with the loss
+statistics during training. A higher loss value on unseen (test) data is expected.
+
+The BERT model trained from scratch had the best performance among the tested models followed closely by the pretrained RoBERTa model. The pretrained
+BERT model used both MLM and NSP pretraining objectives to create the checkpoint used in this thesis. The pretrained RoBERTa model, on the other hand,
+used only the MLM pretraining objective. It is readily apparent from the results of \cite{liu_roberta_2019} that RoBERTa's modified pretraining
+methodology resulted in better model performance for in-domain problems. However, it seems that the pretraining method also allows RoBERTa to
+generalize better than BERT when transfer learned to other NLP domains. It is conceivable that, given additional training data in the future, the
+pretrained RoBERTa model could surpass the performance of the BERT model trained from scratch. However, with the current volume of data available in
+this domain, it is clear from the models' results on the corpus test partition that a BERT model trained from scratch will still be the best option
+for aviation English NLP tasks.
 
@@
 
-With the results and performance of the models established relative to each other, the high perplexity of each model must be addressed.
-While the loss values converge nicely during training, the perplexity is somewhat erratic and consistently high, rarely converging to a consistent curve.
-However, it exhibits a clear trend as training progresses.
-Adjusting the hyperparameters of the optimizer did not solve this issue, and changing the hyperparameters of the model (e.g.~sequence length, number of hidden layers, attention heads, etc.~) was not desirable for performing a direct comparison between the pretrained checkpoints and the base models to study the effectiveness of the models on orthographic transcriptions of aviation English.
-It is difficult to determine what exactly is causing this phenomenon, so several theories were developed in an attempt to explain it (some of which will be elaborated upon in Section \ref{sec:future_work}):
-
-\textbf{Small volume of training data}.
-Especially when compared to \textit{Attention is All You Need}, BERT, and RoBERTa, \cite{vaswani_attention_2017,devlin_bert_2019,liu_roberta_2019} the number of samples used in the training data dwarfs that of the state-of-the-art.
-It has been exhaustively demonstrated that neural networks perform better with large volumes of data.
-The inception of RoBERTa itself is based on the idea that BERT was not trained on enough data \cite{liu_roberta_2019}.
-Increasing the volume of in-domain data used for pretraining (and training) would be the most effective way to immediately improve the performance of the models.
-
-\textbf{Different forms of English}.
-The BERT and RoBERTa pretraining checkpoints were created by training the models on literary works and typed articles such as those from \textit{Wikipedia}.
-The data used in this work are orthographic transcriptions of aviation English, and while the two data sources share the common root of English, transfer learning from literary English to aviation English yields limited results, as evidenced above, especially with relatively small amounts of data.
-
-\textbf{Mixed regional data sources}.
-As described in Section \ref{sec:data_source}, the data aggregated for use here comes from several different regions, namely international airports in the US \cite{godfrey_air_1994} and international and domestic airports in Europe \cite{smidl_air_2019,hofbauer_atcosim_2008,szoke_detecting_2021}.
-During data processing, it was ensured that only English transcriptions were included in the corpus.
-However, proper nouns from each region will be present in the data.
-Additionally, dialectical differences and varying levels of English proficiency will affect the structure of token sequences across regions.
-All of these combined facets may contribute to the high perplexity of the language models relative to other works.
-It should be noted that the varied regional data sources are not necessarily bad.
-Depending on the application and desired robustness of the model, the ability to interpret international sources and contexts can be desirable for model selection; however, it will require significantly more data and proportionally balanced regional contexts within the corpus for the pretraining process to be effective for downstream tasks.
-
-Up to this point, we have followed a procedure for studying the effectiveness of the models that is somewhat symmetrical to that of BERT and RoBERTa, so at this point, the models should be trained for downstream tasks to measure the effectiveness of the pretraining procedures further.
-In the aviation domain and for this particular data, that would mean training for tasks such as callsign identification, speaker identification, flight phase classification, clearance recognition, etc.
-However, labels for these tasks are not present in the corpora used here.
-Corpora with relevant labels for downstream tasks could not be obtained within an appropriate time to run and include the experiments in this work.
-Some potential algorithms were developed to mine labels from the corpora aggregated for this work, which will be expanded upon in section \ref{sec:future_work}.
-For these reasons, testing model performance and effectiveness on downstream tasks must be left for future work.
-
-\section{Conclusion}\label{sec:conclusion}
-Four aviation English corpora are aggregated and unified for use in neural language modeling. A masked language modeling pretraining methodology is
-then used to pretrain two classes of BERT and RoBERTa models (models trained from scratch and pretrained models transfer learned to aviation English).
-
-Contrary to the recommended best practice of using pretrained checkpoints and training for downstream tasks, we find that, in the domain of aviation English, models achieve better performance when pretrained from scratch than model checkpoints transfer learned from the available BERT and RoBERTa pretrained checkpoints.
-Although the pretrained RoBERTa model achieved better performance on the test partition of the corpus than expected, the pretraining methods laid out in this thesis resulted in the BERT model trained from scratch achieving the lowest loss and perplexity overall.
-
-The results from the experiments in this work indicate that models trained from scratch for natural language processing tasks with limited data availability in the aviation domain are more effective than transfer learning pretrained models.
-Thus, models trained from scratch should be fine-tuned and used for downstream tasks over pretrained models.
-The most significant limiting factor to the performance of the models is the small volume of aviation English data available for pretraining and downstream task use.
-Investment in creating expansive, high-quality, openly available aviation English corpora with appropriate formatting for developing machine learning applications is also highly recommended.
+With the results and performance of the models established relative to each other, the high perplexity of each model must be addressed. While the loss
+values converge nicely during training, the perplexity is somewhat erratic and consistently high, rarely converging to a consistent curve. However, it
+exhibits a clear trend as training progresses. Adjusting the hyperparameters of the optimizer did not solve this issue, and changing the
+hyperparameters of the model (e.g.~sequence length, number of hidden layers, attention heads, etc.~) was not desirable for performing a direct
+comparison between the pretrained checkpoints and the base models to study the effectiveness of the models on orthographic transcriptions of aviation
+English. The validation curves for each model (Figures \ref{fig:scratch_validation_loss} and \ref{fig:pretrained_validation_loss}) clearly suggest
+that the language models are rapidly overfitting to the training data, which is what is causing the loss on the validation partition to increase as
+the models are trained further. It is difficult to determine what exactly is causing this phenomenon of the models rapidly and extremely overfitting
+to the training data, so several theories were developed in an attempt to explain it (some of which will be elaborated upon in Section
+\ref{sec:future_work}):
+
+\textbf{Small volume of training data}. Especially when compared to \textit{Attention is All You Need}, BERT, and RoBERTa,
+\cite{vaswani_attention_2017,devlin_bert_2019,liu_roberta_2019} the number of samples used in the training data dwarfs that of the state-of-the-art.
+It has been exhaustively demonstrated that neural networks perform better with large volumes of data. The inception of RoBERTa itself is based on the
+idea that BERT was not trained on enough data \cite{liu_roberta_2019}. Increasing the volume of in-domain data used for pretraining (and training)
+would be the most effective way to immediately improve the performance of the models.
+
+\textbf{Different forms of English}. The BERT and RoBERTa pretraining checkpoints were created by training the models on literary works and typed
+articles such as those from \textit{Wikipedia}. The data used in this work are orthographic transcriptions of aviation English, and while the two data
+sources share the common root of English, transfer learning from literary English to aviation English yields limited results, as evidenced above,
+especially with relatively small amounts of data.
+
+\textbf{Mixed regional data sources}. As described in Section \ref{sec:data_source}, the data aggregated for use here comes from several different
+regions, namely international airports in the US \cite{godfrey_air_1994} and international and domestic airports in Europe
+\cite{smidl_air_2019,hofbauer_atcosim_2008,szoke_detecting_2021}. During data processing, it was ensured that only English transcriptions were
+included in the corpus. However, proper nouns from each region will be present in the data. Additionally, dialectical differences and varying levels
+of English proficiency will affect the structure of token sequences across regions. All of these combined facets may contribute to the high perplexity
+of the language models relative to other works. It should be noted that the varied regional data sources are not necessarily bad. Depending on the
+application and desired robustness of the model, the ability to interpret international sources and contexts can be desirable for model selection;
+however, it will require significantly more data and proportionally balanced regional contexts within the corpus for the pretraining process to be
+effective for downstream tasks.
+
+Up to this point, we have followed a procedure for studying the effectiveness of the models that is somewhat symmetrical to that of BERT and RoBERTa,
+so at this point, the models should be trained for downstream tasks to measure the effectiveness of the pretraining procedures further. In the
+aviation domain and for this particular data, that would mean training for tasks such as callsign identification, speaker identification, flight phase
+classification, clearance recognition, etc. However, labels for these tasks are not present in the corpora used here. Corpora with relevant labels for
+downstream tasks could not be obtained within an appropriate time to run and include the experiments in this work. Some potential algorithms were
+developed to mine labels from the corpora aggregated for this work, which will be expanded upon in section \ref{sec:future_work}. For these reasons,
+testing model performance and effectiveness on downstream tasks must be left for future work.
+
+\section{Conclusions}\label{sec:conclusion}
+Four aviation English corpora are aggregated and unified for use in neural language modeling and speech recognition. A masked language modeling
+pretraining methodology is then used to pretrain two classes of BERT and RoBERTa language models (models trained from scratch and pretrained models
+transfer learned to aviation English). For ASR we used MFCCs, calculated from the audio signals aggregated from the aviation English corpora, and the
+transcripts used for language modeling to train and evaluate the models.
+
+Contrary to the recommended best practice of using pretrained checkpoints and training for downstream tasks, we find that, in the domain of aviation
+English, models achieve better performance when pretrained from scratch than model checkpoints transfer learned from the available BERT and RoBERTa
+pretrained checkpoints. Although the pretrained RoBERTa model achieved better performance on the test partition of the corpus than expected, the
+pretraining methods laid out in this thesis resulted in the BERT model trained from scratch achieving the lowest loss and perplexity overall. This
+trend also emerges in the ASR models; the Jasper model trained from scratch achieved the best performance among the models evaluated, including those
+from pre-existing checkpoints.
+
+The results from the experiments in this work indicate that models trained from scratch for natural language processing tasks with limited data
+availability in the aviation domain are more effective than transfer learning pretrained models. Thus, models trained from scratch should be
+fine-tuned and used for downstream tasks over pretrained models. The most significant limiting factor to the performance of the models is the small
+volume of aviation English data available for pretraining and downstream task use. Investment in creating expansive, high-quality, openly available
+aviation English corpora with appropriate formatting for developing machine learning applications is also highly recommended.
+
+Based on the results from training both ASR and language models from scratch and comparing the results with those fine-tuned from existing
+checkpoints, it is clear that, in both cases, the models trained from scratch not only have better performance overall but will also have a higher
+likelihood of reliable callsign detection/identification performance due to the higher individual performance overall. For future applications
+of NLP tasks in the aviation English domain, we recommend pretraining models from their base initializations using in-domain data and fine-tuning on
+additional labeled in-domain data for downstream tasks.
 
 \section{Future Work}\label{sec:future_work}
-This section is intended to serve as a brief proposal or set of suggestions for future work based on the results of this work.
-The main focus of these suggestions is corpus-specific algorithms and data augmentation techniques to enable testing and experiments to determine the effectiveness of the pretraining procedures covered earlier in this thesis.
-
-\textbf{Next Sentence Prediction (NSP)}.
-This training procedure was introduced by Devlin et al.~\cite{devlin_bert_2019} as a pretraining procedure for BERT and is described in more detail in Section \ref{sec:bert}.
-This can be used on top of masked language modeling for language model pretraining.
-However, it requires a knowledge of which sequences in the corpus appear next to each other.
-For pretraining transformer-based neural language models for the aviation domain (with orthographic transcriptions), this same procedure can be used to classify whether two transmissions occur in the correct order.
-In other words, it is classifying a pilot's response to an air traffic controller directive or an air traffic controller's response to a pilot query.
-Of the four corpora used in this work, ATCC is the only corpus that can be mined for an NSP-like pretraining procedure since the raw transcripts have indicators for time (start and end indicators), transmitting party, and intended receiver.
-With this information, we can estimate the correct order of transmissions between two parties to a reasonable degree of certainty (manually or programmatically).
-
-\textbf{Callsign Identification}.
-This is the process of identifying an aircraft callsign from the transcription of the spoken callsign.
-This is a popular subtask for automatic speech recognition applications in aviation, but postprocessing speech recognition outputs can also be accomplished through a language model.
-ATCC is a good candidate corpus for this task.
-There are two potential methods for implementing this as a downstream task: (1) In transmissions from air-traffic controllers to specific aircraft, use the receiver (``TO'' label) as the label for each sequence or (2) Identify the sequence of tokens, in each transmission, that correspond to the callsign of the aircraft and use those tokens as the labels for each sequence.
-Each of these methods brings its problems.
-The first method effectively creates an ever-expanding classification problem with infinitely many labels, so a tokenization or decoding scheme must be created to address this.
-The second method requires manually labeling every sequence in the corpus, which may not be feasible in all cases.
-Lastly, both methods may lack sufficient context to predict aircraft callsigns correctly.
-For example, general aviation (GA) callsigns in North America typically start with the letter \textit{N} followed by a series of numbers and letters.
-However, the spoken callsign is based on the aircraft manufacturer (e.g.~Cessna, Cirrus, Gulfstream, etc.~), so the spoken callsign will begin with the manufacturer name instead of ``November''.
-
-\textbf{Speaker Identification}.
-Generally, speaker identification is a task to classify specific speakers in a series of communications.
-This subtask usually falls under automatic speech recognition, but it is also feasible for language models under some circumstances.
-In aviation, this could be as simple as identifying the side of the communication, i.e., an air-traffic controller or a pilot, or it can be complex enough to be grouped under callsign identification tasks.
-In the simplest form, this can be accomplished easily using ATCC since controller positions are labeled along with transmitting/receiving aircraft.
-
-\textbf{Speech Recognition}.
-Finally, a sufficiently trained language model can improve speech recognition system performance.
-This approach has been used by several state-of-the-art automatic speech recognition models to improve sequence decoding and predictions.
-NVIDIA's Citrinet and QuartzNet models, for example, both use this approach to reduce their relative word error rates compared to the performance of other models on the same corpora \cite{majumdar_citrinet_2021,kriman_quartznet_2020}.
-Considering the relatively high word error rates of most automatic speech recognition models in the aviation domain, the pretraining procedures introduced here could further reduce these models' error rates.
-This will be elaborated further in Appendix B in Section \ref{sec:appendix_a}.
+This section is intended to serve as a brief proposal or set of suggestions for future work based on the results of this work. The main focus of these
+suggestions is corpus-specific algorithms and data augmentation techniques to enable testing and experiments to determine the effectiveness of the
+pretraining procedures covered earlier in this thesis.
+
+\subsection{Next Sentence Prediction}
+This training procedure was introduced by Devlin et al.~\cite{devlin_bert_2019} as a pretraining procedure for BERT and is described in more detail in
+Section \ref{sec:bert}. This can be used on top of masked language modeling for language model pretraining. However, it requires a knowledge of which
+sequences in the corpus appear next to each other. For pretraining transformer-based neural language models for the aviation domain (with orthographic
+transcriptions), this same procedure can be used to classify whether two transmissions occur in the correct order. In other words, it is classifying a
+pilot's response to an air traffic controller directive or an air traffic controller's response to a pilot query. Of the four corpora used in this
+work, ATCC is the only corpus that can be mined for an NSP-like pretraining procedure since the raw transcripts have indicators for time (start and
+end indicators), transmitting party, and intended receiver. With this information, we can estimate the correct order of transmissions between two
+parties to a reasonable degree of certainty (manually or programmatically).
+
+\subsection{Callsign Detection/Identification}
+This is the process of identifying an aircraft callsign from the transcription of the spoken callsign. This is a popular subtask for automatic speech
+recognition applications in aviation, but postprocessing speech recognition outputs can also be accomplished through a language model. ATCC is a good
+candidate corpus for this task. There are two potential methods for implementing this as a downstream task: (1) In transmissions from air-traffic
+controllers to specific aircraft, use the receiver (the ``TO'' label in ATCC's raw transcriptions) as the label for each sequence or (2) Identify the
+sequence of tokens, in each transmission, that correspond to the callsign of the aircraft and use those tokens as the labels for each sequence. Each
+of these methods brings its problems. The first method effectively creates an ever-expanding classification problem with infinitely many labels, so a
+tokenization or decoding scheme must be created to address this. The second method requires manually labeling every sequence in the corpus, which may
+not be feasible in all cases. Lastly, both methods may lack sufficient context to predict aircraft callsigns correctly. Callsigns are agreed upon
+between air carriers and air traffic control, but the callsign used will not always correspond directly to the air carrier's official three-letter
+designation for flights. For example, United Airlines uses the three-letter designator ``UAL'', however, the agreed upon spoken, and therefore
+transcribed, callsign is ``united'', so the flight ``UAL774'' will use the callsign ``united seven seven four''\footnote{The FAA maintains a list of
+    companies, callsigns, and three-letter designators at \url{https://www.faa.gov/air_traffic/publications/atpubs/cnt_html/chap3_section_4.html}}.
+General aviation (GA) callsigns introduce more variability and do not use previously agreed upon callsigns. Instead, for general aviation (GA)
+aircraft, the spoken information is based on the registration or tail number of the aircraft, but not every element of the registration number is
+necessarily required information in communications. For example, the FAA does not require the make or region of registration of the aircraft in spoken
+communications, however, they can and do appear in communications. The following transcript is taken from ATCC and is a communication from ATC to an
+aircraft with the registration number ``N01C'':
+
+\begin{quote}
+    TWIN CESSNA ZERO ONE CHARLIE TURN RIGHT HEADING ZERO NINER ZERO
+\end{quote}
+
+\noindent
+Note that the make of the aircraft, though not required, is included and the region of registration (the \textit{N} at the beginning of the
+registration number) is excluded. The callsign that should be identified in this case is ``TWIN CESSNA ZERO ONE CHARLIE'' and, ideally, this should
+be mapped to the registration ``N01C'', although this is difficult without additional context such as the region of airport and make of the aircraft.
+Even with this information, it does not guarantee correct identification of the aircraft; only an increased likelihood. This variability of
+information present in callsigns highlights the difficulty of callsign identification tasks. A well-trained language model with post-processing rules
+to map callsigns to three-letter airline designators could achieve exceptionally high performance on this task.
+
+\subsection{Speaker (Role) Identification}
+Generally, speaker identification is a task to classify specific speakers in a series of communications. This subtask
+usually falls under automatic speech recognition, but it is also feasible for language models under some circumstances. In aviation, this could be as
+simple as identifying the side of the communication, i.e., an air-traffic controller or a pilot, or it can be complex enough to be grouped under
+callsign identification tasks. In the simplest form, this can be accomplished easily using ATCC since controller positions are labeled along with
+transmitting/receiving aircraft.
 
@@
 \newpage
-\section{Appendix A: Examples of Samples Detected as Outliers}
-The outlying sample that was removed from the corpus is below\footnote{The preprocessing effects on the string, capitalization, punctuation, etc.~are left
-    unmodified for display purposes}:
+\section{Appendix A: Examples of Samples Detected as Outliers}\label{sec:appendix_a}
+The outlying sample that was removed from the corpus is below\footnote{The preprocessing effects on the string, capitalization, punctuation, etc.~are
+    left unmodified for display purposes}:
 
@@
     \end{tabular}
-    \caption{A few examples of samples detected as outliers that were kept in the aggregated corpus.}
+    \caption{A few examples of samples detected as outliers that were kept in the aggregated corpus. Samples with sequence lengths significantly
+        higher or lower than the average for the aggregated corpus were detected as outliers and flagged for additional analysis.}
     \label{tab:outlier_examples}
@@
 \newpage
-\section{Appendix B: Neural Language Modeling Application to Speech Recognition}\label{sec:appendix_a}
-NVIDIA's NeMo (Neural Modules) framework \cite{kuchaiev_nemo_2019} is a state-of-the-art deep learning framework with implementations for dozens of natural language processing models and utilities.
-Among those are the automatic speech recognition (ASR) model implementations and tools for decoding/augmenting predictions from model outputs.
-
-For ASR models, there are generally two methods of decoding predictions: greedy decoding and decoding with beam search.
-Greedy decoding methods simply use the neuron in the network's last layer (the output layer) with the highest value to determine the prediction for that time step.
-Beam search-based decoding methods use a language model and a beam search algorithm to determine predictions at each time step.
-The most commonly used language model for these types of decoding methods is an N-gram and is what PyTorch's \lstinline|CTCDecoder|\footnote{More specifically, the \lstinline|torchaudio| package of PyTorch} \cite{paszke_pytorch_2019} and NeMo's \lstinline|BeamSearchDecoderWithLM| \cite{kuchaiev_nemo_2019} use to decode predictions.
-
-The purpose of language models in beam search decoding algorithms is to estimate the likelihood of the decoded sequence occurring based on the language models' knowledge of sequence structures.
-The beam search algorithm is then used to select tokens to maximize the likelihood of the sequence occurring.
-As mentioned above in Section \ref{sec:ngram}, N-grams are left-associative language models (assuming the model is being applied to the English language in this case), which is an inherently limiting factor.
-A neural language model capable of representing sequences bidirectionally, such as those covered in this work, is a significantly more robust method of language modeling in speech recognition and has been used in the past to improve word error rates \cite{kriman_quartznet_2020,majumdar_citrinet_2021}.
-At least one instance of a deep learning-based language model being used for beam search decoding in aviation applications.
-However, it did not perform well, likely due to the small size of the dataset that it was trained on \cite{pellegrini_airbus_2019}.
-The models trained in this work used considerably more data and will likely perform better in beam search decoding algorithms.
+\section{Appendix B: Neural Language Modeling Application to Speech Recognition}\label{sec:appendix_b}
+NVIDIA's NeMo (Neural Modules) framework \cite{kuchaiev_nemo_2019} is a state-of-the-art deep learning framework with implementations for dozens of
+natural language processing models and utilities. Among those are the automatic speech recognition (ASR) model implementations and tools for
+decoding/augmenting predictions from model outputs.
+
+For ASR models, there are generally two methods of decoding predictions: greedy decoding and decoding with beam search. Greedy decoding methods simply
+use the neuron in the network's last layer (the output layer) with the highest value to determine the prediction for that time step. Beam search-based
+decoding methods use a language model and a beam search algorithm to determine predictions at each time step. The most commonly used language model
+for these types of decoding methods is an N-gram and is what PyTorch's \lstinline|CTCDecoder|\footnote{More specifically, the \lstinline|torchaudio|
+    package of PyTorch} \cite{paszke_pytorch_2019} and NeMo's \lstinline|BeamSearchDecoderWithLM| \cite{kuchaiev_nemo_2019} use to decode predictions.
+
+The purpose of language models in beam search decoding algorithms is to estimate the likelihood of the decoded sequence occurring based on the
+language models' knowledge of sequence structures. The beam search algorithm is then used to select tokens to maximize the likelihood of the
+sequence occurring. As mentioned above in Section \ref{sec:ngram}, N-grams are left-associative language models (assuming the model is being
+applied to the English language in this case), which is an inherently limiting factor. A neural language model capable of representing sequences
+bidirectionally, such as those covered in this work, is a significantly more robust method of language modeling in speech recognition and has been
+used in the past to improve word error rates \cite{kriman_quartznet_2020,majumdar_citrinet_2021}. At least one instance of a deep learning-based
+language model being used for beam search decoding in aviation applications. However, it did not perform well, likely due to the small size of the
+dataset that it was trained on \cite{pellegrini_airbus_2019}. The models trained in this work used considerably more data and will likely perform
+better in beam search decoding algorithms.
 

### 2023-12-22 thesis/MSECE_thesis.tex
Added input representations and training objective section for the ASR models.

@@
 
+\subsection{Input Representation}\label{sec:asr_input_representation}
 Although the internal processing of data varies between architectures, the format of the models input and output is the same for the models described
@@
 The MFCCs are then used as input to the ASR models. Each model outputs a probability distribution of characters or tokens (depending on the vocabulary
-the ASR model is initialized with) over a segment of audio. All the ASR models used are trained with a Connectionist Temporal Classification (CTC)
-loss function~\cite{graves_connectionist_2006}.
+the ASR model is initialized with) over a segment of audio.
+
+\subsection{Training Objective}\label{sec:asr_training_objective}
+The ASR models listed below both use a Connectionist Temporal Classification (CTC) training objective \cite{graves_connectionist_2006}. For a
+dataset, $D$, made up of inputs and labels, $x$ and $y$, respectively.
+
+\begin{equation}\label{eq:ctc_dataset}
+    D = \{(x_1, y_1), (x_2, y_2), ..., (x_N, y_N)\}
+\end{equation}
+
+\noindent
+Where $N$ is the number of samples in the dataset, $D$. The goal of the ASR model training procedure is to predict the target, $y$, based on the input
+sequence, $x$. This can also be written as maximizing the likelihood of $p(y|x)$. More formally, the CTC objective function can be written as
+
+\begin{equation}\label{eq:ctc_objective}
+    O^{ML}(D, \mathcal{N}) = - \sum_{(x,y) \in D} \ln p(y|x)
+\end{equation}
+
+\noindent
+where $\mathcal{N}$ is the network being optimized to maximize the likelihood of predicting $p(y|x)$. Put simply, the goal of the ASR network is to
+maximize the log-likelihood of predicting a label, $y$, given an input, $x$.
+
+The novelty of the CTC objective is in the fact that the audio signal does not need to be segmented in order to label and train a network. Instead,
+the output layer of the network has $L + 1$ units, where $L$ is the alphabet/vocabulary of the network. At each time step in the audio signal, the
+network will predict either a token from its vocabulary or a blank token, hence the extra unit in the output layer. Repeated predictions and blank
+tokens are removed from the prediction, then scored against the ground truth label for the audio signal \cite{graves_connectionist_2006}.
 

### 2023-12-22 thesis/MSECE_thesis.tex
Last thesis edits.

@@
 language models is not yet sufficient to produce results akin to those in the conversational and/or literary English domains. Additionally, the
-results in Section \ref{sec:pretraining_results} show that language models trained ``from scratch'' on in-domain data perform better than models
+results in Section \ref{sec:results_comparison} show that language models trained ``from scratch'' on in-domain data perform better than models
 trained on out-of-domain data and transfer learned on in-domain data.
@@
 \begin{equation}\label{eq:softmax}
-    \sigma(x) = \frac{\exp(x_i)}{\sum_{j=1}^{N}\exp(x_j)} \quad \forall i=1..N
+    \mbox{softmax}(x) = \frac{\exp(x_i)}{\sum_{j=1}^{N}\exp(x_j)} \quad \forall i=1..N
 \end{equation}
@@
 \begin{equation}\label{eq:scaled_dot_attention}
-    \mbox{Attention}(Q,K,V) = \sigma\left(\frac{QK^T}{\sqrt{d_k}}\right)V
+    \mbox{Attention}(Q,K,V) = \mbox{softmax}\left(\frac{QK^T}{\sqrt{d_k}}\right)V
 \end{equation}
@@
 Considering the success of deep learning models with transformer-based architectures for language modeling and natural language processing (NLP)
-tasks, more generally, in the conversational and literary English domains, it follows that transformer architectures should see widespread use in
+tasks, more generally, in the conversational and written English domains, it follows that transformer architectures should see widespread use in
 aviation English for domain-specific NLP tasks. However, the literature shows transformer-based language models have seen sparse use. The most notable
 example of using a transformer neural network is the application of BERT for knowledge extraction from notice to airmen (NOTAM) messages
-\cite{arnold_knowledge_2022}.
+\cite{arnold_knowledge_2022}. While both BERT and RoBERTa have demonstrated that they perform well on English in general
+\cite{devlin_bert_2019,liu_roberta_2019}, it is expected that the pretrained models will have limited performance due to the highly technical and
+domain-specific nature of aviation English as well as the differing forms of the language; transcribed English speech as opposed to written English.
+We expect the domain-specific terms and pronunciations will lead to a sufficiently different vocabulary from that of written English such that the
+models trained from scratch will achieve better performance than the pretrained models.
 
@@
 articles such as those from \textit{Wikipedia}. The data used in this work are orthographic transcriptions of aviation English, and while the two data
-sources share the common root of English, transfer learning from literary English to aviation English yields limited results, as evidenced above,
-especially with relatively small amounts of data.
+sources share the common root of English, transfer learning from the casual style of written English to the technical, highly specific use of aviation
+English yields limited results, as evidenced above, especially with relatively small amounts of data.
 
@@
 sequences in the corpus appear next to each other. For pretraining transformer-based neural language models for the aviation domain (with orthographic
-transcriptions), this same procedure can be used to classify whether two transmissions occur in the correct order. In other words, it is classifying a
-pilot's response to an air traffic controller directive or an air traffic controller's response to a pilot query. Of the four corpora used in this
-work, ATCC is the only corpus that can be mined for an NSP-like pretraining procedure since the raw transcripts have indicators for time (start and
-end indicators), transmitting party, and intended receiver. With this information, we can estimate the correct order of transmissions between two
-parties to a reasonable degree of certainty (manually or programmatically).
+transcriptions), this same procedure can be used to classify whether two transmissions occur in the correct order. Generalizing a bit, we can consider
+two individual transmissions to be $A$ and $B$; the training objective for the network is then to predict whether transmission $B$ occurs in response to
+transmission $A$. This would be considered a binary classification task; $B$ either occurs in response to $A$ or it does not. In practice, for a true
+label (meaning $B$ does, in fact, occur in response to $A$), transmission $A$, for example, can be a pilot request to ATC and transmission $B$ would
+be ATC's response. A false label (meaning $B$ does not occur in response to $A$) would be made up of transmissions randomly shuffled such that $A$ and
+$B$ do not occur together in the corpus. Note that, since these are only two transmissions, neither $A$ nor $B$ will necessarily be one side of the
+transmission i.e.~$A$ can be either a pilot or ATC. Of the four corpora used in this work, ATCC is the only corpus that can be mined for an NSP-like
+pretraining procedure since the raw transcripts have indicators for time (start and end indicators), transmitting party, and intended receiver. With
+this information, we can estimate the correct order of transmissions between two parties to a reasonable degree of certainty (manually or
+programmatically). For example, using the following transmissions from ATCC corpus, in which the Boston-Logan tower is communicating with PAA540. The
+transmission below can be used as $A$:
+
+\begin{quote}
+    CLIPPER SIXTY FIVE FORTY IS AH FOUR FROM RIPIT CLEARED FOR THE I L S D M E TWO SEVEN APPROACH
+\end{quote}
+
+\noindent
+and the response from PAA540 to the Boston-Logan tower can be used as $B$:
+
+\begin{quote}
+    AH ROGER CLEARED FOR THE APPROACH AH CLIPPER FIVE FORTY THANK YOU
+\end{quote}
+
+\noindent
+In this case, $A$ and $B$ occur in the middle of an interaction. There are two transmissions between PAA540 and the Boston-Logan tower that occur prior
+to the two above. We can consider the two instances of $A$ and $B$ above as a true label, meaning they occur one after the other in the corpus. To
+construct a false label, we can simply reverse the input of $A$ and $B$, i.e.~$A$ becomes $B$ and $B$ becomes $A$, or choose another transmission at
+random from the corpus to replace either $A$ or $B$. For proper construction of the corpus for the NSP pretraining task, the distribution of true and
+false labels should be evenly balanced. If there are 500 samples labeled as true, then there should be 500 samples labeled as false.
 

### 2024-01-02 thesis/MSECE_thesis.tex
Structural edits and proofreading edits.

@@
         \textit{
-            Language Modeling of Aviation Communications\\
+            Spoken Language Processing and Modeling for Aviation Communications\\
         }
@@
 With recent advances in machine learning and deep learning technologies and the creation of larger aviation-specific corpora, applying
-natural language processing technologies to aviation communications, especially those based on transformer neural networks, is becoming
+natural language processing technologies, especially those based on transformer neural networks, to aviation communications is becoming
 increasingly feasible. Previous work has focused on machine learning applications to natural language processing, such as N-grams and
-word lattices. This thesis will lay out a process for pretraining transformer-based language models on aviation English corpora and
+word lattices. This thesis experiments with a process for pretraining transformer-based language models on aviation English corpora and
 compare the effectiveness and performance of language models transfer learned from pretrained checkpoints and those trained from their
-base weight initializations (trained from scratch). The results show that transformer language models trained from scratch outperform
+base weight initializations (trained from scratch). The results suggest that transformer language models trained from scratch outperform
 models fine-tuned from pretrained checkpoints. The work concludes by recommending future work to improve pretraining performance and
@@
 EUROCONTROL simulation centers in Europe to train air traffic controllers and study the potential effects of AI-augmented air traffic management (ATM)
-systems on controller workloads. Preliminary studies have found that AI-augmented ATM systems reduce controller workloads and fuel consumption of
-aircraft by optimizing aircraft routes in the airspace and reducing departure/arrival times, thereby reducing fuel consumption
-\cite{helmke_quantifying_2017}.
+systems on controller workloads \cite{helmke_quantifying_2017,helmke_assistant-based_2015}. Preliminary studies have found that AI-augmented ATM
+systems reduce controller workloads and fuel consumption of aircraft by optimizing aircraft routes in the airspace and reducing departure/arrival
+times, thereby reducing fuel consumption \cite{helmke_quantifying_2017}.
 
@@
     low fidelity medium, depending on the equipment used to transmit and receive communications}, (3) a noisy and active communication environment
-i.e.~an active flight deck under the influence of wind, engine noise, etc., (4) an active Air Traffic Control (ATC) tower and channels, and (5) a
-rapid rate of speech. Pilots are also operating sophisticated equipment during while communicating, requiring significant attention and mental
-resources further contributing to the complexity of the communications \cite{paltridge_handbook_2013}.
+i.e.~an active flight deck under the influence of wind, engine noise, etc., (4) an active Air Traffic Control (ATC) tower and channels contributing to
+increased background noise on the ATC side, and (5) a rapid rate of speech. Pilots are also operating sophisticated equipment during while
+communicating, requiring significant attention and mental resources further contributing to the complexity of the communications
+\cite{paltridge_handbook_2013}.
 
@@
 of the fine-tuned model will be limited \cite{yadlowsky_pretraining_2023}. The main point of comparison in this thesis will be the performance of
-pretrained models fine-tuned on in-domain (aviation) data and models trained from scratch on in-domain data.
-
-\subsection{Notation and Terminology}
+pretrained models fine-tuned on in-domain (aviation) data and models trained from scratch on in-domain data. The experiments and results presented in
+this thesis are preliminary applications of end-to-end deep learning solutions to ASR and NLP problems in aviation. The intention is for this work to
+provide knowledge and possibly checkpoints for the ASR and language models which can be used for downstream tasks such as callsign detection and beam
+search prediction decoding for ASR models.
+
+\subsection{Notation \& Terminology}
 This section is included for the sake of clarity throughout this thesis. In pursuit of that, a few terms and notations are defined below.
@@
 \begin{itemize}
-    \item A \textbf{word} is the representation in writing of a sound or combination of sounds that symbolizes and communicates a meaning
+    \item A \textbf{word} is the representation in writing of a sound or combination of sounds that symbolizes and communicates a meaning.
     \item A \textbf{sentence} is a grammatical unit that is syntactically independent that is expressed or understood and contains at least one finite
-          verb
+          verb.
 \end{itemize}
@@
           characters, punctuation, etc. \cite{devlin_bert_2019,wu_googles_2016,sennrich_neural_2016}
-    \item A \textbf{sequence} shall be defined as a set of tokens (distinct from a sentence)
+    \item A \textbf{sequence} shall be defined as a set of tokens (distinct from a sentence).
 \end{itemize}
@@
 \noindent
-As mentioned above, we consider each sample in a corpus to be a sentence, then as the corpus is collated, processed, and tokenized, each sample will
-be processed into sequences of tokens. These sequences are what will be analyzed and modeled by the language models.
+As mentioned above, we consider each \textbf{sample} in a corpus to be a sentence, then as the corpus is collated, processed, and tokenized, each
+sample will be processed into sequences of tokens. These sequences are what will be analyzed and modeled by the language models.
 
@@
 % transformer history (background)
-The transformer neural network architecture was proposed in 2017 for Neural Machine Translation tasks and immediately achieved state-of-the-art
-performance on language translation tasks \cite{vaswani_attention_2017}. Transformer architectures are extremely effective at learning representations
-and understandings of languages to predict token probabilities instead of transforming one language into another
+The transformer neural network architecture (shown in Figure \ref{fig:transformer_arch}) was proposed in 2017 for Neural Machine Translation tasks and
+immediately achieved state-of-the-art performance on language translation tasks \cite{vaswani_attention_2017}. Transformer architectures are extremely
+effective at learning representations and understandings of languages to predict token probabilities instead of transforming one language into another
 \cite{devlin_bert_2019,liu_roberta_2019}. Transformers have also been found to be very effective at other NLP tasks such as prompt completion
@@
 al.~\cite{liu_roberta_2019} by training longer, with larger batch sizes, and removing the next sentence prediction task.
+
+\begin{figure}[!t]
+    \centering
+    \includegraphics[width=0.5\linewidth]{figures/transformer_arch.png}
+    \caption{The transformer neural network architecture. Copied from Vaswani et al.~\cite{vaswani_attention_2017}}
+    \label{fig:transformer_arch}
+\end{figure}
 
@@
 
-\section{Data Sources}\label{sec:data_source}
-Four corpora are used in this thesis. The datasets are primarily speech corpora intended for automatic speech recognition research in aviation
-communications, linguistics research in aviation English, or both. All data in all four corpora include both recordings and transcriptions, which are
-audio and text, respectively. The transcriptions are used to pretrain and train the language models and tokenizers (described in sections
-\ref{sec:language_models} and \ref{sec:tokenizers}).
+\section{Datasets \& Preparation}\label{sec:data_source}
+Four corpora are combined and used for the experiments in this thesis. The datasets are primarily speech corpora intended for automatic speech
+recognition research in aviation communications, linguistics research in aviation English, or both. All data in all four corpora include both
+recordings and transcriptions, which are audio and text, respectively. The transcriptions are used to pretrain and train the language models and
+tokenizers (described in sections \ref{sec:language_models} and \ref{sec:tokenizers}).
 
@@
 
-\section{Data Preparation}\label{sec:data_preparation}
+\subsection{Data Preparation}\label{sec:data_preparation}
 All corpora listed above (see Section \ref{sec:data_source}) are created at different times, for various purposes, and by multiple authors. As a
@@
 
-\clearpage
 \begin{lstlisting}
@@
 
-\clearpage
 ZCU CZ ATC uses a different nesting and labeling style, like below\footnote{Some information is excluded for readability.}:
@@
 
-% TODO: fix; lost some text somehow
 All data was processed in Python using primarily built-in functions/modules\footnote{The \lstinline|re| and \lstinline|xml| modules were also used,
@@
 format (an ASCII text file with one line in the file corresponding to one sample from the corpus). The resulting output is four corpus files that
-correspond to the original corpora in addition to a fifth corpus file with the samples from all four corpora combined and shuffled for later use.
+correspond to the original corpora in addition to a fifth corpus file with the samples from all four corpora combined and shuffled for later
+use\footnote{See the \lstinline|parse_transcripts| functions in the files at \url{https://github.com/AVanDeBrook/msece-thesis/tree/main/source/data}
+    for implementation details.}.
 
@@
 human error that introduced those errors in the first place. Instead, the frequency of token occurrence in the corpora, combined with manual review
-thereafter, was used to detect and correct the most common errors. The frequency of occurrence of a token is calculated using equation
-(\ref{eq:token_frequency_occurrence}).
+thereafter, was used to detect and correct the most common errors. Tokens with the lowest frequency across corpora (and lowest number of occurrences)
+were manually reviewed and corrected where necessary. Tokens that needed to be corrected typically occurred five times or less per corpus although
+all tokens with occurrences ranging from 1-50 were reviewed. The relative frequency of those tokens varies by corpus (due to the differing amont of
+tokens in total for each corpus) The frequency of occurrence of a token is calculated using equation (\ref{eq:token_frequency_occurrence}), below
 
@@
 \noindent
-Where $f_w$ is the frequency at which a token occurs in a corpus, $\phi_w(w)$ is the number of times a token, $w$, occurs, and $D_{tokens}$ is the
-total number of tokens in the corpus. Tokens with the lowest frequency across corpora (and lowest number of occurrences) were manually reviewed and
-corrected where necessary. An $A \rightarrow B$ mapping system was used to correct the errors, where \(A\) is the erroneous token and \(B\) is the
+where $f_w$ is the frequency at which a token occurs in a corpus, $\phi_w(w)$ is the number of times a token, $w$, occurs, and $D_{tokens}$ is the
+total number of tokens in the corpus.
+
+An $A \rightarrow B$ mapping system was used to correct the errors, where \(A\) is the erroneous token and \(B\) is the
 corrected or replaced token. As the text is parsed from the corpora, each sample is analyzed. If an erroneous token, $A$, is found, it is replaced
@@
 % hesitations, cut off words, different kinds of filled pauses
-The mapping system described above is also used to modify corpora conventions to be consistent across corpora. Three areas are addressed here:
+The mapping system described above is also used to make corpora conventions to be consistent across corpora. Three areas are addressed here:
 hesitations, cut-off, or partial words, and filled pauses within speech. These areas are defined differently between the four corpora, which makes
@@
         \midrule
-        Hesitations              & Instances of hesitation for which the specific realization is not defined (usually some special token is used instead). \\
+        Hesitations              & Instances of spoken words for which the specific realization is not defined (usually some special token is used instead). \\
         \midrule
@@
 % sequence lengths, unique tokens, total number of tokens, total number of samples, breakdown by each dataset
-Relevant corpora statistics were observed or calculated and are presented in Table \ref{tab:corpora_stats}. This includes the number of samples in the
-corpus, the mean sequence length per sample, the number of unique tokens, and the total number of tokens in the corpus. The same statistics were also
-calculated across corpora since all four corpora will be combined for training and testing the language and speech recognition models. Note that this
-is based on the raw data after loading and unifying the corpora before preprocessing for model training/testing and outlier removal.
+To better understand the above four corpora, we calculate and tabulate their relevant corpora statistics in Table \ref{tab:corpora_stats}. This
+includes the number of samples in the corpus, the mean sequence length per sample, the number of unique tokens, and the total number of tokens in the
+corpus. The same statistics were also calculated across corpora since all four corpora will be combined for training and testing the language and
+speech recognition models. Note that this is based on the raw data after loading and unifying the corpora before preprocessing for model
+training/testing and outlier removal.
 
@@
 
+\noindent
+In the equation above, $V_1$ and $V_2$ are the two numbers being compared.
+
 \subsection{Outlier Detection and Analysis}\label{sec:outliers}
@@
 
-Sci-Kit Learn's \cite{pedregosa_scikit-learn_2011} implementation of the LOF algorithm was used with a value of \(K = 35\) for the K-Nearest Neighbors
-algorithm and Minkowski distance for distance metrics/calculations (in this case, since \(p=2\), Euclidean distance is effectively used for distance
-calculations). A total of 47 samples were detected as outliers and removed for analysis. They are summarized in Table \ref{tab:outlier_stats}.
+Sci-Kit Learn's \cite{pedregosa_scikit-learn_2011} implementation of the LOF algorithm was used with a value of $K = 35$ for the K-Nearest Neighbors
+algorithm; the value for $K$ is arbitrarily chosen, partially through trial and error and the size of the clusters for sequence lengths. Minkowski
+distance is used for the distance metrics and is defined as
+
+\begin{equation}\label{eq:minkowski_distance}
+    d(x, y) = (|x - y|^p)^{\frac{1}{p}}
+\end{equation}
+
+\noindent
+where $x$ and $y$ are the points between which the distance is being measured and $p$ is the order on which to calculate the distance\footnote{For
+    the LOF algorithm, $p$ can be arbitrarily chosen.}. $p=2$ is chosen in this case to make the Minkowski distance equivalent to Euclidean distance:
+
+\begin{equation}\label{eq:euclidean_distance}
+    d(x, y) = \sqrt{|x-y|^2}
+\end{equation}
+
+\noindent
+A total of 47 samples were detected as outliers and removed for analysis. They are summarized in Table \ref{tab:outlier_stats}.
 
@@
 for one outlier, all of the samples labeled outlying by the LOF algorithm were kept in the data. Examples of the outliers are listed in Appendix A in
-Section \ref{sec:appendix_a}, including the sample removed.
-
-\section{Model Architectures}
-
-\subsection{Automatic Speech Recognition}\label{sec:automatic_speech_recognition}
+Section \ref{sec:appendix_a}, including the removed sample.
+
+\section{End-to-End ASR Models \& Experiment Results}
 Automatic speech recognition (ASR) or simply speech recognition is the process of translating spoken language to text i.e.~transcribing spoken
 language. There have been a myriad of approaches to ASR problems, including, but not limited to rule-based, machine learning, and deep learning
-methods. For the purposes of this work, we will focus primarily on deep learning approaches to ASR tasks with a specific focus on transcribing
-spoken language in aviation (following from the data accumulated, processed, and analyzed in sections \ref{sec:data_source},
+methods. For the purposes of this work, we will focus primarily on end-to-end deep learning approaches to ASR tasks with a specific focus on
+transcribing spoken language in aviation (following from the data accumulated, processed, and analyzed in sections \ref{sec:data_source},
 \ref{sec:data_preparation}, and \ref{sec:data_analysis}).
@@
 \subsection{Input Representation}\label{sec:asr_input_representation}
-Although the internal processing of data varies between architectures, the format of the models input and output is the same for the models described
-below. ASR models have seen much success and performance gains for speech-related tasks such as speech and speaker recognition through the use of
+Although the internal processing of data varies between architectures, the representation of the models' input for the models described below.
+ASR models have seen much success and performance gains for speech-related tasks such as speech and speaker recognition through the use of
 Mel-frequency Cepstral Coefficients (MFCCs) \cite{sahidullah_design_2012}. Since ASR is not the main focus of this thesis, the process for computing
@@
 \noindent
-The MFCCs are then used as input to the ASR models. Each model outputs a probability distribution of characters or tokens (depending on the vocabulary
-the ASR model is initialized with) over a segment of audio.
+The MFCCs are then used as input to the ASR models. It should be noted that MFCCs are not the only input representation available
+for ASR models, just the most popular choice among the models used here. Each model outputs a probability distribution of characters or tokens
+(depending on the vocabulary the ASR model is initialized with) over a segment of audio.
 
@@
 The ASR models listed below both use a Connectionist Temporal Classification (CTC) training objective \cite{graves_connectionist_2006}. For a
-dataset, $D$, made up of inputs and labels, $x$ and $y$, respectively.
+dataset, $D$, made up of samples and labels, $x$ and $y$, respectively, defined as
 
@@
 \noindent
-Where $N$ is the number of samples in the dataset, $D$. The goal of the ASR model training procedure is to predict the target, $y$, based on the input
+where $N$ is the number of samples in the dataset. The goal of the ASR model training procedure is to predict the target, $y$, based on the input
 sequence, $x$. This can also be written as maximizing the likelihood of $p(y|x)$. More formally, the CTC objective function can be written as
@@
 
-The novelty of the CTC objective is in the fact that the audio signal does not need to be segmented in order to label and train a network. Instead,
-the output layer of the network has $L + 1$ units, where $L$ is the alphabet/vocabulary of the network. At each time step in the audio signal, the
-network will predict either a token from its vocabulary or a blank token, hence the extra unit in the output layer. Repeated predictions and blank
-tokens are removed from the prediction, then scored against the ground truth label for the audio signal \cite{graves_connectionist_2006}.
-
-\subsubsection{Jasper}\label{sec:jasper}
-Jasper is a family of end-to-end convolutional neural networks (CNNs) developed by NVIDIA to replace traditional ASR models that use separately
-learned components for each stage of the pipeline (acoustic, pronunciation, language modeling, etc.) \cite{li_jasper_2019}.
-
-Jasper models are designed with $B$ blocks and $R$ sub-blocks and are named accordingly as \textit{Jasper $B\mbox{x}R$} models. The Jasper
-architecture also contains four extra convolutional blocks, one for preprocessing and three for postprocessing (see figure \ref{fig:jasper}). Each
-convolutional sub-block contains one dimensional convolutions. Li et al.~\cite{li_jasper_2019} trained and tested a Jasper 10x5 model that produced
-state-of-the-art results on the LibriSpeech, Wall Street Journal, and Fisher+Switchboard speech corpora.
+The novelty of the CTC objective is in the fact that the MFCC frames do not need to be aligned with the letter-based tokens to train a network.
+Instead, the output layer of the network has $L + 1$ units, where $L$ is the alphabet/vocabulary of the network. At each time step in the audio
+signal, the network will predict either a token from its vocabulary or a blank token, hence the extra unit in the output layer. Repeated predictions
+and blank tokens are removed from the prediction, then scored against the ground truth label for the audio signal \cite{graves_connectionist_2006}.
+Note that Graves et al.~\cite{graves_connectionist_2006} defined the tokens predicted at each time step as letters and the vocabulary,
+correspondingly, as the letters \textit{A} through \textit{Z} of the alphabet.
+
+\subsection{Jasper}\label{sec:jasper}
+Jasper is a family of deep convolutional neural networks (CNNs) developed by NVIDIA as end-to-end ASR models to replace traditional ASR models that
+use separately learned components for each stage of the pipeline (acoustic, pronunciation, language modeling, etc.) \cite{li_jasper_2019}.
+
+Jasper models, shown in Figure \ref{fig:jasper} are designed with $B$ blocks and $R$ sub-blocks and are named accordingly as
+\textit{Jasper $B\mbox{x}R$} models. The Jasper architecture also contains four extra convolutional blocks, one for preprocessing, shown at the
+bottom of Figure \ref{fig:jasper}, and three for postprocessing, shown at the top of Figure \ref{fig:jasper}. Each convolutional sub-block contains
+one dimensional convolutions. Li et al.~\cite{li_jasper_2019} trained and tested a Jasper 10x5 model that produced state-of-the-art results on the
+LibriSpeech, Wall Street Journal, and Fisher+Switchboard speech corpora.
 
@@
 
-\subsubsection{QuartzNet}\label{sec:quartznet}
+\subsection{QuartzNet}\label{sec:quartznet}
 The QuartzNet architecture is based on Jasper with some key modifications, namely, that the one dimensional convolutions are replaced with ``one
-dimensional time-channel separable convolutions''. As with Jasper, QuartzNet models are denoted as \textit{QuartzNet $B\mbox{x}R$}, however, blocks
-are repeated $S$ times and have a certain number of input and output channels, $c_{in}$ and $c_{out}$, respectively.
+dimensional time-channel separable convolutions'', as shown in Figure \ref{fig:quartznet}. As with Jasper, QuartzNet models are denoted as
+\textit{QuartzNet $B\mbox{x}R$} with $B$ and $R$ representing the number of blocks and sub-blocks, respectively. However, blocks are repeated $S$
+times and have a certain number of input and output channels, $c_{in}$ and $c_{out}$, respectively. Time-channel separable convolutions are separated
+into two components: (1) convolutional layers that operate on each audio channel independently across time frames and (2) convolutional layers that
+operate on time frames independently across all audio channels \cite{kriman_quartznet_2020}.
 
@@
 
-\subsection{Language Models}\label{sec:language_models}
-% bert, roberta
-% n-gram
+\subsection{Experiment Results}
+This section will summarize the experiment setup, evaluation metrics, and final results of the two best performing ASR models. Two Jasper models with 10
+blocks and 5 sub-blocks and two QuartzNet models with 15 blocks and 5 sub-blocks were trained from scratch and fine-tuned from pretrained
+checkpoints; four models were evaluated in total. Each model used a vocabulary of 28 tokens; the English alphabet, an apostrophe, and a blank label.
+
+\subsubsection{Performance Metrics}\label{sec:asr_performance_metrics}
+% word error rate
+This section defines the metrics used to track the performance of the ASR models during training and the performance of the models on the test
+partitions of their datasets.
+
+\textbf{Word Error Rate (WER)}. This is, simply, the number of words predicted incorrectly by the model. WER is based on the edit distance between the
+models prediction and the prediction target \cite{hunt_figures_1990}. Let $N$, $S$, $I$, and $D$ be the number of words, substitutions, insertions,
+and deletions, respectively. Then the WER is calculated as
+
+\begin{equation}\label{eq:word_error_rate}
+    \mbox{WER} = \frac{S + D + I}{N} \times 100
+\end{equation}
+
+\noindent
+In this case, we calculate WER as a percentage by multiplying by 100. Note that it is possible for the WER to be greater than 100\% if
+$S + D + I > N$. This can occur, for example, if the model predicts more words than are actually present in the audio signal.
+
+\subsubsection{Optimizer}\label{sec:asr_optimizer}
+The NovoGrad optimizer \cite{ginsburg_stochastic_2019} is used with the parameters shown in Table \ref{tab:novograd_params}.
+
+\begin{table}[!t]
+    \centering
+    \begin{tabular}{l r}
+        \toprule
+        Parameter     & Value              \\
+        \midrule
+        $\beta_1$     & $0.8$              \\
+        $\beta_2$     & $0.5$              \\
+        $\epsilon$    & $1 \times 10^{-8}$ \\
+        Learning rate & $0.01$             \\
+        Weight Decay  & $0.001$            \\
+        \bottomrule
+    \end{tabular}
+    \caption{NovoGrad optimizer parameters for the Jasper and QuartzNet ASR model training procedures.}
+    \label{tab:novograd_params}
+\end{table}
+
+The Jasper model trained from scratch performed best overall with about a 15\% WER on the test data. The pretrained QuartzNet model that was
+fine-tuned on the aggregated data performed second best among the evaluated models reaching a WER of about 30\%. The best-performing ASR models are
+summarized in Table \ref{tab:asr_performance}. These tests were run using a greedy decoding method, meaning that at each time step, the most active
+neuron i.e., that with the highest value after softmax was chosen as the prediction. These results can likely be improved further by adjusting the
+parameters of the model. However we believe the most significant boost in performance can be achieved using a well-performing language model with a
+beam search decoding algorithm. Beam search-based decoding methods use a language model and a beam search algorithm to determine predictions at each
+time step. The most commonly used language model for these types of decoding methods is an N-gram and is what PyTorch's
+\lstinline|CTCDecoder|\footnote{More specifically, the \lstinline|torchaudio| package of PyTorch.} \cite{paszke_pytorch_2019} and NeMo's
+\lstinline|BeamSearchDecoderWithLM| \cite{kuchaiev_nemo_2019} use to decode predictions. The purpose of language models in beam search decoding
+algorithms is to estimate the likelihood of the decoded sequence occurring based on the language models' knowledge of sequence structures. The beam
+search algorithm is then used to select tokens to maximize the likelihood of the
+sequence occurring. As mentioned above in Section \ref{sec:ngram}, N-grams are left-associative language models (assuming the model is being
+applied to the English language in this case), which is an inherently limiting factor. A neural language model capable of representing sequences
+bidirectionally, such as those covered in this work, is a significantly more robust method of language modeling in speech recognition and has been
+used in the past to improve word error rates \cite{kriman_quartznet_2020,majumdar_citrinet_2021}. At least one instance of a deep learning-based
+language model being used for beam search decoding in aviation applications. However, it did not perform well, likely due to the small size of the
+dataset that it was trained on \cite{pellegrini_airbus_2019}. In the next section, the language model architectures, experiments, and results are defined
+for the purpose of improving ASR model performance through the use of language models and downstream tasks such as callsign detection.
+
+\begin{table}[!t]
+    \centering
+    \begin{tabular}{l r}
+        \toprule
+        Model                  & WER (\%) \\
+        \midrule
+        Jasper (scratch)       & 15.3     \\
+        QuartzNet (pretrained) & 30.2     \\
+        \bottomrule
+    \end{tabular}
+    \caption{WERs of the best-performing ASR models on the test partition of the corpus.}
+    \label{tab:asr_performance}
+\end{table}
+
+\section{Language Models and Experiment Results}\label{sec:language_models}
 This section introduces the chosen language model architectures and their pretrained checkpoints. Two transformer-based architectures, BERT and
@@
 
-\subsubsection{N-Gram}\label{sec:ngram}
+\subsection{N-Gram}\label{sec:ngram}
 An N-gram language model estimates the probability of a token occurring in a sequence based on the $N$ tokens preceding or succeeding that token.
@@
 
-\subsubsection{Transformer}\label{sec:transformer}
+\subsection{Transformer}\label{sec:transformer}
 The transformer neural network uses an Encoder-Decoder architecture with multi-head self-attention layers throughout the network. Details of the
@@
 
-\subsubsection{BERT}\label{sec:bert}
+\subsection{BERT}\label{sec:bert}
 The purpose of BERT (Bidirectional Encoder Representations from Transformers) is to represent tokens in text sequences based on the tokens preceding
@@
 
-\subsubsection{RoBERTa}\label{sec:roberta}
+\subsection{RoBERTa}\label{sec:roberta}
 As the name implies, the Robustly Optimized BERT Pretraining Approach (RoBERTa) is an iteration on BERT (described in Section \ref{sec:bert}, above).
@@
 
-\section{Experiments}\label{sec:experiments}
-This section lays out the details of the setup and results of each set of experiments. The experiments are divided into two sets: ASR model
-experiments and language model experiments. Each section defines the performance metrics used to evaluate the models during the training and testing
-procedures for each experiment as well as the results of the experiments, individually and compared to each other where possible.
-
-\subsection{Automatic Speech Recognition Experiments}\label{sec:asr_experiments}
-This section will summarize the experiment setup, evaluation metrics, and final results of the two best performing ASR models. Two Jasper and
-QuartzNet models were trained from scratch and pretrained checkpoints (four models evaluated in total), similar to how the language models in Section
-\ref{sec:lm_pretraining} were setup.
-
-\subsubsection{Performance Metrics}\label{sec:asr_performance_metrics}
-% word error rate
-This section defines the metrics used to track the performance of the ASR models during training and the performance of the models on the test
-partitions of their datasets.
-
-\textbf{Word Error Rate (WER)}. This is, simply, the number of words predicted incorrectly by the model. WER is based on the edit distance between the
-models prediction and the prediction target \cite{hunt_figures_1990}. It is defined formally as follows:
-
-Let $N$, $S$, $I$, and $D$ be the number of words, substitutions, insertions, and deletions, respectively. Then the WER is calculated as
-
-\begin{equation}\label{eq:word_error_rate}
-    \mbox{WER} = \frac{S + D + I}{N} \times 100
-\end{equation}
-
-\noindent
-In this case, we calculate WER as a percentage by multiplying by 100. Note that it is possible for the WER to be greater than 100\% if
-$S + D + I > N$. This can occur, for example, if the model predicts more words than are actually present in the audio signal.
-
-\subsubsection{Optimizer}\label{sec:asr_optimizer}
-The NovoGrad optimizer \cite{ginsburg_stochastic_2019} is used with the parameters shown in Table \ref{tab:novograd_params}.
-
-\begin{table}[!t]
-    \centering
-    \begin{tabular}{l r}
-        \toprule
-        Parameter     & Value \\
-        \midrule
-        Learning rate & 0.01  \\
-        Weight Decay  & 0.001 \\
-        $\beta_1$     & 0.8   \\
-        $\beta_2$     & 0.5   \\
-        \bottomrule
-    \end{tabular}
-    \caption{NovoGrad optimizer parameters for the Jasper and QuartzNet ASR model training procedures.}
-    \label{tab:novograd_params}
-\end{table}
-
-\subsubsection{Results}\label{sec:asr_results}
-The Jasper model trained from scratch performed best overall with about a 15\% WER on the test data. The pretrained QuartzNet model that was
-fine-tuned on the aggregated data performed second best among the evaluated models reaching a WER of about 30\%. These tests were run using a greedy
-decoding method, meaning that at each time step, the most active neuron or that with the highest value after softmax was chosen as the prediction.
-These results can likely be improved further by using a beam search decoding algorithm with a language model. The beam search decoding algorithm is
-explained further in Appendix B in Section \ref{sec:appendix_b} The best-performing ASR models are summarized in Table \ref{tab:asr_performance}.
-
-\begin{table}[!t]
-    \centering
-    \begin{tabular}{l r}
-        \toprule
-        Model                  & WER (\%) \\
-        \midrule
-        Jasper (scratch)       & 15.3     \\
-        QuartzNet (pretrained) & 30.2     \\
-        \bottomrule
-    \end{tabular}
-    \caption{WERs of the best-performing ASR models on the test partition of the corpus.}
-    \label{tab:asr_performance}
-\end{table}
-
-The other two models that were trained and tested are left out of Table \ref{tab:asr_performance} due to their unusually poor performance on average
-over the test samples in the corpus. The QuartzNet model trained from scratch consistently reached WERs above 100\% and the pretrained Jasper model
-did not perform better than 80\% WER on the test set.
-
-\subsection{Language Model Pretraining Experiments}\label{sec:lm_pretraining}
+\subsection{Experiment Results}
 This section will layout the experiment setup and results for the language modeling pretraining experiments. The language models are divided into two
@@
 
-This sample has a sequence length of 71 tokens, well above the mean of the aggregated corpora. The decision was made to remove this sample, because
-it has little to no relevance to aviation communications, especially the ones being examined by the model in this thesis.
+This sample has a sequence length of 71 tokens, well above the mean of the aggregated corpora. The decision was made to remove this sample because
+it has little relevance to aviation communications, especially the ones being examined by the model in this thesis.
 
@@
 \newpage
-\section{Appendix B: Neural Language Modeling Application to Speech Recognition}\label{sec:appendix_b}
-NVIDIA's NeMo (Neural Modules) framework \cite{kuchaiev_nemo_2019} is a state-of-the-art deep learning framework with implementations for dozens of
-natural language processing models and utilities. Among those are the automatic speech recognition (ASR) model implementations and tools for
-decoding/augmenting predictions from model outputs.
-
-For ASR models, there are generally two methods of decoding predictions: greedy decoding and decoding with beam search. Greedy decoding methods simply
-use the neuron in the network's last layer (the output layer) with the highest value to determine the prediction for that time step. Beam search-based
-decoding methods use a language model and a beam search algorithm to determine predictions at each time step. The most commonly used language model
-for these types of decoding methods is an N-gram and is what PyTorch's \lstinline|CTCDecoder|\footnote{More specifically, the \lstinline|torchaudio|
-    package of PyTorch} \cite{paszke_pytorch_2019} and NeMo's \lstinline|BeamSearchDecoderWithLM| \cite{kuchaiev_nemo_2019} use to decode predictions.
-
-The purpose of language models in beam search decoding algorithms is to estimate the likelihood of the decoded sequence occurring based on the
-language models' knowledge of sequence structures. The beam search algorithm is then used to select tokens to maximize the likelihood of the
-sequence occurring. As mentioned above in Section \ref{sec:ngram}, N-grams are left-associative language models (assuming the model is being
-applied to the English language in this case), which is an inherently limiting factor. A neural language model capable of representing sequences
-bidirectionally, such as those covered in this work, is a significantly more robust method of language modeling in speech recognition and has been
-used in the past to improve word error rates \cite{kriman_quartznet_2020,majumdar_citrinet_2021}. At least one instance of a deep learning-based
-language model being used for beam search decoding in aviation applications. However, it did not perform well, likely due to the small size of the
-dataset that it was trained on \cite{pellegrini_airbus_2019}. The models trained in this work used considerably more data and will likely perform
-better in beam search decoding algorithms.
-
-\newpage
 \bibliographystyle{IEEEtran}

### 2024-01-08 thesis/MSECE_thesis.tex
Final round of edits; update title page.

@@
 \begin{center}
-    \begin{minipage}{3in}
-        \vspace*{0.25in}
-        THESIS COMMITTEE:\\
-        \vspace*{0.25in}
+    \begin{minipage}{4.5in}
+        \vspace*{0.35in}
         \hrule
@@
         Jianhua Liu, Ph.D.\\
-        Committee Chairman\\
-        \vspace*{0.75in}
+        Committee Chairman \& Graduate Program Coordinator \\
+        \vspace*{0.5in}
         \hrule
@@
         Committee Member\\
-        \vspace*{0.75in}
+        \vspace*{0.5in}
         \hrule
@@
         Department Chair, Electrical Engineering \& Computer Science\\
-        \vspace*{0.75in}
+        \vspace*{0.5in}
         \hrule
         \vspace*{2pt}
-        Associate Vice President for Academics
+        James Gregory, Ph.D.\\
+        Dean, College of Engineering\\
+        \vspace*{0.5in}
+        \hrule
+        \vspace*{2pt}
+        Christopher D.~Grant, Ph.D.\\
+        Vice Provost of Academic Support\\
     \end{minipage}
@@
 pretrained models fine-tuned on in-domain (aviation) data and models trained from scratch on in-domain data. The experiments and results presented in
-this thesis are preliminary applications of end-to-end deep learning solutions to ASR and NLP problems in aviation. The intention is for this work to
-provide knowledge and possibly checkpoints for the ASR and language models which can be used for downstream tasks such as callsign detection and beam
-search prediction decoding for ASR models.
+this thesis are preliminary applications of end-to-end deep learning solutions to Automatic Speech Recognition (ASR) and Natural Language Processing
+(NLP) problems in aviation. The intention is for this work to provide knowledge and possibly checkpoints for the ASR and language models which can be
+used for downstream tasks such as callsign detection and beam search prediction decoding for ASR models.
 
@@
     \includegraphics[width=0.5\linewidth]{figures/transformer_arch.png}
-    \caption{The transformer neural network architecture. Copied from Vaswani et al.~\cite{vaswani_attention_2017}}
+    \caption{The transformer neural network architecture. Copied from page 3 of Vaswani et al.~\cite[Fig. 1]{vaswani_attention_2017}}
     \label{fig:transformer_arch}
@@
 search algorithm is then used to select tokens to maximize the likelihood of the
-sequence occurring. As mentioned above in Section \ref{sec:ngram}, N-grams are left-associative language models (assuming the model is being
-applied to the English language in this case), which is an inherently limiting factor. A neural language model capable of representing sequences
-bidirectionally, such as those covered in this work, is a significantly more robust method of language modeling in speech recognition and has been
+sequence occurring. N-grams are left-associative language models (see Section \ref{sec:ngram}; assuming the model is being applied to the English
+language in this case), which is an inherently limiting factor. A neural language model capable of representing sequences bidirectionally, such as
+those covered in this work, is a significantly more robust method of language modeling in speech recognition and has been
 used in the past to improve word error rates \cite{kriman_quartznet_2020,majumdar_citrinet_2021}. At least one instance of a deep learning-based
@@
 
-Table \ref{tab:wordpiece_tokenization_example}, below, shows an example of a WordPiece tokenizer model output on a sentence that was seen during the training procedure (trained on the same data as the WordLevel tokenizer in Section \ref{sec:wordlevel}).
-Since this sentence was in the training data, the segmentation is identical to that of the WordLevel tokenizer in Table \ref{tab:wordlevel_tokenization_example}, however, tokenization of similar words (e.g.~words with shared roots or similar contractions) are handled differently, as in Table \ref{tab:wordpiece_tokenization_shared_roots_example}.
-Since each word has a shared root word in the contractions (as well as an apostrophe separating the contraction), the tokenizer segments each word and represents them as groups of small, more common tokens (thus with higher probabilities of occurring).
-Additionally, the WordPiece algorithm effectively handles OOV words.
-For example, the OOV tokens in Table \ref{tab:wordpiece_unk_word_similar}, with a shared root and similar contractions as those in Table \ref{tab:wordpiece_tokenization_shared_roots_example}, which appear in the training data.
-The base of the contraction, ``you'', occurs in the training data and the apostrophe in the contraction, so the tokenizer segments those parts of the word as tokens.
-The last part of the word, ``ll'', does not appear at all in the training data, so it is broken down into its component characters, which are represented as ``l'' (beginning of a word) and ``\#\#l'' (segmented part of a word that should be concatenated to the previous token when they are recombined).
-Comparing this to the output of the WordLevel tokenizer on the same word in Table \ref{tab:wordlevel_unk_word}, the WordPiece algorithm effectively addresses the OOV problem for words similar to those in the training corpus and, according to the output of the WordPiece tokenizer in Table \ref{tab:wordpiece_unk_word}, for words unlike those in the training corpus.
+Table \ref{tab:wordpiece_tokenization_example} shows an example of a WordPiece tokenizer model output on a sentence that was seen during the training
+procedure (trained on the same data as the WordLevel tokenizer in Section \ref{sec:wordlevel}). Since this sentence was in the training data, the
+segmentation is identical to that of the WordLevel tokenizer in Table \ref{tab:wordlevel_tokenization_example}, however, tokenization of similar
+words (e.g.~words with shared roots or similar contractions) are handled differently, as in Table
+\ref{tab:wordpiece_tokenization_shared_roots_example}. Since each word has a shared root word in the contractions (as well as an apostrophe separating
+the contraction), the tokenizer segments each word and represents them as groups of small, more common tokens (thus with higher probabilities of
+occurring). Additionally, the WordPiece algorithm effectively handles OOV words. For example, the OOV tokens in Table
+\ref{tab:wordpiece_unk_word_similar}, with a shared root and similar contractions as those in Table
+\ref{tab:wordpiece_tokenization_shared_roots_example}, which appear in the training data. The base of the contraction, ``you'', occurs in the training
+data and the apostrophe in the contraction, so the tokenizer segments those parts of the word as tokens. The last part of the word, ``ll'', does not
+appear at all in the training data, so it is broken down into its component characters, which are represented as ``l'' (beginning of a word) and
+``\#\#l'' (segmented part of a word that should be concatenated to the previous token when they are recombined). Comparing this to the output of the
+WordLevel tokenizer on the same word in Table \ref{tab:wordlevel_unk_word}, the WordPiece algorithm effectively addresses the OOV problem for words
+similar to those in the training corpus and, according to the output of the WordPiece tokenizer in Table \ref{tab:wordpiece_unk_word}, for words
+unlike those in the training corpus.
 
@@
 Both classes of models (scratch and pretrained) quickly reach a steady state in terms of loss on the training partition of the corpus, although the
-initial loss values are significantly different. Notably, the pretrained models start with a considerably lower loss, about 1.3, compared to the
+initial loss values are considerably different. Notably, the pretrained models start with a considerably lower loss, about 1.3, compared to the
 models trained from scratch, which began with a loss value of just over 5.0. This indicates that the models are training well and starting to converge
@@
 and somewhat unexpected results. Although the pretrained models have a more stable curve over time, the models trained from scratch result in a
-significantly lower perplexity. The final perplexity values (over the training set) of the models trained from scratch were markedly better than that
+lower perplexity. The final perplexity values (over the training set) of the models trained from scratch were markedly better than that
 of the pretrained models. At the end of training, the BERT and RoBERTa models trained from scratch had perplexity values of $0.52\times 10^7$ and
